### 2018-09-13 analyses/identify-optimal-lbi-parameters/results.tex
Added analysis code, figures, and write up for identification of optimal LBI parameters

\documentclass[12pt]{article}

\usepackage{graphicx}
\usepackage{caption}
\usepackage[font=normalsize]{subcaption}
\usepackage{float}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{wrapfig}

% spacing surround wrapfigs
\setlength{\columnsep}{12pt}

% bold the 'Figure~#' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelsep=period,font=small,labelfont=bf]{caption}

%%% STYLING %%%%%%%%%%%%%%%%%%%%%%%
\usepackage[compact]{titlesec}
\titleformat{\subsection}{\normalfont\fontsize{13}{15}\bfseries}{\thesubsubsection}{1em}{}
\titleformat{\subsubsection}{\normalfont\fontsize{12}{15}\bfseries}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\normalfont\bfseries}{\theparagraph}{3pt}{}[ $\;\cdotp\;$]
\titlespacing{\paragraph}{0pt}{*1}{*0.5}
\usepackage{parskip}
\setlength{\parskip}{0.15cm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{Identifying optimal LBI parameters}

\subsection{Parameter sweep}

To determine the optimal tau and window parameters for the LBI fitness predictor, I ran a parameter sweep for all possible combinations of tau and window values across a range of both parameters.
I tested tau values of 0.03125, 0.0625, 0.125, 0.25, 0.5, 0.75, 1.0 and window values of 0.1, 0.25, 0.75.
I evaluated the accuracy of each combination of tau and window values by running the fitness model with the LBI predictor on a tree with 92 viruses per month across Oct 2006--Apr 2018 and measuring the frequency fold change correlation and Matthew's correlation coefficient (MCC) for growth or decline of clade frequencies.
The first of these two metrics measures how well the model predicts the specific change in frequency for each clade.
The second metric measures how well the model predicts that a clade will grow or decline in frequency.

\begin{figure}[h]
\includegraphics[width=5in]{mcc_by_lbi_parameters.pdf}
\caption{Matthew's correlation coefficient (MCC) by LBI parameters}
\label{mcc-by-lbi-parameters}
\end{figure}

The model performed best by MCC (0.35) with tau of 0.25 and a window of 0.1 or 0.25.
The model performed best by correlation (0.20) with tau of 1.0 and window of 0.1.
Importantly, the current LBI parameters for 12 year trees (tau of 0.75 and window of 0.75) performed poorly by both correlation (0.06) and MCC (0.02).
The longest time window (0.75) produced the worst MCCs for all values of tau.
For tau greater or equal to 0.125, the window sizes of 0.1 and 0.25 were almost equally accurate.

\subsection{Model performance}

Since LBI is not expected to perform well in predicting specific clade frequencies but rather in predicting clade growth or decline, I opted to select the LBI parameters that maximized the MCC (tau of 0.25 and window of 0.1).
I ran the forecasting builds with all predictors and these LBI parameters.
In addition to the standard ``all'' predictor that includes all individual predictors, I added a predictor based on the titer substitution model and LBI and another based on titers, LBI, and non-epitope mutations.
The models with the best correlation and MCC values were consistently those with titer substitution predictors (Fig.~\ref{fig:model-frequency-correlations} and \ref{fig:model-mcc}).

\begin{figure}
\includegraphics[width=5in]{../../figures/frequency_correlation.pdf}
\caption{\label{fig:model-frequency-correlations}Pearson's correlation between observed and predicted frequency fold change by fitness predictors.}
\end{figure}

\begin{figure}
\includegraphics[width=5in]{../../figures/mcc.pdf}
\caption{\label{fig:model-mcc}Matthew's correlation coefficient (MCC) for clade growth or decline predictions by fitness predictors.}
\end{figure}

By itself, the LBI predictor has a small, positive $\beta$ value of 0.08 (Fig.~\ref{fig:model-parameters}) and a relatively low correlation of 0.13 (Fig.~\ref{fig:model-frequency-correlations}).
However, LBI also has the second highest MCC (0.35) after the titer substitution model indicating that it generally does a better job of predicting clade growth or decline (Fig.~\ref{fig:model-mcc}).
In combination with other predictors, LBI has a larger, negative $\beta$ with the largest value of -0.33 in the combined titers and LBI model.
Despite this negative $\beta$ parameter for LBI, its inclusion with the titer substitution model produces a better MCC than either individual predictor.

\begin{figure}
\includegraphics[width=5in]{../../figures/model_parameters.pdf}
\caption{\label{fig:model-parameters}Model parameters for all combinations of fitness predictors.}
\end{figure}

To understand why the LBI predictor had such a low overall correlation between predicted and observed frequency change, I inspected these predicted and observed values by groups of clades at different initial frequencies.
For each group of clades, I recalculated the Pearson's correlation and Matthew's correlation coefficient.
LBI performs best with medium- to large-sized clades with an initial frequency greater than 50\% (Fig.~\ref{fig:model-fold-change-lbi}).
For clades with an initial frequency less than 50\%, LBI is often overly optimistic about future clade growth.

\begin{figure}
\includegraphics[width=5in]{../../model_fold_change/2006-2018/92/lbi/0.pdf}
\caption{\label{fig:model-fold-change-lbi}Observed and predicted frequency fold change for LBI predictor.}
\end{figure}

In contrast with the LBI predictor, the titer substitution model predictor is less likely to predict that clades will grow when they do not (Fig.~\ref{fig:model-fold-change-cTiterSub}).
Combining the LBI and titer substitution predictors together decreases the model's accuracy in predicting clade growth or decline for clades at greater than 50\% initial frequency, but it increases the model's accuracy with clades at less than 50\% frequency (Fig.~\ref{fig:model-fold-change-cTiterSub-lbi}).

\begin{figure}
\includegraphics[width=5in]{../../model_fold_change/2006-2018/92/cTiterSub/0.pdf}
\caption{\label{fig:model-fold-change-cTiterSub}Observed and predicted frequency fold change for titer substitution predictor.}
\end{figure}

\begin{figure}
\includegraphics[width=5in]{../../model_fold_change/2006-2018/92/cTiterSub-lbi/0.pdf}
\caption{\label{fig:model-fold-change-cTiterSub-lbi}Observed and predicted frequency fold change for titer substitution and LBI predictors.}
\end{figure}

\end{document}

### 2018-09-13 analyses/identify-optimal-lbi-parameters/plot_accuracy_by_parameters.py

@@ -1,0 +1,45 @@
+"""Plot model accuracies by LBI parameters.
+"""
+import argparse
+import matplotlib.pyplot as plt
+import pandas as pd
+import seaborn as sns
+
+# Configure matplotlib theme.
+fontsize = 14
+matplotlib_params = {
+    'axes.labelsize': fontsize,
+    'font.size': fontsize,
+    'legend.fontsize': 12,
+    'xtick.labelsize': fontsize,
+    'ytick.labelsize': fontsize,
+    'text.usetex': False,
+    'figure.figsize': [8, 6],
+    'savefig.dpi': 300,
+    'figure.dpi': 300,
+    'text.usetex': False
+}
+
+plt.rcParams.update(matplotlib_params)
+
+# Turn off spines for all plots.
+plt.rc("axes.spines", top=False, right=False)
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("results", help="tab-delimited model results for all LBI parameters")
+    parser.add_argument("correlation", help="plot of correlations by LBI parameters")
+    parser.add_argument("mcc", help="plot of MCCs by LBI parameters")
+
+    args = parser.parse_args()
+
+    results_df = pd.read_table(args.results)
+
+    sns.lmplot(x="tau", y="correlation", hue="time_window", fit_reg=False, data=results_df, aspect=1.5)
+    plt.savefig(args.correlation)
+
+    sns.lmplot(x="tau", y="mcc", hue="time_window", fit_reg=False, data=results_df, aspect=1.5)
+    plt.savefig(args.mcc)
+
+    print(results_df.to_string())

### 2018-09-13 analyses/identify-optimal-lbi-parameters/sweep_parameters.py

@@ -1,0 +1,152 @@
+"""
+Identify optimal LBI parameters
+
+Perform a parameter sweep across a range of values for both the tau and time window parameters, fitting a fitness model for each parameter set, and identify the set that optimizes the fitness model's frequency correlation metric.
+"""
+import argparse
+
+# Configure environment
+
+import os
+import sys
+
+# Project path
+project_path = os.path.dirname(os.path.dirname(os.path.dirname(os.path.abspath(__file__))))
+
+# Path to top-level of the augur source code repository.
+augur_path = os.path.join(project_path, "dist", "augur")
+print(augur_path)
+sys.path.append(augur_path)
+
+import Bio
+import Bio.Phylo
+from collections import defaultdict
+from datetime import datetime, timedelta
+import itertools
+import json
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+from matplotlib import gridspec
+from matplotlib.collections import LineCollection, PatchCollection
+from matplotlib.dates import YearLocator, MonthLocator, DateFormatter
+import matplotlib.patches as mpatches
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+from base.io_util import json_to_tree
+from base.frequencies import KdeFrequencies
+from base.fitness_model import fitness_model as FitnessModel, get_matthews_correlation_coefficient_for_data_frame
+from base.titer_model import TiterCollection
+
+# Configure matplotlib theme.
+fontsize = 14
+matplotlib_params = {
+    'axes.labelsize': fontsize,
+    'font.size': fontsize,
+    'legend.fontsize': 12,
+    'xtick.labelsize': fontsize,
+    'ytick.labelsize': fontsize,
+    'text.usetex': False,
+    'figure.figsize': [8, 6],
+    'savefig.dpi': 300,
+    'figure.dpi': 300,
+    'text.usetex': False
+}
+
+plt.rcParams.update(matplotlib_params)
+
+# Turn off spines for all plots.
+plt.rc("axes.spines", top=False, right=False)
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("tree", help="auspice JSON tree")
+    parser.add_argument("frequencies", help="JSON with frequencies estimated for the given tree")
+    parser.add_argument("results", help="tab-delimited model results for all LBI parameters")
+
+    args = parser.parse_args()
+
+    # Load tree
+    with open(args.tree, "r") as fh:
+        json_tree = json.load(fh)
+
[…]

### 2018-09-13 scripts/plot_model_fold_change.py
Fix augur path for model fold change plotting script

@@ -10,7 +10,8 @@ import seaborn as sns
 import sys
 
 # augur imports.
-augur_path = os.path.join(os.path.dirname(os.path.abspath(__file__)), "dist", "augur")
+project_path = os.path.dirname(os.path.dirname(os.path.abspath(__file__)))
+augur_path = os.path.join(project_path, "dist", "augur")
 sys.path.append(augur_path)
 from base.fitness_model import get_matthews_correlation_coefficient_for_data_frame
 

### 2018-09-13 analyses/identify-optimal-lbi-parameters/results.tex
Added fitness model results including relative fitnesses and frequency projections

@@
 
-\subsection{Parameter sweep}
-
 To determine the optimal tau and window parameters for the LBI fitness predictor, I ran a parameter sweep for all possible combinations of tau and window values across a range of both parameters.
@@
 \begin{figure}[h]
-\includegraphics[width=5in]{mcc_by_lbi_parameters.pdf}
-\caption{Matthew's correlation coefficient (MCC) by LBI parameters}
-\label{mcc-by-lbi-parameters}
+\includegraphics[width=5in]{frequency_correlation_by_lbi_parameters.pdf}
+\caption{\label{fig:frequency-correlation-by-lbi-parameters}Correlation of predicted and observed frequency fold change by LBI parameters}
 \end{figure}
 
-The model performed best by MCC (0.35) with tau of 0.25 and a window of 0.1 or 0.25.
-The model performed best by correlation (0.20) with tau of 1.0 and window of 0.1.
+\begin{figure}[h]
+\includegraphics[width=5in]{mcc_by_lbi_parameters.pdf}
+\caption{\label{fig:mcc-by-lbi-parameters}Matthew's correlation coefficient (MCC) by LBI parameters}
+\end{figure}
+
+The model performed best by correlation (0.20) with tau of 1.0 and window of 0.1 (Fig.~\ref{fig:frequency-correlation-by-lbi-parameters}).
+The model performed best by MCC (0.35) with tau of 0.25 and a window of 0.1 or 0.25 (Fig.~\ref{fig:mcc-by-lbi-parameters}).
 Importantly, the current LBI parameters for 12 year trees (tau of 0.75 and window of 0.75) performed poorly by both correlation (0.06) and MCC (0.02).
@@
 For tau greater or equal to 0.125, the window sizes of 0.1 and 0.25 were almost equally accurate.
+Since LBI is not expected to perform well in predicting specific clade frequencies but rather in predicting clade growth or decline, I opted to select the LBI parameters that maximized the MCC (tau of 0.25 and window of 0.1).
 
-\subsection{Model performance}
+To understand how these new LBI parameters change the annotations of recent clades, I ran LBI with both the standard settings (tau of 0.75, window of 0.75) and the new optimal settings on a tree with 92 viruses per month from 2016--2018 and visualized the results in auspice.
+The standard LBI parameters clearly highlight the rapid emergence of the A2/re clade over the last year (Fig.~\ref{auspice-lbi-with-standard-parameters}).
+In contrast, the optimal LBI parameters for the fitness model identify recent rapid branching in one small subclade of A1b/135K (Fig.~\ref{auspice-lbi-with-optimal-parameters}).
 
-Since LBI is not expected to perform well in predicting specific clade frequencies but rather in predicting clade growth or decline, I opted to select the LBI parameters that maximized the MCC (tau of 0.25 and window of 0.1).
-I ran the forecasting builds with all predictors and these LBI parameters.
+\begin{figure}[h]
+\includegraphics[width=5in]{auspice-lbi-tau-0-75-window-0-75.png}
+\caption{\label{auspice-lbi-with-standard-parameters}Auspice view of recent clades with standard LBI parameters (tau = 0.75, window = 0.75)}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=5in]{auspice-lbi-tau-0-25-window-0-1.png}
+\caption{\label{auspice-lbi-with-optimal-parameters}Auspice view of recent clades with optimal LBI parameters (tau = 0.25, window = 0.1)}
+\end{figure}
+
+After standardization of LBI values across all tips in the most recent timepoint of the fitness model, the A2/re clade has much lower LBI values than the A1b/135K clade (Fig.~\ref{fitness-tree-with-lbi-predictor}).
+
+\begin{figure}[h]
+\includegraphics[width=5in]{h3n2-ha-fitness-phylogeny-lbi_2018-67.png}
+\caption{\label{fitness-tree-with-lbi-predictor}BALTIC view of recent clades with LBI calculated from new optimal parameters and standardized across all tips in the current timepoint.}
+\end{figure}
+
+\section{Model performance}
+
+I ran the forecasting builds with all predictors and the new optimal LBI parameters.
 In addition to the standard ``all'' predictor that includes all individual predictors, I added a predictor based on the titer substitution model and LBI and another based on titers, LBI, and non-epitope mutations.
@@
 
-By itself, the LBI predictor has a small, positive $\beta$ value of 0.08 (Fig.~\ref{fig:model-parameters}) and a relatively low correlation of 0.13 (Fig.~\ref{fig:model-frequency-correlations}).
+By itself, the LBI predictor has a small, positive $\beta$ value of 0.15 (Fig.~\ref{fig:model-parameters}) and a relatively low correlation of 0.16 (Fig.~\ref{fig:model-frequency-correlations}).
 However, LBI also has the second highest MCC (0.35) after the titer substitution model indicating that it generally does a better job of predicting clade growth or decline (Fig.~\ref{fig:model-mcc}).
-In combination with other predictors, LBI has a larger, negative $\beta$ with the largest value of -0.33 in the combined titers and LBI model.
+In combination with other predictors, LBI has a larger, negative $\beta$ with the largest value of -0.34 in the combined titers and LBI model.
 Despite this negative $\beta$ parameter for LBI, its inclusion with the titer substitution model produces a better MCC than either individual predictor.
@@
 
+\begin{figure}
+\includegraphics[width=5in]{../../model_fold_change/2006-2018/92/ne-cTiterSub-lbi/0.pdf}
+\caption{\label{fig:model-fold-change-ne-cTiterSub-lbi}Observed and predicted frequency fold change for non-epitope mutations, titer substitution, and LBI predictors.}
+\end{figure}
+
+\section{Forecasts with LBI predictors}
+
+Using the optimal LBI parameters for the fitness model, the clade A1b/135K has higher standardized LBI values than A2/re (Fig.~\ref{fig:relative-clade-fitnesses}).
+A1b/135K has higher standard values for all three metrics from the best fit model (non-epitopes, titer substitutions, and LBI).
+As a result, this clade also has a higher overall fitness and is projected to increase in frequency by the best fit model (Fig.~\ref{fig:projected-global-clade-frequencies}).
+The only models that do not project a decline in A2/re are the LBI and sequence-only models which project a slight decline or no change over the next year.
+
+\begin{figure}
+\includegraphics[width=5in]{h3n2-ha-relative-clade-fitnesses.png}
+\caption{\label{fig:relative-clade-fitnesses}Standardized fitness metrics by clade and overall fitness based on model parameters for the non-epitope, titer substitution, and LBI model.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=5in]{h3n2-ha-projected-global-frequencies-by-clade.png}
+\caption{\label{fig:projected-global-clade-frequencies}Projected global clade frequencies.}
+\end{figure}
+
 \end{document}

### 2019-06-07 src/fit_model.py
Don't add L1 penalty to final error calculated by the models

@@ -195,7 +195,7 @@ class ExponentialGrowthModel(object):
 
         return projected_frequencies
 
-    def _fit(self, coefficients, X, y):
+    def _fit(self, coefficients, X, y, use_l1_penalty=True):
         """Calculate the error between observed and estimated values for the given
         parameters and data.
 
@@ -250,7 +250,11 @@ class ExponentialGrowthModel(object):
             frequencies["frequency_estimated"],
             initial=frequencies["frequency"]
         )
-        l1_penalty = self.l1_lambda * np.abs(coefficients).sum()
+
+        if use_l1_penalty:
+            l1_penalty = self.l1_lambda * np.abs(coefficients).sum()
+        else:
+            l1_penalty = 0.0
 
         return error + l1_penalty
 
@@ -381,7 +385,7 @@ class ExponentialGrowthModel(object):
         float :
             model error
         """
-        return self._fit(self.coef_, X, y)
+        return self._fit(self.coef_, X, y, use_l1_penalty=False)
 
 
 class DistanceExponentialGrowthModel(ExponentialGrowthModel):
@@ -389,7 +393,7 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
         super().__init__(predictors, delta_time, l1_lambda, cost_function)
         self.distances = distances
 
-    def _fit(self, coefficients, X, y):
+    def _fit(self, coefficients, X, y, use_l1_penalty=True):
         """Calculate the error between observed and estimated values for the given
         parameters and data.
 
@@ -457,7 +461,12 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
         #     y_diff=targets["y_diff"],
 
         # )
-        l1_penalty = self.l1_lambda * np.abs(coefficients).sum()
+        #error = error / float(count)
+
+        if use_l1_penalty:
+            l1_penalty = self.l1_lambda * np.abs(coefficients).sum()
+        else:
+            l1_penalty = 0.0
 
         return error + l1_penalty
 

### 2019-06-10 manuscript/manuscript.tex

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.png,.png,.jpg}

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}
\usepackage{microtype}
\usepackage{url}
\usepackage{hyperref}

% text layout
\usepackage{geometry}
\geometry{textwidth=17cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

\title{Integrative forecasting of seasonal influenza A/H3N2 by genotype and phenotype}
%\title{Experimentally informed forecasts of seasonal influenza A/H3N2}
%\title{Long-term forecasts of seasonal influenza reveal historical contingency of fitness metrics}

\author[1,2]{John Huddleston}
\author[2]{Richard A.\ Neher}
\author[1]{Trevor Bedford}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Moleculary and Cell Biology, University of Washington, Seattle, WA, USA}
\affil[3]{Biozentrum, University of Basel, Basel, Switzerland}

\begin{document}

\maketitle

\section{Abstract}

\section{Introduction}

Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}.
Vaccination remains the most effective public health response available.
However, frequent viral mutation results in viruses that escape previously acquired human immunity.
The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
Current vaccine predictions favor viruses that are distinct from prior vaccine viruses in the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
The hemagglutination inhibition (HI) assay \cite{hirst1943studies} is used to measure the degree of cross-reactivity between pairs of circulating viruses.
HI assays are fundamental for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.

Despite the promise of these sequence-only models, they explicitly omit experimental measurements of antigenic or functional phenotypes.
Recent developments in computational methods and influenza virology have made it feasible to integrate these important metrics of influenza fitness into a single predictive model.
For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \cite{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning experiments \cite{Lee2018}.
We describe an approach to integrate previously disparate sequence-only models of influenza evolution with high-quality experimental measurements of antigenic drift and functional constraint.

\section{Results}

\subsection{Recapitulation of Luksza and Lassig 2014}

We implemented a predictive model inspired by the basic framework of {\L}uksza and L\"assig \cite{Luksza:2014hj}.

\subsection{Forecasts with experimental and phylogenetic fitness metrics}

\subsection{Historical contingency of fitness metrics}

\subsection{Forecasts of other influenza genes and subtypes by A/H3N2 hemagglutinin models}

\subsection{Comparison of chosen and optimal vaccine strains}

\subsection{Forecasts for February 2020}

\section{Discussion}

\section{Methods}

\subsection{Strain selection}

\subsection{Multiple sequence alignment}

\subsection{Phylogenetic interference}

\subsection{Frequency estimation}

\subsection{Fitness model}

\subsection{Model fitting and evaluation}

\subsubsection{Model targets}

\subsubsection{Time-series cross-validation}

\subsection{Fitness metrics}

\subsubsection{Antigenic drift}

\subsubsection{Functional constraint}

\subsubsection{Clade growth}

\clearpage

\bibliographystyle{nih}
\bibliography{manuscript}

\end{document}

### 2019-07-04 scripts/node_data_to_table.py
Add support for true fitness predictor from SANTA-SIM

@@ -13,6 +13,7 @@ if __name__ == '__main__':
         formatter_class=argparse.ArgumentDefaultsHelpFormatter
     )
     parser.add_argument("--tree", required=True, help="Newick file for the tree used to construct the given node data JSONs")
+    parser.add_argument("--metadata", help="file with metadata associated with viral sequences, one for each segment")
     parser.add_argument("--jsons", nargs="+", required=True, help="node data JSON(s) from augur")
     parser.add_argument("--annotations", nargs="+", help="additional annotations to add to the output table in the format of 'key=value' pairs")
     parser.add_argument("--excluded-fields", nargs="+", help="names of columns to omit from output table")
@@ -23,6 +24,9 @@ if __name__ == '__main__':
     # Load tree.
     tree = Bio.Phylo.read(args.tree, "newick")
 
+    # Load metadata for samples.
+    metadata = pd.read_csv(args.metadata, sep="\t")
+
     # Load one or more node data JSONs into a single dictionary indexed by node name.
     node_data = read_node_data(args.jsons)
 
@@ -30,6 +34,9 @@ if __name__ == '__main__':
     # Data are initially loaded with one column per node.
     # Transposition converts the table to the expected one row per node format.
     df = pd.DataFrame(node_data["nodes"]).T.rename_axis("strain").reset_index()
+
+    # Annotate node data with per sample metadata.
+    df = df.merge(metadata, on="strain", suffixes=["", "_metadata"])
 
     # Remove excluded fields if they are in the data frame.
     df = df.drop(columns=[field for field in args.excluded_fields if field in df.columns])

### 2019-07-04 src/fit_model.py
Flush debugging output to stdout when fitting models

@@ -641,7 +641,8 @@ def cross_validate(model, data, targets, train_validate_timepoints, coefficients
                 validation_error,
                 null_validation_error,
                 model.coef_[0]
-            )
+            ),
+            flush=True
         )
 
         # Get the estimated frequencies for training and validation sets to export.

### 2019-07-05 manuscript/main.tex
Update manuscript structure and organization
The new organization reflects the plan for the manuscript outline and the
structure more closely aligns this paper with others from the lab.

--- 2019-06-10 manuscript/manuscript.tex
@@
-\documentclass[11pt,oneside,letterpaper]{article}
-
-% graphicx package, useful for including eps and pdf graphics
-\usepackage{graphicx}
-\DeclareGraphicsExtensions{.png,.png,.jpg}
-
-% basic packages
-\usepackage{color}
-\usepackage{parskip}
-\usepackage{float}
-\usepackage{microtype}
-\usepackage{url}
-\usepackage{hyperref}
-
-% text layout
-\usepackage{geometry}
-\geometry{textwidth=17cm} % 15.25cm for single-space, 16.25cm for double-space
-\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space
-
-% helps to keep figures from being orphaned on a page by themselves
-\renewcommand{\topfraction}{0.85}
-\renewcommand{\textfraction}{0.1}
-
-% bold the 'Figure #' in the caption and separate it with a period
-% Captions will be left justified
-\usepackage[labelfont=bf,labelsep=period,font=small]{caption}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-\usepackage{authblk}
-\renewcommand\Authands{ \& }
-\renewcommand\Authfont{\normalsize \bf}
-\renewcommand\Affilfont{\small \normalfont}
-\makeatletter
-\renewcommand\AB@affilsepx{, \protect\Affilfont}
-\makeatother
-
-\title{Integrative forecasting of seasonal influenza A/H3N2 by genotype and phenotype}
-%\title{Experimentally informed forecasts of seasonal influenza A/H3N2}
-%\title{Long-term forecasts of seasonal influenza reveal historical contingency of fitness metrics}
-
-\author[1,2]{John Huddleston}
-\author[2]{Richard A.\ Neher}
-\author[1]{Trevor Bedford}
-
-\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
-\affil[2]{Moleculary and Cell Biology, University of Washington, Seattle, WA, USA}
-\affil[3]{Biozentrum, University of Basel, Basel, Switzerland}
-
-\begin{document}
-
-\maketitle
-
-\section{Abstract}
-
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-\section{Results}
+\textit{Brief summary of model implementation here.}
 
-\subsection{Recapitulation of Luksza and Lassig 2014}
+\section*{Results}
 
-We implemented a predictive model inspired by the basic framework of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+\subsection*{Models accurately forecast evolution of simulated populations of A/H3N2-like viruses}
 
-\subsection{Forecasts with experimental and phylogenetic fitness metrics}
+To test our model framework, we simulated the long-term evolution of five independent populations of A/H3N2-like viruses for 5,000 generations each using SANTA-SIM \cite{Jariani2019}.
+We seeded each simulated population with the full length HA from A/Beijing/32/1992 such that all simulated sequences contained the signal peptide, HA1, and HA2 domains.
+We defined purifying selection across all three domains, allowing the preferred amino acid at each site to change at a fixed rate over time.
+We additionally defined exposure-dependent selection for 129 putative epitope sites in HA1 \cite{Wolf:2006da} to impose an effect of cross-immunity that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
+These selective constraints produce phylogenetic structures and accumulation of epitope and non-epitope mutations that are consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
 
-\subsection{Historical contingency of fitness metrics}
+Next, we fit models to 20 years of simulated sequences corresponding to generations 1000--3000 where 100 generations represents one year.
+In the absence of any experimental data for these simulated sequences, we only fit models with sequence-based fitness metrics including epitope cross-immunity, non-epitope mutations, LBI, and delta frequency.
 
-\subsection{Forecasts of other influenza genes and subtypes by A/H3N2 hemagglutinin models}
+\textit{Model results here for sequence-based metrics relative to naive model}
 
-\subsection{Comparison of chosen and optimal vaccine strains}
+\textit{Repeat simulations with multiple epochs to show model coefficients can reflect these underlying changes to evolutionary constraints.}
 
-\subsection{Forecasts for February 2020}
+\subsection*{Models based on X metrics most accurately forecast natural populations of A/H3N2 viruses}
 
-\section{Discussion}
+\subsection*{Model accuracy and coefficients reflect historical patterns of A/H3N2 evolution}
 
-\section{Methods}
+\subsection*{Composite models outperform models with individual fitness metrics}
 
-\subsection{Strain selection}
+\subsection*{Models enable selection of vaccine candidate strains}
 
-\subsection{Multiple sequence alignment}
+\subsection*{Forecasts predict the rise of 3c3.A and A1b clades in February 2020}
 
-\subsection{Phylogenetic interference}
+\section*{Discussion}
 
-\subsection{Frequency estimation}
+\section*{Methods}
 
-\subsection{Fitness model}
+\subsection*{Strain selection}
 
-\subsection{Model fitting and evaluation}
+\subsection*{Multiple sequence alignment}
 
-\subsubsection{Model targets}
+\subsection*{Phylogenetic interference}
 
-\subsubsection{Time-series cross-validation}
+\subsection*{Frequency estimation}
 
-\subsection{Fitness metrics}
+\subsection*{Fitness model}
 
-\subsubsection{Antigenic drift}
+\subsection*{Model fitting and evaluation}
 
-\subsubsection{Functional constraint}
+\subsubsection*{Model targets}
 
-\subsubsection{Clade growth}
+\subsubsection*{Time-series cross-validation}
 
-\clearpage
+\subsection*{Fitness metrics}
 
-\bibliographystyle{nih}
-\bibliography{manuscript}
+\subsubsection*{Antigenic drift}
 
-\end{document}
+\subsubsection*{Functional constraint}
+
+\subsubsection*{Clade growth}

### 2019-07-05 manuscript/manuscript.tex
Update manuscript structure and organization
The new organization reflects the plan for the manuscript outline and the
structure more closely aligns this paper with others from the lab.

@@
-\documentclass[11pt,oneside,letterpaper]{article}
+\documentclass[12pt]{article}
 
@@
 
+\begin{abstract}
+\input{abstract.tex}
+\end{abstract}
+
 \maketitle
 
-\section{Abstract}
-
-\section{Introduction}
-
-Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}.
-Vaccination remains the most effective public health response available.
-However, frequent viral mutation results in viruses that escape previously acquired human immunity.
-The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
-Current vaccine predictions favor viruses that are distinct from prior vaccine viruses in the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
-The hemagglutination inhibition (HI) assay \cite{hirst1943studies} is used to measure the degree of cross-reactivity between pairs of circulating viruses.
-HI assays are fundamental for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
-As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.
-
-Despite the promise of these sequence-only models, they explicitly omit experimental measurements of antigenic or functional phenotypes.
-Recent developments in computational methods and influenza virology have made it feasible to integrate these important metrics of influenza fitness into a single predictive model.
-For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \cite{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning experiments \cite{Lee2018}.
-We describe an approach to integrate previously disparate sequence-only models of influenza evolution with high-quality experimental measurements of antigenic drift and functional constraint.
-
-\section{Results}
-
-\subsection{Recapitulation of Luksza and Lassig 2014}
-
-We implemented a predictive model inspired by the basic framework of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
-
-\subsection{Forecasts with experimental and phylogenetic fitness metrics}
-
-\subsection{Historical contingency of fitness metrics}
-
-\subsection{Forecasts of other influenza genes and subtypes by A/H3N2 hemagglutinin models}
-
-\subsection{Comparison of chosen and optimal vaccine strains}
-
-\subsection{Forecasts for February 2020}
-
-\section{Discussion}
-
-\section{Methods}
-
-\subsection{Strain selection}
-
-\subsection{Multiple sequence alignment}
-
-\subsection{Phylogenetic interference}
-
-\subsection{Frequency estimation}
-
-\subsection{Fitness model}
-
-\subsection{Model fitting and evaluation}
-
-\subsubsection{Model targets}
-
-\subsubsection{Time-series cross-validation}
-
-\subsection{Fitness metrics}
-
-\subsubsection{Antigenic drift}
-
-\subsubsection{Functional constraint}
-
-\subsubsection{Clade growth}
+\input{main.tex}
+\bibliographystyle{nih}
+\bibliography{manuscript}
 
@@
 
-\bibliographystyle{nih}
-\bibliography{manuscript}
+\input{supplement.tex}
 

### 2019-07-05 manuscript/supplement.tex
Update manuscript structure and organization
The new organization reflects the plan for the manuscript outline and the
structure more closely aligns this paper with others from the lab.

\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\Roman{table}}

\section*{Supplemental Material}

\subsection*{Simulations of A/H3N2-like populations}

\begin{figure*}[h]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/simulated-h3n2-ha-phylogeny.png}
  \caption{Phylogeny of simulated A/H3N2-like HA sequences. The phylogenetic structure and rate of accumulated epitope and non-epitope mutations match patterns observed in phylogenies of natural sequences.}
  \label{sup_fig:simulated_h3n2_ha_phylogeny}
  \end{center}
\end{figure*}

### 2019-07-06 analyses/simulations/run.sh
Increase sampling frequency, decrease total samples, export fitness
Uses a modified branch of SANTA-SIM that exports the true fitness of each sample
in the FASTA header of the alignment output. This commit also makes SANTA-SIM
emit samples more often between generations to get the effect of multiple sample
timepoints within a single month. This change should be helpful for Nextstrain
pipelines that are used to working with real (noisier) data.

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-#java -jar ~/projects/santa-sim/santa.jar influenza_h3n2_ha.xml
+java -jar ~/projects/santa-sim/dist/santa.jar influenza_h3n2_ha.xml
 
 #seqtk sample simulated_HA_sequences.fasta 0.05 > simulated_HA_sequences_5pct.fasta
 seqtk sample simulated_HA_sequences.fasta 0.2 > simulated_HA_sequences_20pct.fasta

### 2019-07-06 analyses/simulations/run.sh
Build test trees for simulated populations with 5% of all sequences
We don't need to build very densely sampled trees to determine whether the
resulting structures and other features are correct.

@@ -2,11 +2,11 @@
 
 java -jar ~/projects/santa-sim/dist/santa.jar influenza_h3n2_ha.xml
 
-#seqtk sample simulated_HA_sequences.fasta 0.05 > simulated_HA_sequences_5pct.fasta
-seqtk sample simulated_HA_sequences.fasta 0.2 > simulated_HA_sequences_20pct.fasta
+seqtk sample simulated_HA_sequences.fasta 0.05 > simulated_HA_sequences_5pct.fasta
+#seqtk sample simulated_HA_sequences.fasta 0.2 > simulated_HA_sequences_20pct.fasta
 
-#augur parse --sequences simulated_HA_sequences_5pct.fasta --output-sequences sequences.fasta --output-metadata original_metadata.tsv --fields strain generation
-augur parse --sequences simulated_HA_sequences_20pct.fasta --output-sequences sequences.fasta --output-metadata original_metadata.tsv --fields strain generation
+augur parse --sequences simulated_HA_sequences_5pct.fasta --output-sequences sequences.fasta --output-metadata original_metadata.tsv --fields strain generation fitness
+#augur parse --sequences simulated_HA_sequences_20pct.fasta --output-sequences sequences.fasta --output-metadata original_metadata.tsv --fields strain generation
 
 # Manual python step to add dates to metadata from generations
 # from augur.frequency_estimators import float_to_datestring

### 2019-07-09 manuscript/main.tex
Write up initial simulation results and include figures
The numbers will probably change a little, but the general results will be the
same for now.

@@
 
-\textit{Brief summary of model implementation here.}
+\textit{Brief summary of model implementation here including reference to Fig.~\ref{fig:model}.}
 
@@
 
-\subsection*{Models accurately forecast evolution of simulated populations of A/H3N2-like viruses}
+\subsection*{Models accurately forecast evolution of simulated populations of {A/H3N2-like viruses}}
 
-To test our model framework, we simulated the long-term evolution of five independent populations of A/H3N2-like viruses for 5,000 generations each using SANTA-SIM \cite{Jariani2019}.
-We seeded each simulated population with the full length HA from A/Beijing/32/1992 such that all simulated sequences contained the signal peptide, HA1, and HA2 domains.
+The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions at epitopes that enable escape from adaptive immunity and purifying selection on domains required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we first simulated the long-term evolution of five independent populations of A/H3N2-like viruses for 5,000 generations each using SANTA-SIM \cite{Jariani2019}.
+We seeded each simulated population with the full length HA from A/Beijing/32/1992 such that all simulated sequences contained signal peptide, HA1, and HA2 domains.
 We defined purifying selection across all three domains, allowing the preferred amino acid at each site to change at a fixed rate over time.
 We additionally defined exposure-dependent selection for 129 putative epitope sites in HA1 \cite{Wolf:2006da} to impose an effect of cross-immunity that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
-These selective constraints produce phylogenetic structures and accumulation of epitope and non-epitope mutations that are consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
+These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
 
-Next, we fit models to 20 years of simulated sequences corresponding to generations 1000--3000 where 100 generations represents one year.
-In the absence of any experimental data for these simulated sequences, we only fit models with sequence-based fitness metrics including epitope cross-immunity, non-epitope mutations, LBI, and delta frequency.
+Under the evolutionary constraints of our simulations, we expected the models to learn positive coefficients for cross-immunity and negative coefficients for non-epitope mutations corresponding to the fitness benefits of novel epitope sequences and costs of accumulating deleterious mutations.
+We reasoned that because LBI and delta frequency measure the recent success of samples without any mechanistic explanation for that success, these two models should be assigned positive coefficients and should outperform the individual mutation models by being able to simultaneously represent benefits of epitope mutations and costs of non-epitope mutations with a single metric.
+To achieve the same accuracy from mutation-only models, we anticipated that these two models would need to be combined.
+To test these hypotheses, we trained models for each individual fitness metric on 20 years of simulated sequences corresponding to generations 1000--3000 where 100 generations represents one year.
+As a positive control, we trained a model on the true fitness of each sample as measured by the simulator.
+To control for variability in distances between seasons, we evaluated all models relative to the observed distance between seasons (the adjusted distance between seasons).
+This distance corresponded to a model without an exponential growth parameter which we labeled as the ``naive'' model.
 
-\textit{Model results here for sequence-based metrics relative to naive model}
+The average distance per year between populations averaged 18 +/- 5 amino acids in the naive model (Supplemental Fig.~\ref{sup_fig:weighted_distance_between_timepoints}).
+On average, all models reduced this distance with LBI performing the best by reducing this distance to the future by over four amino acids (Fig.~\ref{fig:model_accuracy_and_coefficients_for_simulated_populations}).
+As expected, the true fitness metric always outperformed the naive model.
+The non-epitope metric also consistently outperformed the naive model, predicting the future nearly as well as the true fitness.
+Surprisingly, cross-immunity did not accurately forecast future populations, receiving instead an average coefficient near zero in most training windows.
+While the delta frequency metric was not as accurate as LBI, it was considerably less variable with an average adjusted distance similar to the non-epitope metric.
+Interestingly, the coefficients for non-epitope mutations, LBI, and delta frequency were stably non-zero across all training windows with standard deviations ranging from 0.16 to 0.42.
+In contrast, the true fitness metric's coefficients gradually varied more over time with a standard deviation of 0.62.
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/model-accuracy-and-coefficients-for-simulated-populations.png}
+  \caption{Model a) accuracy and b) coefficients for simulated populations of A/H3N2-like viruses.}
+  \label{fig:model_accuracy_and_coefficients_for_simulated_populations}
+  \end{center}
+\end{figure*}
+
+We tested our hypothesis that composite models of cross-immunity and non-epitope mutations would perform as well as LBI or delta frequency.
+We further tested the relationship between LBI and these two individual metrics by fitting composite models with all combinations of these three predictors.
+Surprisingly, cross-immunity provided no additional accuracy to the non-epitope metric, while neither of these mutation-based metrics provided any improvement over LBI (Supplemental Fig.~\ref{sup_fig:composite_model_accuracy_and_coefficients_for_simulated_populations}).
+In the latter models, LBI maintained a consistent coefficient across all training windows while the models assigned both cross-immunity and non-epitope metrics coefficients near zero.
+These results suggest that non-epitope mutations and LBI are mutually exclusive and either metric is sufficient to forecast the evolution of these simulated populations.
 

### 2019-07-09 manuscript/supplement.tex
Write up initial simulation results and include figures
The numbers will probably change a little, but the general results will be the
same for now.

@@
   \caption{Phylogeny of simulated A/H3N2-like HA sequences.
-The phylogenetic structure and rate of accumulated epitope and non-epitope mutations match patterns observed in phylogenies of natural sequences.}
+The phylogenetic structure and rate of accumulated epitope and non-epitope mutations match patterns observed in phylogenies of natural sequences.
+Sample dates were annotated as the generation in the simulation divided by 100 and added to 2000, to acquire realistic date ranges that were compatible with our modeling machinery.}
   \label{sup_fig:simulated_h3n2_ha_phylogeny}
@@
 \end{figure*}
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/composite-model-accuracy-and-coefficients-for-simulated-populations.png}
+  \caption{Composite model a) accuracy and b) coefficients for simulated populations of A/H3N2-like viruses.}
+  \label{sup_fig:composite_model_accuracy_and_coefficients_for_simulated_populations}
+  \end{center}
+\end{figure*}

### 2019-07-09 src/fit_model.py
Corrected docstring

@@ -76,8 +76,8 @@ class ExponentialGrowthModel(object):
         X : pandas.DataFrame
             standardized tip attributes by timepoint
 
-        predictors : ndarray
-            predictor values per sample (n x p matrix for p predictors and n samples)
+        predictors : list
+            names of predictors for which mean standard deviations should be calculated
 
         Returns
         -------

### 2019-07-09 src/fit_model.py
Fit models with average weighted distance to the future
This approach turns out to work about as well as EMD, it runs a lot faster, and
it is a lot easier to understand and explain to others.

@@ -393,7 +393,7 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
         super().__init__(predictors, delta_time, l1_lambda, cost_function)
         self.distances = distances
 
-    def _fit(self, coefficients, X, y, use_l1_penalty=True):
+    def _fit_emd(self, coefficients, X, y, use_l1_penalty=True):
         """Calculate the error between observed and estimated values for the given
         parameters and data.
 
@@ -439,6 +439,8 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
                 self.distances
             ).astype(np.float32)
 
+            # Estimate the distance between the model's estimated future and the
+            # observed future populations.
             model_emd, _, model_flow = cv2.EMD(
                 sample_a_frequencies,
                 sample_b_frequencies,
@@ -446,6 +448,8 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
                 cost=distance_matrix
             )
 
+            # Estimate the distance between the current and future populations
+            # (the naive model of no frequency changes).
             null_emd, _, null_flow = cv2.EMD(
                 sample_a_initial_frequencies,
                 sample_b_frequencies,
@@ -453,11 +457,13 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
                 cost=distance_matrix
             )
 
-            # print("EMD for %s: %.2f with %i of %i nonzero flow cells" % (timepoint, model_emd, (model_flow != 0).sum(), model_flow.flatten().shape[0]), file=sys.stderr)
-            # print("Null EMD for %s: %.2f with %i of %i nonzero flow cells" % (timepoint, null_emd, (null_flow != 0).sum(), null_flow.flatten().shape[0]), file=sys.stderr)
+            #print("EMD for %s: %.2f with %i of %i nonzero flow cells" % (timepoint, model_emd, (model_flow != 0).sum(), model_flow.flatten().shape[0]), file=sys.stderr)
+            #print("Null EMD for %s: %.2f with %i of %i nonzero flow cells\n" % (timepoint, null_emd, (null_flow != 0).sum(), null_flow.flatten().shape[0]), file=sys.stderr)
+
             # print(model_flow[model_flow != 0].flatten())
             # print(null_flow[null_flow != 0].flatten())
-            error += (model_emd)
+            #error += ((model_emd - null_emd) / null_emd) * 100
+            error += model_emd
             count += 1
 
         # # Merge estimated and observed target values.
[…]

### 2019-07-16 manuscript/main.tex
Start writing methods

@@
 
-\subsection*{Strain selection}
+\subsection*{Simulation of influenza A/H3N2-like populations}
 
-\subsection*{Multiple sequence alignment}
+\subsection*{Strain selection for natural populations}
+
+For model training and validation, we selected XX HA sequences of length >=1,700 nucleotides that were sampled between October 1994 and October 2015.
+To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 10 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
+We excluded all samples with ambiguous year, month, and day annotations and prioritized samples with more available HI titer measurements.
+
+For model testing, we extended the set of training and validation sequences to include HA sequences that were sampled between October 2015 and April 2019.
+We used these test sequences to evaluate the out-of-sample error of fixed model parameters learned during training and validation.
 
@@
 
+For each timepoint in model training, validation, and testing, we selected the subsampled HA sequences with collection dates up to that timepoint.
+We aligned sequences with the augur align command \cite{Hadfield2018} and MAFFT v7.407 \cite{Katoh2002}.
+We inferred initial phylogenies for HA sequences at each timepoint with IQ-TREE v1.6.10 \cite{Nguyen2014}.
+To reconstruct time-resolved phylogenies, we applied TreeTime v0.5.6 \cite{Sagulenko2018} with the augur refine command.
+
 \subsection*{Frequency estimation}
 
-\subsection*{Fitness model}
+To account for uncertainty in collection date and sampling error, we applied a kernel density estimation (KDE) approach to calculate global sample frequencies.
+Specifically, we constructed a Gaussian kernel for each sample with the mean at the reported collection date and a variance of one month, roughly corresponding to the average lag time between sample collection and submission to the GISAID database.
+We estimated the frequency of each sample at each timepoint by calculating the cumulative density function of each KDE between the current and previous timepoint and normalizing the resulting values to sum to one.
+We implemented this logic in the augur frequencies command.
 
@@
 
-\subsubsection*{Model targets}
+As in \cite{Luksza:2014hj}, we assumed that influenza A/H3N2 populations can be described by an exponential growth model.
+Under this model, we estimated the future frequency of the global population, $\hat{X}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
+
+$$
+\hat{X}(t + \Delta{t}) = \frac{1}{Z(t)}\sum_{i}x_{i}(t)\exp(f_{i}(t))
+$$
+
+We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
+For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the mean weighted Hamming distance between the estimated and observed future populations as defined by,
+
+$$
+\Delta(X(t + \Delta{t}), \hat{X}(t + \Delta{t})) = \frac{(\sum_{i}x_{i}\exp(f_{i}(t))(\sum_{j}x_{j}d_{i,j})) - (\sum_{j}x_{j}(\sum_{j'}x_{j'}d_{j,j'}))}{(\sum_{i}x_{i}(\sum_{j}x_{j}d_{i,j}))},
+$$
+
+where the first term in the numerator measures the mean distance between estimated and observed populations, the second term accounts for the diversity of the future population, and the denominator accounts for the diversity between the current and future populations.
+When the sequence diversity of the estimated and observed future populations are equal, the metric evaluates to zero.
+When the estimated sequence diversity is closer to the center of the future population's observed diversity, the metric evaluates to a negative proportion.
+We applied the Nelder-Mead minimization algorithm as implemented in SciPy \cite{SciPy} to learn fitness coefficients that minimize the average of this distance metric over all timepoints in a given training window.
 

### 2019-07-16 manuscript/main.tex
Describe time-series cross-validation

@@
 
+\subsubsection*{Fitness model}
+
 As in \cite{Luksza:2014hj}, we assumed that influenza A/H3N2 populations can be described by an exponential growth model.
@@
 We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
+
+\subsubsection*{Model target}
+
 For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the mean weighted Hamming distance between the estimated and observed future populations as defined by,
@@
 
+To obtain unbiased estimates for the out-of-sample errors of our models, we adopted the standard cross-validation strategy of training, validation, and testing.
+We divided our available data into an initial training and validation set spanning October 1994 to October 2015 and an additional testing set spanning October 2015 to April 2019.
+We partitioned our training and validation data into six month seasons corresponding to winter in the Northern Hemisphere (October--April) and the Southern Hemisphere (April--October) and trained models to estimate frequencies of populations one year into the future from each season in six-year sliding windows.
+To calculate validation error for each training window, we applied the resulting model coefficients to estimate the future frequencies for the year after the last timepoint in the training window.
+These validation errors informed our tuning of hyperparameters including a L1 regularization of the fitness coefficients, the LBI neighborhood parameter $\tau$, and the length of the training window itself.
+Finally, we fixed the coefficients for each model at the mean values across all training windows and applied these fixed models to the test data to estimate the true forecasting accuracy of each model on previously unobserved data.
+
 \subsection*{Fitness metrics}

### 2019-07-16 manuscript/main.tex
Update figures for simulation results, add figures for natural populations
Includes rough text describing results.

@@
 We describe an approach to integrate previously disparate sequence-only models of influenza evolution with high-quality experimental measurements of antigenic drift and functional constraint.
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/model.png}
+  \caption{Schematic representation of the fitness model.}
+  \label{fig:model}
+  \end{center}
+\end{figure*}
 
@@
 
-The average distance per year between populations averaged 18 +/- 5 amino acids in the naive model (Supplemental Fig.~\ref{sup_fig:weighted_distance_between_timepoints}).
-On average, all models reduced this distance with LBI performing the best by reducing this distance to the future by over four amino acids (Fig.~\ref{fig:model_accuracy_and_coefficients_for_simulated_populations}).
+The average distance per year between populations under the naive model of no frequency change was 0.10 +/- 0.17 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+The naive model accurately approximated the future in five out of 23 validation timepoints (22\%).
+All biologically-informed models consistently reduced this distance with estimates by LBI accurately estimating the future in 22 of 23 timepoints (96\%) with an average distance of -0.16 +/- 0.16 (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
 As expected, the true fitness metric always outperformed the naive model.
 The non-epitope metric also consistently outperformed the naive model, predicting the future nearly as well as the true fitness.
-Surprisingly, cross-immunity did not accurately forecast future populations, receiving instead an average coefficient near zero in most training windows.
-While the delta frequency metric was not as accurate as LBI, it was considerably less variable with an average adjusted distance similar to the non-epitope metric.
-Interestingly, the coefficients for non-epitope mutations, LBI, and delta frequency were stably non-zero across all training windows with standard deviations ranging from 0.16 to 0.42.
-In contrast, the true fitness metric's coefficients gradually varied more over time with a standard deviation of 0.62.
+Surprisingly, cross-immunity did not accurately forecast future populations and was consistently assigned a weakly negative coefficient.
+While the delta frequency metric was not as accurate as LBI, it was consistently more accurate than the naive model with an average distance similar to the non-epitope metric.
+The coefficients for all metrics except for the true fitness were stably non-zero across all training windows.
+In contrast, the true fitness metric's coefficients gradually decreased over time.
 
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/model-accuracy-and-coefficients-for-simulated-populations.png}
+  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
   \caption{Model a) accuracy and b) coefficients for simulated populations of A/H3N2-like viruses.}
-  \label{fig:model_accuracy_and_coefficients_for_simulated_populations}
+  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
   \end{center}
@@
 We further tested the relationship between LBI and these two individual metrics by fitting composite models with all combinations of these three predictors.
-Surprisingly, cross-immunity provided no additional accuracy to the non-epitope metric, while neither of these mutation-based metrics provided any improvement over LBI (Supplemental Fig.~\ref{sup_fig:composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Surprisingly, cross-immunity provided no additional accuracy to the non-epitope metric, while neither of these mutation-based metrics provided any improvement over LBI (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 In the latter models, LBI maintained a consistent coefficient across all training windows while the models assigned both cross-immunity and non-epitope metrics coefficients near zero.
@@
 
-\subsection*{Models based on X metrics most accurately forecast natural populations of A/H3N2 viruses}
+\subsection*{Phylogenetic metrics most accurately forecast natural populations of A/H3N2 viruses}
+
+The average distance per year between natural populations was twice that of simulated populations at 0.20 +/- 0.16 (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
+Consistent with the simulated population results, almost all biologically-informed metrics performed better than the naive model for natural populations with the exception of the cross-immunity metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+The two phylogenetic metrics, LBI and delta frequency, performed consistently better than all other metrics with average distances to the future of -0.06 +/- 0.20 and 0.00 +/- 0.12, respectively.
+Importantly, HI phenotypes produced more accurate estimates of the future than the other antigenic metric of cross-immunity.
+Variability in the coefficient for the HI phenotypes metric was consistent with relatively sparse HI titer measurements prior to 2011.
+In contrast, the sequence-based metric for functional constraint, non-epitope mutations, outperformed the DMS-based phenotypic metric.
+The gradual decline of the DMS metric's coefficient around 2009 is consistent with overfitting models to the original background strain of the DMS experiments, A/Perth/16/2009.
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
+  \caption{Model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.}
+  \label{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}
+  \end{center}
+\end{figure*}
+
+The results suggested that combinations of the best individual metrics like LBI, delta frequency, non-epitope mutations, and HI phenotypes could produce even more accurate models.
+To this end, we fit models for X combinations of these metrics and evaluated their distance to the future and coefficients relative to the corresponding individual metrics.
+We found results (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
 

### 2019-07-16 manuscript/build.sh
Add build script for manuscript

@@ -1,0 +1,16 @@
+TITLE=manuscript
+
+rm -f *.log
+rm -f *.aux
+rm -f *.out
+rm -f *.bbl
+rm -f *.bcf
+rm -f *.blg
+rm -f .DS_Store
+rm -f ${TITLE}.pdf
+
+pdflatex ${TITLE}.tex
+bibtex ${TITLE}
+pdflatex ${TITLE}.tex
+pdflatex ${TITLE}.tex
+pdflatex ${TITLE}.tex

### 2019-07-16 src/fit_model.py
Modify model target to be scaled distance to the future as in the LBI paper

@@ -450,12 +450,12 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
 
             # Estimate the distance between the current and future populations
             # (the naive model of no frequency changes).
-            null_emd, _, null_flow = cv2.EMD(
-                sample_a_initial_frequencies,
-                sample_b_frequencies,
-                cv2.DIST_USER,
-                cost=distance_matrix
-            )
+            # null_emd, _, null_flow = cv2.EMD(
+            #     sample_a_initial_frequencies,
+            #     sample_b_frequencies,
+            #     cv2.DIST_USER,
+            #     cost=distance_matrix
+            # )
 
             #print("EMD for %s: %.2f with %i of %i nonzero flow cells" % (timepoint, model_emd, (model_flow != 0).sum(), model_flow.flatten().shape[0]), file=sys.stderr)
             #print("Null EMD for %s: %.2f with %i of %i nonzero flow cells\n" % (timepoint, null_emd, (null_flow != 0).sum(), null_flow.flatten().shape[0]), file=sys.stderr)
@@ -522,17 +522,25 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
         count = 0
         for timepoint, timepoint_df in y_hat.groupby("timepoint"):
             samples_a = timepoint_df["strain"]
-            sample_a_initial_frequencies = timepoint_df["frequency"].values.astype(np.float32)
-            sample_a_frequencies = timepoint_df["projected_frequency"].values.astype(np.float32)
+            sample_a_initial_frequencies = timepoint_df["frequency"].values
+            sample_a_frequencies = timepoint_df["projected_frequency"].values
             sample_a_weighted_distance_to_future = timepoint_df["weighted_distance_to_future"].values
+
+            future_timepoint_df = y[y["timepoint"] == timepoint]
+            assert future_timepoint_df.shape[0] > 0
+
+            samples_b = future_timepoint_df["strain"]
+            sample_b_frequencies = future_timepoint_df["frequency"].values
+            sample_b_weighted_distance_to_present = future_timepoint_df["weighted_distance_to_present"].values
 
             d_t_u = (sample_a_initial_frequencies * sample_a_weighted_distance_to_future).sum()
             d_u_hat_u = (sample_a_frequencies * sample_a_weighted_distance_to_future).sum()
+            d_u_u = (sample_b_frequencies * sample_b_weighted_distance_to_present).sum()
 
-            error += d_u_hat_u
[…]

### 2019-07-16 src/fit_model.py
Do not include raw tip attributes in model output by default to save space

@@ -645,7 +645,8 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
         return estimated_targets
 
 
-def cross_validate(model, data, targets, train_validate_timepoints, coefficients=None, group_by="clade_membership"):
+def cross_validate(model, data, targets, train_validate_timepoints, coefficients=None, group_by="clade_membership",
+                   include_attributes=False):
     """Calculate cross-validation scores for the given data and targets across the
     given train/validate timepoints.
 
@@ -670,6 +671,15 @@ def cross_validate(model, data, targets, train_validate_timepoints, coefficients
         an optional array of fixed coefficients for the given model's predictors
         to use when calculating cross-validation error for specific models
         (e.g., naive forecasts)
+
+    group_by : string
+        column of the tip attributes by which they should be grouped to
+        calculate the total number of samples in the model (e.g., group by clade
+        or strain)
+
+    include_attributes : boolean
+        specifies whether tip attribute data used to train/validate models
+        should be included in the output per training window
 
     Returns
     -------
@@ -731,10 +741,9 @@ def cross_validate(model, data, targets, train_validate_timepoints, coefficients
                     df[column] = df[column].dt.strftime("%Y-%m-%d")
 
         # Store training results, beta coefficients, and validation results.
-        results.append({
+        result = {
             "predictors": model.predictors,
             "training_data": {
-                "X": training_X.to_dict(orient="records"),
                 "y": training_y.to_dict(orient="records"),
                 "y_hat": training_y_hat.to_dict(orient="records")
             },
@@ -743,7 +752,6 @@ def cross_validate(model, data, targets, train_validate_timepoints, coefficients
             "coefficients": model.coef_.tolist(),
             "mean_stds": model.mean_stds_.tolist(),
             "validation_data": {
-                "X": validation_X.to_dict(orient="records"),
                 "y": validation_y.to_dict(orient="records"),
                 "y_hat": validation_y_hat.to_dict(orient="records")
             },
@@ -751,7 +759,14 @@ def cross_validate(model, data, targets, train_validate_timepoints, coefficients
             "validation_error": validation_error,
             "last_training_timepoint": training_timepoints[-1].strftime("%Y-%m-%d"),
             "validation_timepoint": validation_timepoint.strftime("%Y-%m-%d")
-        })
+        }
+
+        # Include tip attributes, if requested.
+        if include_attributes:
[…]

### 2019-07-16 src/fit_model.py
Report debugging output with higher precision floats to support new proportion-based distance metric

@@ -716,9 +716,8 @@ def cross_validate(model, data, targets, train_validate_timepoints, coefficients
         # Calculate the model score for the validation data.
         validation_error = model.score(validation_X, validation_y)
         null_validation_error = model._fit(np.zeros_like(model.coef_), validation_X, validation_y)
-
         print(
-            "%s\t%s\t%.1f\t%.1f\t%.1f\t%.1f\t%.2f" % (
+            "%s\t%s\t%.2f\t%.2f\t%.2f\t%.2f\t%.2f" % (
                 training_timepoints[-1].strftime("%Y-%m"),
                 validation_timepoint.strftime("%Y-%m"),
                 training_error,

### 2019-07-16 src/weighted_distances.py
Always calculate the average distance to the present
Even if there is no future to calculate an average distance to.

@@ -91,24 +91,28 @@ if __name__ == '__main__':
     # Find valid timepoints for calculating distances to the future.
     timepoints = tips["timepoint"].drop_duplicates()
     last_timepoint = timepoints.max() - pd.DateOffset(months=args.delta_months)
-    valid_timepoints = timepoints[timepoints <= last_timepoint]
 
     # Calculate weighted distance to the present and future for each sample at a
     # given timepoint.
     weighted_distances = []
-    for timepoint in valid_timepoints:
+    for timepoint in timepoints:
         future_timepoint = timepoint + pd.DateOffset(months=args.delta_months)
         timepoint_tips = tips[tips["timepoint"] == timepoint]
         future_timepoint_tips = tips[tips["timepoint"] == future_timepoint]
 
         for current_tip, current_tip_frequency in timepoint_tips.loc[:, ["strain", "frequency"]].values:
+            # Calculate the distance to the present for all timepoints.
             weighted_distance_to_present = 0.0
             for other_current_tip, other_current_tip_frequency in timepoint_tips.loc[:, ["strain", "frequency"]].values:
                 weighted_distance_to_present += other_current_tip_frequency * distances_by_sample_names[current_tip][other_current_tip]
 
-            weighted_distance_to_future = 0.0
-            for future_tip, future_tip_frequency in future_timepoint_tips.loc[:, ["strain", "frequency"]].values:
-                weighted_distance_to_future += future_tip_frequency * distances_by_sample_names[current_tip][future_tip]
+            # Calculate the distance to the future only for valid timepoints (those with future information).
+            if timepoint <= last_timepoint:
+                weighted_distance_to_future = 0.0
+                for future_tip, future_tip_frequency in future_timepoint_tips.loc[:, ["strain", "frequency"]].values:
+                    weighted_distance_to_future += future_tip_frequency * distances_by_sample_names[current_tip][future_tip]
+            else:
+                weighted_distance_to_future = np.nan
 
             weighted_distances.append({
                 "timepoint": timepoint,

### 2019-07-16 scripts/calculate_delta_frequency.py
Calculate delta frequencies with diffusion-based frequencies and without using arbitrary clade designations

@@ -16,6 +16,7 @@ if __name__ == "__main__":
     )
     parser.add_argument("--tree", required=True, help="Newick tree")
     parser.add_argument("--frequencies", required=True, help="frequencies JSON")
+    parser.add_argument("--frequency-method", required=True, choices=["kde", "diffusion"], help="method used to estimate frequencies")
     parser.add_argument("--clades", required=True, help="JSON of clade annotations for nodes in the given tree")
     parser.add_argument("--delta-pivots", type=int, default=1, help="number of frequency pivots to look back in time for change in frequency calculation")
     parser.add_argument("--output", required=True, help="JSON of delta frequency annotations for nodes in the given tree")
@@ -29,47 +30,58 @@ if __name__ == "__main__":
     with open(args.frequencies, "r") as fh:
         frequencies_json = json.load(fh)
 
-    frequencies = TreeKdeFrequencies.from_json(frequencies_json)
+    if args.frequency_method == "kde":
+        kde_frequencies = TreeKdeFrequencies.from_json(frequencies_json)
+        frequencies = kde_frequencies.frequencies
 
-    # Load clades.
-    with open(args.clades, "r") as fh:
-        clades_json = json.load(fh)
+        # Load clades.
+        with open(args.clades, "r") as fh:
+            clades_json = json.load(fh)
 
-    clades_by_node = {
-        key: value["clade_membership"]
-        for key, value in clades_json["nodes"].items()
-    }
-
-    # Calculate the total frequency per clade at the most recent timepoint and
-    # requested timepoint in the past using non-zero tip frequencies.
-    current_clade_frequencies = defaultdict(float)
-    previous_clade_frequencies = defaultdict(float)
-
-    for tip in tree.find_clades(terminal=True):
-        # Add tip to current clade frequencies if it is currently alive.
-        if frequencies.frequencies[tip.name][-1] > 0.0:
-            current_clade_frequencies[clades_by_node[tip.name]] += frequencies.frequencies[tip.name][-1]
-
-        # Add tip to previous clade frequencies if it was alive in the last timepoint.
-        if frequencies.frequencies[tip.name][-(args.delta_pivots + 1)] > 0.0:
-            previous_clade_frequencies[clades_by_node[tip.name]] += frequencies.frequencies[tip.name][-(args.delta_pivots + 1)]
-
-    # Determine the total time that elapsed between the current and past timepoint.
-    delta_time = frequencies.pivots[-1] - frequencies.pivots[-(args.delta_pivots + 1)]
-
[…]

### 2019-07-16 scripts/filter_translations.py
Added script to support filtering translations

@@ -1,0 +1,33 @@
+"""Filter translation sequences by date
+"""
+import argparse
+from augur.reconstruct_sequences import load_alignments
+from augur.utils import read_node_data
+from Bio import AlignIO, SeqIO, Seq, SeqRecord
+import pandas as pd
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--alignment", nargs="+", help="amino acid sequences to filter", required=True)
+    parser.add_argument("--branch-lengths", help="JSON with branch lengths and internal node dates estimated by TreeTime", required=True)
+    parser.add_argument("--min-date", help="minimum date for sequences to emit", required=True)
+    parser.add_argument("--output", nargs="+", help="filtered amino acid sequences (one per input)", required=True)
+    args = parser.parse_args()
+
+    # Get min date.
+    min_date = pd.to_datetime(args.min_date)
+
+    # Load branch lengths.
+    node_data = read_node_data(args.branch_lengths)
+
+    # Write alignments to file.
+    for alignment_file, output_file in zip(args.alignment, args.output):
+        alignments = AlignIO.read(alignment_file, "fasta")
+        new_alignments = []
+        for alignment in alignments:
+            date = pd.to_datetime(node_data["nodes"][alignment.id]["date"])
+            if date >= min_date:
+                new_alignments.append(alignment)
+
+        SeqIO.write(new_alignments, output_file, "fasta")

### 2019-07-16 manuscript/supplement.tex

@@
 \section*{Supplemental Material}
-
-\subsection*{Simulations of A/H3N2-like populations}
 
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/composite-model-accuracy-and-coefficients-for-simulated-populations.png}
-  \caption{Composite model a) accuracy and b) coefficients for simulated populations of A/H3N2-like viruses.}
-  \label{sup_fig:composite_model_accuracy_and_coefficients_for_simulated_populations}
+  \includegraphics[width=\textwidth]{figures/distance-of-simulated-populations-between-timepoints.png}
+  \caption{Average weighted distance per year between simulated populations of viruses under the ``naive'' model.}
+  \label{sup_fig:distance_of_simulated_populations_between_timepoints}
   \end{center}
 \end{figure*}
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.png}
+  \caption{Composite model a) accuracy and b) coefficients for simulated populations of A/H3N2-like viruses.}
+  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}
+  \end{center}
+\end{figure*}
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/distance-of-natural-populations-between-timepoints.png}
+  \caption{Average weighted distance per year between natural populations of viruses under the ``naive'' model.}
+  \label{sup_fig:distance_of_natural_populations_between_timepoints}
+  \end{center}
+\end{figure*}
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations.png}
+  \caption{Composite model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.}
+  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}
+  \end{center}
+\end{figure*}

### 2019-07-17 manuscript/main.tex
Revise summary of models for simulated populations

@@
 The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions at epitopes that enable escape from adaptive immunity and purifying selection on domains required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we first simulated the long-term evolution of five independent populations of A/H3N2-like viruses for 5,000 generations each using SANTA-SIM \cite{Jariani2019}.
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of five independent populations of A/H3N2-like viruses for 4,000 generations each using SANTA-SIM \cite{Jariani2019}.
 We seeded each simulated population with the full length HA from A/Beijing/32/1992 such that all simulated sequences contained signal peptide, HA1, and HA2 domains.
@@
 To achieve the same accuracy from mutation-only models, we anticipated that these two models would need to be combined.
-To test these hypotheses, we trained models for each individual fitness metric on 20 years of simulated sequences corresponding to generations 1000--3000 where 100 generations represents one year.
+To test these hypotheses, we trained models for each individual fitness metric on 20 years of simulated sequences corresponding to generations 1000--3000 where 100 generations represent one year.
 As a positive control, we trained a model on the true fitness of each sample as measured by the simulator.
-To control for variability in distances between seasons, we evaluated all models relative to the observed distance between seasons (the adjusted distance between seasons).
-This distance corresponded to a model without an exponential growth parameter which we labeled as the ``naive'' model.
+As a negative control, we evaluated a ``naive''  model under which there was no change in population composition over the one year forecasting period.
+To control for genetic diversity within and between seasons, we evaluated all model forecasts by the average distance between the estimated and observed future population sequences (Methods).
 
-The average distance per year between populations under the naive model of no frequency change was 0.10 +/- 0.17 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
-The naive model accurately approximated the future in five out of 23 validation timepoints (22\%).
-All biologically-informed models consistently reduced this distance with estimates by LBI accurately estimating the future in 22 of 23 timepoints (96\%) with an average distance of -0.16 +/- 0.16 (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-As expected, the true fitness metric always outperformed the naive model.
-The non-epitope metric also consistently outperformed the naive model, predicting the future nearly as well as the true fitness.
-Surprisingly, cross-immunity did not accurately forecast future populations and was consistently assigned a weakly negative coefficient.
-While the delta frequency metric was not as accurate as LBI, it was consistently more accurate than the naive model with an average distance similar to the non-epitope metric.
-The coefficients for all metrics except for the true fitness were stably non-zero across all training windows.
-In contrast, the true fitness metric's coefficients gradually decreased over time.
+As expected, the true fitness model consistently estimated a distance to the future of zero or less than zero, with the exception of two timepoints that almost all models failed to estimate properly (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+The naive model consistently underestimated the observed change in simulated populations, accurately approximating the future in just five of 23 timepoints (22\%) with an average distance of 0.10 +/- 0.17 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+In contrast, all biologically-informed models except for epitope cross-immunity accurately estimated the distance to the future with average distances less than zero.
+Consistent with the importance of purifying selection in the simulation parameters, the non-epitope mutations model had an average distance to the future of -0.01 +/- 0.18 and received an average coefficient of -2.43 +/- 0.53.
+Surprisingly, epitope cross-immunity did not perform well and received a negative average coefficient, despite the anticipated benefit of epitope mutations that could escape exposure-dependent fitness penalties.
+LBI outperformed all other models including the true fitness model, by accurately estimating the future in 22 of 23 timepoints (96\%) with an average distance of -0.16 +/- 0.16.
+The improved performance of LBI and delta frequency over the individual mutation-based metrics matched our expectations that a single metric could represent the fitness of multiple evolutionary constraints.
 
@@
 
-We tested our hypothesis that composite models of cross-immunity and non-epitope mutations would perform as well as LBI or delta frequency.
-We further tested the relationship between LBI and these two individual metrics by fitting composite models with all combinations of these three predictors.
-Surprisingly, cross-immunity provided no additional accuracy to the non-epitope metric, while neither of these mutation-based metrics provided any improvement over LBI (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-In the latter models, LBI maintained a consistent coefficient across all training windows while the models assigned both cross-immunity and non-epitope metrics coefficients near zero.
-These results suggest that non-epitope mutations and LBI are mutually exclusive and either metric is sufficient to forecast the evolution of these simulated populations.
+To test our hypothesis that composite models of epitope cross-immunity and non-epitope mutations would perform as well as LBI or delta frequency, we learned parameters for this composite model.
+Surpisingly, we found that the composite mutation model did not perform better than the individual mutation models and actually performed slightly worse than the non-epitope model with more variability in estimated distance to the future (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+
+To determine which evolutionary constraints LBI represented, we fit two additional composite models including LBI and non-epitope mutations and LBI with both mutation metrics.
+We reasoned that the coefficients learned by these composite models should reflect the relative contribution of each individual metric to the model accuracy.
+For example, if LBI measures antigenic advance, we would expect the values of the LBI and cross-immunity coefficients to be mutually exclusive such that increase in LBI coefficient is associated with decrease in the cross-immunity coefficient (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Interestingly, we found that the coefficients of the mutation metrics were almost always zero while the LBI coefficients were nearly identical to the coefficients of the individual LBI model.
+These results indicate that the predictive contributions of both cross-immunity and non-epitope mutations can be accounted for by LBI alone.
 

### 2019-07-18 manuscript/abstract.tex
Add key points for manuscript to abstract

Key points:

\begin{itemize}
\item{Phenotypic measurements of antigenic advance outperform genetic estimates from epitope mutations in long-term forecasts}
\item{DMS measurements enable more precise identification of mutationally intolerant sites to replace previous sites based on all non-epitope regions}
\item{Phylogenetic fitness metrics outperform and appear to be mutually exclusive of mechanistic metrics}
\item{Distance-based model target performs as well or better than clade-based targets and enables future forecasting efforts where phylogenies are not feasible}
\item{We provide (the first?) open source, extensible forecasting framework for influenza}
\end{itemize}

### 2019-07-19 manuscript/main.tex
Clarify equation for distance to the future
Use more standard representation of double sums, use time parameters to
indicate which timepoints each set of strains come from, and add delta time to
the fitness parameter.

@@
 
-For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the mean weighted Hamming distance between the estimated and observed future populations as defined by,
+For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the mean weighted Hamming distance between the estimated and observed future populations at time $u = t + \Delta{t}$ as defined by,
 
 $$
-\Delta(X(t + \Delta{t}), \hat{X}(t + \Delta{t})) = \frac{(\sum_{i}x_{i}\exp(f_{i}(t))(\sum_{j}x_{j}d_{i,j})) - (\sum_{j}x_{j}(\sum_{j'}x_{j'}d_{j,j'}))}{(\sum_{i}x_{i}(\sum_{j}x_{j}d_{i,j}))},
+\Delta(X(u), \hat{X}(u)) = \frac{\sum_{i}\sum_{j}x_{i}(t)\exp(f_{i}(t)\Delta{t})x_{j}(u)d_{i,j} - \sum_{j}\sum_{k}x_{j}(u)x_{k}(u)d_{j,k}}{\sum_{i}\sum_{j}x_{i}(t)x_{j}(u)d_{i,j}}.
 $$
 
-where the first term in the numerator measures the mean distance between estimated and observed populations, the second term accounts for the diversity of the future population, and the denominator accounts for the diversity between the current and future populations.
-When the sequence diversity of the estimated and observed future populations are equal, the metric evaluates to zero.
+The first term in the numerator measures the mean distance between estimated and observed future populations, the second term measures the average distance of the future population to itself, and the denominator measures the average distance between the current and future populations.
+When a model accurately estimates the average distance of the future population to itself, the metric evaluates to zero.
 When the estimated sequence diversity is closer to the center of the future population's observed diversity, the metric evaluates to a negative proportion.

### 2019-08-25 scripts/pairwise_titer_tree_distances.py
Add titer tree model cross-immunity predictor, cTiter_x
For fair comparisons between different antigenic metrics, we need a
cross-immunity metric for the titer tree model to correspond with existing
substitution model and epitope mutation cross-immunity metrics. Since pairwise
tree comparisons are slower than pairwise sequence comparisons (due to tree
traversal time), the pairwise tree titer distance script accepts a maximum
number of past samples to compare with exist current sample.

@@ -1,0 +1,101 @@
+"""Calculate the distance between sequences between seasons.
+"""
+import argparse
+from augur.frequency_estimators import TreeKdeFrequencies
+from augur.utils import annotate_parents_for_tree, read_node_data, write_json
+import Bio.Phylo
+import json
+import numpy as np
+
+
+def get_titer_distance_between_nodes(tree, past_node, current_node, titer_attr="dTiter"):
+    ancestor = tree.common_ancestor([past_node, current_node])
+    nodes_between_tips = ancestor.get_path(past_node) + ancestor.get_path(current_node)
+    return sum([node.attr[titer_attr] for node in nodes_between_tips])
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--tree", help="Newick tree", required=True)
+    parser.add_argument("--frequencies", help="frequencies JSON", required=True)
+    parser.add_argument("--attribute-name", help="name to store distances associated with the given distance map; multiple attribute names are linked to corresponding positional comparison method and distance map arguments", required=True)
+    parser.add_argument("--model", help="JSON providing the titer tree model", required=True)
+    parser.add_argument("--date-annotations", help="JSON of branch lengths and date annotations from augur refine for samples in the given tree; required for comparisons to earliest or latest date", required=True)
+    parser.add_argument("--years-back-to-compare", type=int, help="number of years prior to the current season to search for samples to calculate pairwise comparisons with", required=True)
+    parser.add_argument("--max-past-samples", type=int, default=500, help="maximum number of past samples to randomly select for comparison to current samples")
+    parser.add_argument("--output", help="JSON file with calculated distances stored by node name and attribute name", required=True)
+
+    args = parser.parse_args()
+
+    # Load tree and annotate parents.
+    tree = Bio.Phylo.read(args.tree, "newick")
+    tree = annotate_parents_for_tree(tree)
+
+    # Load frequencies.
+    with open(args.frequencies, "r") as fh:
+        frequencies_json = json.load(fh)
+
+    frequencies = TreeKdeFrequencies.from_json(frequencies_json)
+    pivots = frequencies.pivots
+
+    # Identify pivots that belong within our search window for past samples.
+    past_pivot_indices = (pivots < pivots[-1]) & (pivots >= pivots[-1] - args.years_back_to_compare)
+
+    # Load date and titer model annotations and annotate tree with them.
[…]

### 2019-08-25 scripts/partition_strains_by_timepoint.py
Fix a time boundary bug in partitioning of strains by timepoint
Fixes a floating point rounding bug that allowed strains with collection dates
equal to the current timepoint to be included in a timepoint's build even though
it would not get an estimate frequency or be included in the downstream
analysis. The fix here is to round the numeric dates passed to the partitioning
script to a monthly floating point resolution instead of the daily resolution of
the TreeTime numeric_date function.

@@ -24,7 +24,7 @@ if __name__ == '__main__':
 
     # Convert date string to a datetime instance.
     timepoint = pd.to_datetime(args.timepoint)
-    numeric_timepoint = numeric_date(timepoint)
+    numeric_timepoint = np.around(numeric_date(timepoint), 2)
 
     # Load metadata with strain names and dates.
     metadata, columns = read_metadata(args.metadata)
@@ -43,13 +43,13 @@ if __name__ == '__main__':
     # earliest date to accept for strains is.
     if args.years_back is not None:
         earliest_timepoint = timepoint - pd.DateOffset(years=args.years_back)
-        numeric_earliest_timepoint = numeric_date(earliest_timepoint)
+        numeric_earliest_timepoint = np.around(numeric_date(earliest_timepoint), 2)
 
         # If reference strains are provided, calculate the earliest date to
         # accept those strains.
         if len(reference_strains) > 0:
             earliest_reference_timepoint = earliest_timepoint - pd.DateOffset(years=args.additional_years_back_for_references)
-            numeric_earliest_reference_timepoint = numeric_date(earliest_reference_timepoint)
+            numeric_earliest_reference_timepoint = np.around(numeric_date(earliest_reference_timepoint), 2)
 
     # Find strains sampled prior to the current timepoint. Strains may have
     # multiple numerical dates, so we filter on the latest (maximum) observed

### 2019-08-26 manuscript/main.tex
Update figures and results text for simulated populations
Simulated populations now use 30 viruses per month to match the higher density
of the natural population builds.

@@
 The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions at epitopes that enable escape from adaptive immunity and purifying selection on domains required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of five independent populations of A/H3N2-like viruses for 4,000 generations each using SANTA-SIM \cite{Jariani2019}.
-We seeded each simulated population with the full length HA from A/Beijing/32/1992 such that all simulated sequences contained signal peptide, HA1, and HA2 domains.
-We defined purifying selection across all three domains, allowing the preferred amino acid at each site to change at a fixed rate over time.
-We additionally defined exposure-dependent selection for 129 putative epitope sites in HA1 \cite{Wolf:2006da} to impose an effect of cross-immunity that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under purifying and positive selection for 30 years (Methods).
 These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
 
-Under the evolutionary constraints of our simulations, we expected the models to learn positive coefficients for cross-immunity and negative coefficients for non-epitope mutations corresponding to the fitness benefits of novel epitope sequences and costs of accumulating deleterious mutations.
-We reasoned that because LBI and delta frequency measure the recent success of samples without any mechanistic explanation for that success, these two models should be assigned positive coefficients and should outperform the individual mutation models by being able to simultaneously represent benefits of epitope mutations and costs of non-epitope mutations with a single metric.
-To achieve the same accuracy from mutation-only models, we anticipated that these two models would need to be combined.
-To test these hypotheses, we trained models for each individual fitness metric on 20 years of simulated sequences corresponding to generations 1000--3000 where 100 generations represent one year.
-As a positive control, we trained a model on the true fitness of each sample as measured by the simulator.
-As a negative control, we evaluated a ``naive''  model under which there was no change in population composition over the one year forecasting period.
+We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while the non-epitope mutations metric would be assigned a negative coefficient.
+We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
+Correspondingly, we expected that a combined model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics.
+In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which there was no estimated change in population composition over the one year forecasting period.
 To control for genetic diversity within and between seasons, we evaluated all model forecasts by the average distance between the estimated and observed future population sequences (Methods).
-
-As expected, the true fitness model consistently estimated a distance to the future of zero or less than zero, with the exception of two timepoints that almost all models failed to estimate properly (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-The naive model consistently underestimated the observed change in simulated populations, accurately approximating the future in just five of 23 timepoints (22\%) with an average distance of 0.10 +/- 0.17 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
-In contrast, all biologically-informed models except for epitope cross-immunity accurately estimated the distance to the future with average distances less than zero.
-Consistent with the importance of purifying selection in the simulation parameters, the non-epitope mutations model had an average distance to the future of -0.01 +/- 0.18 and received an average coefficient of -2.43 +/- 0.53.
-Surprisingly, epitope cross-immunity did not perform well and received a negative average coefficient, despite the anticipated benefit of epitope mutations that could escape exposure-dependent fitness penalties.
-LBI outperformed all other models including the true fitness model, by accurately estimating the future in 22 of 23 timepoints (96\%) with an average distance of -0.16 +/- 0.16.
-The improved performance of LBI and delta frequency over the individual mutation-based metrics matched our expectations that a single metric could represent the fitness of multiple evolutionary constraints.
 
@@
 
-To test our hypothesis that composite models of epitope cross-immunity and non-epitope mutations would perform as well as LBI or delta frequency, we learned parameters for this composite model.
-Surpisingly, we found that the composite mutation model did not perform better than the individual mutation models and actually performed slightly worse than the non-epitope model with more variability in estimated distance to the future (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+As expected, the true fitness model was the most consistently accurate, estimating distances to the future less than or equal to zero at 21 of 23 (91\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+The naive model was the least accurate, with only seven estimates of zero or less (30\% of timepoints) and an average estimated distance of 0.08 +/- 0.13 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+Three of the four biologically-informed models performed as expected with the exception of epitope cross-immunity.
+LBI and delta frequency both received positive coefficients and outperformed the mechanistic metrics on average.
+The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average distance less than zero.
+Surprisingly, epitope cross-immunity performed no better than the naive model and received a negative average coefficient, despite the anticipated benefit of epitope mutations that could escape exposure-dependent fitness penalties.
+Similarly, the combined model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+From these results, we concluded that our models accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions is dominated by purifying selection rather than by positive selection at epitope sites.
 
@@
 
+Discussion.
+
 \section*{Methods}
@@
 \subsection*{Simulation of influenza A/H3N2-like populations}
+
+We simulated the long-term evolution of A/H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 40 years where 100 generations was equivalent to 1 year.
+We discarded the first 10 years as a burn-in period, selected the next 20 years for model fitting and validation, and held out the last 10 years as out-of-sample data for model testing.
+Each simulated population was seeded with the full length HA from A/Beijing/32/1992 (NCBI accession: U26830.1) such that all simulated sequences contained signal peptide, HA1, and HA2 domains.
+We defined purifying selection across all three domains, allowing the preferred amino acid at each site to change at a fixed rate over time.
+We additionally defined exposure-dependent selection for 49 putative epitope sites in HA1 \cite{Luksza:2014hj} to impose an effect of cross-immunity that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
+We modified the SANTA-SIM source code to enable the inclusion of true fitness values for each sample in the FASTA header of the sampled sequences from each generation.
+This modified implementation is available at https://github.com/huddlej/santa-sim/tree/emit-fitness.
 

### 2019-08-27 src/forecast_model.py
Add optional table output to forecasting script
While the primary use case of the forecast model script is to produce
auspice-compatible JSONs, it is also helpful for downstream analyses to have
access to the complete forecast data frames for all delta month projection
points. This commit adds an optional `--output-table` flag to provide this
table.

@@ -18,6 +18,7 @@ if __name__ == "__main__":
     parser.add_argument("--delta-months", required=True, type=int, nargs="+", help="number of months to project clade frequencies into the future")
     parser.add_argument("--output-node-data", required=True, help="node data JSON of forecasts for the given tips")
     parser.add_argument("--output-frequencies", help="frequencies JSON extended with forecasts for the given tips")
+    parser.add_argument("--output-table", help="table of forecasts for the given tips")
 
     args = parser.parse_args()
 
@@ -95,6 +96,7 @@ if __name__ == "__main__":
     pivots = frequencies.pop("pivots")
     projection_pivot = pivots[-1]
 
+    forecasts = []
     for delta_month in args.delta_months:
         delta_time = delta_month / 12.0
         delta_offset = pd.DateOffset(months=delta_month)
@@ -111,6 +113,7 @@ if __name__ == "__main__":
 
         # collect fitness and projection
         forecasts_df = model.predict(tips)
+        forecasts_df["future_timepoint"] = forecasts_df["timepoint"] + delta_offset
 
         # collect dicts from dataframe
         strain_to_projected_frequency = {}
@@ -128,6 +131,10 @@ if __name__ == "__main__":
         # extend pivots
         pivots.append(projection_pivot + delta_time)
 
+        # Collect forecast data frames, if requested.
+        if args.output_table:
+            forecasts.append(forecasts_df)
+
     # reconnect pivots and label projection pivot
     frequencies['pivots'] = pivots
     frequencies['projection_pivot'] = projection_pivot
@@ -135,3 +142,8 @@ if __name__ == "__main__":
     # output to file
     with open(args.output_frequencies, "w") as jsonfile:
         json.dump(frequencies, jsonfile, indent=1)
+
+    # Save forecasts table, if requested.
+    if args.output_table:
+        all_forecasts = pd.concat(forecasts, ignore_index=True)
+        all_forecasts.to_csv(args.output_table, sep="\t", index=False, header=True, na_rep="N/A")

### 2019-08-28 manuscript/main.tex
Add results subsection briefly describing model
Adds a rough draft section about the distance model and revises results for
simulated populations.

@@
 
+\section*{Results}
+
+\subsection*{A distance-based model of seasonal influenza evolution}
+
 \begin{figure*}[t]
@@
 
-\textit{Brief summary of model implementation here including reference to Fig.~\ref{fig:model}.}
-
-\section*{Results}
+Here, we present a model of seasonal influenza evolution inspired by the exponential growth model of \cite{Luksza:2014hj}.
+As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying an exponential growth factor to each virus sample scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
+We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
+We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four novel metrics including hemagglutination inhibition (HI) cross-immunity, deep mutational scanning (DMS) mutational effects, local branching index (LBI), and change in clade frequency over time (delta frequency).
+We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
+After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
+Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model.
+With this approach, we avoid the intrinsic variability of ancestral sequence reconstruction associated with previous clade-based models.
+However, we retain the benefits of fitting models to highly similar samples found within clades.
 
@@
 These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
+We fit models to these simulated populations using all sequence-only fitness metrics.
 
@@
 Correspondingly, we expected that a combined model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics.
-In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which there was no estimated change in population composition over the one year forecasting period.
-To control for genetic diversity within and between seasons, we evaluated all model forecasts by the average distance between the estimated and observed future population sequences (Methods).
+In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which the exponential growth factor is set to one and populations do not change composition over the one year forecasting period.
+The distances estimated under the naive model represent the average distance between the current and future timepoints.
 
@@
 As expected, the true fitness model was the most consistently accurate, estimating distances to the future less than or equal to zero at 21 of 23 (91\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-The naive model was the least accurate, with only seven estimates of zero or less (30\% of timepoints) and an average estimated distance of 0.08 +/- 0.13 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
-Three of the four biologically-informed models performed as expected with the exception of epitope cross-immunity.
+The naive model was the least accurate with an average estimated distance of 0.08 +/- 0.13 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average distance less than zero.
 LBI and delta frequency both received positive coefficients and outperformed the mechanistic metrics on average.
-The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average distance less than zero.
+LBI performed as well as the true fitness predictor, on average, but suffered from higher variability in distance estimates.
 Surprisingly, epitope cross-immunity performed no better than the naive model and received a negative average coefficient, despite the anticipated benefit of epitope mutations that could escape exposure-dependent fitness penalties.
 Similarly, the combined model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-From these results, we concluded that our models accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions is dominated by purifying selection rather than by positive selection at epitope sites.
+From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.
 
-To determine which evolutionary constraints LBI represented, we fit two additional composite models including LBI and non-epitope mutations and LBI with both mutation metrics.
-We reasoned that the coefficients learned by these composite models should reflect the relative contribution of each individual metric to the model accuracy.
-For example, if LBI measures antigenic advance, we would expect the values of the LBI and cross-immunity coefficients to be mutually exclusive such that increase in LBI coefficient is associated with decrease in the cross-immunity coefficient (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-Interestingly, we found that the coefficients of the mutation metrics were almost always zero while the LBI coefficients were nearly identical to the coefficients of the individual LBI model.
-These results indicate that the predictive contributions of both cross-immunity and non-epitope mutations can be accounted for by LBI alone.
-
-\textit{Repeat simulations with multiple epochs to show model coefficients can reflect these underlying changes to evolutionary constraints.}
+To determine which evolutionary constraints the LBI model represented, we fit an additional composite models including LBI and both mutation metrics.
+We expected the coefficients from these composite models to reveal which individual metrics were mutually exclusive or beneficial.
+For example, if LBI accounts for antigenic drift, the coefficients for LBI and epitope cross-immunity should exhibit negatively correlated changes through time.
+Interestingly, we found that this composite model converged to an LBI-only model with zero-valued coefficients for both mutation metrics in the last half of the simulated timepoints (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+These results indicate that LBI alone can account for the predictive contributions of both cross-immunity and non-epitope mutations.
 
@@
 
+Next, we trained and validated models with 21 years of natural A/H3N2 populations spanning from October 1994 to October 2015.
+We held out data subsequent to October 2015 for model testing.
 The average distance per year between natural populations was twice that of simulated populations at 0.20 +/- 0.16 (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).

### 2019-09-02 manuscript/main.tex
Update figures and text for densely sampled populations
Includes results for simulated and natural populations sampled at 90 viruses per
month, a new mockup of the how the model works for Figure 1, and some additional
text in the introduction to contextualize our results.

@@
 
+The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods into an extensible, open source framework that can be used by professional virologists in their vaccine design process \cite{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
+Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, these methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
+Here, we provide the first such open source long-term forecasting framework for seasonal influenza.
+With this framework, we show that HI phenotypes enable more accurate long-term forecasts of A/H3N2 populations compared to previous metrics based on epitope mutations alone.
+However, we also find that phylogenetic fitness metrics based on recent growth of circulating clades consistently outperform any combination of genotypic or phenotypic metrics, suggesting that existing mechanistic models of seasonal influenza evolution are missing critical components.
+
 \section*{Results}
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/model.png}
-  \caption{Schematic representation of the fitness model.}
+  \includegraphics[width=\columnwidth]{figures/distance-based-fitness-model.png}
+  \caption{
+    Schematic representation of the fitness model wherein the fitness of samples at timepoint $t$ is used to minimize the distance between projected samples from that timepoint and the observed population at timepoint $u$ one year in the future.
+    A) A phylogeny at timepoint $u=$April 2004 shows the observed population at that timepoint.
+    The vertical dashed line delineates the previous timepoint of $t=$April 2003 where the size and color of each sample indicates the projected frequency of those samples one year in the future based on the composite model of delta frequency, HI cross-immunity, and non-epitope mutations.
+    Bright yellow samples indicate those with the highest projected frequency and closest to the observed future population one year in the future.
+    B) The density of average distances between samples at different timepoints show how the model estimates the future population from the current population.
+    The blue line indicates the distance between populations at timepoints $t$ and $u$, or the best estimate by a naive model that projects no change in sample frequencies.
+    The orange line shows the distance between the observed future population at timepoint $u$ and the estimated population at that time ($\hat{u}$) produced by projecting the frequency of samples at timepoint $t$ based on estimates of sample fitness.
+    The green line shows the distance between the observed future population and itself.
+    Notably, the estimated future population shifts the distribution of the naive model closer to zero.
+  }
   \label{fig:model}
@@
 We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
-We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four novel metrics including hemagglutination inhibition (HI) cross-immunity, deep mutational scanning (DMS) mutational effects, local branching index (LBI), and change in clade frequency over time (delta frequency).
+We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
 We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
@@
 With this approach, we avoid the intrinsic variability of ancestral sequence reconstruction associated with previous clade-based models.
-However, we retain the benefits of fitting models to highly similar samples found within clades.
+However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.
 
@@
 As expected, the true fitness model was the most consistently accurate, estimating distances to the future less than or equal to zero at 21 of 23 (91\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-The naive model was the least accurate with an average estimated distance of 0.08 +/- 0.13 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+The naive model was the least accurate with an average estimated distance of 0.07 +/- 0.10 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
 The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average distance less than zero.
@@
 We expected the coefficients from these composite models to reveal which individual metrics were mutually exclusive or beneficial.
-For example, if LBI accounts for antigenic drift, the coefficients for LBI and epitope cross-immunity should exhibit negatively correlated changes through time.
-Interestingly, we found that this composite model converged to an LBI-only model with zero-valued coefficients for both mutation metrics in the last half of the simulated timepoints (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+For example, if LBI accounted for antigenic drift, the coefficients for LBI and epitope cross-immunity should exhibit negatively correlated changes through time.
+Interestingly, we found that this composite model nearly converged to a LBI-only model with small or zero-valued coefficients for both mutation metrics in the last half of the simulated timepoints (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 These results indicate that LBI alone can account for the predictive contributions of both cross-immunity and non-epitope mutations.
+In contrast, when we fit a composite model of delta frequency, cross-immunity, and non-epitope mutations, both delta frequency and non-epitope mutations maintained their respective positive and negative coefficients across all timepoints.
+This result suggests that LBI and delta frequency measure qualitatively different components of fitness despite both being similar growth metrics.
 
-\subsection*{Phylogenetic metrics most accurately forecast natural populations of A/H3N2 viruses}
+\subsection*{Model accuracy and coefficients reflect historical patterns of A/H3N2 evolution}
 
-Next, we trained and validated models with 21 years of natural A/H3N2 populations spanning from October 1994 to October 2015.
-We held out data subsequent to October 2015 for model testing.
-The average distance per year between natural populations was twice that of simulated populations at 0.20 +/- 0.16 (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
-Consistent with the simulated population results, almost all biologically-informed metrics performed better than the naive model for natural populations with the exception of the cross-immunity metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-The two phylogenetic metrics, LBI and delta frequency, performed consistently better than all other metrics with average distances to the future of -0.06 +/- 0.20 and 0.00 +/- 0.12, respectively.
-Importantly, HI phenotypes produced more accurate estimates of the future than the other antigenic metric of cross-immunity.
-Variability in the coefficient for the HI phenotypes metric was consistent with relatively sparse HI titer measurements prior to 2011.
-In contrast, the sequence-based metric for functional constraint, non-epitope mutations, outperformed the DMS-based phenotypic metric.
-The gradual decline of the DMS metric's coefficient around 2009 is consistent with overfitting models to the original background strain of the DMS experiments, A/Perth/16/2009.
+Next, we trained and validated models for individual fitness predictors using 23 years of natural A/H3N2 populations spanning from October 1992 to October 2015.
+We held out samples collected between October 2015 and April 2019 for model testing.
+In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes from HI and DMS assays.
+We hypothesized that both HI and DMS phenotype metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
+As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope cross-immunity and HI phenotypes would be individually more predictive than non-epitope mutations or DMS phenotypes.
+Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
 
@@
 
+The average distance per year between natural populations was 0.21 +/- 0.16, indicating that the diversity of natural populations was three times that of simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
+Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS phenotypes metrics (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+Surprisingly, neither antigenic fitness metric outperformed the best functional constraint metric of non-epitope mutations.
+Indeed, epitope cross-immunity only outperformed the naive model at 11 of 29 timepoints (34\%), while HI phenotypes outperformed the naive model at 24 timepoints (83\%).
+Epitope cross-immunity was also the only metric whose coefficient started at a positive value and transitioned to a negative value through the validation period.
+This change in coefficient and the greater accuracy of epitope mutations in the earliest timepoints suggested that positive selection may have weakened at these epitope sites over time.
+In contrast, HI phenotypes maintained a positive coefficient across most timepoints and consistently outperformed the epitope model.
+The HI phenotypes metric may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
+
+Non-epitope mutations also outperformed the DMS phenotypes metric with an average distance to the future of 0.11 +/- 0.16 compared to 0.21 +/- 0.19.
+In contrast to \cite{Luksza:2014hj}, where the coefficient of the non-epitope mutations metric was hardcoded to -0.5, our model learned a consistently stronger coefficient of -2.03 +/- 0.33.
+Notably, the DMS metric performed slightly better for timepoints before April 2010 and performed no better than the naive model thereafter.
+This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments, A/Perth/16/2009.
+Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
+These results suggest that any mutation at a non-epitope site is deleterious for globally circulating strains, despite the more fine-grained fitness constraints of within-host infections.
+
+LBI was the best individual fitness metric and the only metric to estimate an average distance to the future that was less than zero (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+Delta frequency performed as well as HI cross-immunity and non-epitope mutations.
+While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
+Both of these phylogenetic growth metrics strongly depend on a sufficient density of samples, as revealed by the weaker model coefficients for these metrics in the earliest validation timepoints when there are relatively few samples.
+
+\subsection*{Composite models outperform models with individual fitness metrics}
+
 The results suggested that combinations of the best individual metrics like LBI, delta frequency, non-epitope mutations, and HI phenotypes could produce even more accurate models.
@@
 We found results (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
-
-\subsection*{Model accuracy and coefficients reflect historical patterns of A/H3N2 evolution}
-
-\subsection*{Composite models outperform models with individual fitness metrics}
 

### 2019-09-02 manuscript/supplement.tex
Update figures and text for densely sampled populations
Includes results for simulated and natural populations sampled at 90 viruses per
month, a new mockup of the how the model works for Figure 1, and some additional
text in the introduction to contextualize our results.

@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations.png}
+  \includegraphics[width=\textwidth]{figures/unadjusted-DMS-model-accuracy-and-coefficients-for-natural-populations.png}
+  \caption{
+  Primary and alternate fitness metrics based on deep mutational scanning (DMS) preferences for seasonal influenza A/H3N2 strain A/Perth/16/2009.
+  Metrics are increasingly less dependent on the sequence content of the DMS strain's genetic background from top to bottom.
+  The most structurally naive metric of non-epitope mutations outperforms all of the DMS metrics, indicating that these metrics still overfit to the DMS strain's genetic background.
+  }
+  \label{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}
+  \end{center}
+\end{figure*}
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/best-unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
   \caption{Composite model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.}

### 2019-09-02 manuscript/main.tex
Add placeholder forecast figure from Nextstrain

@@
   \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
-  \caption{Model a) accuracy and b) coefficients for simulated populations of A/H3N2-like viruses.}
+  \caption{Model a) coefficients and b) accuracy for simulated populations of A/H3N2-like viruses.}
   \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
@@
   \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
-  \caption{Model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.}
+  \caption{Model a) coefficients and b) accuracy for natural populations of A/H3N2 viruses.}
   \label{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}
@@
 
-\subsection*{Forecasts predict the rise of 3c3.A and A1b clades in February 2020}
+\subsection*{Forecasts predict the rise of A1b/131K and A1b/135K sub-clades in September 2020}
+
+See Fig.~\ref{fig:nextstrain_forecasts}.
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/nextstrain-forecasts-for-september-2020.png}
+  \caption{Live forecasts on nextstrain.org from our best model for September 2020.}
+  \label{fig:nextstrain_forecasts}
+  \end{center}
+\end{figure*}
 

### 2019-09-05 scripts/pairwise_titer_tree_distances.py
Reduce run time for pairwise titer tree distances
Reduces the maximum number of past samples to consider for comparison to current
samples and also introduces a minimum frequency that a sample must be or have
been in the present or past to be included in a calculation. This latter change
helps us avoid calculations for samples that do not actually contribute to
forecasts.

@@ -22,7 +22,8 @@ if __name__ == "__main__":
     parser.add_argument("--model", help="JSON providing the titer tree model", required=True)
     parser.add_argument("--date-annotations", help="JSON of branch lengths and date annotations from augur refine for samples in the given tree; required for comparisons to earliest or latest date", required=True)
     parser.add_argument("--years-back-to-compare", type=int, help="number of years prior to the current season to search for samples to calculate pairwise comparisons with", required=True)
-    parser.add_argument("--max-past-samples", type=int, default=500, help="maximum number of past samples to randomly select for comparison to current samples")
+    parser.add_argument("--max-past-samples", type=int, default=200, help="maximum number of past samples to randomly select for comparison to current samples")
+    parser.add_argument("--min-frequency", type=float, default=1e-5, help="minimum frequency to consider a sample alive")
     parser.add_argument("--output", help="JSON file with calculated distances stored by node name and attribute name", required=True)
 
     args = parser.parse_args()
@@ -57,10 +58,10 @@ if __name__ == "__main__":
         # Samples with nonzero frequencies in the last timepoint are current
         # samples. Those with one or more nonzero frequencies in the search
         # window of the past timepoints are past samples.
-        if frequencies.frequencies[tip.name][-1] > 0:
+        if frequencies.frequencies[tip.name][-1] > args.min_frequency:
             current_samples.append(tip.name)
             tips_by_sample[tip.name] = tip
-        elif (frequencies.frequencies[tip.name][past_pivot_indices] > 0).sum() > 0:
+        elif (frequencies.frequencies[tip.name][past_pivot_indices] > args.min_frequency).sum() > 0:
             past_samples.append(tip.name)
             tips_by_sample[tip.name] = tip
 

### 2019-09-05 manuscript/main.tex
Clean up math notation
Convert capital X's to lowercase bolded x's, etc. Still to do is replacement of
these characters with custom math terms.

@@
 As in \cite{Luksza:2014hj}, we assumed that influenza A/H3N2 populations can be described by an exponential growth model.
-Under this model, we estimated the future frequency of the global population, $\hat{X}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
+Under this model, we estimated the future frequency of the global population, $\mathbf{\hat{x}}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
 
 $$
-\hat{X}(t + \Delta{t}) = \frac{1}{Z(t)}\sum_{i}x_{i}(t)\exp(f_{i}(t))
+\mathbf{\hat{x}}(t + \Delta{t}) = \frac{1}{Z(t)}\sum_{i}x_{i}(t)\exp(f_{i}(t))
 $$
@@
 We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
+For example, the following equation estimates fitness per sample by epitope cross-immunity ($\mathrm{ep}$), non-epitope mutations ($\mathrm{ne}$), and local branching index ($\mathrm{lbi}$).
+
+$$
+f_{i}(t) = \beta_{\mathrm{ep}}f_{i, \mathrm{ep}}(t) + \beta_{\mathrm{ne}}f_{i, \mathrm{ne}}(t) + \beta_{\mathrm{lbi}}f_{i, \mathrm{lbi}}(t)
+$$
 
@@
 $$
-\Delta(X(u), \hat{X}(u)) = \frac{\sum_{i}\sum_{j}x_{i}(t)\exp(f_{i}(t)\Delta{t})x_{j}(u)d_{i,j} - \sum_{j}\sum_{k}x_{j}(u)x_{k}(u)d_{j,k}}{\sum_{i}\sum_{j}x_{i}(t)x_{j}(u)d_{i,j}}.
+\Delta\mathbf{d} = \frac{\mathbf{d}(\hat{u}, u) - \mathbf{d}(u, u)}{\mathbf{d}(t, u)}
+                                           = \frac{\sum_{i}\sum_{j}x_{i}(t)\exp(f_{i}(t)\Delta{t})x_{j}(u)d_{i,j} - \sum_{j}\sum_{k}x_{j}(u)x_{k}(u)d_{j,k}}{\sum_{i}\sum_{j}x_{i}(t)x_{j}(u)d_{i,j}}.
 $$

### 2019-09-05 analyses/simulations/run.sh
Simulate flu-like populations with changing evolutionary constraints in different epochs

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-java -jar ~/projects/santa-sim/dist/santa.jar influenza_h3n2_ha.xml
+java -jar ~/projects/santa-sim/dist/santa.jar influenza_h3n2_ha_epochs.xml
 
 seqtk sample simulated_HA_sequences.fasta 0.05 > simulated_HA_sequences_5pct.fasta
 #seqtk sample simulated_HA_sequences.fasta 0.2 > simulated_HA_sequences_20pct.fasta

### 2019-09-05 manuscript/abstract.tex
Update main figures and key points in abstract outline

@@
 \item{Phenotypic measurements of antigenic advance outperform genetic estimates from epitope mutations in long-term forecasts}
-\item{DMS measurements enable more precise identification of mutationally intolerant sites to replace previous sites based on all non-epitope regions}
 \item{Phylogenetic fitness metrics outperform and appear to be mutually exclusive of mechanistic metrics}
+\item{Fitness metrics based on epitope mutations are not robust to long-term changes in genetic background of successful populations}
+\item{Fitness metrics based on DMS measurements overfit to the genetic background of the experimental strain}
 \item{Distance-based model target performs as well or better than clade-based targets and enables future forecasting efforts where phylogenies are not feasible}

### 2019-09-06 manuscript/build.sh
Update TeX compile script and rename manuscript file
1. You can compile much faster with `-draftmode` for the first two runs
of pdflatex. And you only need pdflatex -> bibtex -> pdflatex -> pdflatex.

2. I prefer to name my manuscript files something other than manuscript.pdf
because this often gets shared to Slack etc... Nice to have a name with
context.

@@ -1,4 +1,4 @@
-TITLE=manuscript
+TITLE=flu_forecasting
 
 rm -f *.log
 rm -f *.aux
@@ -9,8 +9,7 @@ rm -f *.blg
 rm -f .DS_Store
 rm -f ${TITLE}.pdf
 
-pdflatex ${TITLE}.tex
+pdflatex -draftmode ${TITLE}
 bibtex ${TITLE}
-pdflatex ${TITLE}.tex
-pdflatex ${TITLE}.tex
-pdflatex ${TITLE}.tex
+pdflatex -draftmode ${TITLE}
+pdflatex ${TITLE}

### 2019-09-21 src/fit_model.py
Stabilize random components of model optimization
Now that model coefficients are initialized by random floating point values, we
need to set numpy's random seed to ensure consistent results for identical input
data. This commit also initializes coefficients from prior timepoint
coefficients during the cross-validation analysis. This prior information is a
more reasonable default for the optimizer than a random number.

@@ -16,6 +16,8 @@ from weighted_distances import get_distances_by_sample_names, get_distance_matri
 MAX_PROJECTED_FREQUENCY = 1e3
 FREQUENCY_TOLERANCE = 1e-3
 
+np.random.seed(314159)
+
 
 def sum_of_differences(observed, estimated, y_diff, **kwargs):
     """
@@ -279,7 +281,11 @@ class ExponentialGrowthModel(object):
         self.mean_stds_ = self.calculate_mean_stds(X, self.predictors)
 
         # Find coefficients that minimize the model's cost function.
-        initial_coefficients = np.random.random(len(self.predictors))
+        if hasattr(self, "coef_"):
+            initial_coefficients = self.coef_
+        else:
+            initial_coefficients = np.random.random(len(self.predictors))
+
         results = minimize(
             self._fit,
             initial_coefficients,
@@ -652,9 +658,13 @@ def cross_validate(model_class, model_kwargs, data, targets, train_validate_time
     """
     results = []
     differences_of_model_and_naive_errors = []
+    previous_coefficients = None
 
     for timepoints in train_validate_timepoints:
         model = model_class(**model_kwargs)
+
+        if previous_coefficients is not None:
+            model.coef_ = previous_coefficients
 
         # Get training and validation timepoints.
         training_timepoints = pd.to_datetime(timepoints["train"])
@@ -667,6 +677,7 @@ def cross_validate(model_class, model_kwargs, data, targets, train_validate_time
         # Fit a model to the training data.
         if coefficients is None:
             training_error = model.fit(training_X, training_y)
+            previous_coefficients = model.coef_
             null_training_error = model._fit(np.zeros_like(model.coef_), training_X, training_y)
         else:
             model.coef_ = coefficients

### 2019-09-21 src/fit_model.py
Include proportion of timepoints model is better than naive for debugging

@@ -694,14 +694,15 @@ def cross_validate(model_class, model_kwargs, data, targets, train_validate_time
         null_validation_error = model._fit(np.zeros_like(model.coef_), validation_X, validation_y)
         differences_of_model_and_naive_errors.append(validation_error - null_validation_error)
         print(
-            "%s\t%s\t%.2f\t%.2f\t%.2f\t%.2f\t%s" % (
+            "%s\t%s\t%.2f\t%.2f\t%.2f\t%.2f\t%s\t%.2f" % (
                 training_timepoints[-1].strftime("%Y-%m"),
                 validation_timepoint.strftime("%Y-%m"),
                 training_error,
                 null_training_error,
                 validation_error,
                 null_validation_error,
-                model.coef_
+                model.coef_,
+                (np.array(differences_of_model_and_naive_errors) < 0).sum() / float(len(differences_of_model_and_naive_errors))
             ),
             flush=True
         )

### 2019-09-23 manuscript/main.tex
Forecasts are not live when they are in the manuscript
Embed dynamically generated tables in manuscript
Adds a dynamically generated table summarizing model accuracy for simulated
populations and the associated pandas code in a Jupyter notebook to generate
this table. Also adds the corresponding latex code to include the table in the
appropriate environment.
Update distance-based model schematic
Simplifies the caption for this figure by removing description of weighted
distance-based model for now. This simplification also allows the new figure
mockup to fit in the correct place of the text.

@@
     Bright yellow samples indicate those with the highest projected frequency and closest to the observed future population one year in the future.
-    B) The density of average distances between samples at different timepoints show how the model estimates the future population from the current population.
-    The blue line indicates the distance between populations at timepoints $t$ and $u$, or the best estimate by a naive model that projects no change in sample frequencies.
-    The orange line shows the distance between the observed future population at timepoint $u$ and the estimated population at that time ($\hat{u}$) produced by projecting the frequency of samples at timepoint $t$ based on estimates of sample fitness.
-    The green line shows the distance between the observed future population and itself.
-    Notably, the estimated future population shifts the distribution of the naive model closer to zero.
   }
@@
 \end{figure*}
+
+\begin{table*}[h]
+  \begin{center}
+    \input{tables/simulated_model_selection.tex}
+    \caption{Simulated population model accuracy relative to the naive model}
+    \label{table_simulated_model_selection}
+  \end{center}
+\end{table*}
 
@@
   \includegraphics[width=\textwidth]{figures/nextstrain-forecasts-for-september-2020.png}
-  \caption{Live forecasts on nextstrain.org from our best model for September 2020.}
+  \caption{Forecasts on nextstrain.org from our best model for September 2020.}
   \label{fig:nextstrain_forecasts}

### 2019-09-24 src/forecast_model.py
Add rule to forecast tips from all timepoints
Uses the best model for a given build to create a table of forecasts for all
tips at all timepoints. This table is used for model validation/testing. This
commit also modifies the forecasting script to make frequency inputs and node
data and frequencies outputs optional. The script will complain loudly with an
error if the user doesn't request any of the three possible inputs.

@@ -4,6 +4,7 @@ import argparse
 import json
 import numpy as np
 import pandas as pd
+import sys
 
 from fit_model import DistanceExponentialGrowthModel
 from weighted_distances import get_distances_by_sample_names
@@ -16,11 +17,22 @@ if __name__ == "__main__":
     parser.add_argument("--frequencies", help="JSON representing historical frequencies to project from")
     parser.add_argument("--model", required=True, help="JSON representing the model fit with training and cross-validation results, beta coefficients for predictors, and summary statistics")
     parser.add_argument("--delta-months", required=True, type=int, nargs="+", help="number of months to project clade frequencies into the future")
-    parser.add_argument("--output-node-data", required=True, help="node data JSON of forecasts for the given tips")
+    parser.add_argument("--output-node-data", help="node data JSON of forecasts for the given tips")
     parser.add_argument("--output-frequencies", help="frequencies JSON extended with forecasts for the given tips")
     parser.add_argument("--output-table", help="table of forecasts for the given tips")
 
     args = parser.parse_args()
+
+    # Confirm that at least one output file has been specified.
+    outputs = [
+        args.output_node_data,
+        args.output_frequencies,
+        args.output_table
+    ]
+    outputs_missing =[output is None for output in outputs]
+    if all(outputs_missing):
+        print("ERROR: No output files were specified", file=sys.stderr)
+        sys.exit(1)
 
     # Load standardized tip attributes subsetting to tip name, clade, frequency,
     # and requested predictors.
@@ -74,27 +86,31 @@ if __name__ == "__main__":
         strain_to_projected_frequency[row['strain']] = row['projected_frequency']
         strain_to_weighted_distance_to_future[row['strain']] = row['y']
 
-    # populate node data
-    node_data = {}
-    strains = list(tips['strain'])
-    for strain in strains:
-        node_data[strain] = {
-            "fitness": strain_to_fitness[strain],
-            "future_timepoint": strain_to_future_timepoint[strain],
-            "projected_frequency": strain_to_projected_frequency[strain],
-            "weighted_distance_to_future": strain_to_weighted_distance_to_future[strain]
-        }
-
     # output to file
-    with open(args.output_node_data, "w") as jsonfile:
-        json.dump({"nodes": node_data}, jsonfile, indent=1)
+    if args.output_node_data:
+        # populate node data
+        node_data = {}
+        strains = list(tips['strain'])
+        for strain in strains:
[…]

### 2019-09-25 scripts/plot_tree.py
Plot Newick trees for rapid debugging of tree topology
The original tree plotting tool expected an auspice tree JSON as input which is
handy for plotting trees colored by specific tip attributes, but these JSONs are
almost literally the last output from the entire forecasting pipeline. This
commit adds support for plotting from Newick files created by augur refine, so
we can more quickly detect and resolve topological issues witih trees.

@@ -1,5 +1,6 @@
 import argparse
-from augur.utils import json_to_tree
+from augur.utils import json_to_tree, read_tree
+import Bio.Phylo
 import json
 import matplotlib as mpl
 mpl.use("Agg")
@@ -164,7 +165,7 @@ def plot_tree(tree, figure_name, color_by_trait, initial_branch_width, tip_size,
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser()
-    parser.add_argument("tree", help="auspice tree JSON")
+    parser.add_argument("tree", help="auspice tree JSON or Newick tree")
     parser.add_argument("output", help="plotted tree figure")
     parser.add_argument("--colorby", help="trait in tree to color by", default="num_date")
     parser.add_argument("--branch_width", help="initial branch width", type=int, default=10)
@@ -174,20 +175,27 @@ if __name__ == "__main__":
     parser.add_argument("--include-color-bar", action="store_true", help="display a color bar for the color by option at the bottom of the plot")
     args = parser.parse_args()
 
-    with open(args.tree, "r") as json_fh:
-        json_tree = json.load(json_fh)
+    if args.tree.endswith(".json"):
+        with open(args.tree, "r") as json_fh:
+            json_tree = json.load(json_fh)
 
-    # Convert JSON tree layout to a Biopython Clade instance.
-    tree = json_to_tree(json_tree)
+        # Convert JSON tree layout to a Biopython Clade instance.
+        tree = json_to_tree(json_tree)
 
-    # Plot the tree.
-    plot_tree(
-        tree,
-        args.output,
-        args.colorby,
-        args.branch_width,
-        args.tip_size,
-        args.start_date,
-        args.end_date,
-        args.include_color_bar
-    )
+        # Plot the tree.
+        plot_tree(
+            tree,
+            args.output,
+            args.colorby,
+            args.branch_width,
+            args.tip_size,
+            args.start_date,
+            args.end_date,
+            args.include_color_bar
+        )
+    else:
+        tree = read_tree(args.tree)
+        tree.ladderize()
+        fig, ax = plt.subplots(1, 1, figsize=(8, 8))
+        Bio.Phylo.draw(tree, axes=ax, label_func=lambda node: "", show_confidence=False)
+        plt.savefig(args.output)

### 2019-09-26 manuscript/main.tex
Add figures/tables for EMD-based models with dense builds
Also, updates schematic figure for how the model works.
Update figures/text to reflect natural population EMD-based results

@@
 
-\subsection*{Models accurately forecast evolution of simulated populations of {A/H3N2-like viruses}}
+\subsection*{Models accurately forecast evolution of A/H3N2-like viruses}
 
 The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions at epitopes that enable escape from adaptive immunity and purifying selection on domains required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under purifying and positive selection for 30 years (Methods).
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under purifying and positive selection for 40 years (Methods).
 These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
@@
 
-\begin{figure*}[t]
+\begin{figure*}[ht]
   \begin{center}
@@
 
-\begin{table*}[h]
+\begin{table*}[ht]
   \begin{center}
@@
 
-As expected, the true fitness model was the most consistently accurate, estimating distances to the future less than or equal to zero at 21 of 23 (91\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-The naive model was the least accurate with an average estimated distance of 0.07 +/- 0.10 (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
-The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average distance less than zero.
-LBI and delta frequency both received positive coefficients and outperformed the mechanistic metrics on average.
-LBI performed as well as the true fitness predictor, on average, but suffered from higher variability in distance estimates.
-Surprisingly, epitope cross-immunity performed no better than the naive model and received a negative average coefficient, despite the anticipated benefit of epitope mutations that could escape exposure-dependent fitness penalties.
-Similarly, the combined model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+The average distance between yearly populations, as measured by the naive model, was 8.97 +/- 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.17 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection} and Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+With the exception of epitope cross-immunity, all biologically-informed models also outperformed the naive model.
+LBI was the best of these models, reducing the distance between populations by 1.41 amino acids on average.
+Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
+The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average improvement of 0.71 amino acids.
+Surprisingly, the combined model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.
 
-To determine which evolutionary constraints the LBI model represented, we fit an additional composite models including LBI and both mutation metrics.
-We expected the coefficients from these composite models to reveal which individual metrics were mutually exclusive or beneficial.
-For example, if LBI accounted for antigenic drift, the coefficients for LBI and epitope cross-immunity should exhibit negatively correlated changes through time.
-Interestingly, we found that this composite model nearly converged to a LBI-only model with small or zero-valued coefficients for both mutation metrics in the last half of the simulated timepoints (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-These results indicate that LBI alone can account for the predictive contributions of both cross-immunity and non-epitope mutations.
-In contrast, when we fit a composite model of delta frequency, cross-immunity, and non-epitope mutations, both delta frequency and non-epitope mutations maintained their respective positive and negative coefficients across all timepoints.
-This result suggests that LBI and delta frequency measure qualitatively different components of fitness despite both being similar growth metrics.
+We hypothesized that a combined model of mutually beneficial metrics could better approximate the true fitness of simulated viruses.
+To this end, we fit an additional combined model including both LBI and non-epitope mutations.
+This combined model outperformed all individual metrics, reducing the distance between populations by 1.76 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+Interestingly, we found that the coefficients for LBI and non-epitope mutations remained relatively stable across all validation timepoints.
+These results support our hypothesis that multiple complementary metrics can produce more accurate models.
 
-\subsection*{Model accuracy and coefficients reflect historical patterns of A/H3N2 evolution}
+Finally, we sought to validate and test the best performing model by two metrics relevant for practical influenza forecasting and vaccine design efforts.
+First, we measured the ability of the best model to accurately estimate clade dynamics by correlating estimated and observed clade growth rates.
+Under these simulated conditions, our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.71$ $(p < 0.001)$ and overall accuracies for clade growth and decline predictions of 66\% and 61\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+Next, we tested how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
+The estimated best strain was in the top 4th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 82\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+
+\subsection*{Model coefficients and performance reflect historical patterns of A/H3N2 evolution}
 
 Next, we trained and validated models for individual fitness predictors using 23 years of natural A/H3N2 populations spanning from October 1992 to October 2015.
-We held out samples collected between October 2015 and April 2019 for model testing.
+We held out samples collected between October 2015 and October 2019 for model testing.
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes from HI and DMS assays.
@@
 
-The average distance per year between natural populations was 0.21 +/- 0.16, indicating that the diversity of natural populations was three times that of simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/natural_model_selection.tex}
+    \caption{Natural population model accuracy relative to the naive model}
+    \label{table_natural_model_selection}
+  \end{center}
+\end{table*}
+
+The average distance per year between natural populations was 6.50 +/- 1.42 amino acids or 72\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
 Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS phenotypes metrics (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-Surprisingly, neither antigenic fitness metric outperformed the best functional constraint metric of non-epitope mutations.
-Indeed, epitope cross-immunity only outperformed the naive model at 11 of 29 timepoints (34\%), while HI phenotypes outperformed the naive model at 24 timepoints (83\%).
+Surprisingly, the best antigenic fitness metric of HI phenotypes performed only slightly better the best functional constraint metric of non-epitope mutations.
+Indeed, epitope cross-immunity only outperformed the naive model at one of 19 timepoints (5\%), while HI phenotypes outperformed the naive model at 15 (79\%) timepoints (Table~\ref{table_natural_model_selection}).
 Epitope cross-immunity was also the only metric whose coefficient started at a positive value and transitioned to a negative value through the validation period.
-This change in coefficient and the greater accuracy of epitope mutations in the earliest timepoints suggested that positive selection may have weakened at these epitope sites over time.
-In contrast, HI phenotypes maintained a positive coefficient across most timepoints and consistently outperformed the epitope model.
+This change in coefficient suggests that positive selection may have weakened at these epitope sites over time.
+In contrast, HI phenotypes maintained a positive coefficient across most timepoints.
 The HI phenotypes metric may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
 
-Non-epitope mutations also outperformed the DMS phenotypes metric with an average distance to the future of 0.11 +/- 0.16 compared to 0.21 +/- 0.19.
-In contrast to \cite{Luksza:2014hj}, where the coefficient of the non-epitope mutations metric was hardcoded to -0.5, our model learned a consistently stronger coefficient of -2.03 +/- 0.33.
-Notably, the DMS metric performed slightly better for timepoints before April 2010 and performed no better than the naive model thereafter.
-This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments, A/Perth/16/2009.
+Non-epitope mutations also outperformed the DMS phenotypes metric, reducing the distance to the future by 0.40 amino acids on average compared to 0.19 amino acids, respectively.
+In contrast to \cite{Luksza:2014hj}, where the coefficient of the non-epitope mutations metric was hardcoded to -0.5, our model learned a consistently stronger coefficient of -1.45 +/- 0.24.
+Notably, the DMS metric only performed noticeably better than the naive model at the timepoint immediately preceding when the background strain of the DMS experiments, A/Perth/16/2009, was sampled.
+This result is consistent with the DMS model overfitting to the evolutionary history of the DMS background strain used to perform the DMS experiments.
 Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
@@
 
-LBI was the best individual fitness metric and the only metric to estimate an average distance to the future that was less than zero (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-Delta frequency performed as well as HI cross-immunity and non-epitope mutations.
+LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+However, LBI only outperformed the naive model 63\% of the time (Table~\ref{table_natural_model_selection}).
+These results suggest that when LBI forecasts correctly, it does so better than other metrics, but that it is prone to overfitting.
+Delta frequency did not surpass LBI in average estimated distance to the future and performed worse than both the HI phenotypes and non-epitope mutation models.
 While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
-Both of these phylogenetic growth metrics strongly depend on a sufficient density of samples, as revealed by the weaker model coefficients for these metrics in the earliest validation timepoints when there are relatively few samples.
 
@@
 
-The results suggested that combinations of the best individual metrics like LBI, delta frequency, non-epitope mutations, and HI phenotypes could produce even more accurate models.
-To this end, we fit models for X combinations of these metrics and evaluated their distance to the future and coefficients relative to the corresponding individual metrics.
-We found results (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
+To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
+Specifically, we fit models based on HI phenotypes and non-epitope mutations, LBI and non-epitope mutations, and all three of these metrics together.
+We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.
+
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/best-composite-unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
+  \caption{Composite model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.}
+  \label{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}
+  \end{center}
+\end{figure*}
+
+As expected, all of these combined models performed better than any of their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
+The model with all three metrics performed the best, reducing the distance to the future by 1.21 amino acids on average.
+However, the composite of LBI and non-epitope mutations performed nearly as well with an average reduced distance of 1.14.
+Surprisingly, the combination of HI phenotypes and non-epitope mutations did not perform better than LBI by itself, despite the stability of the coefficients for both metrics across all timepoints.
 
 \subsection*{Models enable selection of vaccine candidate strains}
+
+As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
+We find that the combined model of LBI and non-epitope mutations accurately captures clade decline and modestly reflects observed clade growth (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
+
+\begin{figure*}[ht]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
+  \caption{
+  Validation of best model for natural populations of A/H3N2 viruses, the combined model of LBI and non-epitope mutations.
+  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  B) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
+  }
+  \label{fig:validation_of_best_model_for_natural_populations}
+  \end{center}
+\end{figure*}
 
@@
 
-We simulated the long-term evolution of A/H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 40 years where 100 generations was equivalent to 1 year.
-We discarded the first 10 years as a burn-in period, selected the next 20 years for model fitting and validation, and held out the last 10 years as out-of-sample data for model testing.
+We simulated the long-term evolution of A/H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 50 years where 200 generations was equivalent to 1 year.
+We discarded the first 10 years as a burn-in period, selected the next 30 years for model fitting and validation, and held out the last 10 years as out-of-sample data for model testing.
 Each simulated population was seeded with the full length HA from A/Beijing/32/1992 (NCBI accession: U26830.1) such that all simulated sequences contained signal peptide, HA1, and HA2 domains.

### 2019-09-26 manuscript/supplement.tex
Add figures/tables for EMD-based models with dense builds
Also, updates schematic figure for how the model works.
Update figures/text to reflect natural population EMD-based results

@@
   \includegraphics[width=\textwidth]{figures/simulated-h3n2-ha-phylogeny.png}
-  \caption{Phylogeny of simulated A/H3N2-like HA sequences.
+  \caption{
+  Phylogeny of A/H3N2-like HA sequences sampled between the 24th and 30th years of simulated evolution.
   The phylogenetic structure and rate of accumulated epitope and non-epitope mutations match patterns observed in phylogenies of natural sequences.
-Sample dates were annotated as the generation in the simulation divided by 100 and added to 2000, to acquire realistic date ranges that were compatible with our modeling machinery.}
+  Sample dates were annotated as the generation in the simulation divided by 200 and added to 2000, to acquire realistic date ranges that were compatible with our modeling machinery.
+  }
   \label{sup_fig:simulated_h3n2_ha_phylogeny}
@@
   \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-simulated-populations.png}
+  \caption{
+  Validation of best model for simulated populations of A/H3N2-like viruses.
+  a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  b) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
+  }
+  \label{sup_fig:validation_of_best_model_for_simulated_populations}
+  \end{center}
+\end{figure*}
+
+\begin{figure*}[t]
+  \begin{center}
   \includegraphics[width=\textwidth]{figures/distance-of-natural-populations-between-timepoints.png}
@@
 \end{figure*}
-
-\begin{figure*}[t]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/best-unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
-  \caption{Composite model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.}
-  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}
-  \end{center}
-\end{figure*}

### 2019-09-27 manuscript/main.tex
Use better metric names, flesh out results for natural populations
Consistently uses "HI cross-immunity" instead of "HI phenotypes" and "DMS
mutational effect" instead of "DMS phenotypes".

@@
 
-Here, we present a model of seasonal influenza evolution inspired by the exponential growth model of \cite{Luksza:2014hj}.
-As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying an exponential growth factor to each virus sample scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
+Here, we present a model of seasonal influenza evolution inspired by the exponential growth model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
 We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
@@
 After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
-Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model.
-With this approach, we avoid the intrinsic variability of ancestral sequence reconstruction associated with previous clade-based models.
+Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model (Fig.~\ref{fig:model}).
+With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
 However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.
@@
 
-The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions at epitopes that enable escape from adaptive immunity and purifying selection on domains required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under purifying and positive selection for 40 years (Methods).
+The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods).
 These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
@@
 
-We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while the non-epitope mutations metric would be assigned a negative coefficient.
+We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like non-epitope mutations, would receive negative coefficients.
 We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
-Correspondingly, we expected that a combined model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics.
+Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics.
 In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which the exponential growth factor is set to one and populations do not change composition over the one year forecasting period.
@@
 Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
-The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average improvement of 0.71 amino acids.
-Surprisingly, the combined model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+The non-epitope mutations metric received a consistently negative coefficient with an average improvement of 0.71 amino acids.
+Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.
 
-We hypothesized that a combined model of mutually beneficial metrics could better approximate the true fitness of simulated viruses.
-To this end, we fit an additional combined model including both LBI and non-epitope mutations.
-This combined model outperformed all individual metrics, reducing the distance between populations by 1.76 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses.
+To this end, we fit an additional model including both LBI and non-epitope mutations.
+This composite model outperformed all individual metrics, reducing the distance between populations by 1.76 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
 Interestingly, we found that the coefficients for LBI and non-epitope mutations remained relatively stable across all validation timepoints.
@@
 Finally, we sought to validate and test the best performing model by two metrics relevant for practical influenza forecasting and vaccine design efforts.
-First, we measured the ability of the best model to accurately estimate clade dynamics by correlating estimated and observed clade growth rates.
-Under these simulated conditions, our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.71$ $(p < 0.001)$ and overall accuracies for clade growth and decline predictions of 66\% and 61\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+First, we measured the ability of the true fitness model to accurately estimate clade dynamics by correlating estimated and observed clade growth rates.
+Under these simulated conditions, our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.71 (p < 0.001)$ and overall accuracies for clade growth and decline predictions of 66\% and 61\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
 Next, we tested how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
 The estimated best strain was in the top 4th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 82\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
 
@@
 We held out samples collected between October 2015 and October 2019 for model testing.
-In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes from HI and DMS assays.
-We hypothesized that both HI and DMS phenotype metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
-As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope cross-immunity and HI phenotypes would be individually more predictive than non-epitope mutations or DMS phenotypes.
+In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
+We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
+As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than non-epitope mutations or DMS mutational effects.
 Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
@@
 The average distance per year between natural populations was 6.50 +/- 1.42 amino acids or 72\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
-Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS phenotypes metrics (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-Surprisingly, the best antigenic fitness metric of HI phenotypes performed only slightly better the best functional constraint metric of non-epitope mutations.
-Indeed, epitope cross-immunity only outperformed the naive model at one of 19 timepoints (5\%), while HI phenotypes outperformed the naive model at 15 (79\%) timepoints (Table~\ref{table_natural_model_selection}).
+Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+Surprisingly, the best antigenic fitness metric of HI cross-immunity performed only slightly better the best functional constraint metric of non-epitope mutations.
+Indeed, epitope cross-immunity only outperformed the naive model at one of 19 timepoints (5\%), while HI cross-immunity outperformed the naive model at 15 (79\%) timepoints (Table~\ref{table_natural_model_selection}).
 Epitope cross-immunity was also the only metric whose coefficient started at a positive value and transitioned to a negative value through the validation period.
 This change in coefficient suggests that positive selection may have weakened at these epitope sites over time.
-In contrast, HI phenotypes maintained a positive coefficient across most timepoints.
-The HI phenotypes metric may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
+In contrast, HI cross-immunity maintained a positive coefficient across most timepoints.
+The HI cross-immunity may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
 
-Non-epitope mutations also outperformed the DMS phenotypes metric, reducing the distance to the future by 0.40 amino acids on average compared to 0.19 amino acids, respectively.
-In contrast to \cite{Luksza:2014hj}, where the coefficient of the non-epitope mutations metric was hardcoded to -0.5, our model learned a consistently stronger coefficient of -1.45 +/- 0.24.
-Notably, the DMS metric only performed noticeably better than the naive model at the timepoint immediately preceding when the background strain of the DMS experiments, A/Perth/16/2009, was sampled.
-This result is consistent with the DMS model overfitting to the evolutionary history of the DMS background strain used to perform the DMS experiments.
+Non-epitope mutations also outperformed the DMS mutational effects, reducing the distance to the future by 0.40 amino acids on average compared to 0.19 amino acids, respectively.
+In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -1.45 +/- 0.24.
+Notably, DMS mutational effects only performed noticeably better than the naive model at the timepoint immediately preceding when the background strain of the DMS experiments, A/Perth/16/2009, was sampled.
+This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
 Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
-These results suggest that any mutation at a non-epitope site is deleterious for globally circulating strains, despite the more fine-grained fitness constraints of within-host infections.
+Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects.
 
@@
 These results suggest that when LBI forecasts correctly, it does so better than other metrics, but that it is prone to overfitting.
-Delta frequency did not surpass LBI in average estimated distance to the future and performed worse than both the HI phenotypes and non-epitope mutation models.
+Delta frequency did not surpass LBI in average estimated distance to the future and performed worse than both the HI cross-immunity and non-epitope mutation models.
 While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
@@
 To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
-Specifically, we fit models based on HI phenotypes and non-epitope mutations, LBI and non-epitope mutations, and all three of these metrics together.
+Specifically, we fit models based on HI cross-immunity and non-epitope mutations, LBI and non-epitope mutations, and all three of these metrics together.
 We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.
@@
 
-As expected, all of these combined models performed better than any of their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
+As expected, all of these composite models performed better than any of their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
 The model with all three metrics performed the best, reducing the distance to the future by 1.21 amino acids on average.
+This 19\% reduction of distance to the future relative to the naive model was nearly equal to the 20\% relative reduction of the best composite model for simulated populations.
 However, the composite of LBI and non-epitope mutations performed nearly as well with an average reduced distance of 1.14.
-Surprisingly, the combination of HI phenotypes and non-epitope mutations did not perform better than LBI by itself, despite the stability of the coefficients for both metrics across all timepoints.
+Both of these models were more accurate than the naive model for 79\% of validation timepoints.
+Although the composite of HI cross-immunity and non-epitope mutations outperformed the naive model more often than LBI, this composite model did not reduce the estimated distance to the future better on average than the individual LBI metric.
+Indeed, we observed that the contribution of HI cross-immunity to the best composite model declined after October 2010 when that metric's coefficient converged to zero (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
+These results reinforce the historical importance of HI assays in measuring viral fitness, while also indicating that these assays are no longer as predictive as sequence-only metrics.
 
@@
 As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
-We find that the combined model of LBI and non-epitope mutations accurately captures clade decline and modestly reflects observed clade growth (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
+We find that the composite model of LBI and non-epitope mutations accurately captures clade decline and modestly reflects observed clade growth (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
 
@@
   \caption{
-  Validation of best model for natural populations of A/H3N2 viruses, the combined model of LBI and non-epitope mutations.
+  Validation of best model for natural populations of A/H3N2 viruses, the composite model of LBI and non-epitope mutations.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.

### 2019-09-27 manuscript/main.tex
Expand methods and figures captions, update forecast example figure

@@
   \caption{
-    Schematic representation of the fitness model wherein the fitness of samples at timepoint $t$ is used to minimize the distance between projected samples from that timepoint and the observed population at timepoint $u$ one year in the future.
-    A) A phylogeny at timepoint $u=$April 2004 shows the observed population at that timepoint.
-    The vertical dashed line delineates the previous timepoint of $t=$April 2003 where the size and color of each sample indicates the projected frequency of those samples one year in the future based on the composite model of delta frequency, HI cross-immunity, and non-epitope mutations.
-    Bright yellow samples indicate those with the highest projected frequency and closest to the observed future population one year in the future.
+    Schematic representation of the fitness model for simulated A/H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
+    Samples are colored by their amino acid sequence composition such that genetically similar samples have similar colors.
+    A) Samples at timepoint $t$ are shown in their phylogenetic context and sized by their frequency at that timepoint.
+    The estimated future population at timepoint $u$ is projected to the right with samples scaled in size by their projected frequency based on the true fitness model.
+    B) The frequency trajectories of samples at timepoint $t$ to $u$ represent the predicted the growth of the orange samples to the detriment of the purple samples.
+    C) Samples at timepoint $u$ are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
+    D) The observed frequency trajectories of samples at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate (e.g., the emergence of the yellow cluster from within the successful orange cluster).
   }
@@
   \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
-  \caption{Model a) coefficients and b) accuracy for simulated populations of A/H3N2-like viruses.}
+  \caption{
+    Model a) coefficients and b) accuracy for simulated populations of A/H3N2-like viruses.
+    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) with the mean $\pm$ standard deviation in the top-left corner of each panel.
+    Distances to the future in amino acids are shown per validation timepoint for each individual model (black) in the context of the observed distance between timepoints from the naive model (grey).
+    Models outperform the naive model when the model's distance to the future is less than the naive model.
+    The mean $\pm$ standard deviation of amino acids per timepoint are shown in the top-left of each panel for both the model and the naive model.
+  }
   \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
@@
 
-The average distance between yearly populations, as measured by the naive model, was 8.97 +/- 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+The average distance between yearly populations, as measured by the naive model, was 8.97 $\pm$ 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
 As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.17 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection} and Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
@@
 Next, we trained and validated models for individual fitness predictors using 23 years of natural A/H3N2 populations spanning from October 1992 to October 2015.
-We held out samples collected between October 2015 and October 2019 for model testing.
+We held out samples collected between October 2015 and April 2019 for model testing.
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
@@
 
-\begin{figure*}[t]
+\begin{figure*}[ht]
   \begin{center}
   \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
-  \caption{Model a) coefficients and b) accuracy for natural populations of A/H3N2 viruses.}
+  \caption{
+    Model a) coefficients and b) accuracy for natural populations of A/H3N2 viruses.
+    As in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, coefficients are shown for each individual fitness metric per validation timepoint (N=19) with the mean $\pm$ standard deviation in the top-left corner of each panel.
+    Distances to the future in amino acids are shown per validation timepoint for each individual model (black) in the context of the observed distance between timepoints from the naive model (grey).
+    Models outperform the naive model when the model's distance to the future is less than the naive model.
+    The mean $\pm$ standard deviation of amino acids per timepoint are shown in the top-left of each panel for both the model and the naive model.
+  }
   \label{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}
@@
 
-The average distance per year between natural populations was 6.50 +/- 1.42 amino acids or 72\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
+The average distance per year between natural populations was 6.50 $\pm$ 1.42 amino acids or 72\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
 Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
@@
 Non-epitope mutations also outperformed the DMS mutational effects, reducing the distance to the future by 0.40 amino acids on average compared to 0.19 amino acids, respectively.
-In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -1.45 +/- 0.24.
+In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -1.45 $\pm$ 0.24.
 Notably, DMS mutational effects only performed noticeably better than the naive model at the timepoint immediately preceding when the background strain of the DMS experiments, A/Perth/16/2009, was sampled.
@@
 
-\begin{figure*}[t]
+\begin{figure*}[ht]
   \begin{center}
   \includegraphics[width=\textwidth]{figures/best-composite-unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
-  \caption{Composite model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.}
+  \caption{
+    Composite model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.
+    As in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}, coefficients are shown per timepoint for each individual fitness metric by the corresponding color of the metric.
+  }
   \label{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}
@@
 As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
-We find that the composite model of LBI and non-epitope mutations accurately captures clade decline and modestly reflects observed clade growth (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
+We find that the composite model of LBI and non-epitope mutations capture clade decline with 94\% accuracy and modestly reflects observed clade growth with 66\% accuracy (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
 
@@
 
-\begin{figure*}[t]
+\begin{figure*}[ht]
   \begin{center}
   \includegraphics[width=\textwidth]{figures/nextstrain-forecasts-for-september-2020.png}
-  \caption{Forecasts on nextstrain.org from our best model for September 2020.}
+  \caption{
+    Snapshot of live forecasts on nextstrain.org from our best model for September 2020.
+    The observed frequency trajectories for currently circulating clades are shown up to September 2019.
+    Our model favors the A1b/131K subclade, A1b/197R, to grow over the next year and estimates that the A1b/135K subclade, A1b/137F, will remain stable.
+  }
   \label{fig:nextstrain_forecasts}
@@
 
-Discussion.
+\begin{itemize}
+\item{Our model accurate forecasts evolution of simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
+\item{The combination of sequence-only metrics outperforms all experimentally-informed metrics}
+\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
+\item{Epitope mutations are an inadequate predictor of viral success compared to other antigenic or phylogenetic metrics}
+\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
+\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
+\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
+\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
+\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
+\end{itemize}
 
@@
 We modified the SANTA-SIM source code to enable the inclusion of true fitness values for each sample in the FASTA header of the sampled sequences from each generation.
-This modified implementation is available at https://github.com/huddlej/santa-sim/tree/emit-fitness.
+This modified implementation is available at \url{https://github.com/huddlej/santa-sim/tree/emit-fitness}.
+For our full analysis of model performance, we sampled 90 viruses per month to match the sampling density of natural populations.
+For tuning of hyperparameters, we sampled 10 viruses per month to enable rapid exploration of hyperparameter space.
+
+\subsection*{Hyperparameter tuning with simulated populations}
+
+To avoid overfitting our models to the relatively limited data from natural populations, we used simulated A/H3N2-like populations to tune hyperparameters including the KDE bandwidth for frequency estimates and the L1 penalty for model coefficients.
+We simulated populations, as described above, and fit models for each parameter value using the true fitness of samples from the simulator.
+
+We identified the optimal KDE bandwidth for frequencies as the value that minimized the difference between the mean distances to the future from the true fitness model and the naive model.
+We set the L1 lambda penalty to zero, to reduce variables in the analysis and avoid interactions between the coefficients and the KDE bandwidths.
+Higher bandwidths completely wash out dynamics of populations by making all samples appear to exist for long time periods.
+This flattening of frequency trajectories means that as bandwidths increase, the naive model gets more accurate and less informative.
+Given this behavior, we found the bandwidth that produced the minimum difference between distances to the future for the true fitness and naive models instead of the bandwidth that produced the minimum mean model distance.
+Based on this analysis, we identified an optimal bandwidth of $\frac{2}{12}$ or the equivalent of 2-months for floating point dates.
+Next, we identified an L1 penalty of 0.1 for model coefficients that minimized the mean distance to the future for the true fitness model.
 
@@
 
-For model training and validation, we selected XX HA sequences of length >=1,700 nucleotides that were sampled between October 1994 and October 2015.
-To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 10 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
-We excluded all samples with ambiguous year, month, and day annotations and prioritized samples with more available HI titer measurements.
-
-For model testing, we extended the set of training and validation sequences to include HA sequences that were sampled between October 2015 and April 2019.
+For model training and validation, we selected 13,568 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1992 and October 1, 2015.
+To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 90 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
+We excluded all egg-passaged samples and all samples with ambiguous year, month, and day annotations.
+We prioritized samples with more available HI titer measurements.
+For model testing, we selected an additional XX HA sequences that were sampled between October 1, 2015 and April 1, 2019.
 We used these test sequences to evaluate the out-of-sample error of fixed model parameters learned during training and validation.
@@
 To account for uncertainty in collection date and sampling error, we applied a kernel density estimation (KDE) approach to calculate global sample frequencies.
-Specifically, we constructed a Gaussian kernel for each sample with the mean at the reported collection date and a variance of one month, roughly corresponding to the average lag time between sample collection and submission to the GISAID database.
-We estimated the frequency of each sample at each timepoint by calculating the cumulative density function of each KDE between the current and previous timepoint and normalizing the resulting values to sum to one.
-We implemented this logic in the augur frequencies command.
+Specifically, we constructed a Gaussian kernel for each sample with the mean at the reported collection date and a variance (or KDE bandwidth) of two months.
+The bandwidth was identified by cross-validation, as described above.
+This bandwidth also roughly corresponds to the median lag time between sample collection and submission to the GISAID database.
+We estimated the frequency of each sample at each timepoint by calculating the probabilitiy density function of each KDE at that timepoint and normalizing the resulting values to sum to one.
+We implemented this frequency estimation logic in the augur frequencies command.
 
@@
 
-As in \cite{Luksza:2014hj}, we assumed that influenza A/H3N2 populations can be described by an exponential growth model.
+We assumed that the evolution seasonal influenza A/H3N2 populations can be represented by an exponential growth model, as previously described \cite{Luksza:2014hj}.
 Under this model, we estimated the future frequency of the global population, $\mathbf{\hat{x}}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
@@
 
-For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the mean weighted Hamming distance between the estimated and observed future populations at time $u = t + \Delta{t}$ as defined by,
+For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the weighted Hamming distance between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of samples at time $t$ by their estimated fitnesses.
+We measured distance between populations using Earth Mover's Distance (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{Rubner1998,Kusner2015}.
+Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
+This solution requires both a ``ground distance'' between pairs fo samples from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.
 
-$$
-\Delta\mathbf{d} = \frac{\mathbf{d}(\hat{u}, u) - \mathbf{d}(u, u)}{\mathbf{d}(t, u)}
-                                           = \frac{\sum_{i}\sum_{j}x_{i}(t)\exp(f_{i}(t)\Delta{t})x_{j}(u)d_{i,j} - \sum_{j}\sum_{k}x_{j}(u)x_{k}(u)d_{j,k}}{\sum_{i}\sum_{j}x_{i}(t)x_{j}(u)d_{i,j}}.
-$$
-
-The first term in the numerator measures the mean distance between estimated and observed future populations, the second term measures the average distance of the future population to itself, and the denominator measures the average distance between the current and future populations.
-When a model accurately estimates the average distance of the future population to itself, the metric evaluates to zero.
-When the estimated sequence diversity is closer to the center of the future population's observed diversity, the metric evaluates to a negative proportion.
+For each timepoint $t$ and corresponding timepoint $u = t + 1$, we defined the ground distance as the Hamming distance between HA amino acid sequences for all pairs of samples between timepoints.
+For samples with less than full length nucleotide sequences, we inferred missing nucleotides through TreeTime's ancestral sequence reconstruction analysis.
+We defined weights for samples at timepoint $t$ based on their projected future frequencies.
+We defined weights for samples at timepoint $u$ based on their observed frequencies.
+We then identified the fitness coefficients that provided projected future frequencies that minimized the EMD between the estimated and observed future populations.
+With this metric, an optimal estimate of the future would produce a distance of zero.
+However, the inevitable accumulation of substitutions between the two populations prevents this outcome.
+We calculated EMD with the Python bindings for the OpenCV 3.4.1 implementation \cite{opencv_library}.
 We applied the Nelder-Mead minimization algorithm as implemented in SciPy \cite{SciPy} to learn fitness coefficients that minimize the average of this distance metric over all timepoints in a given training window.
@@
 To obtain unbiased estimates for the out-of-sample errors of our models, we adopted the standard cross-validation strategy of training, validation, and testing.
-We divided our available data into an initial training and validation set spanning October 1994 to October 2015 and an additional testing set spanning October 2015 to April 2019.
+We divided our available data into an initial training and validation set spanning October 1992 to October 2015 and an additional testing set spanning October 2015 to April 2019.
 We partitioned our training and validation data into six month seasons corresponding to winter in the Northern Hemisphere (October--April) and the Southern Hemisphere (April--October) and trained models to estimate frequencies of populations one year into the future from each season in six-year sliding windows.

### 2019-09-30 manuscript/abstract.tex
Update key findings, fix figure placement, specify test set size

--- 2019-07-18 manuscript/abstract.tex
@@
-Key points:
+\section*{Key findings}
 
 \begin{itemize}
-\item{Phenotypic measurements of antigenic advance outperform genetic estimates from epitope mutations in long-term forecasts}
-\item{DMS measurements enable more precise identification of mutationally intolerant sites to replace previous sites based on all non-epitope regions}
-\item{Phylogenetic fitness metrics outperform and appear to be mutually exclusive of mechanistic metrics}
-\item{Distance-based model target performs as well or better than clade-based targets and enables future forecasting efforts where phylogenies are not feasible}
-\item{We provide (the first?) open source, extensible forecasting framework for influenza}
+\item{Our model accurately forecasts clade frequency trajectories and identifies optimal potential vaccine strains for both simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
+\item{Experimental measurements of antigenic drift based on HI assays are more robust and predictive of viral success than epitope mutations}
+\item{The combination of LBI and non-epitope mutation metrics outperforms all experimentally-informed metrics}
+\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
+\item{DMS measurements appear to be too background-specific to be predictive for global viral success}
+\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
+\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
+\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
+\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
+\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
 \end{itemize}
+
+\section*{Open questions}
+
+\begin{itemize}
+\item{How should forecast uncertainty be calculated?}
+\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance?}
+\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness?}
+\item{How does our model performance compare with the original clade-based models?}
+\item{Should we report both validation and test results? Should we report them together or separate?}
+\item{Should we report difference from naive as a primary statistic? As naive - model or as model - naive?}
+\end{itemize}

### 2019-09-30 manuscript/main.tex
Update key findings, fix figure placement, specify test set size

@@
 
-\begin{figure*}[ht]
+\begin{figure*}[ht!]
   \begin{center}
@@
 
+Key findings:
+
 \begin{itemize}
-\item{Our model accurate forecasts evolution of simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
+\item{Our model accurately forecasts clade frequency trajectories and identifies optimal potential vaccine strains for both simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
+\item{Experimental measurements of antigenic drift based on HI assays are more predictive of viral success than epitope mutations}
 \item{The combination of sequence-only metrics outperforms all experimentally-informed metrics}
 \item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
-\item{Epitope mutations are an inadequate predictor of viral success compared to other antigenic or phylogenetic metrics}
+\item{DMS measurements appear to be too background-specific to be predictive for global viral success}
 \item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
@@
 We prioritized samples with more available HI titer measurements.
-For model testing, we selected an additional XX HA sequences that were sampled between October 1, 2015 and April 1, 2019.
+For model testing, we selected an additional 20,609 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and April 1, 2019.
+This increased number of sequences corresponds with the trend of increased sample sequencing and deposition in the last decade.
 We used these test sequences to evaluate the out-of-sample error of fixed model parameters learned during training and validation.

### 2019-10-09 src/fit_model.py
Make delta months argument optional for model testing

@@ -963,7 +963,7 @@ if __name__ == "__main__":
     parser.add_argument("--tip-attributes", required=True, help="tab-delimited file describing tip attributes at all timepoints with standardized predictors")
     parser.add_argument("--output", required=True, help="JSON representing the model fit with training and cross-validation results, beta coefficients for predictors, and summary statistics")
     parser.add_argument("--predictors", nargs="+", help="tip attribute columns to use as predictors of final clade frequencies; optional if a fixed model is provided")
-    parser.add_argument("--delta-months", required=True, type=int, help="number of months to project clade frequencies into the future")
+    parser.add_argument("--delta-months", type=int, help="number of months to project clade frequencies into the future")
     parser.add_argument("--target", required=True, choices=["clades", "distances"], help="target for models to fit")
     parser.add_argument("--final-clade-frequencies", help="tab-delimited file of clades per timepoint and their corresponding tips and tip frequencies at the given delta time in the future")
     parser.add_argument("--distances", help="tab-delimited file of distances between pairs of samples")
@@ -1041,7 +1041,6 @@ if __name__ == "__main__":
         # make them appropriate as targets for the model.
         targets = tips.loc[:, ["strain", "timepoint", "frequency", "weighted_distance_to_present", "weighted_distance_to_future", "y"]].copy()
         targets["future_timepoint"] = targets["timepoint"]
-        targets["timepoint"] = targets["timepoint"] - pd.DateOffset(months=args.delta_months)
 
         model_class = DistanceExponentialGrowthModel
         distances = pd.read_csv(args.distances, sep="\t")
@@ -1084,6 +1083,7 @@ if __name__ == "__main__":
 
         # Calculate test errors/scores for the given coefficients and data at
         # the identified test timepoints.
+        targets["timepoint"] = targets["timepoint"] - pd.DateOffset(months=delta_months)
         scores = test(
             model_class,
             model_kwargs,
@@ -1133,6 +1133,7 @@ if __name__ == "__main__":
         else:
             coefficients = None
 
+        targets["timepoint"] = targets["timepoint"] - pd.DateOffset(months=delta_months)
         scores = cross_validate(
             model_class,
             model_kwargs,

### 2019-10-11 manuscript/main.tex
Fix typo in methods section heading
Start adding methods for antigenic drift fitness metrics

@@
 
-\subsection*{Phylogenetic interference}
+\subsection*{Phylogenetic inference}
 
@@
 
+We defined the following fitness metrics per strain and timepoint.
+
 \subsubsection*{Antigenic drift}
+
+We estimated antigenic drift for each strain using either genetic or HI data.
+To estimate antigenic drift with genetic data, we implemented the cross-immunity metrics originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+Briefly, for each pair of strains in adjacent seasons, we counted the number of amino acid differences between the strains' HA sequences at 49 epitope sites.
+We limited these pairwise comparisons to all strains sampled within the last five years from each timepoint.
+For each individual strain at each timepoint, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as follows.
+
+$$
+f_{i,\mathrm{ep}}(t) = \sum_{j: t_{j} < t_{i}}{x_{j}\exp{(-D_{\mathrm{ep}}(a_{i}, a_{j}) / D_{0})}}
+$$
+
+To estimate antigenic drift with HI data, we first applied the titer tree model to the phylogeny at a given timepoint and the corresponding HI data for its strains, as previously described by Neher et al.
+2016 \cite{Neher:2016hy}.
+This method effectively estimated the antigenic drift associated with each branch in the phylogeny.
+Next, we calculated the pairwise antigenic distance between strains as the sum of antigenic drift weights per branch on the phylogenetic path between each pair of strains.
+Finally, we calculated each strain's ability to escape cross-immunity by applying the same equation as above for epitope mutations where $D_{\mathrm{ep}}(a_{i}, a_{j})$ was replaced by the pairwise antigenic distance from the tree.
 

### 2019-10-15 manuscript/main.tex
Use equation blocks and start writing functional constraints methods

@@
 
-$$
+\begin{equation}
     \mathbf{\hat{x}}(t + \Delta{t}) = \frac{1}{Z(t)}\sum_{i}x_{i}(t)\exp(f_{i}(t))
-$$
+    \label{equation_exponential_growth_model}
+\end{equation}
 
 We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
-For example, the following equation estimates fitness per sample by epitope cross-immunity ($\mathrm{ep}$), non-epitope mutations ($\mathrm{ne}$), and local branching index ($\mathrm{lbi}$).
+For example, Equation~\ref{equation_fitness_estimation} estimates fitness per sample by non-epitope mutations ($\mathrm{ne}$) and local branching index ($\mathrm{lbi}$).
 
-$$
-f_{i}(t) = \beta_{\mathrm{ep}}f_{i, \mathrm{ep}}(t) + \beta_{\mathrm{ne}}f_{i, \mathrm{ne}}(t) + \beta_{\mathrm{lbi}}f_{i, \mathrm{lbi}}(t)
-$$
+\begin{equation}
+    f_{i}(t) = \beta_{\mathrm{ne}}f_{i, \mathrm{ne}}(t) + \beta_{\mathrm{lbi}}f_{i, \mathrm{lbi}}(t)
+    \label{equation_fitness_estimation}
+\end{equation}
 
@@
 We estimated antigenic drift for each strain using either genetic or HI data.
-To estimate antigenic drift with genetic data, we implemented the cross-immunity metrics originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+To estimate antigenic drift with genetic data, we implemented the cross-immunity metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
 Briefly, for each pair of strains in adjacent seasons, we counted the number of amino acid differences between the strains' HA sequences at 49 epitope sites.
 We limited these pairwise comparisons to all strains sampled within the last five years from each timepoint.
-For each individual strain at each timepoint, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as follows.
+For each individual strain $i$ at each timepoint $t$, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as follows.
 
-$$
+\begin{equation}
     f_{i,\mathrm{ep}}(t) = \sum_{j: t_{j} < t_{i}}{x_{j}\exp{(-D_{\mathrm{ep}}(a_{i}, a_{j}) / D_{0})}}
-$$
+    \label{equation_epitope_cross_immunity}
+\end{equation}
 
@@
 2016 \cite{Neher:2016hy}.
-This method effectively estimated the antigenic drift associated with each branch in the phylogeny.
+This method effectively estimates the antigenic drift per branch in units of $log_{2}$ titer change.
 Next, we calculated the pairwise antigenic distance between strains as the sum of antigenic drift weights per branch on the phylogenetic path between each pair of strains.
-Finally, we calculated each strain's ability to escape cross-immunity by applying the same equation as above for epitope mutations where $D_{\mathrm{ep}}(a_{i}, a_{j})$ was replaced by the pairwise antigenic distance from the tree.
+Finally, we calculated each strain's ability to escape cross-immunity using Equation~\ref{equation_epitope_cross_immunity} with the pairwise distances between epitope sequences replaced with pairwise antigenic distance from HI data.
+As with the original epitope cross-immunity described above, this HI cross-immunity metric produces higher values for strains that are more antigenically distinct from previously circulating strains.
 
@@
 
+We estimated functional constraint for each strain using either genetic or deep mutational scanning (DMS) data.
+To estimate functional constraint with genetic data, we implemented the non-epitope mutation metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+This metric counts the number of amino acid differences at 517 non-epitope sites in HA sequences between each strain $i$ at timepoint $t$ and that strain's most recent inferred ancestral sequence in the previous season ($t - 1$).
+
+We estimated functional constraint using mutational preferences from DMS data as previously defined \cite{Lee2018}.
+Briefly, mutational effects were defined as the log ratio of DMS preferences, $\pi$, at site $r$ for the derived amino acid, $a_{i}$, and the ancestral amino acid, $a_{j}$.
+
+\begin{equation}
+    f_{i,\mathrm{DMS}}(t) = \sum_{r \in r,a_{i} != r,a{j}}\log_{2}\frac{\pi_{r,a_{i}}}{\pi_{r,a_{j}}}
+    \label{equation_mutational_preference}
+\end{equation}
+
 \subsubsection*{Clade growth}

### 2019-10-15 manuscript/main.tex
Add/update figures and text for simulated and natural populations
Includes updated results for natural populations after removing outliers.

@@
 We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
-Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics.
+Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.
 In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which the exponential growth factor is set to one and populations do not change composition over the one year forecasting period.
-The distances estimated under the naive model represent the average distance between the current and future timepoints.
+The distances estimated under the naive model represent the observed distance between the current and future timepoints.
 
@@
   \caption{
-    Model a) coefficients and b) accuracy for simulated populations of A/H3N2-like viruses.
-    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) with the mean $\pm$ standard deviation in the top-left corner of each panel.
-    Distances to the future in amino acids are shown per validation timepoint for each individual model (black) in the context of the observed distance between timepoints from the naive model (grey).
-    Models outperform the naive model when the model's distance to the future is less than the naive model.
-    The mean $\pm$ standard deviation of amino acids per timepoint are shown in the top-left of each panel for both the model and the naive model.
+    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for simulated populations of A/H3N2-like viruses.
+    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) in solid circles with the mean $\pm$ standard deviation in the top-left corner of each panel.
+    Distances closer to the future in amino acids are shown per validation or test timepoint for each individual model relative to the observed distance between timepoints.
+    Models outperform the naive model when the distances closer to the future are greater than zero.
+    The mean $\pm$ standard deviation of amino acids per validation timepoint are shown in the top-left of each panel.
+    The same values are shown per test timepoint in the top-right of each panel.
+    Solid circles indicate validation timepoints and open circles indicate test timepoints when coefficients are fixed to their mean values from training/validation and applied to out-of-sample test data.
   }
@@
     \input{tables/simulated_model_selection.tex}
-    \caption{Simulated population model accuracy relative to the naive model}
+    \caption{
+      Model performance with A/H3N2-like populations relative to the naive model.
+      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
+      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
+      Better models produce estimates that are closer to the future population than the naive model.
+    }
     \label{table_simulated_model_selection}
@@
 The average distance between yearly populations, as measured by the naive model, was 8.97 $\pm$ 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
-As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.17 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection} and Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-With the exception of epitope cross-immunity, all biologically-informed models also outperformed the naive model.
-LBI was the best of these models, reducing the distance between populations by 1.41 amino acids on average.
+As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.16 $\pm$ 1.19 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection}).
+With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+LBI was the best of these models, reducing the distance between populations by 1.41 $\pm$ 1.30 amino acids on average.
 Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
-The non-epitope mutations metric received a consistently negative coefficient with an average improvement of 0.71 amino acids.
+The non-epitope mutations metric received a consistently negative coefficient with an average improvement of 0.71 $\pm$ 0.58 amino acids.
 Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric.
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.
 
-We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses.
-To this end, we fit an additional model including both LBI and non-epitope mutations.
-This composite model outperformed all individual metrics, reducing the distance between populations by 1.76 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-Interestingly, we found that the coefficients for LBI and non-epitope mutations remained relatively stable across all validation timepoints.
+We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses than models based on individual metrics.
+To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: non-epitope mutations and LBI.
+This composite model outperformed all individual metrics, reducing the distance between populations by 1.74 $\pm$ 1.02 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+The coefficients for LBI and non-epitope mutations remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
 These results support our hypothesis that multiple complementary metrics can produce more accurate models.
 
-Finally, we sought to validate and test the best performing model by two metrics relevant for practical influenza forecasting and vaccine design efforts.
-First, we measured the ability of the true fitness model to accurately estimate clade dynamics by correlating estimated and observed clade growth rates.
-Under these simulated conditions, our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.71 (p < 0.001)$ and overall accuracies for clade growth and decline predictions of 66\% and 61\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
-Next, we tested how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
+\begin{figure*}[ht]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.png}
+  \caption{
+  Test of best model for simulated populations of A/H3N2-like viruses using 10 years previously unobserved test data and fixed model coefficients.
+  a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  b) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
+  }
+  \label{fig:test_of_best_model_for_simulated_populations}
+  \end{center}
+\end{figure*}
+
+We sought to validate the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
+First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by correlating estimated and observed clade growth rates.
+Our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.72$~$(p < 0.001)$ and overall accuracies for clade growth and decline predictions of 87\% and 58\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
 The estimated best strain was in the top 4th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 82\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+
+Finally, we tested the true fitness metric model on out-of-sample data.
+Specifically, we fixed the coefficient of the true fitness model to the average value across the training/validation period and applied the resulting model to the next 10 years of previously unobserved simulated data.
+A standard expectation from machine learning is that models will perform worse on test data than on validation data.
+As expected, four of our five individual fitness metrics performed slightly worse on test data with non-epitope mutations as the exception (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+The true fitness metric remained the best individual model with an average distance closer to the future of 1.58 $\pm$ 1.01 amino acids.
+However, the composite model of non-epitope mutations and LBI outperformed the true fitness model on test data with an average distance closer to the future of 1.81 $\pm$ 1.19 amino acids.
+Next, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
+We observed a growth rate correlation of $R = 0.37$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 52\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
+The estimated best strain was in the top 5th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 70\% of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
 These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
+These results also highlight a well-established yet critical point that true forecasts on out-of-sample data are less accurate than those based on training and validation data.
 
-\subsection*{Model coefficients and performance reflect historical patterns of A/H3N2 evolution}
+\subsection*{Models reflect historical patterns of A/H3N2 evolution}
 
-Next, we trained and validated models for individual fitness predictors using 23 years of natural A/H3N2 populations spanning from October 1992 to October 2015.
-We held out samples collected between October 2015 and April 2019 for model testing.
+Next, we trained and validated models for individual fitness predictors using 25 years of natural A/H3N2 populations spanning from October 1, 1990 to October 1, 2015.
+We held out samples collected between October 1, 2015 and October 1, 2019 for model testing.
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
@@
   \caption{
-    Model a) coefficients and b) accuracy for natural populations of A/H3N2 viruses.
-    As in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, coefficients are shown for each individual fitness metric per validation timepoint (N=19) with the mean $\pm$ standard deviation in the top-left corner of each panel.
-    Distances to the future in amino acids are shown per validation timepoint for each individual model (black) in the context of the observed distance between timepoints from the naive model (grey).
-    Models outperform the naive model when the model's distance to the future is less than the naive model.
-    The mean $\pm$ standard deviation of amino acids per timepoint are shown in the top-left of each panel for both the model and the naive model.
+    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for natural populations of A/H3N2 viruses.
+    As in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, coefficients are shown for each individual fitness metric per validation timepoint (N=23) in solid circles with the mean $\pm$ standard deviation in the top-left corner of each panel.
+    The mean $\pm$ standard deviation of amino acids per validation and test timepoint are shown in the top-left and top-right of each panel, respectively.
   }
@@
     \input{tables/natural_model_selection.tex}
-    \caption{Natural population model accuracy relative to the naive model}
+    \caption{
+      Model performance with natural A/H3N2 populations relative to the naive model.
+      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
+      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
+      Better models produce estimates that are closer to the future population than the naive model.
+    }
     \label{table_natural_model_selection}
@@
 
-The average distance per year between natural populations was 6.50 $\pm$ 1.42 amino acids or 72\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
+The average distance per year between natural populations was 6.40 $\pm$ 1.36 amino acids or 71\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
 Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+The average improvement of these models over the naive model was consistently lower than the same models in simulated populations.
+This reduced performance may be caused by both the reduced diversity between years in natural populations and the increased complexity of the fitness constraints on these real populations.
+
 Surprisingly, the best antigenic fitness metric of HI cross-immunity performed only slightly better the best functional constraint metric of non-epitope mutations.
-Indeed, epitope cross-immunity only outperformed the naive model at one of 19 timepoints (5\%), while HI cross-immunity outperformed the naive model at 15 (79\%) timepoints (Table~\ref{table_natural_model_selection}).
+Indeed, epitope cross-immunity only outperformed the naive model at six of 19 timepoints (26\%), while HI cross-immunity outperformed the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
 Epitope cross-immunity was also the only metric whose coefficient started at a positive value and transitioned to a negative value through the validation period.
@@
 The HI cross-immunity may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
+Importantly, the 49 epitope sites were originally identified from a historical survey of sites with successful mutations up through 2005 \cite{Shih:2007bd}.
+The success of previous epitope models based on these sites could be due to overfitting to these historical data.
 
-Non-epitope mutations also outperformed the DMS mutational effects, reducing the distance to the future by 0.40 amino acids on average compared to 0.19 amino acids, respectively.
-In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -1.45 $\pm$ 0.24.
-Notably, DMS mutational effects only performed noticeably better than the naive model at the timepoint immediately preceding when the background strain of the DMS experiments, A/Perth/16/2009, was sampled.
+Non-epitope mutations also outperformed the DMS mutational effects, reducing the distance to the future by 0.26 $\pm$ 0.56 amino acids on average compared to -0.35 $\pm$ 1.65 amino acids, respectively.
+In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
+Notably, the best performance of the DMS mutational effects model was forecasting from April 2007 to April 2008 when the major clade containing A/Perth/16/2009 was first emerging.
 This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
@@
 
-LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-However, LBI only outperformed the naive model 63\% of the time (Table~\ref{table_natural_model_selection}).
-These results suggest that when LBI forecasts correctly, it does so better than other metrics, but that it is prone to overfitting.
-Delta frequency did not surpass LBI in average estimated distance to the future and performed worse than both the HI cross-immunity and non-epitope mutation models.
+LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and non-epitope mutations for the most timepoints outperforming the naive model of 17 (74\%) (Table~\ref{table_natural_model_selection}).
+Delta frequency performed worse than LBI and HI cross-immunity and only performed slightly better than non-epitope mutations.
 While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
@@
 
-As expected, all of these composite models performed better than any of their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
-The model with all three metrics performed the best, reducing the distance to the future by 1.21 amino acids on average.
-This 19\% reduction of distance to the future relative to the naive model was nearly equal to the 20\% relative reduction of the best composite model for simulated populations.
-However, the composite of LBI and non-epitope mutations performed nearly as well with an average reduced distance of 1.14.
-Both of these models were more accurate than the naive model for 79\% of validation timepoints.
-Although the composite of HI cross-immunity and non-epitope mutations outperformed the naive model more often than LBI, this composite model did not reduce the estimated distance to the future better on average than the individual LBI metric.
-Indeed, we observed that the contribution of HI cross-immunity to the best composite model declined after October 2010 when that metric's coefficient converged to zero (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
-These results reinforce the historical importance of HI assays in measuring viral fitness, while also indicating that these assays are no longer as predictive as sequence-only metrics.
+Both two-metric composite models performed better than their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
+Although the three-metric composite model performed slightly worse than the individual LBI model, as measured by distance closer to the future, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
+The composite of LBI and non-epitope mutations performed the best by distance closer to the future at 0.96 $\pm$ 1.42 amino acids.
+
+The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
+In contrast, the three-metric model strongly preferred the HI cross-immunity and non-epitope metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
+In the last two validation timepoints, this composite model converged to a nearly LBI-only model with the coefficients for the other two metrics converging to near-zero values.
+These last two timepoints span the period when an HA1:160T substitution swept through the global population.
+This substitution introduced a new glycosylation motif and dramatically reduced the efficiency of HI assays \cite{Zost2017}.
+These results reinforce the historical importance of HI assays in measuring viral fitness, while also indicating that these assays may no longer be as predictive as sequence-only metrics.
 
@@
 As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
-We find that the composite model of LBI and non-epitope mutations capture clade decline with 94\% accuracy and modestly reflects observed clade growth with 66\% accuracy (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
+The composite model of non-epitope mutations and LBI effectively captured clade growth and decline with 87\% and 89\% accuracy, respectively (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
+The estimated best strain from this model was in the top 13th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 70\% of timepoints (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}B).
 
@@
   \caption{
-  Validation of best model for natural populations of A/H3N2 viruses, the composite model of LBI and non-epitope mutations.
+  Validation of best model for natural populations of A/H3N2 viruses, the composite model of non-epitope mutations and LBI.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
 
+Finally, we tested the performance of this same best model on out-of-sample data collected from October 1, 2015 through April 1, 2019.
+
 \subsection*{Forecasts predict the rise of A1b/131K and A1b/135K sub-clades in September 2020}
 
-See Fig.~\ref{fig:nextstrain_forecasts}.
+After identifying the best model to forecast natural populations, we integrated our forecasting model into our existing open source pathogen surveillance framework, Nextstrain \cite{Hadfield2018}, and used our model to forecast A/H3N2 populations in September 2020.
+Our combined non-epitope mutations and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
 
@@
 
-For model training and validation, we selected 13,568 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1992 and October 1, 2015.
+For model training and validation, we selected 13,601 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1990 and October 1, 2015.
 To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 90 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
@@
 We prioritized samples with more available HI titer measurements.
-For model testing, we selected an additional 20,609 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and April 1, 2019.
+For model testing, we selected an additional 20,719 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and October 1, 2019.
 This increased number of sequences corresponds with the trend of increased sample sequencing and deposition in the last decade.
@@
 To obtain unbiased estimates for the out-of-sample errors of our models, we adopted the standard cross-validation strategy of training, validation, and testing.
-We divided our available data into an initial training and validation set spanning October 1992 to October 2015 and an additional testing set spanning October 2015 to April 2019.
+We divided our available data into an initial training and validation set spanning October 1990 to October 2015 and an additional testing set spanning October 2015 to April 2019.
 We partitioned our training and validation data into six month seasons corresponding to winter in the Northern Hemisphere (October--April) and the Southern Hemisphere (April--October) and trained models to estimate frequencies of populations one year into the future from each season in six-year sliding windows.

### 2019-10-15 manuscript/main.tex
Continue to write fitness metric methods
Adds very rough descriptions of clade growth metrics. All of the math in these
sections needs more attention.

@@
 Briefly, mutational effects were defined as the log ratio of DMS preferences, $\pi$, at site $r$ for the derived amino acid, $a_{i}$, and the ancestral amino acid, $a_{j}$.
+As with the non-epitope mutation metric above, we considered only substitutions in HA between each strain $i$ and that strain's most recent inferred ancestral sequence in the previous season.
+We calculated the total effect of these substitutions as the sum of the mutational preferences for each substitution, as in Equation~\ref{equation_mutational_preference}.
 
 \begin{equation}
-    f_{i,\mathrm{DMS}}(t) = \sum_{r \in r,a_{i} != r,a{j}}\log_{2}\frac{\pi_{r,a_{i}}}{\pi_{r,a_{j}}}
+    f_{i,\mathrm{DMS}}(t) = \sum_{r \in r,a_{i} != r,a_{j}}\log_{2}\frac{\pi_{r,a_{i}}}{\pi_{r,a_{j}}}
     \label{equation_mutational_preference}
@@
 \subsubsection*{Clade growth}
+
+We estimated clade growth for each strain using local branching index (LBI) and the change in frequency over time (delta frequency).
+To calculate LBI for each strain at each timepoint, we applied the LBI heuristic algorithm as originally described \cite{Neher:2014eu} to the phylogenetic tree constructed at each timepoint.
+We set the neighborhood parameter, $\tau$, to 0.3 and only considered viruses sampled in the last 6 months of each phylogeny as contributing to recent clade growth.
+
+We estimated the change in frequency over time by calculating clade frequencies under a Brownian motion diffusion process as previously described \cite{Lee2018}.
+These frequency calculations allowed us to assign a partial clade frequency to each strain within nested clades.
+We calculated the delta frequency as the change in frequency for each strain between the most recent timepoint in a given phylogeny and six months prior to that timepoint divided by 0.5 years.

### 2019-10-16 manuscript/main.tex
Add and reference complete model performance table for natural pops
Also clarifies that we tested all models with out-of-sample data and not just
our best model.

@@
 
-Finally, we tested the true fitness metric model on out-of-sample data.
-Specifically, we fixed the coefficient of the true fitness model to the average value across the training/validation period and applied the resulting model to the next 10 years of previously unobserved simulated data.
+Finally, we tested all of our models on out-of-sample data.
+Specifically, we fixed the coefficients of each model to the average values across the training/validation period and applied the resulting models to the next 10 years of previously unobserved simulated data.
 A standard expectation from machine learning is that models will perform worse on test data than on validation data.
@@
 However, the composite model of non-epitope mutations and LBI outperformed the true fitness model on test data with an average distance closer to the future of 1.81 $\pm$ 1.19 amino acids.
-Next, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
+Indeed, both composite models that we tested performed better on out-of-sample data than they did on validation data (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+
+As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
 We observed a growth rate correlation of $R = 0.37$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 52\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
@@
 This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
-Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
+Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection} and Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
 Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects.

### 2019-10-16 manuscript/supplement.tex
Add and reference complete model performance table for natural pops
Also clarifies that we tested all models with out-of-sample data and not just
our best model.

@@
 
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/complete_natural_model_selection.tex}
+    \caption{
+      Model performance with natural A/H3N2 populations relative to the naive model for all models tested.
+      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
+      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
+      Better models produce estimates that are closer to the future population than the naive model.
+    }
+    \label{sup_table:complete_natural_model_selection}
+  \end{center}
+\end{table*}
+
 \begin{figure*}[t]

### 2019-10-22 manuscript/abstract.tex
Address two previously open questions
We decided to report both validation and test results and to report the
difference between models and the naive model as the primary statistic.
Include edits and comments

@@
 \begin{itemize}
-\item{How should forecast uncertainty be calculated?}
-\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance?}
-\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness?}
-\item{How does our model performance compare with the original clade-based models?}
-\item{Should we report both validation and test results? Should we report them together or separate?}
-\item{Should we report difference from naive as a primary statistic? As naive - model or as model - naive?}
+\item{How should forecast uncertainty be calculated? \tbc{Important and will be commented on by reviewers. I think this can be carried by purely retrospective comparisons however.}}
+\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance? \tbc{I'm quite happy with Hamming distance.}}
+\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness?
+\tbc{This is still concerning.
+I wonder about doing a quick calculation of a tree that we output to count epitope mutations and non-epitope mutations to trunk vs side-branches.
+We should be able to replicate the previous results.
+Katherine was able to do this just now with modern data with the Wolf epitope sites.}}
+\item{How does our model performance compare with the original clade-based models? \tbc{I think the clade performance figure is enough here.}}
+\item{\tbc{Do we need to calculate LBI via a region-weighted scheme? May be better left for a later update? But seems important given centrality of LBI to our predictions.}}
 \end{itemize}

### 2019-10-22 manuscript/main.tex
Include edits and comments

@@
 The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
-Current vaccine predictions favor viruses that are distinct from prior vaccine viruses in the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
+Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
 The hemagglutination inhibition (HI) assay \cite{hirst1943studies} is used to measure the degree of cross-reactivity between pairs of circulating viruses.
@@
     C) Samples at timepoint $u$ are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
-    D) The observed frequency trajectories of samples at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate (e.g., the emergence of the yellow cluster from within the successful orange cluster).
+    D) The observed frequency trajectories of samples at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate, e.g.
+the emergence of the yellow cluster from within the successful orange cluster.
+    \tbc{I'd label left and right columns as `Forecast' and `Retrospective' of something to this effect.}
   }
@@
 
-Here, we present a model of seasonal influenza evolution inspired by the exponential growth model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+Here, we present a model of seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
 As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
@@
 Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model (Fig.~\ref{fig:model}).
+\tbc{I'd generally use the term `fitness model' instead of `exponential growth' model. Exponential growth makes me think of actual exponential growth rather than Malthusian growth.}
 With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
@@
   \caption{
-    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for simulated populations of A/H3N2-like viruses.
+    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for simulated populations.
     Coefficients are shown for each individual fitness metric per validation timepoint (N=33) in solid circles with the mean $\pm$ standard deviation in the top-left corner of each panel.
@@
     \caption{
-      Model performance with A/H3N2-like populations relative to the naive model.
+      Model performance with simulated populations relative to the naive model.
       Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
@@
       Better models produce estimates that are closer to the future population than the naive model.
+      \tbc{Confused as to why `epitope cross-immunity' isn't working in the simulated example.}
     }
@@
   \caption{
-  Test of best model for simulated populations of A/H3N2-like viruses using 10 years previously unobserved test data and fixed model coefficients.
+  Test of best model for simulated populations using 10 years previously unobserved test data and fixed model coefficients.
   a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
 Our combined non-epitope mutations and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
+\tbc{Include figure showing predicted match of strains to future population. Possibly compare Hamming match to FRA match.}
 

### 2019-10-23 manuscript/main.tex

@@
 
+\rnc{A paragraph explaining the metrics would help.}
+
 We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like non-epitope mutations, would receive negative coefficients.
@@
 The average distance between yearly populations, as measured by the naive model, was 8.97 $\pm$ 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
+\rnc{would be good to compare the year-on-year distance to diversity. that would give some sense whether a 2aa reduction is big or small.}
 As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.16 $\pm$ 1.19 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection}).

### 2019-10-25 manuscript/supplement.tex
Export mutations by trunk status as a table for manuscript

@@
   \includegraphics[width=\textwidth]{figures/distance-of-simulated-populations-between-timepoints.png}
-  \caption{Average weighted distance per year between simulated populations of viruses under the ``naive'' model.}
+  \caption{
+  Average weighted distance per year between simulated populations of viruses under the ``naive'' model.
+  \tbc{I'm not sure this figure is worth including.}
+  }
   \label{sup_fig:distance_of_simulated_populations_between_timepoints}
@@
 
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/mutations_by_trunk_status.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status.
+    Epitope sites were defined previously described \cite{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.
+    }
+    \label{sup_table:mutations_by_trunk_status}
+  \end{center}
+\end{table*}
+
 \begin{figure*}[t]

### 2019-10-25 manuscript/main.tex
Explicitly list the coordinates of epitope sites
Rather than providing these as a supplemental table with a single column, this
commit includes the coordinates directly in the methods.

@@
 Briefly, for each pair of strains in adjacent seasons, we counted the number of amino acid differences between the strains' HA sequences at 49 epitope sites.
-We limited these pairwise comparisons to all strains sampled within the last five years from each timepoint.
+The one-based coordinates of these sites relative to the start of the HA1 segment were 50, 53, 54, 121, 122, 124, 126, 131, 133, 135, 137, 142, 143, 144, 145, 146, 155, 156, 157, 158, 159, 160, 163, 164, 172, 173, 174, 186, 188, 189, 190, 192, 193, 196, 197, 201, 207, 213, 217, 226, 227, 242, 244, 248, 275, 276, 278, 299, and 307.
+We limited pairwise comparisons to all strains sampled within the last five years from each timepoint.
 For each individual strain $i$ at each timepoint $t$, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as follows.

### 2019-10-28 manuscript/main.tex
Added missing delta time parameter in forecasting formula
Thanks to Tal Einav for catching this error.
Add cross-validation schemes for simulated and natural data
Note that the cross-validation figure for natural data still only reflects
validation timepoints, as we have not yet analyzed the test data for these
populations.

@@
 The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods).
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Figure~\ref{sup_fig:cross_validation_for_simulated_populations}).
 These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
@@
 Next, we trained and validated models for individual fitness predictors using 25 years of natural A/H3N2 populations spanning from October 1, 1990 to October 1, 2015.
-We held out samples collected between October 1, 2015 and October 1, 2019 for model testing.
+We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Figure~\ref{sup_fig:cross_validation_for_natural_populations}).
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
@@
 \begin{equation}
-    \mathbf{\hat{x}}(t + \Delta{t}) = \frac{1}{Z(t)}\sum_{i}x_{i}(t)\exp(f_{i}(t))
+    \mathbf{\hat{x}}(t + \Delta{t}) = \frac{1}{Z(t)}\sum_{i}x_{i}(t)\exp(f_{i}(t)\Delta{t})
     \label{equation_exponential_growth_model}

### 2019-10-28 manuscript/supplement.tex
Add cross-validation schemes for simulated and natural data
Note that the cross-validation figure for natural data still only reflects
validation timepoints, as we have not yet analyzed the test data for these
populations.

@@
 \section*{Supplemental Material}
+
+\begin{figure*}[h]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/cross-validation-for-simulated-populations.png}
+  \caption{
+  Time-series cross-validation scheme for simulated populations.
+  Models were trained in six-year sliding windows (grey lines) and validated on out-of-sample data from validation timepoints (filled circles).
+  Validation results from 30 years of data were used to iteratively tune model hyperparameters.
+  After fixing hyperparameters, model coefficients were fixed at the mean values across all training windows.
+  Fixed coefficients were applied to 10 years of new out-of-sample test data (open circles) to estimate true forecast errors.
+  }
+  \label{sup_fig:cross_validation_for_simulated_populations}
+  \end{center}
+\end{figure*}
 
@@
   \label{sup_fig:validation_of_best_model_for_simulated_populations}
+  \end{center}
+\end{figure*}
+
+\begin{figure*}[h]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/cross-validation-for-natural-populations.png}
+  \caption{
+  Time-series cross-validation scheme for natural populations.
+  Models were trained in six-year sliding windows (grey lines) and validated on out-of-sample data from validation timepoints (filled circles).
+  Validation results from 25 years of data were used to iteratively tune model hyperparameters.
+  After fixing hyperparameters, model coefficients were fixed at the mean values across all training windows.
+  Fixed coefficients were applied to four years of new out-of-sample test data (open circles) to estimate true forecast errors.
+  }
+  \label{sup_fig:cross_validation_for_natural_populations}
   \end{center}

### 2019-10-28 manuscript/supplement.tex
Add epitope mutations by trunk status for simulated populations
To better understand why the epitope cross-immunity metric is not strongly
predictive of simulated flu evolution by itself, this commit includes the same
analysis I performed earlier for the natural populations where we count the
number of epitope and non-epitope mutations per branch based on the trunk or
side branch status of those branches. The resulting table gives us a sense of
how the relative importance of epitope mutations on the trunk vs. the side
branches (which we expect to be high) and also the absolute number of these

@@
     \caption{
-    Number of epitope and non-epitope mutations per branch by trunk or side branch status.
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
     Epitope sites were defined previously described \cite{Luksza:2014hj}.
@@
     \label{sup_table:mutations_by_trunk_status}
+  \end{center}
+\end{table*}
+
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/mutations_by_trunk_status_for_simulated_populations.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
+    Epitope sites were defined previously described \cite{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
+    }
+    \label{sup_table:mutations_by_trunk_status_for_simulated_populations}
   \end{center}

### 2019-10-28 manuscript/abstract.tex
Address some open questions and comments about terminology
Addressed open questions where Trevor signed off on existing content and also
question about substitution-based metrics where we have added supplemental
analyses to address the question. This commit also updates references to a
"fitness model" instead of an "exponential growth model".

@@
 \item{How should forecast uncertainty be calculated? \tbc{Important and will be commented on by reviewers. I think this can be carried by purely retrospective comparisons however.}}
-\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance? \tbc{I'm quite happy with Hamming distance.}}
-\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness?
-\tbc{This is still concerning.
-I wonder about doing a quick calculation of a tree that we output to count epitope mutations and non-epitope mutations to trunk vs side-branches.
-We should be able to replicate the previous results.
-Katherine was able to do this just now with modern data with the Wolf epitope sites.}}
-\item{How does our model performance compare with the original clade-based models? \tbc{I think the clade performance figure is enough here.}}
 \item{\tbc{Do we need to calculate LBI via a region-weighted scheme? May be better left for a later update? But seems important given centrality of LBI to our predictions.}}

### 2019-10-28 manuscript/main.tex
Address some open questions and comments about terminology
Addressed open questions where Trevor signed off on existing content and also
question about substitution-based metrics where we have added supplemental
analyses to address the question. This commit also updates references to a
"fitness model" instead of an "exponential growth model".

@@
 After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
-Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model (Fig.~\ref{fig:model}).
-\tbc{I'd generally use the term `fitness model' instead of `exponential growth' model. Exponential growth makes me think of actual exponential growth rather than Malthusian growth.}
+Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
 With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
@@
 While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric.
-From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.
+From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection rather than by positive selection at epitope sites.
 
@@
 
-We assumed that the evolution seasonal influenza A/H3N2 populations can be represented by an exponential growth model, as previously described \cite{Luksza:2014hj}.
+We assumed that the evolution seasonal influenza A/H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \cite{Luksza:2014hj}.
 Under this model, we estimated the future frequency of the global population, $\mathbf{\hat{x}}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.

### 2019-10-29 manuscript/main.tex
Address comments by Richard and Trevor
Includes relabeling columns in frequency forecasts of figure 1, addition of
paragraph describing fitness metrics, and addition of a figure showing weighted
distance to the future on Nextstrain.

@@
 the emergence of the yellow cluster from within the successful orange cluster.
-    \tbc{I'd label left and right columns as `Forecast' and `Retrospective' of something to this effect.}
   }
@@
 Here, we present a model of seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
-As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
+As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
 We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
+
 We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
+All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.
+The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating samples based on either genetic distance at epitope sites or HI distance.
+The functional constraint metrics consider effects of mutations accumulated since the previous season either for deleterious (non-epitope) mutations or all mutations based on their DMS preferences.
+The growth metrics estimate how successful populations of samples have been in the last six months based on either rapid branching in the phylogeny (LBI) or the change in clade frequencies over time (delta frequency).
+
 We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
-After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
+After selecting optimal models from training and validation, we evaluated the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
 Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
@@
 We fit models to these simulated populations using all sequence-only fitness metrics.
-
-\rnc{A paragraph explaining the metrics would help.}
+As a positive control for our model framework, we also fit a model based on the true fitness of each sample as measured by the simulator.
+We evaluated the performance of each model relative to the observed distance between timepoints.
 
@@
 Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.
-In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which the exponential growth factor is set to one and populations do not change composition over the one year forecasting period.
-The distances estimated under the naive model represent the observed distance between the current and future timepoints.
 
@@
 Our combined non-epitope mutations and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
-\tbc{Include figure showing predicted match of strains to future population. Possibly compare Hamming match to FRA match.}
+To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
 
@@
   \label{fig:nextstrain_forecasts}
+  \end{center}
+\end{figure*}
+
+\begin{figure*}[ht]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/nextstrain-weighted-distance-to-future-per-strain.png}
+  \caption{
+    Snapshot of the last two years of seasonal influenza A/H3N2 evolution on nextstrain.org showing the estimated distance per sample to the future population.
+    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating samples weighted by the other sample's projected frequencies under the best fitness model.
+  }
+  \label{fig:nextstrain_distance_to_future}
   \end{center}

### 2019-11-06 manuscript/supplement.tex
Removed unnecessary supplemental figures
Removes histograms of distances between timepoints

@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/distance-of-simulated-populations-between-timepoints.png}
-  \caption{
-  Average weighted distance per year between simulated populations of viruses under the ``naive'' model.
-  \tbc{I'm not sure this figure is worth including.}
-  }
-  \label{sup_fig:distance_of_simulated_populations_between_timepoints}
-  \end{center}
-\end{figure*}
-
-\begin{figure*}[t]
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.png}
@@
   \label{sup_fig:cross_validation_for_natural_populations}
-  \end{center}
-\end{figure*}
-
-\begin{figure*}[t]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/distance-of-natural-populations-between-timepoints.png}
-  \caption{Average weighted distance per year between natural populations of viruses under the ``naive'' model.}
-  \label{sup_fig:distance_of_natural_populations_between_timepoints}
   \end{center}

### 2019-11-06 manuscript/main.tex
Address confusion about model selection metrics for simulations
Reformats the model selection table for simulation results to report
coefficients, distances to the future, approximations of the future's diversity,
and how often the models are better than the naive model. The naive model is now
explicitly included in this table. The combination of these reported metrics
allows the text to be updated to clarify how much better some models are than
others based on different targets (absolute distance to future, diversity of
future, etc.).

This commit also updated the model performance figures fo

@@
 As a positive control for our model framework, we also fit a model based on the true fitness of each sample as measured by the simulator.
-We evaluated the performance of each model relative to the observed distance between timepoints.
+We evaluated the performance of each model by a) the absolute distance between its projected future populations and the observed future populations, b) the approximation of the observed future population's diversity, and c) how often each model outperformed a naive model that estimated no change in populations between years.
+
+\begin{figure*}[ht!]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations-controls.png}
+  \caption{
+    Model a) coefficients and b) distances between the model's projected population and the observed future population as measured in amino acids (AAs) for simulated populations.
+    Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
+    For model testing, coefficients were fixed to their mean values from training/validation (open circles) and applied to out-of-sample test data.
+    Distances between estimated and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
+    The mean $\pm$ standard deviation of distances per validation timepoint are shown in the top-left of each panel.
+    Corresponding values per test timepoint are in the top-right of each panel.
+    For reference, the naive model's distance to the future is also shown per validation and test timepoint in grey.
+  }
+  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}
+  \end{center}
+\end{figure*}
 
@@
 Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.
+
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/simulated_model_selection.tex}
+    \caption{
+      Model coefficients and validation performance for simulated populations ordered from best to worst by distance to the future.
+      Coefficients are the mean $\pm$ standard deviation for each metric in a given model across 33 training windows.
+      Distance to the future measures the distance in amino acids between estimated and observed future populations.
+      Approximation of the future diversity measures the difference between the average Hamming distance between the estimated and observed future populations and the future population's average Hamming distance to itself (i.e., the diversity of the future populations).
+      Models with values that are closer to or less than zero produce better approximations of the future.
+      The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
+    }
+    \label{table_simulated_model_selection}
+  \end{center}
+\end{table*}
+
+As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 of 33 (97\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
+Importantly, the true fitness model accurately approximated the diversity of future populations coming within -0.29 $\pm$ 2.01 AAs of observed future diversity.
+In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future and an average of 2.26 $\pm$ 1.75 AAs from the future's observed diversity.
+With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
+Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
+The non-epitope mutations metric received a consistently negative coefficient with an average distance of 8.27 $\pm$ 1.35 AAs.
+Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric.
+From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection rather than by positive selection at epitope sites.
 
@@
 \end{figure*}
-
-\begin{table*}[ht]
-  \begin{center}
-    \input{tables/simulated_model_selection.tex}
-    \caption{
-      Model performance with simulated populations relative to the naive model.
-      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
-      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
-      Better models produce estimates that are closer to the future population than the naive model.
-      \tbc{Confused as to why `epitope cross-immunity' isn't working in the simulated example.}
-    }
-    \label{table_simulated_model_selection}
-  \end{center}
-\end{table*}
-
-The average distance between yearly populations, as measured by the naive model, was 8.97 $\pm$ 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
-\rnc{would be good to compare the year-on-year distance to diversity. that would give some sense whether a 2aa reduction is big or small.}
-As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.16 $\pm$ 1.19 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection}).
-With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-LBI was the best of these models, reducing the distance between populations by 1.41 $\pm$ 1.30 amino acids on average.
-Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
-The non-epitope mutations metric received a consistently negative coefficient with an average improvement of 0.71 $\pm$ 0.58 amino acids.
-Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric.
-From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection rather than by positive selection at epitope sites.
 
@@
 
-The average distance per year between natural populations was 6.40 $\pm$ 1.36 amino acids or 71\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
+The average distance per year between natural populations was 6.40 $\pm$ 1.36 amino acids or 71\% of the distance between yearly simulated populations (Table~\ref{table_natural_model_selection}).
 Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).

### 2019-11-07 manuscript/main.tex
Simplify captions, update natural population tables and figures

@@
 Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric.
+While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection rather than by positive selection at epitope sites.
@@
   \caption{
-    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for simulated populations.
-    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) in solid circles with the mean $\pm$ standard deviation in the top-left corner of each panel.
-    Distances closer to the future in amino acids are shown per validation or test timepoint for each individual model relative to the observed distance between timepoints.
-    Models outperform the naive model when the distances closer to the future are greater than zero.
-    The mean $\pm$ standard deviation of amino acids per validation timepoint are shown in the top-left of each panel.
-    The same values are shown per test timepoint in the top-right of each panel.
-    Solid circles indicate validation timepoints and open circles indicate test timepoints when coefficients are fixed to their mean values from training/validation and applied to out-of-sample test data.
+    Model a) coefficients and b) distances to the future for individual biologically-informed fitness metrics and the best composite model fit to simulated populations.
+    Coefficients and distances are shown per validation and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }
@@
   \caption{
-    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for natural populations of A/H3N2 viruses.
-    As in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, coefficients are shown for each individual fitness metric per validation timepoint (N=23) in solid circles with the mean $\pm$ standard deviation in the top-left corner of each panel.
-    The mean $\pm$ standard deviation of amino acids per validation and test timepoint are shown in the top-left and top-right of each panel, respectively.
+    Model a) coefficients and b) distances to the future for individual biologically-informed fitness metrics fit to natural populations.
+    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=XX) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }
@@
     \caption{
-      Model performance with natural A/H3N2 populations relative to the naive model.
-      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
-      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
-      Better models produce estimates that are closer to the future population than the naive model.
+      Model coefficients and validation performance for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Results are based on training and validation across 23 training windows.
     }
@@
   \caption{
-    Composite model a) accuracy and b) coefficients for natural populations of A/H3N2 viruses.
-    As in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}, coefficients are shown per timepoint for each individual fitness metric by the corresponding color of the metric.
+    Model a) coefficients and b) distances to the future for composite fitness metrics fit to natural populations.
+    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=XX) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }

### 2019-11-07 manuscript/supplement.tex
Simplify captions, update natural population tables and figures

@@
     \caption{
-      Model performance with natural A/H3N2 populations relative to the naive model for all models tested.
-      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
-      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
-      Better models produce estimates that are closer to the future population than the naive model.
+      All model coefficients and validation performance for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Results are based on training and validation across 23 training windows.
+      Model results for additional variants of fitness metrics including those based on epitope mutations and DMS preferences are included for reference.
     }

### 2019-11-08 manuscript/main.tex
Switch text and figures to absolute distances to the future
It is easier to communicate absolute distances to the future in amino acids
instead of values relative to the naive model. This commit includes text
describing how well models approximate future diversity, to address Richard's
comment about how we know whether any of the absolute distances to the future
are better than others.

This commit also includes an expanded discussion of the epitope cross-immunity
results for simulated and natural populations to address why these metrics are
less predictive than we expect (and for different re

@@
 These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
+
 We fit models to these simulated populations using all sequence-only fitness metrics.
 As a positive control for our model framework, we also fit a model based on the true fitness of each sample as measured by the simulator.
-We evaluated the performance of each model by a) the absolute distance between its projected future populations and the observed future populations, b) the approximation of the observed future population's diversity, and c) how often each model outperformed a naive model that estimated no change in populations between years.
+We evaluated the performance of each model by the absolute distance between projected and observed future populations and how often it outperformed a naive model that estimated no change in populations between years.
+As each model only projects existing sequences forward one year without trying to predict which mutations occur in the future, the model's absolute distance to the future can never be zero.
+For this reason, we additionally measured how well each model approximated the diversity of the future population.
+We measured this approximation at each timepoint as the average Hamming distance between projected and observed populations minus the average Hamming distance of the observed population to itself.
 
@@
   \caption{
-    Model a) coefficients and b) distances between the model's projected population and the observed future population as measured in amino acids (AAs) for simulated populations.
+    Model a) coefficients and b) distances between projected and observed future populations as measured in amino acids (AAs) for simulated populations.
     Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
     For model testing, coefficients were fixed to their mean values from training/validation (open circles) and applied to out-of-sample test data.
-    Distances between estimated and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
+    Distances between projected and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
     The mean $\pm$ standard deviation of distances per validation timepoint are shown in the top-left of each panel.
-    Corresponding values per test timepoint are in the top-right of each panel.
+    Corresponding values per test timepoint are in the top-right.
     For reference, the naive model's distance to the future is also shown per validation and test timepoint in grey.
@@
 In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future and an average of 2.26 $\pm$ 1.75 AAs from the future's observed diversity.
-With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
 LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
@@
 The non-epitope mutations metric received a consistently negative coefficient with an average distance of 8.27 $\pm$ 1.35 AAs.
-Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
-From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection rather than by positive selection at epitope sites.
 
@@
 
+Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
+The cross-immunity fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
+To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
+Although this linear fitness metric produced a slightly stronger coefficient, it did not perform better than the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
+However, when we fit a composite model of the epitope ancestor and non-epitope mutations metrics, this model outperformed the individual non-epitope mutations model and produced a stronger coefficient for epitope mutations of 0.35 $\pm$ 0.07 (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly by a linear effect of positive selection at epitope sites.
+
 We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses than models based on individual metrics.
 To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: non-epitope mutations and LBI.
-This composite model outperformed all individual metrics, reducing the distance between populations by 1.74 $\pm$ 1.02 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+This composite model outperformed all individual metrics with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
 The coefficients for LBI and non-epitope mutations remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
@@
   \caption{
-  Test of best model for simulated populations using 10 years previously unobserved test data and fixed model coefficients.
+  Test of best model for simulated populations (true fitness) using 10 years previously unobserved test data and fixed model coefficients.
   a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
 A standard expectation from machine learning is that models will perform worse on test data than on validation data.
-As expected, four of our five individual fitness metrics performed slightly worse on test data with non-epitope mutations as the exception (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-The true fitness metric remained the best individual model with an average distance closer to the future of 1.58 $\pm$ 1.01 amino acids.
-However, the composite model of non-epitope mutations and LBI outperformed the true fitness model on test data with an average distance closer to the future of 1.81 $\pm$ 1.19 amino acids.
-Indeed, both composite models that we tested performed better on out-of-sample data than they did on validation data (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+The composite model of non-epitope mutations and LBI outperformed the true fitness metric with average distance to the future of 7.09 $\pm$ 1.15 compared to 7.32 $\pm$ 1.80, respectively (Supplemental Table~\ref{table_simulated_model_test}).
+This composite model also estimated a much less diverse future population on average than the observed future populations by 1.28 $\pm$ 1.96 AAs.
+In contrast, the true fitness model estimated future populations that were closer to the observed diversity.
+Both the composite and true fitness models outperformed the naive model at 95\% and 90\% of test timepoints, respectively.
 
@@
 \subsection*{Models reflect historical patterns of A/H3N2 evolution}
+
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/natural_model_selection.tex}
+    \caption{
+      Model coefficients and validation performance for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Results are based on training and validation across 23 training windows.
+    }
+    \label{table_natural_model_selection}
+  \end{center}
+\end{table*}
 
@@
 
-\begin{table*}[ht]
-  \begin{center}
-    \input{tables/natural_model_selection.tex}
-    \caption{
-      Model coefficients and validation performance for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
-      Results are based on training and validation across 23 training windows.
-    }
-    \label{table_natural_model_selection}
-  \end{center}
-\end{table*}
-
-The average distance per year between natural populations was 6.40 $\pm$ 1.36 amino acids or 71\% of the distance between yearly simulated populations (Table~\ref{table_natural_model_selection}).
+The naive model measured a distance of 6.40 $\pm$ 1.36 AAs between natural populations or 71\% of the distance between yearly simulated populations (Table~\ref{table_natural_model_selection}).
 Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
@@
 Surprisingly, the best antigenic fitness metric of HI cross-immunity performed only slightly better the best functional constraint metric of non-epitope mutations.
-Indeed, epitope cross-immunity only outperformed the naive model at six of 19 timepoints (26\%), while HI cross-immunity outperformed the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
-Epitope cross-immunity was also the only metric whose coefficient started at a positive value and transitioned to a negative value through the validation period.
+Indeed, epitope cross-immunity only outperformed the naive model at six of 23 timepoints (26\%), while HI cross-immunity outperformed the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
+Epitope cross-immunity was also the only metric whose coefficient started at a positive value (1.45 $\pm$ 0.10 on average prior to October 2009) and transitioned to a near-zero value through the validation period (-0.13 $\pm$ 0.32 on average for October 2009 and after).
 This change in coefficient suggests that positive selection may have weakened at these epitope sites over time.
@@
 The success of previous epitope models based on these sites could be due to overfitting to these historical data.
+For example, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict from October 2004 to October 2005, at which point it begins to overfit to the training data.
 
-Non-epitope mutations also outperformed the DMS mutational effects, reducing the distance to the future by 0.26 $\pm$ 0.56 amino acids on average compared to -0.35 $\pm$ 1.65 amino acids, respectively.
+Non-epitope mutations outperformed the DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
 In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
@@
 Both two-metric composite models performed better than their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
-Although the three-metric composite model performed slightly worse than the individual LBI model, as measured by distance closer to the future, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
-The composite of LBI and non-epitope mutations performed the best by distance closer to the future at 0.96 $\pm$ 1.42 amino acids.
+Although the three-metric composite model performed slightly worse than the individual LBI model, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
+The composite of LBI and non-epitope mutations performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs and an approximation of future diversity near zero (-0.02 $\pm$ 1.50 AAs).
 

### 2019-11-08 manuscript/supplement.tex
Switch text and figures to absolute distances to the future
It is easier to communicate absolute distances to the future in amino acids
instead of values relative to the naive model. This commit includes text
describing how well models approximate future diversity, to address Richard's
comment about how we know whether any of the absolute distances to the future
are better than others.

This commit also includes an expanded discussion of the epitope cross-immunity
results for simulated and natural populations to address why these metrics are
less predictive than we expect (and for different re

@@
 \end{figure*}
+
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/simulated_model_test.tex}
+    \caption{
+      Model coefficients and test performance for simulated populations ordered from best to worst by distance to the future.
+      Coefficients shown here are the mean values across all training and validation windows which were used to test models on out-of-sample data.
+      Distance to the future, approximation of future diversity, and number of times model outperforms the naive model are as in Table~\ref{table_simulated_model_selection} for all test timepoints (N=20).
+    }
+    \label{table_simulated_model_test}
+  \end{center}
+\end{table*}
 

### 2019-11-12 analyses/simulations/correct_metadata.py
Set higher lower bound for HA2 mutations
Improve reliability of the simulator's fitness values by preventing the best
residue at any HA2 site from being a low fitness value (e.g., 0.8). Also, fix
auspice exports to work with latest version of augur and slightly adjust the
speed of evolution from 100 generations per year to 150.

@@ -2,6 +2,6 @@ from augur.frequency_estimators import float_to_datestring
 import pandas as pd
 
 df = pd.read_table("original_metadata.tsv")
-df["num_date"] = 2000.0 + (df["generation"] / 100.0)
+df["num_date"] = 2000.0 + (df["generation"] / 150.0)
 df["date"] = df["num_date"].apply(float_to_datestring)
 df.to_csv("metadata.tsv", sep="\t", header=True, index=False)

### 2019-11-12 analyses/simulations/run.sh

@@ -34,6 +34,6 @@ augur reconstruct-sequences --tree tree.nwk --gene HA2 --mutations aa_muts.json 
 
 augur distance --tree tree.nwk --alignment aa-seq_SigPep.fasta aa-seq_HA1.fasta aa-seq_HA2.fasta --gene-names SigPep HA1 HA2 --attribute-name ep ne --compare-to root root --map ../../config/distance_maps/h3n2/ha/wolf.json ../../config/distance_maps/h3n2/ha/wolf_nonepitope.json --output distances.json
 
-augur export --tree tree.nwk --metadata metadata.tsv --auspice-config ../../config/auspice_config.json --colors ../../config/colors.tsv --output-tree test_tree.json --output-meta test_meta.json --node-data branch_lengths.json nt_muts.json aa_muts.json lbi.json distances.json --output-sequence test_seq.json
+augur export v1 --tree tree.nwk --metadata metadata.tsv --auspice-config ../../config/auspice_config.json --colors ../../config/colors.tsv --output-tree test_tree.json --output-meta test_meta.json --node-data branch_lengths.json nt_muts.json aa_muts.json lbi.json distances.json --output-sequence test_seq.json
 
 augur frequencies --method kde --metadata metadata.tsv --tree tree.nwk --output test_tip-frequencies.json

### 2019-11-13 manuscript/main.tex
Add references to trunk/side branch tables, remove DMS result figure
Tables with mutations in trunk vs. side branch are useful and should be included
as evidence of how the simulated populations match the natural
populations. There is no need for a separate figure showing DMS metric
coefficients and errors through time since we already have a table that
summarizes these.

@@
 To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Figure~\ref{sup_fig:cross_validation_for_simulated_populations}).
-These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
+These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}, Supplemental Tables~\ref{sup_table:mutations_by_trunk_status_for_simulated_populations} and \ref{sup_table:mutations_by_trunk_status}).
 
@@
 This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
-Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection} and Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
+Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
 Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects.

### 2019-11-13 manuscript/supplement.tex
Add references to trunk/side branch tables, remove DMS result figure
Tables with mutations in trunk vs. side branch are useful and should be included
as evidence of how the simulated populations match the natural
populations. There is no need for a separate figure showing DMS metric
coefficients and errors through time since we already have a table that
summarizes these.

@@
 \end{figure*}
+
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/mutations_by_trunk_status_for_simulated_populations.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
+    Epitope sites were defined previously described \cite{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
+    }
+    \label{sup_table:mutations_by_trunk_status_for_simulated_populations}
+  \end{center}
+\end{table*}
+
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/mutations_by_trunk_status.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
+    Epitope sites were defined previously described \cite{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.
+    }
+    \label{sup_table:mutations_by_trunk_status}
+  \end{center}
+\end{table*}
 
@@
 \end{table*}
-
-\begin{table*}[ht]
-  \begin{center}
-    \input{tables/mutations_by_trunk_status.tex}
-    \caption{
-    Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
-    Epitope sites were defined previously described \cite{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
-    Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.
-    }
-    \label{sup_table:mutations_by_trunk_status}
-  \end{center}
-\end{table*}
-
-\begin{table*}[ht]
-  \begin{center}
-    \input{tables/mutations_by_trunk_status_for_simulated_populations.tex}
-    \caption{
-    Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
-    Epitope sites were defined previously described \cite{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
-    Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
-    }
-    \label{sup_table:mutations_by_trunk_status_for_simulated_populations}
-  \end{center}
-\end{table*}
-
-\begin{figure*}[t]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-DMS-model-accuracy-and-coefficients-for-natural-populations.png}
-  \caption{
-  Primary and alternate fitness metrics based on deep mutational scanning (DMS) preferences for seasonal influenza A/H3N2 strain A/Perth/16/2009.
-  Metrics are increasingly less dependent on the sequence content of the DMS strain's genetic background from top to bottom.
-  The most structurally naive metric of non-epitope mutations outperforms all of the DMS metrics, indicating that these metrics still overfit to the DMS strain's genetic background.
-  }
-  \label{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}
-  \end{center}
-\end{figure*}

### 2019-11-13 scripts/annotate_model_tables.py
Annotate model JSONs with tip attributes
Collects redundant logic for annotating model JSONs into a Python script that
also supports annotating models by tip attribute data. Specifically, we annotate
the weighted distance to the future for each timepoint. This information is not
available for all timepoints in the validation/test scores, so we add it here.

@@ -1,0 +1,90 @@
+import argparse
+import json
+import pandas as pd
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--tip-attributes", required=True, help="tab-delimited file describing tip attributes at all timepoints with standardized predictors and weighted distances to the future")
+    parser.add_argument("--model", required=True, help="JSON representing the model fit with training and cross-validation results, beta coefficients for predictors, and summary statistics")
+    parser.add_argument("--errors-by-timepoint", help="data frame of cross-validation errors by validation timepoint")
+    parser.add_argument("--coefficients-by-timepoint", help="data frame of coefficients by validation timepoint")
+    parser.add_argument("--annotated-errors-by-timepoint", help="annotated model errors by timepoint")
+    parser.add_argument("--annotated-coefficients-by-timepoint", help="annotated model coefficients by timepoint")
+    parser.add_argument("--delta-months", type=int, help="number of months to project clade frequencies into the future")
+    parser.add_argument("--annotations", nargs="+", help="additional annotations to add to the output table in the format of 'key=value' pairs")
+
+    args = parser.parse_args()
+
+    # Load tip attributes to calculate within-timepoint diversity.
+    tips = pd.read_csv(args.tip_attributes, sep="\t", parse_dates=["timepoint"])
+
+    # Calculate weighted distance of each timepoint to itself.
+    tips["average_distance_to_present"] = tips["weighted_distance_to_present"] * tips["frequency"]
+    distances_to_present_per_timepoint = tips.groupby("timepoint")["average_distance_to_present"].sum().reset_index()
+
+    print(distances_to_present_per_timepoint.head())
+
+    # Load the model JSON to get access to projected frequencies for tips.
+    with open(args.model, "r") as fh:
+        model = json.load(fh)
+
+    # Collect all projected frequencies and weighted distances, to enable
+    # calculation of weighted average distances within and between seasons.
+    df = pd.concat([
+        pd.DataFrame(scores["validation_data"]["y_hat"])
+        for scores in model["scores"]
+    ])
+
+    # Prepare to calculate weighted distance of each timepoint's projected
+    # future to its observed future timepoint.
+    df["average_distance_to_future"] = df["weighted_distance_to_future"] * df["projected_frequency"]
+
+    # Sum the scaled weighted distances to get average distances per timepoint.
[…]

### 2019-11-18 manuscript/main.tex
Add analysis for historical contingency of epitope sites
Includes a description of and results for a new fitness metric, oracle
cross-immunity, based on knowledge from the future of which sites contain
beneficial mutations across the entire training/validation period. Results
include a supplemental figure comparing epitope cross-immunity and oracle
cross-immunity coefficients and distances through time and a new line in the
complete table of models fit to natural populations.

Based on the results of this new metric, this commit also substantially rewrites
the description of antigenic model pe

@@
 To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
-Although this linear fitness metric produced a slightly stronger coefficient, it did not perform better than the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
-However, when we fit a composite model of the epitope ancestor and non-epitope mutations metrics, this model outperformed the individual non-epitope mutations model and produced a stronger coefficient for epitope mutations of 0.35 $\pm$ 0.07 (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+This linear fitness metric did not outperform the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
+However, when we fit a composite model of the epitope ancestor and non-epitope mutations metrics, this model outperformed both the individual non-epitope mutations model and the composite cross-immunity and non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly by a linear effect of positive selection at epitope sites.
@@
 The average improvement of these models over the naive model was consistently lower than the same models in simulated populations.
-This reduced performance may be caused by both the reduced diversity between years in natural populations and the increased complexity of the fitness constraints on these real populations.
+This reduced performance may have been caused by both the reduced diversity between years in natural populations and the increased complexity of the fitness constraints on these real populations.
 
-Surprisingly, the best antigenic fitness metric of HI cross-immunity performed only slightly better the best functional constraint metric of non-epitope mutations.
-Indeed, epitope cross-immunity only outperformed the naive model at six of 23 timepoints (26\%), while HI cross-immunity outperformed the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
-Epitope cross-immunity was also the only metric whose coefficient started at a positive value (1.45 $\pm$ 0.10 on average prior to October 2009) and transitioned to a near-zero value through the validation period (-0.13 $\pm$ 0.32 on average for October 2009 and after).
-This change in coefficient suggests that positive selection may have weakened at these epitope sites over time.
-In contrast, HI cross-immunity maintained a positive coefficient across most timepoints.
-The HI cross-immunity may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
-Importantly, the 49 epitope sites were originally identified from a historical survey of sites with successful mutations up through 2005 \cite{Shih:2007bd}.
-The success of previous epitope models based on these sites could be due to overfitting to these historical data.
-For example, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict from October 2004 to October 2005, at which point it begins to overfit to the training data.
+Of the two metrics for antigenic drift, HI cross-immunity consistently outperformed epitope cross-immunity (Table~\ref{table_natural_model_selection}).
+HI cross-immunity estimated an average distance to the future of 6.04 $\pm$ 1.57 AAs and outperformed the naive model at 17 of 23 timepoints (74\%).
+After the first three training windows where HI measurements were relatively sparse, the HI cross-immunity coefficient was stable across all remaining timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+In contrast, epitope cross-immunity estimated a distance of 7.06 $\pm$ 1.37 AAs and only outperformed the naive model at six timepoints (26\%).
+Epitope cross-immunity was also the only metric whose coefficient started at a positive value (1.45 $\pm$ 0.10 on average prior to October 2009) and transitioned to a negative value through the validation period (-0.13 $\pm$ 0.32 on average for October 2009 and after).
+This strong coefficient for the first half of training windows indicated that, unlike the results for simulated populations, the nonlinear cross-immunity metric was historically an effective measure of antigenic drift.
+The historical importance of the epitope sites used for this metric was further supported by the relative enrichment of mutations at these sites for the most successful ``trunk'' lineages of natural populations compared to side branch lineages (Supplemental Table~\ref{sup_table:mutations_by_trunk_status}).
 
-Non-epitope mutations outperformed the DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
+These results led us to hypothesize that positive selection at these specific epitope sites has weakened over time.
+Importantly, these 49 epitope sites were originally identified by a historical survey from 1968 through 2005 of sites with successful mutations \cite{Shih:2007bd}.
+If the beneficial effects of mutations at these sites were due to historical contingency rather than a constant contribution to antigenic drift, we would expect models based on these sites to perform well until 2005 and then overfit relative to future data.
+Indeed, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict to October 2005, at which point it begins to overfit to the training data.
+To further test our hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
+Following the original approach of Shih et al. \cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population.
+Only 12 (48\%) of these sites also appeared in {\L}uksza and L\"assig's \cite{Luksza:2014hj} original set of 49 sites.
+We fit a cross-immunity model to these sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefitted from knowledge of the future in its forecasts.
+In contrast to the epitope cross-immunity model, our oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
+These results support our hypothesis that the success of previous epitope models based on a fixed set of 49 sites is due to historical contingency rather than a constant benefit of mutations at these sites.
+We suspect that our HI cross-immunity model benefits from its ability to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
+
+Of the two metrics for functional constraint, non-epitope mutations outperformed DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
 In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
@@
 Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
-Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects.
+Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects of a single strain.
 
@@
 
+To test the historical contingency of the epitope sites defined above, we additionally identified a new set of sites with beneficial mutations across the training/validation period of October 1990 through October 2015.
+Following the general approach of Shih et al. \cite{Shih:2007bd}, we manually identified 25 sites in HA1 where mutations rapidly swept through the global population.
+We required mutations to emerge from below 5\% global frequency and reach >90\% frequency.
+Although we did not require sweeps to complete within a fixed amount of time, we observed that they required no longer than 1-3 years to complete.
+To minimize false positives, we eliminated any sites where one or more mutations rose above 20\% frequency and subsequently died out.
+If two or more sites had redundant sweep dynamics (mutations emerging and fixing at the same times), we retained the site with the most mutational sweeps.
+Based on this requirements, we defined our final collection of ``oracle'' sites in HA1 coordinates as 3, 45, 48, 50, 75, 140, 145, 156, 158, 159, 173, 186, 189, 193, 198, 202, 212, 222, 223, 225, 226, 227, 278, 311, and 312.
+
 To estimate antigenic drift with HI data, we first applied the titer tree model to the phylogeny at a given timepoint and the corresponding HI data for its strains, as previously described by Neher et al.

### 2019-11-18 manuscript/supplement.tex
Add analysis for historical contingency of epitope sites
Includes a description of and results for a new fitness metric, oracle
cross-immunity, based on knowledge from the future of which sites contain
beneficial mutations across the entire training/validation period. Results
include a supplemental figure comparing epitope cross-immunity and oracle
cross-immunity coefficients and distances through time and a new line in the
complete table of models fit to natural populations.

Based on the results of this new metric, this commit also substantially rewrites
the description of antigenic model pe

@@
 \end{table*}
+
+\begin{figure*}[h]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.png}
+  \caption{
+  Model a) coefficients and b) distances to the future for cross-immunity models fit to natural populations.
+  The epitope cross-immunity model relies on previously published epitope sites \cite{Luksza:2014hj}.
+  The ``oracle'' cross-immunity model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period.
+  The improved performance of the ``oracle'' model indicates that the cross-immunity can be effective when sites of beneficial mutations are known prior to forecasting.
+  Coefficients and distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+  }
+  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
+  \end{center}
+\end{figure*}

### 2019-11-18 manuscript/main.tex
Clarify figure captions for model and validation results

@@
   a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  b) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
+  b) The percentile rank in observed distance to the future of the model's estimated best strain for 20 test timepoints.
+  The estimated best strain is in the top 20th percentile of observed closest strains for 90\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
   }

### 2019-11-18 manuscript/supplement.tex
Clarify figure captions for model and validation results

@@
   \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.png}
-  \caption{Composite model a) accuracy and b) coefficients for simulated populations of A/H3N2-like viruses.}
+  \caption{
+  Composite model a) coefficients and b) distances to the future for models fit to simulated populations.
+  Coefficients and distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+  }
   \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}
@@
   a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  b) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
+  b) The percentile rank in observed distance to the future of the model's estimated best strain for 33 validation timepoints.
+  The estimated best strain is in the top 20th percentile of observed closest strains for 97\% of timepoints.
   }

### 2019-11-18 manuscript/main.tex
Update best model validation/test rank results after fixing bug
After fixing figures in an earlier commit where the estimated best strain at
each timepoint was being selected by fitness instead of estimated distance to
the future, this commit updates the corresponding text and also improves the
aesthetics of the y-axis tick labels for the simulated population test results
figure.

@@
 Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
-The estimated best strain was in the top 4th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 82\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+The estimated best strain was in the top 2nd percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 97\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
 
@@
 We observed a growth rate correlation of $R = 0.37$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 52\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
-The estimated best strain was in the top 5th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 70\% of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
+The estimated best strain was in the top 2nd percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 90\% of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
 These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.

### 2019-11-18 manuscript/main.tex
Update text to reflect use of Oct 2019 endpoint for testing
Now that there are enough samples up through October 2019, we can use this
timepoint for testing models fit to natural populations. This commit also
reduces some verbosity in the main text.

@@
 
-We sought to validate the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
+We validated the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
 First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by correlating estimated and observed clade growth rates.
@@
 
-Finally, we tested the performance of this same best model on out-of-sample data collected from October 1, 2015 through April 1, 2019.
+Finally, we tested the performance of this same best model on out-of-sample data collected from October 1, 2015 through October 1, 2019.
 
@@
 We prioritized samples with more available HI titer measurements.
-For model testing, we selected an additional 20,719 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and October 1, 2019.
+For model testing, we selected an additional 20,803 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and October 1, 2019.
 This increased number of sequences corresponds with the trend of increased sample sequencing and deposition in the last decade.
@@
 To obtain unbiased estimates for the out-of-sample errors of our models, we adopted the standard cross-validation strategy of training, validation, and testing.
-We divided our available data into an initial training and validation set spanning October 1990 to October 2015 and an additional testing set spanning October 2015 to April 2019.
+We divided our available data into an initial training and validation set spanning October 1990 to October 2015 and an additional testing set spanning October 2015 to October 2019.
 We partitioned our training and validation data into six month seasons corresponding to winter in the Northern Hemisphere (October--April) and the Southern Hemisphere (April--October) and trained models to estimate frequencies of populations one year into the future from each season in six-year sliding windows.

### 2019-11-18 manuscript/main.tex
Clarify more subtle details of the epitope cross-immunity analysis
We aren't technically measuring positive selection, so refer to antigenic drift
instead. We also aren't following the Shih et al. method exactly as we don't
need to recreate the specifics of that entire paper in this one, but our
approach is inspired by theirs.

@@
 
-These results led us to hypothesize that positive selection at these specific epitope sites has weakened over time.
+These results led us to hypothesize that the contribution of these specific epitope sites to antigenic drift has weakened over time.
 Importantly, these 49 epitope sites were originally identified by a historical survey from 1968 through 2005 of sites with successful mutations \cite{Shih:2007bd}.
@@
 To further test our hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
-Following the original approach of Shih et al. \cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population.
+Inspired by the original approach of Shih et al. \cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population.
 Only 12 (48\%) of these sites also appeared in {\L}uksza and L\"assig's \cite{Luksza:2014hj} original set of 49 sites.
-We fit a cross-immunity model to these sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefitted from knowledge of the future in its forecasts.
-In contrast to the epitope cross-immunity model, our oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
-These results support our hypothesis that the success of previous epitope models based on a fixed set of 49 sites is due to historical contingency rather than a constant benefit of mutations at these sites.
+We fit a cross-immunity model to these 25 sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefitted from knowledge of the future in its forecasts.
+In contrast to the epitope cross-immunity model, the oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
+These results support our hypothesis that the success of previous epitope models based on a fixed set of 49 sites is due to historical contingency rather than a constant contribution of these sites to antigenic drift.
 We suspect that our HI cross-immunity model benefits from its ability to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.

### 2019-11-18 src/fit_model.py
Reduce memory required to read in pairwise distances
Instead of reading distances into memory as a DataFrame and then converting them
in memory to a dictionary with redundant pairwise information, read distances
with a simple iterator and convert them to a nonredundant dictionary.

@@ -1,6 +1,7 @@
 """Fit a model for the given data using the requested predictors and evaluate the model by time series cross-validation.
 """
 import argparse
+import csv
 import cv2
 import json
 import numpy as np
@@ -1043,8 +1044,14 @@ if __name__ == "__main__":
         targets["future_timepoint"] = targets["timepoint"]
 
         model_class = DistanceExponentialGrowthModel
-        distances = pd.read_csv(args.distances, sep="\t")
-        distances_by_sample_names = get_distances_by_sample_names(distances)
+
+        with open(args.distances, "r") as fh:
+            print("Read distances", flush=True)
+            reader = csv.DictReader(fh, delimiter="\t")
+            print("Get distances by sample names", flush=True)
+            distances_by_sample_names = get_distances_by_sample_names(reader)
+            print("Data loaded", flush=True)
+
         model_kwargs = {"distances": distances_by_sample_names}
         group_by_attribute = "strain"
 

### 2019-11-18 src/forecast_model.py

@@ -1,6 +1,7 @@
 """Forecast given tip data into future using the given previously trained model.
 """
 import argparse
+import csv
 import json
 import numpy as np
 import pandas as pd
@@ -43,8 +44,9 @@ if __name__ == "__main__":
     )
 
     # Load distances.
-    distances = pd.read_csv(args.distances, sep="\t")
-    distances_by_sample_names = get_distances_by_sample_names(distances)
+    with open(args.distances, "r") as fh:
+        reader = csv.DictReader(fh, delimiter="\t")
+        distances_by_sample_names = get_distances_by_sample_names(reader)
 
     # Load model details
     with open(args.model, "r") as fh:

### 2019-11-18 src/weighted_distances.py

@@ -1,6 +1,8 @@
 """Calculate weighted distances between samples in a given timepoint and both other samples in that timepoint and samples from a timepoint at a given delta time in the future.
 """
 import argparse
+from collections import defaultdict
+import csv
 import numpy as np
 import pandas as pd
 import sys
@@ -11,24 +13,20 @@ def get_distances_by_sample_names(distances):
 
     Parameters
     ----------
-    distances : pandas.DataFrame
-        data frame with the columns distance, sample, and other_sample
+    distances : iterator
+        an iterator of dictionaries with keys of distance, sample, and other_sample
 
     Returns
     -------
     dict :
         dictionary of distances by pairs of sample names
     """
-    distances_by_sample_names = {}
-    for distance, sample_b, sample_a in distances.loc[:, ["distance", "other_sample", "sample"]].values:
-        if sample_a not in distances_by_sample_names:
-            distances_by_sample_names[sample_a] = {}
-
-        if sample_b not in distances_by_sample_names:
-            distances_by_sample_names[sample_b] = {}
-
+    distances_by_sample_names = defaultdict(dict)
+    for record in distances:
+        sample_a = record["sample"]
+        sample_b = record["other_sample"]
+        distance = int(record["distance"])
         distances_by_sample_names[sample_a][sample_b] = distance
-        distances_by_sample_names[sample_b][sample_a] = distance
 
     return distances_by_sample_names
 
@@ -83,10 +81,11 @@ if __name__ == '__main__':
     tips = pd.read_csv(args.tip_attributes, sep="\t", parse_dates=["timepoint"])
 
     # Load distances.
-    distances = pd.read_csv(args.distances, sep="\t")
+    with open(args.distances, "r") as fh:
+        reader = csv.DictReader(fh, delimiter="\t")
 
-    # Map distances by sample names.
-    distances_by_sample_names = get_distances_by_sample_names(distances)
+        # Map distances by sample names.
+        distances_by_sample_names = get_distances_by_sample_names(reader)
 
     # Find valid timepoints for calculating distances to the future.
     timepoints = tips["timepoint"].drop_duplicates()

### 2019-11-18 src/fit_model.py
Initialize coefficients from prior values plus a random offset
If no coefficients have been fit for a model yet, use a random value between
-0.5 and 0.5 instead of a random value between 0 and 1. If coefficients have
been fit, use that prior information plus a small random offset. This latter
change speeds up convergence for slow-running processes but also prevents models
from getting stuck in local maxima.

@@ -283,9 +283,12 @@ class ExponentialGrowthModel(object):
 
         # Find coefficients that minimize the model's cost function.
         if hasattr(self, "coef_"):
-            initial_coefficients = self.coef_
+            # Use the previous coefficients +/- a small random offset (+/- 0.05)
+            # to prevent getting stuck in local minima.
+            initial_coefficients = self.coef_ + (0.1 * np.random.random(len(self.predictors)) - 0.05)
         else:
-            initial_coefficients = np.random.random(len(self.predictors))
+            # If no previous coefficients exist, sample random values between -0.5 and 0.5.
+            initial_coefficients = np.random.random(len(self.predictors)) - 0.5
 
         results = minimize(
             self._fit,

### 2019-11-18 src/fit_model.py
Include time to fit each timepoint in debugging output

@@ -8,6 +8,7 @@ import numpy as np
 import pandas as pd
 from scipy.optimize import minimize
 import sys
+import time
 
 from forecast.fitness_model import get_train_validate_timepoints
 from forecast.metrics import add_pseudocounts_to_frequencies, negative_information_gain
@@ -680,10 +681,13 @@ def cross_validate(model_class, model_kwargs, data, targets, train_validate_time
 
         # Fit a model to the training data.
         if coefficients is None:
+            start_time = time.time()
             training_error = model.fit(training_X, training_y)
+            end_time = time.time()
             previous_coefficients = model.coef_
             null_training_error = model._fit(np.zeros_like(model.coef_), training_X, training_y)
         else:
+            start_time = end_time = time.time()
             model.coef_ = coefficients
             model.mean_stds_ = model.calculate_mean_stds(training_X, model.predictors)
             training_error = model.score(training_X, training_y)
@@ -698,7 +702,7 @@ def cross_validate(model_class, model_kwargs, data, targets, train_validate_time
         null_validation_error = model._fit(np.zeros_like(model.coef_), validation_X, validation_y)
         differences_of_model_and_naive_errors.append(validation_error - null_validation_error)
         print(
-            "%s\t%s\t%.2f\t%.2f\t%.2f\t%.2f\t%s\t%.2f" % (
+            "%s\t%s\t%.2f\t%.2f\t%.2f\t%.2f\t%s\t%.2f\t%.2f" % (
                 training_timepoints[-1].strftime("%Y-%m"),
                 validation_timepoint.strftime("%Y-%m"),
                 training_error,
@@ -706,7 +710,8 @@ def cross_validate(model_class, model_kwargs, data, targets, train_validate_time
                 validation_error,
                 null_validation_error,
                 model.coef_,
-                (np.array(differences_of_model_and_naive_errors) < 0).sum() / float(len(differences_of_model_and_naive_errors))
+                (np.array(differences_of_model_and_naive_errors) < 0).sum() / float(len(differences_of_model_and_naive_errors)),
+                end_time - start_time
             ),
             flush=True
         )

### 2019-12-05 manuscript/abstract.tex
Add results on test data to key points in abstract

@@
 \item{Experimental measurements of antigenic drift based on HI assays are more robust and predictive of viral success than epitope mutations}
-\item{The combination of LBI and non-epitope mutation metrics outperforms all experimentally-informed metrics}
+\item{Models based on a predefined set of epitope sites risk overfitting to historical data}
+\item{The combination of non-epitope mutation and LBI metrics outperforms all experimentally-informed metrics on validation data (Oct 1990--Oct 2015) but fails to account for reassortment events in test data (Oct 2015--Oct 2019)}
+\item{The combination of HI assays and non-epitope mutations performs best on test data and is among the best models for validation data}
 \item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}

### 2019-12-05 manuscript/main.tex
Update model results tables including test results
Removes approximation of future diversity from tables and text in favor of a
later summary of these statistics in the corresponding results sections. Also
adds the files for model test results on natural populations.

@@
       Distance to the future measures the distance in amino acids between estimated and observed future populations.
-      Approximation of the future diversity measures the difference between the average Hamming distance between the estimated and observed future populations and the future population's average Hamming distance to itself (i.e., the diversity of the future populations).
-      Models with values that are closer to or less than zero produce better approximations of the future.
       The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
@@
 Although the three-metric composite model performed slightly worse than the individual LBI model, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
-The composite of LBI and non-epitope mutations performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs and an approximation of future diversity near zero (-0.02 $\pm$ 1.50 AAs).
+The composite of LBI and non-epitope mutations performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
 

### 2019-12-05 manuscript/supplement.tex
Update model results tables including test results
Removes approximation of future diversity from tables and text in favor of a
later summary of these statistics in the corresponding results sections. Also
adds the files for model test results on natural populations.

@@
 \end{figure*}
+
+\begin{table*}[ht]
+  \begin{center}
+    \input{tables/complete_natural_model_test.tex}
+    \caption{
+      All model coefficients and test performance for natural populations ordered from best to worst by distance to the future, as in Supplemental Table~\ref{table_simulated_model_test}.
+      Results are based on 8 test timepoints.
+    }
+    \label{sup_table:complete_natural_model_test}
+  \end{center}
+\end{table*}

### 2019-12-05 manuscript/main.tex
Clarification of text and correction of summary statistics
Attempts to clarify descriptions of fitness metrics including how epitope
sites were originally designed and fix structure of equations.

@@
 All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.
-The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating samples based on either genetic distance at epitope sites or HI distance.
-The functional constraint metrics consider effects of mutations accumulated since the previous season either for deleterious (non-epitope) mutations or all mutations based on their DMS preferences.
+The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating samples based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
+Increased antigenic drift relative to previously circulating samples is expected to correspond to increased viral fitness.
+Non-epitope mutations are the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each sample since their ancestor in the previous season.
+DMS mutational effects provide a more comprehensive biophysical model of functional constraint by measuring the beneficial or deleterious effect of each possible single amino acid mutation in HA from the background of a previous vaccine strain, A/Perth/16/2009.
 The growth metrics estimate how successful populations of samples have been in the last six months based on either rapid branching in the phylogeny (LBI) or the change in clade frequencies over time (delta frequency).
@@
 However, when we fit a composite model of the epitope ancestor and non-epitope mutations metrics, this model outperformed both the individual non-epitope mutations model and the composite cross-immunity and non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly by a linear effect of positive selection at epitope sites.
+From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
 
@@
 Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-The average improvement of these models over the naive model was consistently lower than the same models in simulated populations.
-This reduced performance may have been caused by both the reduced diversity between years in natural populations and the increased complexity of the fitness constraints on these real populations.
+The average improvement of the sequence-only models over the naive model was consistently lower than the same models in simulated populations.
+This reduced performance may have been caused by both the relatively reduced diversity between years in natural populations and the increased complexity of the fitness constraints on these real populations.
 
@@
 These results led us to hypothesize that the contribution of these specific epitope sites to antigenic drift has weakened over time.
-Importantly, these 49 epitope sites were originally identified by a historical survey from 1968 through 2005 of sites with successful mutations \cite{Shih:2007bd}.
+Importantly, these 49 epitope sites were originally selected by {\L}uksza and L\"assig \cite{Luksza:2014hj} from a previous historical survey of sites with beneficial mutations between 1968--2005 \cite{Shih:2007bd}.
 If the beneficial effects of mutations at these sites were due to historical contingency rather than a constant contribution to antigenic drift, we would expect models based on these sites to perform well until 2005 and then overfit relative to future data.
-Indeed, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict to October 2005, at which point it begins to overfit to the training data.
+Indeed, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict to October 2005.
 To further test our hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
-Inspired by the original approach of Shih et al. \cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population.
-Only 12 (48\%) of these sites also appeared in {\L}uksza and L\"assig's \cite{Luksza:2014hj} original set of 49 sites.
-We fit a cross-immunity model to these 25 sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefitted from knowledge of the future in its forecasts.
+Inspired by the original approach of Shih et al.
+\cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population, including 12 that were also present in the original set of 49 sites.
+We fit a cross-immunity model to these 25 sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefited from knowledge of the future in its forecasts.
 In contrast to the epitope cross-immunity model, the oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
@@
 The composite model of non-epitope mutations and LBI effectively captured clade growth and decline with 87\% and 89\% accuracy, respectively (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
-The estimated best strain from this model was in the top 13th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 70\% of timepoints (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}B).
+The estimated best strain from this model was in the top 11th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 78\% of timepoints (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}B).
 
@@
 We assumed that the evolution seasonal influenza A/H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \cite{Luksza:2014hj}.
-Under this model, we estimated the future frequency of the global population, $\mathbf{\hat{x}}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
+Under this model, we estimated the future frequency, $\hat{x}_{i}(t + \Delta{t})$, of each sample $i$ from the sample's current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
 
 \begin{equation}
-    \mathbf{\hat{x}}(t + \Delta{t}) = \frac{1}{Z(t)}\sum_{i}x_{i}(t)\exp(f_{i}(t)\Delta{t})
+    \hat{x}_{i}(t + \Delta{t}) = \frac{1}{Z(t)}x_{i}(t)\exp(f_{i}(t)\Delta{t})
     \label{equation_exponential_growth_model}
@@
 Following the general approach of Shih et al. \cite{Shih:2007bd}, we manually identified 25 sites in HA1 where mutations rapidly swept through the global population.
-We required mutations to emerge from below 5\% global frequency and reach >90\% frequency.
+We required mutations to emerge from below 5\% global frequency and reach $>$90\% frequency.
 Although we did not require sweeps to complete within a fixed amount of time, we observed that they required no longer than 1-3 years to complete.

### 2019-12-05 manuscript/main.tex
Add summary of naive models to table captions

@@
     For reference, the naive model's distance to the future is also shown per validation and test timepoint in grey.
+    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 8.90 $\pm$ 1.69 AAs for test timepoints.
   }
@@
     Coefficients and distances are shown per validation and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 8.90 $\pm$ 1.69 AAs for test timepoints.
   }
@@
     Model a) coefficients and b) distances to the future for individual biologically-informed fitness metrics fit to natural populations.
-    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=XX) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    The naive model's distance to the future was 6.40 $\pm$ 1.36 AAs for validation timepoints and 6.82 $\pm$ 1.74 AAs for test timepoints.
   }
@@
     Model a) coefficients and b) distances to the future for composite fitness metrics fit to natural populations.
-    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=XX) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }

### 2019-12-09 manuscript/main.tex
Correct test results for simulated population with 18 timepoints
Previous results were based on an accidental inclusion of two timepoints in the
test data that did not actually include any new simulated samples. This issue
was caused by the fact that I only simulated 10,000 generations or 50 years of
data, but the assumption of 10 years of test timepoints would have required 51
years of data. This commit updates the test results based on the 18 valid
timepoints.

@@
     Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
-    For model testing, coefficients were fixed to their mean values from training/validation (open circles) and applied to out-of-sample test data.
+    For model testing, coefficients were fixed to their mean values from training/validation (open circles, N=18) and applied to out-of-sample test data.
     Distances between projected and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
@@
     For reference, the naive model's distance to the future is also shown per validation and test timepoint in grey.
-    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 8.90 $\pm$ 1.69 AAs for test timepoints.
+    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 9.07 $\pm$ 1.71 AAs for test timepoints.
   }
@@
   \caption{
-  Test of best model for simulated populations (true fitness) using 10 years previously unobserved test data and fixed model coefficients.
+  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
   a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  b) The percentile rank in observed distance to the future of the model's estimated best strain for 20 test timepoints.
-  The estimated best strain is in the top 20th percentile of observed closest strains for 90\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  b) The percentile rank in observed distance to the future of the model's estimated best strain for 18 test timepoints.
+  The estimated best strain is in the top 20th percentile of observed closest strains for 89\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
   }
@@
 Finally, we tested all of our models on out-of-sample data.
-Specifically, we fixed the coefficients of each model to the average values across the training/validation period and applied the resulting models to the next 10 years of previously unobserved simulated data.
+Specifically, we fixed the coefficients of each model to the average values across the training/validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
 A standard expectation from machine learning is that models will perform worse on test data than on validation data.
 Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-The composite model of non-epitope mutations and LBI outperformed the true fitness metric with average distance to the future of 7.09 $\pm$ 1.15 compared to 7.32 $\pm$ 1.80, respectively (Supplemental Table~\ref{table_simulated_model_test}).
-This composite model also estimated a much less diverse future population on average than the observed future populations by 1.28 $\pm$ 1.96 AAs.
-In contrast, the true fitness model estimated future populations that were closer to the observed diversity.
-Both the composite and true fitness models outperformed the naive model at 95\% and 90\% of test timepoints, respectively.
+The composite model of non-epitope mutations and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_test}).
+Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.
 
 As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
-We observed a growth rate correlation of $R = 0.37$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 52\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
-The estimated best strain was in the top 2nd percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 90\% of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
+We observed a growth rate correlation of $R = 0.38$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
+The estimated best strain was in the top 2nd percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 of 18 (89\%) of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
 These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
@@
 
-We simulated the long-term evolution of A/H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 50 years where 200 generations was equivalent to 1 year.
-We discarded the first 10 years as a burn-in period, selected the next 30 years for model fitting and validation, and held out the last 10 years as out-of-sample data for model testing.
+We simulated the long-term evolution of A/H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 10,000 generations or 50 years where 200 generations was equivalent to 1 year.
+We discarded the first 10 years as a burn-in period, selected the next 30 years for model fitting and validation, and held out the last 9 years as out-of-sample data for model testing.
 Each simulated population was seeded with the full length HA from A/Beijing/32/1992 (NCBI accession: U26830.1) such that all simulated sequences contained signal peptide, HA1, and HA2 domains.

### 2019-12-09 manuscript/supplement.tex
Correct test results for simulated population with 18 timepoints
Previous results were based on an accidental inclusion of two timepoints in the
test data that did not actually include any new simulated samples. This issue
was caused by the fact that I only simulated 10,000 generations or 50 years of
data, but the assumption of 10 years of test timepoints would have required 51
years of data. This commit updates the test results based on the 18 valid
timepoints.

@@
       Coefficients shown here are the mean values across all training and validation windows which were used to test models on out-of-sample data.
-      Distance to the future, approximation of future diversity, and number of times model outperforms the naive model are as in Table~\ref{table_simulated_model_selection} for all test timepoints (N=20).
+      Distance to the future, approximation of future diversity, and number of times model outperforms the naive model are as in Table~\ref{table_simulated_model_selection} for all test timepoints (N=18).
     }

### 2019-12-11 scripts/plot_validation_figure_by_population.py
Add four-panel validation figure with clade errors
Adds a new Jupyter notebook with the simplest logic required to plot a single
four-panel validation figure for simulated populations. This figure includes two
additional panels showing absolute clade errors with a LOESS fit and the
correlation of all estimated and observed ranks for all strains across all
timepoints.

Also adds a generic script version of the above notebook that takes the
appropriate input files and outputs one of the validation figures for the
corresponding data. This script will allow these figures to be automatically
produc

@@ -1,0 +1,447 @@
+"""
+Produce all validation/test figures for all populations in a single notebook.
+"""
+import argparse
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+import matplotlib.gridspec as gridspec
+import numpy as np
+import pandas as pd
+from scipy.stats import pearsonr, spearmanr
+import seaborn as sns
+import statsmodels.api as sm
+
+
+PLOT_THEME_ATTRIBUTES = {
+    "axes.labelsize": 14,
+    "font.size": 18,
+    "legend.fontsize": 12,
+    "xtick.labelsize": 14,
+    "ytick.labelsize": 14,
+    "figure.figsize": [6.0, 4.0],
+    "savefig.dpi": 200,
+    "figure.dpi": 200,
+    "axes.spines.top": False,
+    "axes.spines.right": False,
+    "font.weight": 300,
+    "axes.labelweight": 300,
+    "text.usetex": False
+}
+
+
+def matthews_correlation_coefficient(tp, tn, fp, fn):
+    """Return Matthews correlation coefficient for values from a confusion matrix.
+    Implementation is based on the definition from wikipedia:
+
+    https://en.wikipedia.org/wiki/Matthews_correlation_coefficient
+    """
+    numerator = (tp * tn) - (fp * fn)
+    denominator = np.sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))
+    if denominator == 0:
+            denominator = 1
+
+    return float(numerator) / denominator
+
+
+def get_matthews_correlation_coefficient_for_data_frame(freq_df, return_confusion_matrix=False):
+        """Calculate Matthew's correlation coefficient from a given pandas data frame
+        with columns for initial, observed, and predicted frequencies.
+        """
+        observed_growth = (freq_df["frequency_final"] > freq_df["frequency"])
+        predicted_growth = (freq_df["projected_frequency"] > freq_df["frequency"])
+        true_positives = ((observed_growth) & (predicted_growth)).sum()
+        false_positives= ((~observed_growth) & (predicted_growth)).sum()
+
+        observed_decline = (freq_df["frequency_final"] < freq_df["frequency"])
+        predicted_decline = (freq_df["projected_frequency"] < freq_df["frequency"])
+        true_negatives = ((observed_decline) & (predicted_decline)).sum()
+        false_negatives = ((~observed_decline) & (predicted_decline)).sum()
+
+        mcc = matthews_correlation_coefficient(
[…]

### 2019-12-11 manuscript/main.tex
Include notes to self and update provisional manuscript title

@@
 We measured this approximation at each timepoint as the average Hamming distance between projected and observed populations minus the average Hamming distance of the observed population to itself.
+%% Insert summary statistics of future diversity without describing average Hamming distance
+%% and remove approximation of future diversity text.
 
@@
 \item{Experimental measurements of antigenic drift based on HI assays are more predictive of viral success than epitope mutations}
-\item{The combination of sequence-only metrics outperforms all experimentally-informed metrics}
+\item{Models based on a predefined set of epitope sites risk overfitting to historical data}
+\item{The combination of sequence-only metrics outperforms all experimentally-informed metrics across validation data but fails to account for reassortment events in test data}
 \item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}

### 2019-12-11 src/weighted_distances.py
Support symmetric lookup of pairwise distances
Distances between any pair of strains are only stored once in the lookup
dictionary. This can result in key errors with the corresponding symmetric
lookup. This commit adds a doctest for this use case and the code to make this
test pass.

@@ -55,13 +55,17 @@ def get_distance_matrix_by_sample_names(samples_a, samples_b, distances):
     >>> get_distance_matrix_by_sample_names(samples_a, samples_b, distances)
     array([[1., 2.],
            [3., 4.]])
-    >>>
-
+    >>> get_distance_matrix_by_sample_names(samples_b, samples_a, distances)
+    array([[1., 3.],
+           [2., 4.]])
     """
     matrix = np.zeros((len(samples_a), len(samples_b)))
     for i, sample_a in enumerate(samples_a):
         for j, sample_b in enumerate(samples_b):
-            matrix[i, j] = distances[sample_a][sample_b]
+            try:
+                matrix[i, j] = distances[sample_a][sample_b]
+            except KeyError:
+                matrix[i, j] = distances[sample_b][sample_a]
 
     return matrix
 

### 2019-12-12 scripts/plot_validation_figure_by_population.py
Add Snakemake rules and config to plot validation figures
Includes a new rule to make validation/test figures for each build. This rule
requires additional functions and configuration variables to link the build in
question to other "full tree" builds. Running the script on our cluster also
requires seaborn and modification of matplotlib config.

@@ -23,8 +23,6 @@ PLOT_THEME_ATTRIBUTES = {
     "figure.dpi": 200,
     "axes.spines.top": False,
     "axes.spines.right": False,
-    "font.weight": 300,
-    "axes.labelweight": 300,
     "text.usetex": False
 }
 

### 2019-12-13 manuscript/main.tex
Merge validation and test results into a single table per population
Enables comparisons of validation and test results by putting them in the same
table and placing relevant measures next to each other.

Closes #57.

@@
   \begin{center}
+    \scalebox{0.9}{
       \input{tables/simulated_model_selection.tex}
+    }
     \caption{
-      Model coefficients and validation performance for simulated populations ordered from best to worst by distance to the future.
+      Model coefficients and performance on validation and test data for simulated populations ordered from best to worst by distance to the future in the validation analysis.
       Coefficients are the mean $\pm$ standard deviation for each metric in a given model across 33 training windows.
-      Distance to the future measures the distance in amino acids between estimated and observed future populations.
+      Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
       The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
+      Test results are based on 18 timepoints not observed during model training and validation.
     }
@@
 Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-The composite model of non-epitope mutations and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_test}).
+The composite model of non-epitope mutations and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_selection}).
 Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.
@@
   \begin{center}
+    \scalebox{0.9}{
       \input{tables/natural_model_selection.tex}
+    }
     \caption{
-      Model coefficients and validation performance for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
-      Results are based on training and validation across 23 training windows.
+      Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Validation results are based on 23 timepoints.
+      Test results are based on 8 timepoints not observed during model training and validation.
     }

### 2019-12-13 manuscript/supplement.tex
Merge validation and test results into a single table per population
Enables comparisons of validation and test results by putting them in the same
table and placing relevant measures next to each other.

Closes #57.

--- 2019-11-18 manuscript/supplement.tex
@@
 
-\begin{table*}[ht]
-  \begin{center}
-    \input{tables/simulated_model_test.tex}
-    \caption{
-      Model coefficients and test performance for simulated populations ordered from best to worst by distance to the future.
-      Coefficients shown here are the mean values across all training and validation windows which were used to test models on out-of-sample data.
-      Distance to the future, approximation of future diversity, and number of times model outperforms the naive model are as in Table~\ref{table_simulated_model_selection} for all test timepoints (N=20).
-    }
-    \label{table_simulated_model_test}
-  \end{center}
-\end{table*}
-
 \begin{figure*}[h]
@@
     \caption{
-      All model coefficients and validation performance for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
-      Results are based on training and validation across 23 training windows.
+      All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Validation results are based on 23 timepoints.
+      Test results are based on 8 timepoints not observed during model training and validation.
       Model results for additional variants of fitness metrics including those based on epitope mutations and DMS preferences are included for reference.

### 2019-12-13 manuscript/main.tex
Rename "non-epitope mutations" metric to "mutational load"
Updates figures and main text.

Closes #64.

@@
 
-We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
+We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
 All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.
@@
 Increased antigenic drift relative to previously circulating samples is expected to correspond to increased viral fitness.
-Non-epitope mutations are the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each sample since their ancestor in the previous season.
+Mutational load is the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each sample since their ancestor in the previous season.
 DMS mutational effects provide a more comprehensive biophysical model of functional constraint by measuring the beneficial or deleterious effect of each possible single amino acid mutation in HA from the background of a previous vaccine strain, A/Perth/16/2009.
@@
 
-We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like non-epitope mutations, would receive negative coefficients.
+We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like mutational load, would receive negative coefficients.
 We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
-Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.
+Correspondingly, we expected that a composite model of epitope cross-immunity and mutational load would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.
 
@@
 Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
-The non-epitope mutations metric received a consistently negative coefficient with an average distance of 8.27 $\pm$ 1.35 AAs.
+The mutational load metric received a consistently negative coefficient with an average distance of 8.27 $\pm$ 1.35 AAs.
 
@@
 
-Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Surprisingly, the composite model of epitope cross-immunity and mutational load did not perform better than the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
@@
 This linear fitness metric did not outperform the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
-However, when we fit a composite model of the epitope ancestor and non-epitope mutations metrics, this model outperformed both the individual non-epitope mutations model and the composite cross-immunity and non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+However, when we fit a composite model of the epitope ancestor and mutational load metrics, this model outperformed both the individual mutational load model and the composite cross-immunity and mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
@@
 We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses than models based on individual metrics.
-To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: non-epitope mutations and LBI.
+To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
 This composite model outperformed all individual metrics with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
-The coefficients for LBI and non-epitope mutations remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
+The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
 These results support our hypothesis that multiple complementary metrics can produce more accurate models.
@@
 Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-The composite model of non-epitope mutations and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_selection}).
+The composite model of mutational load and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_selection}).
 Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.
@@
 We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
-As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than non-epitope mutations or DMS mutational effects.
+As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than mutational load or DMS mutational effects.
 Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
@@
 
-Of the two metrics for functional constraint, non-epitope mutations outperformed DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
-In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
+Of the two metrics for functional constraint, mutational load outperformed DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
+In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the mutational load metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
 Notably, the best performance of the DMS mutational effects model was forecasting from April 2007 to April 2008 when the major clade containing A/Perth/16/2009 was first emerging.
 This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
-Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
+Alternate implementations of less background-dependent DMS metrics never performed better than the mutational load metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
 Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects of a single strain.
 
-LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and non-epitope mutations for the most timepoints outperforming the naive model of 17 (74\%) (Table~\ref{table_natural_model_selection}).
-Delta frequency performed worse than LBI and HI cross-immunity and only performed slightly better than non-epitope mutations.
+LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and mutational load for the most timepoints outperforming the naive model of 17 (74\%) (Table~\ref{table_natural_model_selection}).
+Delta frequency performed worse than LBI and HI cross-immunity and only performed slightly better than mutational load.
 While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
@@
 To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
-Specifically, we fit models based on HI cross-immunity and non-epitope mutations, LBI and non-epitope mutations, and all three of these metrics together.
+Specifically, we fit models based on HI cross-immunity and mutational load, LBI and mutational load, and all three of these metrics together.
 We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.
@@
 Although the three-metric composite model performed slightly worse than the individual LBI model, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
-The composite of LBI and non-epitope mutations performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
+The composite of mutational load and LBI performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
 
@@
 As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
-The composite model of non-epitope mutations and LBI effectively captured clade growth and decline with 87\% and 89\% accuracy, respectively (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
+The composite model of mutational load and LBI effectively captured clade growth and decline with 87\% and 89\% accuracy, respectively (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
 The estimated best strain from this model was in the top 11th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 78\% of timepoints (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}B).
@@
   \caption{
-  Validation of best model for natural populations of A/H3N2 viruses, the composite model of non-epitope mutations and LBI.
+  Validation of best model for natural populations of A/H3N2 viruses, the composite model of mutational load and LBI.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
 After identifying the best model to forecast natural populations, we integrated our forecasting model into our existing open source pathogen surveillance framework, Nextstrain \cite{Hadfield2018}, and used our model to forecast A/H3N2 populations in September 2020.
-Our combined non-epitope mutations and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
+Our combined mutational load and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
 To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
@@
 We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
-For example, Equation~\ref{equation_fitness_estimation} estimates fitness per sample by non-epitope mutations ($\mathrm{ne}$) and local branching index ($\mathrm{lbi}$).
+For example, Equation~\ref{equation_fitness_estimation} estimates fitness per sample by mutational load ($\mathrm{ml}$) and local branching index ($\mathrm{lbi}$).
 
 \begin{equation}
-    f_{i}(t) = \beta_{\mathrm{ne}}f_{i, \mathrm{ne}}(t) + \beta_{\mathrm{lbi}}f_{i, \mathrm{lbi}}(t)
+    f_{i}(t) = \beta_{\mathrm{ne}}f_{i, \mathrm{ml}}(t) + \beta_{\mathrm{lbi}}f_{i, \mathrm{lbi}}(t)
     \label{equation_fitness_estimation}

### 2019-12-16 scripts/plot_validation_figure_by_population.py
Rank best strains with distinct amino acid sequences
Instead of ranking all strains by observed or estimated distance to the future,
collapse redundant amino acid sequences such that only one strain represents
each sequence prior to ranking. There is no effective difference between strains
with identical amino acid sequences, but including these strains can artifically
inflate strain ranks.

@@ -12,6 +12,8 @@ import seaborn as sns
 import statsmodels.api as sm
 
 
+np.random.seed(314159)
+
 PLOT_THEME_ATTRIBUTES = {
     "axes.labelsize": 14,
     "font.size": 18,
@@ -110,9 +112,10 @@ if __name__ == "__main__":
         args.tip_attributes,
         sep="\t",
         parse_dates=["timepoint"],
-        usecols=["strain", "timepoint", "frequency"]
+        usecols=["strain", "timepoint", "frequency", "aa_sequence"]
     )
     tips = tips.query("timepoint >= '%s'" % first_validation_timepoint).copy()
+    distinct_tips_with_sequence = tips.groupby(["timepoint", "aa_sequence"]).first().reset_index()
 
     # Load mapping of tips to clades based on a single tree that included all of
     # the tips in the given tip attributes table.
@@ -258,6 +261,12 @@ if __name__ == "__main__":
     sorted_df = full_forecasts.dropna().sort_values(
         ["timepoint"]
     ).copy()
+
+    # Filter sorted records by strains with distinct amino acid sequences.
+    sorted_df = sorted_df.merge(
+        distinct_tips_with_sequence,
+        on=["timepoint", "strain"]
+    )
 
     # First, calculate the rank per strain by observed distance to the future.
     sorted_df["timepoint_rank"] = sorted_df.groupby("timepoint")["weighted_distance_to_future"].rank(pct=True)

### 2019-12-18 scripts/plot_validation_figure_by_population.py
Increase size of validation figures for improved spacing between panels

@@ -292,7 +292,7 @@ if __name__ == "__main__":
     # Summarize model fit by clade frequencies and strain ranks.
     #
 
-    fig = plt.figure(figsize=(8, 8), facecolor='w')
+    fig = plt.figure(figsize=(10, 10), facecolor='w')
     gs = gridspec.GridSpec(2, 2, width_ratios=[1, 1], height_ratios=[1, 1], wspace=0.1)
 
     ticks = np.array([0, 0.2, 0.4, 0.6, 0.8, 1.0])

### 2019-12-18 manuscript/main.tex
Compare vaccine strains with observed and estimated best strains
Adds a figure comparing the vaccine strain selections per timepoint to the
observed closest strain to the future and the models' estimated closest
strain. Also, updates the corresponding Jupyter notebook to resize the image to
so the text annotations don't get cropped from the final image on the right side.

Closes #59.

@@
 
+\begin{figure*}[ht]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/vaccine-comparison.png}
+  \caption{
+    Observed distance to natural A/H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated (orange) closest strains to the future at the corresponding timepoints.
+    Vaccine strains were assigned to their closest timepoint in the validation or test time periods and their distance to the future was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
+    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (titer cross-immunity and mutational load).
+  }
+  \label{fig:vaccine_comparison}
+  \end{center}
+\end{figure*}
+
 \subsection*{Forecasts predict the rise of A1b/131K and A1b/135K sub-clades in September 2020}

### 2019-12-19 manuscript/abstract.tex
Close out open questions and add notes to self in results/discussion

@@
 \end{itemize}
-
-\section*{Open questions}
-
-\begin{itemize}
-\item{How should forecast uncertainty be calculated? \tbc{Important and will be commented on by reviewers. I think this can be carried by purely retrospective comparisons however.}}
-\item{\tbc{Do we need to calculate LBI via a region-weighted scheme? May be better left for a later update? But seems important given centrality of LBI to our predictions.}}
-\end{itemize}

### 2019-12-19 manuscript/main.tex
Close out open questions and add notes to self in results/discussion
Add test results for natural populations
Includes new text describing results for natural populations on out-of-sample
data, updated captions for the validation/test figures describing the four-panel
figures, more consistent references to supplemental figures, and clearer
annotation for composite models in the results tables.

@@
 The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Figure~\ref{sup_fig:cross_validation_for_simulated_populations}).
-These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}, Supplemental Tables~\ref{sup_table:mutations_by_trunk_status_for_simulated_populations} and \ref{sup_table:mutations_by_trunk_status}).
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Fig.~\ref{sup_fig:cross_validation_for_simulated_populations}).
+These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Fig.
+\ref{sup_fig:simulated_h3n2_ha_phylogeny}, Supplemental Tables~\ref{sup_table:mutations_by_trunk_status_for_simulated_populations} and \ref{sup_table:mutations_by_trunk_status}).
 
@@
 We measured this approximation at each timepoint as the average Hamming distance between projected and observed populations minus the average Hamming distance of the observed population to itself.
-%% Insert summary statistics of future diversity without describing average Hamming distance
-%% and remove approximation of future diversity text.
+\jhc{Insert summary statistics of future diversity without describing average Hamming distance and remove approximation of future diversity text.}
 
@@
 
-\begin{figure*}[ht]
+\begin{figure*}[p]
   \begin{center}
@@
   Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
-  a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  b) The percentile rank in observed distance to the future of the model's estimated best strain for 18 test timepoints.
-  The estimated best strain is in the top 20th percentile of observed closest strains for 89\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  B) The rank of the estimated best strain based on its distance to the future in the best model for 18 timepoints shows how often the model makes a good choice when forced to select a single representative strain for the future population.
+  The estimated best strain was in the top 20th percentile of observed closest strains for 89\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
   }
@@
 First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by correlating estimated and observed clade growth rates.
-Our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.72$~$(p < 0.001)$ and overall accuracies for clade growth and decline predictions of 87\% and 58\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+Our model recapitulated clade dynamics with a Pearson's correlation of $R^2 = 0.52$~$(p < 0.001)$ between observed and estimated growth rates and overall accuracies for clade growth and decline predictions of 87\% and 58\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+As expected, our model produced more accurate forecasts for clades with higher initial frequencies (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}C).
 Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
-The estimated best strain was in the top 2nd percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 97\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+The estimated best strain was in the top 1st percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, $p < 0.001$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
 
@@
 As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
-We observed a growth rate correlation of $R = 0.38$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
-The estimated best strain was in the top 2nd percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 of 18 (89\%) of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
+We observed a growth rate correlation of $R^2 = 0.14$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
+Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
+The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 of 18 (89\%) of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
+Observed and estimated strain ranks remained strongly correlated (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
 These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
@@
 Next, we trained and validated models for individual fitness predictors using 25 years of natural A/H3N2 populations spanning from October 1, 1990 to October 1, 2015.
-We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Figure~\ref{sup_fig:cross_validation_for_natural_populations}).
+We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
@@
 We fit a cross-immunity model to these 25 sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefited from knowledge of the future in its forecasts.
-In contrast to the epitope cross-immunity model, the oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
+In contrast to the epitope cross-immunity model, the oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
 These results support our hypothesis that the success of previous epitope models based on a fixed set of 49 sites is due to historical contingency rather than a constant contribution of these sites to antigenic drift.
@@
 Alternate implementations of less background-dependent DMS metrics never performed better than the mutational load metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
-Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects of a single strain.
+Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive biophysical model based on measured mutational effects of a single strain.
 
-LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and mutational load for the most timepoints outperforming the naive model of 17 (74\%) (Table~\ref{table_natural_model_selection}).
+LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
 Delta frequency performed worse than LBI and HI cross-immunity and only performed slightly better than mutational load.
 While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
-
-\subsection*{Composite models outperform models with individual fitness metrics}
-
-To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
-Specifically, we fit models based on HI cross-immunity and mutational load, LBI and mutational load, and all three of these metrics together.
-We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.
 
@@
 
+To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
+Specifically, we fit models based on HI cross-immunity and mutational load, mutational load and LBI, and all three of these metrics together.
+We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.
+
 Both two-metric composite models performed better than their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
-Although the three-metric composite model performed slightly worse than the individual LBI model, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
+Although the three-metric composite model performed slightly worse than the individual LBI model based on average distance to the future, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
 The composite of mutational load and LBI performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
@@
 
-\subsection*{Models enable selection of vaccine candidate strains}
+As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
+The composite model of mutational load and LBI effectively captured clade dynamics with a growth correlation of $R^2 = 0.35$~$(p < 0.001)$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
+Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}C).
+The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}B).
+This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, $p < 0.001$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}D).
 
-As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
-The composite model of mutational load and LBI effectively captured clade growth and decline with 87\% and 89\% accuracy, respectively (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
-The estimated best strain from this model was in the top 11th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 78\% of timepoints (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}B).
-
-\begin{figure*}[ht]
+\begin{figure*}[p]
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
+  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-natural-populations.png}
   \caption{
-  Validation of best model for natural populations of A/H3N2 viruses, the composite model of mutational load and LBI.
+  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity and mutational load.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
+  B) The rank of the estimated best strain based on its distance to the future for 8 timepoints.
+  The estimated best strain was in the top 20th percentile of observed closest strains for 88\% of timepoints.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
   }
-  \label{fig:validation_of_best_model_for_natural_populations}
+  \label{fig:test_of_best_model_for_natural_populations}
   \end{center}
@@
 
-Finally, we tested the performance of this same best model on out-of-sample data collected from October 1, 2015 through October 1, 2019.
+Finally, we tested the performance of all models on out-of-sample data collected from October 1, 2015 through October 1, 2019.
+We anticipated that most models would perform worse on truly out-of-sample data than on validation data.
+As expected, all models estimated higher average distances to the future across these eight test timepoints (Table~\ref{table_natural_model_selection}).
+However, these increases were mostly consistent with an increased average distance between years of 6.82 $\pm$ 1.74 AAs as measured by the naive model.
+The composite of HI cross-immunity, mutational load, and LBI outperformed all other models on the test data with an average distance of 6.12 $\pm$ 1.42 AAs.
+Surprisingly, the best model for the validation data -- mutational load and LBI -- was one of the worst models for the test data with an average distance to the future of 7.70 $\pm$ 3.53 AAs.
+The individual LBI model was the worst model, while mutational load continued to perform well with test data.
+LBI performed especially poorly in the last two test timepoints of April and October 2018 (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+These timepoints correspond to the dominance and sudden decline of a reassortant clade named A2/re \cite{Potter2019}.
+By April 2018, the A2/re clade had risen to a global frequency over 50\% from less than 15\% the previous year, despite an absence of antigenic drift.
+However, by October 2018, this clade had declined in frequency to approximately 30\% and, by October 2019, it had gone extinct.
+These results highlight a major limitation of growth-based fitness metrics like LBI and a corresponding benefit of more mechanistic metrics that explicitly measure antigenic drift and functional constraint.
+
+After identifying the composite HI cross-immunity, mutational load, and LBI model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
+The composite model effectively captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$~$(p < 0.001)$ between observed and estimated growth rates and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
+The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Fig.~\ref{fig:test_of_best_model_for_natural_populations}C).
+The estimated best strain from this model was in the top second percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}B).
+Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.69$, $p < 0.001$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
 
@@
     Vaccine strains were assigned to their closest timepoint in the validation or test time periods and their distance to the future was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
-    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (titer cross-immunity and mutational load).
+    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI cross-immunity and mutational load).
   }
@@
 \item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
+\item{Region-aware LBI could immediate integrate geography into models to account for issues like the A2/re forecast errors}
 \item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}

### 2019-12-19 manuscript/supplement.tex
Add test results for natural populations
Includes new text describing results for natural populations on out-of-sample
data, updated captions for the validation/test figures describing the four-panel
figures, more consistent references to supplemental figures, and clearer
annotation for composite models in the results tables.

@@
   Validation of best model for simulated populations of A/H3N2-like viruses.
-  a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  b) The percentile rank in observed distance to the future of the model's estimated best strain for 33 validation timepoints.
-  The estimated best strain is in the top 20th percentile of observed closest strains for 97\% of timepoints.
+  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
+  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
   }
@@
 \end{figure*}
+
+\begin{figure*}[ht]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
+  \caption{
+  Validation of best model for natural populations of A/H3N2 viruses, the composite model of mutational load and LBI.
+  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.
+  The estimated best strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+  }
+  \label{sup_fig:validation_of_best_model_for_natural_populations}
+  \end{center}
+\end{figure*}

### 2019-12-19 manuscript/main.tex
Summarize vaccine comparison results
Closes #60.

@@
     Vaccine strains were assigned to their closest timepoint in the validation or test time periods and their distance to the future was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
-    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI cross-immunity and mutational load).
+    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI cross-immunity, mutational load, and LBI).
   }
@@
 
+We further evaluated our best model's apparent success at estimating the closest strain to the next season's A/H3N2 population by comparing our model's selection to the WHO's vaccine strain selection.
+For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed and estimated closest strains.
+Our model selected a strain that was closer to the future than the selected vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
+In the last three seasons, the model's selection and the vaccine strain converged to similar distances.
+
 \subsection*{Forecasts predict the rise of A1b/131K and A1b/135K sub-clades in September 2020}
 
-After identifying the best model to forecast natural populations, we integrated our forecasting model into our existing open source pathogen surveillance framework, Nextstrain \cite{Hadfield2018}, and used our model to forecast A/H3N2 populations in September 2020.
-Our combined mutational load and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
+To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}, and used our best model to forecast A/H3N2 populations in September 2020.
+Our combined HI cross-immunity, mutational load, and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
 To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).

### 2019-12-19 manuscript/main.tex
Update results for best model in natural population test data
After updating to the model with the absolute lowest mean distance to the future
in the test data, this commit updates the corresponding text and figures to
reflect this model's performance. This commit also corrects a typo in the rank
results for the test data caused by a copy/paste of prose.

@@
 After identifying the composite HI cross-immunity, mutational load, and LBI model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
-The composite model effectively captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$~$(p < 0.001)$ between observed and estimated growth rates and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
+The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.50$~$(p < 0.001)$ between observed and estimated growth rates and growth and decline accuracies of 57\% and 58\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
 The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Fig.~\ref{fig:test_of_best_model_for_natural_populations}C).
-The estimated best strain from this model was in the top second percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}B).
+The estimated best strain from this model was in the top second percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for seven (88\%) of eight timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}B).
 Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.69$, $p < 0.001$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
@@
 For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed and estimated closest strains.
-Our model selected a strain that was closer to the future than the selected vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
-In the last three seasons, the model's selection and the vaccine strain converged to similar distances.
+Our models selected a strain that was closer to the future than the selected vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
+In the two seasons where the vaccine strain was closer to the future than the model's selection, the difference between the two strains' distances to the future was less than one amino acid.
 

### 2019-12-20 manuscript/main.tex
Fixed a reference to the best model missed in the last commit
Add methods for alternate DMS metrics and rename primary metric in tables

@@
   \caption{
-  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity and mutational load.
+  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity, mutational load, and LBI.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
 
+To determine whether DMS preferences could be used to define fitness metrics that were less dependent on the historical context of the background strain, we implemented two additional DMS-based metrics: ``DMS entropy'' and ``DMS mutational load''.
+For both metrics, we calculated the distance between HA amino acid sequences of each strain and its ancestral sequence in the previous season, to enable comparison of these metrics with the DMS mutational effects and mutational load metrics.
+For the ``DMS entropy'' metric, we calculated the distance between sequences such that each mismatch was weighted by the inverse entropy of DMS preferences at the site of the mismatch.
+We expected this metric to produce a negative coefficient similar to the mutational load metric, as higher values will result from mutations at sites with lower entropy and, thus, lower tolerance for mutations.
+For the ``DMS mutational load'' metric, we defined a novel set of non-epitope sites corresponding to each position in HA with a standardized entropy less than zero.
+With this metric, we sought to identify more highly conserved sites without weighting any one site differently from others.
+We anticipated that this lack of site-specific weighting would make the DMS mutational load metric even less background-dependent than the DMS entropy and DMS mutational effect metrics.
+
 \subsubsection*{Clade growth}

### 2019-12-20 manuscript/main.tex
Edit main text, add methods for sequence clustering
Primarily includes rephrasing of results text and figure captions. Also adds
citations and method for clustering analysis used in the model schematic
figure.

@@
     Schematic representation of the fitness model for simulated A/H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
-    Samples are colored by their amino acid sequence composition such that genetically similar samples have similar colors.
-    A) Samples at timepoint $t$ are shown in their phylogenetic context and sized by their frequency at that timepoint.
-    The estimated future population at timepoint $u$ is projected to the right with samples scaled in size by their projected frequency based on the true fitness model.
-    B) The frequency trajectories of samples at timepoint $t$ to $u$ represent the predicted the growth of the orange samples to the detriment of the purple samples.
-    C) Samples at timepoint $u$ are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
+    Samples are colored by their amino acid sequence composition such that genetically similar samples have similar colors (Methods).
+    A) Samples at timepoint $t$, $\mathbf{x}(t)$, are shown in their phylogenetic context and sized by their frequency at that timepoint.
+    The estimated future population at timepoint $u$, $\mathbf{\hat{x}}(u)$, is projected to the right with samples scaled in size by their projected frequency based on the true fitness model.
+    B) The frequency trajectories of samples at timepoint $t$ to $u$ represent the predicted the growth of the dark blue samples to the detriment of the pink samples.
+    C) Samples at timepoint $u$, $\mathbf{x}(u)$, are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
     D) The observed frequency trajectories of samples at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate, e.g.
-the emergence of the yellow cluster from within the successful orange cluster.
+the emergence of the light blue cluster from within the successful dark blue cluster.
   }
@@
 As a positive control for our model framework, we also fit a model based on the true fitness of each sample as measured by the simulator.
-We evaluated the performance of each model by the absolute distance between projected and observed future populations and how often it outperformed a naive model that estimated no change in populations between years.
+We evaluated the performance of each model by the absolute distance between projected and observed future populations.
 As each model only projects existing sequences forward one year without trying to predict which mutations occur in the future, the model's absolute distance to the future can never be zero.
-For this reason, we additionally measured how well each model approximated the diversity of the future population.
-We measured this approximation at each timepoint as the average Hamming distance between projected and observed populations minus the average Hamming distance of the observed population to itself.
-\jhc{Insert summary statistics of future diversity without describing average Hamming distance and remove approximation of future diversity text.}
+For this reason, we additionally calculated how often each model outperformed a naive model that estimated no change in populations between years and, therefore, measured the observed distance between annual populations.
 
@@
   \caption{
-    Model a) coefficients and b) distances between projected and observed future populations as measured in amino acids (AAs) for simulated populations.
-    Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
-    For model testing, coefficients were fixed to their mean values from training/validation (open circles, N=18) and applied to out-of-sample test data.
-    Distances between projected and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
+    Model coefficients and distances between projected and observed future populations as measured in amino acids (AAs) for simulated populations.
+    A) Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
+    For model testing, coefficients were fixed to their mean values from training/validation and applied to out-of-sample test data (open circles, N=18).
+    B) Distances between projected and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
     The mean $\pm$ standard deviation of distances per validation timepoint are shown in the top-left of each panel.
@@
 
-We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like mutational load, would receive negative coefficients.
+We hypothesized that fitness metrics associated with viral success such as true fitness, epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like mutational load, would receive negative coefficients.
 We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
@@
 
-As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 of 33 (97\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
-Importantly, the true fitness model accurately approximated the diversity of future populations coming within -0.29 $\pm$ 2.01 AAs of observed future diversity.
-In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future and an average of 2.26 $\pm$ 1.75 AAs from the future's observed diversity.
+As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 (97\%) of 33 timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
+In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future.
 With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
@@
   \caption{
-    Model a) coefficients and b) distances to the future for individual biologically-informed fitness metrics and the best composite model fit to simulated populations.
-    Coefficients and distances are shown per validation and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    Model coefficients and distances to the future for individual biologically-informed fitness metrics and the best composite model fit to simulated populations.
+    A) Coefficients and B) distances are shown per validation and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
     The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 8.90 $\pm$ 1.69 AAs for test timepoints.
@@
 To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
-This linear fitness metric did not outperform the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
+By itself, this linear fitness metric did not outperform the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
 However, when we fit a composite model of the epitope ancestor and mutational load metrics, this model outperformed both the individual mutational load model and the composite cross-immunity and mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
@@
 To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
-This composite model outperformed all individual metrics with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
+This composite model outperformed all individual metrics with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
 The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
@@
 We validated the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
-First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by correlating estimated and observed clade growth rates.
-Our model recapitulated clade dynamics with a Pearson's correlation of $R^2 = 0.52$~$(p < 0.001)$ between observed and estimated growth rates and overall accuracies for clade growth and decline predictions of 87\% and 58\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
-As expected, our model produced more accurate forecasts for clades with higher initial frequencies (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}C).
+First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed and estimated clade growth rates.
+The model's estimated growth rates correlated well with observed growth rates (Pearson's $R^2 = 0.52$, $p < 0.001$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
+Model forecasts were increasingly more accurate with increasing initial clade frequencies (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}C).
 Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
-The estimated best strain was in the top 1st percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
 Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, $p < 0.001$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
@@
 As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
-We observed a growth rate correlation of $R^2 = 0.14$~$(p < 0.001)$ and clade growth and decline prediction accuracies of 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
+While observed and estimated growth rates correlated more weakly for test data (Pearson's $R^2 = 0.14$, $p < 0.001$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
 Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
-The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 of 18 (89\%) of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
-Observed and estimated strain ranks remained strongly correlated (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
+The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
+Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
 These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
-These results also highlight a well-established yet critical point that true forecasts on out-of-sample data are less accurate than those based on training and validation data.
+These results also highlight a well-established yet critical point that forecasts on truly out-of-sample data are less accurate than those based on training and validation data.
 
@@
       Validation results are based on 23 timepoints.
-      Test results are based on 8 timepoints not observed during model training and validation.
+      Test results are based on eight timepoints not observed during model training and validation.
     }
@@
   \caption{
-    Model a) coefficients and b) distances to the future for individual biologically-informed fitness metrics fit to natural populations.
-    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    Model coefficients and distances to the future for individual biologically-informed fitness metrics fit to natural populations.
+    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
     The naive model's distance to the future was 6.40 $\pm$ 1.36 AAs for validation timepoints and 6.82 $\pm$ 1.74 AAs for test timepoints.
@@
 
-The naive model measured a distance of 6.40 $\pm$ 1.36 AAs between natural populations or 71\% of the distance between yearly simulated populations (Table~\ref{table_natural_model_selection}).
-Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+The naive model measured a distance of 6.40 $\pm$ 1.36 AAs between natural populations or 71\% of the average distance between yearly simulated populations (Table~\ref{table_natural_model_selection}).
+Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
 The average improvement of the sequence-only models over the naive model was consistently lower than the same models in simulated populations.
@@
 This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
-Alternate implementations of less background-dependent DMS metrics never performed better than the mutational load metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
+Alternate implementations of less background-dependent DMS metrics never performed better than the mutational load metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}, Methods).
 Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive biophysical model based on measured mutational effects of a single strain.
@@
   \caption{
-    Model a) coefficients and b) distances to the future for composite fitness metrics fit to natural populations.
-    Coefficients and distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    Model coefficients and distances to the future for composite fitness metrics fit to natural populations.
+    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }
@@
 We calculated the delta frequency as the change in frequency for each strain between the most recent timepoint in a given phylogeny and six months prior to that timepoint divided by 0.5 years.
+
+\subsection*{Clustering of amino acid sequences for visualization}
+
+For the purpose of visualizing related amino acid sequences in Fig.~\ref{fig:model}, we applied dimensionality reduction to pairwise amino acid distances followed by hierarchical clustering.
+Specifically, we selected a representative tree from our simulated population of viruses at month 10 of year 30.
+From this tree, we selected all samples with a collection date in the previous two years.
+We calculated the pairwise Hamming distance between the full-length HA amino acid sequences for all selected samples and applied t-SNE dimensionality reduction \cite{vanDerMaaten2008} to the resulting distance matrix (n=2 components, perplexity=30.0, and learning rate=400).
+We assigned each sample to a cluster based on its two-dimensional t-SNE embedding using DBSCAN \cite{Ester1996} with a maximum neighborhood distance of 10 AAs and a minimum of 20 samples per cluster.
+Despite known limitations of applying hierarchical clustering to manifold projections that do not preserve sample density, this approach allowed us to effectively assign samples to qualitative genetic clusters for the purposes of visualization.

### 2019-12-20 manuscript/supplement.tex
Edit main text, add methods for sequence clustering
Primarily includes rephrasing of results text and figure captions. Also adds
citations and method for clustering analysis used in the model schematic
figure.

@@
   \caption{
-  Composite model a) coefficients and b) distances to the future for models fit to simulated populations.
-  Coefficients and distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+  Composite model coefficients and distances to the future for models fit to simulated populations.
+  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }
@@
   \begin{center}
+    \scalebox{0.9}{
         \input{tables/complete_natural_model_selection.tex}
+    }
     \caption{
@@
   \caption{
-  Model a) coefficients and b) distances to the future for cross-immunity models fit to natural populations.
+  Model coefficients and distances to the future for cross-immunity models fit to natural populations.
+  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   The epitope cross-immunity model relies on previously published epitope sites \cite{Luksza:2014hj}.
-  The ``oracle'' cross-immunity model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period.
-  The improved performance of the ``oracle'' model indicates that the cross-immunity can be effective when sites of beneficial mutations are known prior to forecasting.
-  Coefficients and distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+  The ``oracle'' cross-immunity model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
+  The improved performance of the ``oracle'' model indicates that the sequence-based cross-immunity metric can be effective when sites of beneficial mutations are known prior to forecasting.
   }

### 2019-12-20 manuscript/main.tex
Add summary of coefficients learned across natural test data
Includes an updated analysis notebook, a new supplemental figure, and a brief
summary in the main results text.

@@
 To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}, and used our best model to forecast A/H3N2 populations in September 2020.
-Our combined HI cross-immunity, mutational load, and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
+Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning coefficients per timepoint from the test data.
+None of the three models performed better across the test data than the original best model, indicating that the coefficients for our selected best model would be most accurate for live forecasts (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
+With the original coefficients, our combined HI cross-immunity, mutational load, and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
 To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).

### 2019-12-20 manuscript/supplement.tex
Add summary of coefficients learned across natural test data
Includes an updated analysis notebook, a new supplemental figure, and a brief
summary in the main results text.

@@
 \end{figure*}
+
+\begin{figure*}[ht]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.png}
+  \caption{
+    Model coefficients and distances to the future for best composite models fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    A) Coefficients and B) distances are shown per test timepoint (N=8).
+    In contrast to the results for these models based on fixed coefficients from training/validation, these coefficients were learned for each six-year window prior to the corresponding test timepoint.
+    The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
+    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.
+  }
+  \label{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
+  \end{center}
+\end{figure*}

### 2019-12-21 manuscript/main.tex
Trying to start the discussion
This is slow going

@@
 
+We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza A/H3N2 by estimating the sequence composition of future populations.
+The best models from this framework effectively capture clade dynamics and accurately identify optimal vaccine candidates without relying on phylogenetic clades as model targets.
+
+Our evaluation of models by time-series cross-validation and true out-of-sample tests reveal substantial potential for overfitting models to historical data.
+The clearest examples of this overfitting were found in the definition of antigenic drift by historically relevant epitope sites and the definition of functional constraint by DMS experiments against a single viral background.
+
+Indeed, the modest improvement of our best models over a naive model of no change between years highlights the inherent difficulty and potentially limited ability to make precise forecasts for such a complex, dynamical system.
+
 Key findings:
@@
 \item{DMS measurements appear to be too background-specific to be predictive for global viral success}
+\item{\rnc{The overall gain from prediction is low (1aa), but this has to be seen in the light of the short prediction horizon (1y) and the larger coalescence time of about 3-8y.
+Hence we expect diversity to persist and any gain has to come from clade frequencies.}}
+\item{\rnc{The future tends to anchor deeper in the tree away from the tips.
+This may be the primary reason for the success of the LBI and non-epitope.
+Both of these metrics are lower at the tips than they are at viruses close to the base of the tree.}}
 \item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
@@
 \item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
+\item{\jhc{Cite Bell et al. 2019 and ML-based forecasting work for flu}}
 \end{itemize}

### 2019-12-24 manuscript/main.tex
Continue writing discussion and try a better title

@@
 
+\pagebreak
+
 \section*{Discussion}
@@
 We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza A/H3N2 by estimating the sequence composition of future populations.
-The best models from this framework effectively capture clade dynamics and accurately identify optimal vaccine candidates without relying on phylogenetic clades as model targets.
+A key innovation of this framework is its ability to directly compare viral populations between seasons using the Earth Mover's Distance metric \cite{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
+The best models from this framework still effectively capture clade dynamics and accurately identify optimal vaccine candidates from simulated and natural A/H3N2 populations without relying on clades as model targets.
+We have further introduced novel fitness metrics based on experimental measurements of antigenic drift and functional constraint.
+We demonstrated how these metrics can be effectively integrated with previously published sequence-only metrics.
+We have added this framework as a component of seasonal influenza analyses on nextstrain.org where it provides real-time forecasts for influenza researchers, decision makers, and the public.
 
-Our evaluation of models by time-series cross-validation and true out-of-sample tests reveal substantial potential for overfitting models to historical data.
-The clearest examples of this overfitting were found in the definition of antigenic drift by historically relevant epitope sites and the definition of functional constraint by DMS experiments against a single viral background.
+\subsection*{Integration of genotypic and phenotypic metrics minimizes overfitting}
 
+Our evaluation of models by time-series cross-validation and true out-of-sample forecasts revealed substantial potential for model overfitting.
+We observed overfitting to both specific genetic backgrounds and general historical contexts.
+A clear example of the former was the poor performance of our DMS-based fitness metric compared to a simpler mutational load metric.
+Although the DMS experiments provided detailed estimates of which amino acids were preferred at which positions in HA, these measurements were specific to a single strain, A/Perth/16/2009 \cite{Lee2018}.
+When we applied these measurements to predict the success of global populations, they were less informative on average than the naive model.
+To benefit from the more comprehensive fitness costs measured by DMS data, future models will need to synthesize DMS measurements across multiple A/H3N2 strains from distinct historical contexts.
+We anticipate that these measurements could be used to define and continually update a modern set of sites contributing to mutational load in natural populations.
+This set of sites could replace the statically defined set of ``non-epitope'' sites we use to estimate mutational load here.
+
+We observed two different cases of overfitting to general historical contexts.
+In the first case, the model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
+Specifically, we found that this model failed to predict the sudden decline of a dominant reassortant clade, A2/re, in 2019.
+Despite this model's excellent performance historically, it was unable to account for rare yet important events such as reassortment.
+To avoid such overfitting in the future, models will need fitness metrics that can represent the evolutionary contributions of neuraminidase, specifically, or of reassortment, generally.
+
+In the second case, we observed overfitting of sites associated with antigenic drift to historical contexts.
+The fitness benefit of mutations that lead to antigenic drift in A/H3N2 is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}
+Although the antigenic importance of seven specific sites in HA has been experimentally validated \cite{Koel:2013jz}, these sites do not explain all antigenic drift observed in natural populations \cite{Neher:2016hy}.
+Other attempts to define these so-called ``epitope sites'' have relied on either aggregation of results from antigenic escape assays \cite{Wolf:2006da} or retrospective computational analyses of sites with beneficial mutations \cite{Shih:2007bd,Luksza:2014hj}.
+We found that all of these specific definitions suffer from overfitting to the historical context in which they were discovered (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
+These results suggest that the set of sites that contribute to antigenic drift at any given time may depend on both the fitness landscape of currently circulating strains and the immune landscape of the hosts these strains need to infect.
+Recent experimental mapping of antigenic escape mutations in A/H3N2 HA with human sera show that the specific sites that confer antigenic escape can vary dramatically between individuals based on their exposure history \cite{Lee2019}.
+In contrast to models based on predefined ``epitope sites'', our model based on experimental measurements of antigenic drift did not suffer from overfitting in the validation or test periods.
+We suspect that this model was able to minimize overfitting by continuously updating its antigenic model with recent experimental data and assigning antigenic weight to branches of a phylogeny rather than specific positions in HA.
+
+Finally, we observed that composite models of multiple orthogonal fitness metrics often outperformed models based on their individual components.
+These results are consistent with previous work that found improved performance by integrating components of antigenic drift, functional constraint, and clade growth \cite{Luksza:2014hj}.
+However, the convergence of our three-metric model to a LBI-only model during the validation period (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}) reveals the limitations of our current additive approach to composite models.
+The recent success of weighted ensembles for short-term influenza forecasting \cite{Reich:2019bq} suggests that long-term forecasting may benefit from a similar approach.
+
+\subsection*{Practical recommendations for forecasting}
+
+The best model in the validation period got 0.96 AAs closer to the future than the naive model.
+The best model in the test period got 0.66 AAs closer to the future on average.
+At a historical substitution rate of 0.0043 substitutions per site per year, we anticipate an average accumulation of 2.4 AAs per season.
 Indeed, the modest improvement of our best models over a naive model of no change between years highlights the inherent difficulty and potentially limited ability to make precise forecasts for such a complex, dynamical system.

### 2019-12-24 manuscript/main.tex
Updates the main text to refer to the simpler best model
Uses simpler model name and test results. Also updates a remaining reference
from "non-epitope" to "mutational load".

@@
 The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
-In contrast, the three-metric model strongly preferred the HI cross-immunity and non-epitope metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
+In contrast, the three-metric model strongly preferred the HI cross-immunity and mutational load metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
 In the last two validation timepoints, this composite model converged to a nearly LBI-only model with the coefficients for the other two metrics converging to near-zero values.
@@
   \caption{
-  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity, mutational load, and LBI.
+  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity and mutational load.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
 The composite of HI cross-immunity, mutational load, and LBI outperformed all other models on the test data with an average distance of 6.12 $\pm$ 1.42 AAs.
+The simpler composite of HI cross-immunity and mutational load performed nearly as well, however, with more stable coefficients per metric over time.
+For these reasons, we selected this two-metric model for our practical forecasts.
 Surprisingly, the best model for the validation data -- mutational load and LBI -- was one of the worst models for the test data with an average distance to the future of 7.70 $\pm$ 3.53 AAs.
@@
 
-After identifying the composite HI cross-immunity, mutational load, and LBI model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
-The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.50$~$(p < 0.001)$ between observed and estimated growth rates and growth and decline accuracies of 57\% and 58\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
+After identifying the composite HI cross-immunity and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
+The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$~$(p < 0.001)$ between observed and estimated growth rates and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
 The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Fig.~\ref{fig:test_of_best_model_for_natural_populations}C).
@@
     Vaccine strains were assigned to their closest timepoint in the validation or test time periods and their distance to the future was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
-    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI cross-immunity, mutational load, and LBI).
+    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI cross-immunity and mutational load).
   }
@@
 None of the three models performed better across the test data than the original best model, indicating that the coefficients for our selected best model would be most accurate for live forecasts (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
-With the original coefficients, our combined HI cross-immunity, mutational load, and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
+With the original coefficients, our combined HI cross-immunity and mutational load model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
 To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).

### 2019-12-26 manuscript/main.tex
Expand on consequences for practical forecasting

@@
 
-\subsection*{Practical recommendations for forecasting}
+\subsection*{Forecasting framework aids practical forecasts}
 
-The best model in the validation period got 0.96 AAs closer to the future than the naive model.
-The best model in the test period got 0.66 AAs closer to the future on average.
-At a historical substitution rate of 0.0043 substitutions per site per year, we anticipate an average accumulation of 2.4 AAs per season.
-Indeed, the modest improvement of our best models over a naive model of no change between years highlights the inherent difficulty and potentially limited ability to make precise forecasts for such a complex, dynamical system.
+By forecasting the composition of future A/H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
+While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AA per year for validation data and 0.66 AA per year for test data.
+These results are consistent with the observed dynamics of A/H3N2.
+First, the one-year forecast horizon is a fraction of the average coalescence time for A/H3N2 populations of about 3--8 years \cite{}.
+Hence, we expect the diversity of circulating strains to persist between seasons.
+Second, the average substitution rate for A/H3N2's HA is 0.0043 substititions per site per year or an average accumulation of 2.4 AAs per year.
+This accumulation of substitutions contributes to the distance between annual populations observed by the naive model.
+In this context, our model gains of 0.96 and 0.66 AAs per year correspond to an explanation of 40\% and 28\% of the expected additional distance between annual populations, respectively.
+This modest improvement over the naive model highlights the inherent difficulty and potentially limited ability to make precise forecasts for such a complex, dynamical system with the current framework.
 

### 2019-12-27 manuscript/main.tex
Finish drafting discussion, stub out remaining meta-sections
Includes acknowledgements, author contributions, etc. and some methods sections
that should be added or expanded.

@@
     Observed distance to natural A/H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated (orange) closest strains to the future at the corresponding timepoints.
-    Vaccine strains were assigned to their closest timepoint in the validation or test time periods and their distance to the future was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
+    Vaccine strains were assigned to the validation or test timepoint closest to the date they were selected by the WHO.
+    The distance to the future of each vaccine strain was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
     The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI cross-immunity and mutational load).
@@
 In this context, our model gains of 0.96 and 0.66 AAs per year correspond to an explanation of 40\% and 28\% of the expected additional distance between annual populations, respectively.
-This modest improvement over the naive model highlights the inherent difficulty and potentially limited ability to make precise forecasts for such a complex, dynamical system with the current framework.
 
-Key findings:
+There are still several clear opportunities to improve forecasts.
+Integration of more recent experimental data may improve estimates of antigenic drift.
+Influenza collaborating centers now routinely measure antigenic drift by focus reduction assays (FRAs), as an alternative to hemagglutination inhibition (HI) assays that is less affected by glycosylation motifs that alter binding efficacy \cite{Zost2017}.
+While there are not enough historical FRA measurements to validate and test a FRA cross-immunity metric, it may be possible to reuse coefficients from the HI cross-immunity models with modern FRA data.
+In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
+These assays reveal the specific residues that confer antigenic escape at specific sites.
+A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape A/H3N2 strains must navigate to be globally successful.
 
-\begin{itemize}
-\item{Our model accurately forecasts clade frequency trajectories and identifies optimal potential vaccine strains for both simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
-\item{Experimental measurements of antigenic drift based on HI assays are more predictive of viral success than epitope mutations}
-\item{Models based on a predefined set of epitope sites risk overfitting to historical data}
-\item{The combination of sequence-only metrics outperforms all experimentally-informed metrics across validation data but fails to account for reassortment events in test data}
-\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
-\item{DMS measurements appear to be too background-specific to be predictive for global viral success}
-\item{\rnc{The overall gain from prediction is low (1aa), but this has to be seen in the light of the short prediction horizon (1y) and the larger coalescence time of about 3-8y.
-Hence we expect diversity to persist and any gain has to come from clade frequencies.}}
-\item{\rnc{The future tends to anchor deeper in the tree away from the tips.
-This may be the primary reason for the success of the LBI and non-epitope.
-Both of these metrics are lower at the tips than they are at viruses close to the base of the tree.}}
-\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
-\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
-\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
-\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
-\item{Region-aware LBI could immediate integrate geography into models to account for issues like the A2/re forecast errors}
-\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
-\item{\jhc{Cite Bell et al. 2019 and ML-based forecasting work for flu}}
-\end{itemize}
+Finally, forecasting models need to account for the geographic distribution of viruses.
+Globally successful clades of influenza A/H3N2 most commonly emerge from Southern Asia, while viruses emerging from Africa, North America, and Europe are less likely to fix globally \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
+Explicitly accounting for these source/sink dynamics would allow models to weight forecasts based on both viral fitness and transmission.
+For example, a geographically-aware implementation of LBI would most likely have not incorrectly predicted the success of the predominantly North American and European reassortant clade A2/re in 2019.
 
-\section*{Methods}
+\subsection*{Conclusions}
+
+Antigenic drift and mutational load remain the best predictors of future success for seasonal influenza A/H3N2 populations.
+Our framework enables real-time practical forecasts of these populations by leveraging historical and modern experimental assays and gene sequences.
+By releasing our framework as an open source tool based on modern data science standards like tidy data frames, we hope to encourage continued development by the influenza research community.
+We additionally anticipate that the ability to forecast the sequence composition of populations with Earth Mover's Distance will enable future forecasting research with pathogens whose genomes cannot be analyzed by traditional phylogenetic methods including recombinant viruses, bacteria, and fungi.
+
+\subsection*{Model sharing and extensions}
+
+We have provided all source code, configuration files, and datasets at https://github.com/blab/flu-forecasting.
+
+\section*{Materials and methods}
+
+\subsection*{Data}
+
+\subsubsection*{Sequences}
+
+\jhc{Document HA sequences from GISAID (reference acknowledgements table).}
+
+\subsubsection*{Titers}
+
+\jhc{Document HI titers from influenza CCs.}
 
@@
 
+\subsubsection*{Strain-specific distance to the future}
+
+The weighted Hamming distance to the future of each strain was calculated from the strain's HA amino acid sequence and the frequencies and sequences of the corresponding population one year in the future.
+\jhc{Add equation for weighted Hamming distance here.}
+
 \subsubsection*{Time-series cross-validation}
@@
 Despite known limitations of applying hierarchical clustering to manifold projections that do not preserve sample density, this approach allowed us to effectively assign samples to qualitative genetic clusters for the purposes of visualization.
+
+\subsection*{Data and software availability}
+
+All source code, configuration files, and datasets are available at https://github.com/blab/flu-forecasting.
+
+\section*{Acknowledgements}
+
+We thank Jesse Bloom, Erick Matsen, Harmit Malik, Sidney Bell, Alli Black, Lola Arakaki, and members of the Bedford lab for useful advice and discussions.
+JH is a Graduate Research Fellow and is supported by the National Institute Of Allergy And Infectious Diseases under NIH F31AI140714.
+TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.
+\jhc{Add RN support here.}
+
+\section*{Author contributions}
+
+JH.
+RN.
+TB.

### 2019-12-28 manuscript/main.tex
Clarify points in discussion
This wraps up the first draft of the discussion.

Closes #61.

@@
 While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AA per year for validation data and 0.66 AA per year for test data.
-These results are consistent with the observed dynamics of A/H3N2.
-First, the one-year forecast horizon is a fraction of the average coalescence time for A/H3N2 populations of about 3--8 years \cite{}.
+However, these results are consistent with the observed dynamics of A/H3N2.
+First, the one-year forecast horizon is a fraction of the average coalescence time for A/H3N2 populations of about 3--8 years \cite{Rambaut:2008ew}.
 Hence, we expect the diversity of circulating strains to persist between seasons.
-Second, the average substitution rate for A/H3N2's HA is 0.0043 substititions per site per year or an average accumulation of 2.4 AAs per year.
+Second, the average substitution rate for A/H3N2's HA is 0.0043 substitutions per site per year or an average accumulation of 2.4 AAs per year \cite{}.
 This accumulation of substitutions contributes to the distance between annual populations observed by the naive model.
@@
 Influenza collaborating centers now routinely measure antigenic drift by focus reduction assays (FRAs), as an alternative to hemagglutination inhibition (HI) assays that is less affected by glycosylation motifs that alter binding efficacy \cite{Zost2017}.
-While there are not enough historical FRA measurements to validate and test a FRA cross-immunity metric, it may be possible to reuse coefficients from the HI cross-immunity models with modern FRA data.
+While there are not enough historical FRA measurements to extensively validate and test a FRA cross-immunity metric, it may be possible to reuse coefficients from the HI cross-immunity models with modern FRA data.
+Further experiments will be required to fully understand how HI and FRA measurements of the same pairs of strains correlate with each other.
 In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
-These assays reveal the specific residues that confer antigenic escape at specific sites.
+These assays reveal the specific sites and residues that confer antigenic escape from polyclonal sera obtained from individual humans.
 A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape A/H3N2 strains must navigate to be globally successful.
@@
 Our framework enables real-time practical forecasts of these populations by leveraging historical and modern experimental assays and gene sequences.
-By releasing our framework as an open source tool based on modern data science standards like tidy data frames, we hope to encourage continued development by the influenza research community.
+By releasing our framework as an open source tool based on modern data science standards like tidy data frames, we hope to encourage continued development of this tool by the influenza research community.
 We additionally anticipate that the ability to forecast the sequence composition of populations with Earth Mover's Distance will enable future forecasting research with pathogens whose genomes cannot be analyzed by traditional phylogenetic methods including recombinant viruses, bacteria, and fungi.

### 2019-12-28 manuscript/main.tex
Add GISAID acknowledgements table and correct counts in methods

@@
 
-\jhc{Document HA sequences from GISAID (reference acknowledgements table).}
-
 \subsubsection*{Titers}
@@
 
-For model training and validation, we selected 13,601 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1990 and October 1, 2015.
+Prior to our analyses, we downloaded all HA sequences and metadata from GISAID \cite{Elbe2017}.
+For model training and validation, we selected 15,583 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1990 and October 1, 2015.
 To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 90 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
@@
 We prioritized samples with more available HI titer measurements.
-For model testing, we selected an additional 20,803 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and October 1, 2019.
-This increased number of sequences corresponds with the trend of increased sample sequencing and deposition in the last decade.
+For model testing, we selected an additional 7,171 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and October 1, 2019.
 We used these test sequences to evaluate the out-of-sample error of fixed model parameters learned during training and validation.
+Supplemental File S1 describes contributing laboratories for all 22,754 validation and test strains.
 

### 2019-12-28 manuscript/main.tex
Add brief description of HI data and prosaic author contributions
Both of HI data description and author contributions are placeholders until we
have discussed what else should be included here.

@@
 
-\subsection*{Data}
-
-\subsubsection*{Sequences}
-
-\subsubsection*{Titers}
-
-\jhc{Document HI titers from influenza CCs.}
-
 \subsection*{Simulation of influenza A/H3N2-like populations}
@@
 Next, we identified an L1 penalty of 0.1 for model coefficients that minimized the mean distance to the future for the true fitness model.
+
+\subsection*{Antigenic data}
+
+Hemagglutination inhibition (HI) measurements were provided by collaborating centers from the WHO Global Influenza Surveillance and Response System.
+We converted these raw two-fold dilution measurements to $log_{2}$ titer drops normalized by the corresponding $log_{2}$ autologous measurements as previously described \cite{Neher:2016hy}.
 
@@
 
-JH.
-RN.
-TB.
+JH planned experiments, implemented the final forecasting framework, analyzed results, and wrote the manuscript.
+RN planned experiments and reviewed the manuscript.
+TB planned experiments, implemented the initial forecasting framework, and reviewed the manuscript.

### 2019-12-30 manuscript/abstract.tex
Write first draft of abstract
Closes #62.

@@
-\section*{Key findings}
-
-\begin{itemize}
-\item{Our model accurately forecasts clade frequency trajectories and identifies optimal potential vaccine strains for both simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
-\item{Experimental measurements of antigenic drift based on HI assays are more robust and predictive of viral success than epitope mutations}
-\item{Models based on a predefined set of epitope sites risk overfitting to historical data}
-\item{The combination of non-epitope mutation and LBI metrics outperforms all experimentally-informed metrics on validation data (Oct 1990--Oct 2015) but fails to account for reassortment events in test data (Oct 2015--Oct 2019)}
-\item{The combination of HI assays and non-epitope mutations performs best on test data and is among the best models for validation data}
-\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
-\item{DMS measurements appear to be too background-specific to be predictive for global viral success}
-\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
-\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
-\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
-\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
-\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
-\end{itemize}
+Seasonal influenza virus A/H3N2 is a major cause of illness and death globally.
+While vaccination is the most effective way to prevent infection, rapid accumulation of mutations in the surface protein hemagglutinin (HA) allows viruses to escape previous adaptive immunity due to vaccination or natural infection.
+This antigenic drift requires vaccines to be updated regularly.
+Due to a nearly one-year lag between the selection of vaccine strains and deployment of the vaccine to the public, effective vaccine strain selection requires predictions about future influenza populations.
+Historically, experts predicted successful future viruses through manual inspection of hemagglutination inhibition (HI) assays, an experimental measure of antigenic drift.
+Modern strain selection is increasingly informed by modeling of influenza fitness using HA sequences to estimate antigenic drift, mutational load, and recent population growth.
+The models rely entirely on genotypic information without accounting for the gold standard phenotypic measures of antigenic drift provided by HI assays.
+We developed a novel open source forecasting framework based on these original fitness models that explicitly integrates genotypic and phenotypic measures of antigenic drift as well as modern experimental measurements of mutational load from deep mutational scanning (DMS) experiments.
+In contrast with previous work, our models minimize the distance between observed and estimated future populations using the Earth Mover's Distance metric instead of minimizing errors in frequencies of phylogenetic clades.
+We show that this integrated approach to forecasting can accurately estimate the sequence compositions of simulated and natural populations.
+These forecasts capture dynamics of influenza clades and enable the identification of optimal individual strains for vaccine composition.
+Using this framework and a modern machine learning approach to model testing, we found that phenotypic measures of antigenic drift were more consistently predictive of future influenza populations than sequence-only measures while the opposite was true for measures of mutational load.
+Importantly, the combination of antigenic drift and mutational load was the most predictive of future success.
+We provide real-time forecasts of seasonal influenza A/H3N2 populations based on our combined model of antigenic drift and mutational load as part of Nextstrain.
+As our open source framework allows definition of fitness metrics per sample using tidy data frames and does not require phylogenetic clades for forecasts, we hope it paves the way for future forecasting efforts with pathogens that cannot be readily analyzed by standard phylogenetic methods including recombinant viruses, bacteria, and fungi.

### 2019-12-30 manuscript/main.tex
Add equation for weighted Hamming distance, define D0 constant

@@
 
-The weighted Hamming distance to the future of each strain was calculated from the strain's HA amino acid sequence and the frequencies and sequences of the corresponding population one year in the future.
-\jhc{Add equation for weighted Hamming distance here.}
+We calculated the weighted Hamming distance to the future of each strain from the strain's HA amino acid sequence and the frequencies and sequences of the corresponding population one year in the future.
+Specifically, the distance between any strain $i$ from timepoint $t$ to the future timepoint $u$ was the Hamming distance, $h$, between strain $i$'s amino acid sequence, $s_{i}$, each future strain $j$'s amino acid sequence, $s_{j}$, and the frequency of strain $j$ in the future timepoint, $x_{j}(u)$.
+
+\begin{equation}
+  d_{i}(u) = \sum_{j \in s(u)}{x_{j}(u)h(s_{i}, s_{j})}
+  \label{equation_observed_weighted_distance_to_future}
+\end{equation}
+
+We calculated the estimated distance to the future for live forecasts with the same approach, replacing the observed future population frequencies and sequences with the estimated population based on our models.
+
+\begin{equation}
+  d_{i}(\hat{u}) = \sum_{j \in s(\hat{u})}{x_{j}(\hat{u})h(s_{i}, s_{j})}
+  \label{equation_estimated_weighted_distance_to_future}
+\end{equation}
 
@@
 We limited pairwise comparisons to all strains sampled within the last five years from each timepoint.
-For each individual strain $i$ at each timepoint $t$, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as follows.
+For each individual strain $i$ at each timepoint $t$, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as in Equation~\ref{equation_epitope_cross_immunity}.
+We defined the constant $D_{0} = 14$, as in the original definition of cross-immunity \cite{Luksza:2014hj}.
 

### 2019-12-30 manuscript/main.tex
Corrected statement about overfitting of epitope sites
The seven Koel sites did not overfit to historical data.

@@
 Other attempts to define these so-called ``epitope sites'' have relied on either aggregation of results from antigenic escape assays \cite{Wolf:2006da} or retrospective computational analyses of sites with beneficial mutations \cite{Shih:2007bd,Luksza:2014hj}.
-We found that all of these specific definitions suffer from overfitting to the historical context in which they were discovered (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
+We found that models based on all of these definitions except for the seven Koel epitope sites overfit to the historical context from which they were identified (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
 These results suggest that the set of sites that contribute to antigenic drift at any given time may depend on both the fitness landscape of currently circulating strains and the immune landscape of the hosts these strains need to infect.

### 2019-12-30 manuscript/abstract.tex
Copyedit main and abstract text
Correct spelling and clarify wording.

@@
 Historically, experts predicted successful future viruses through manual inspection of hemagglutination inhibition (HI) assays, an experimental measure of antigenic drift.
-Modern strain selection is increasingly informed by modeling of influenza fitness using HA sequences to estimate antigenic drift, mutational load, and recent population growth.
+Modern strain selection is increasingly informed by modeling of influenza fitness using HA sequences to estimate antigenic drift, functional constraint, and recent population growth.
 The models rely entirely on genotypic information without accounting for the gold standard phenotypic measures of antigenic drift provided by HI assays.
-We developed a novel open source forecasting framework based on these original fitness models that explicitly integrates genotypic and phenotypic measures of antigenic drift as well as modern experimental measurements of mutational load from deep mutational scanning (DMS) experiments.
-In contrast with previous work, our models minimize the distance between observed and estimated future populations using the Earth Mover's Distance metric instead of minimizing errors in frequencies of phylogenetic clades.
+We developed a novel open source forecasting framework based on these original fitness models that explicitly integrates genotypic and phenotypic measures of antigenic drift as well as modern experimental measurements of functional constraint from deep mutational scanning (DMS) experiments.
+In contrast with previous work, our models minimize the distance between observed and estimated future populations using the Earth Mover's Distance metric instead of minimizing errors between observed and estimated phylogenetic clade frequencies.
 We show that this integrated approach to forecasting can accurately estimate the sequence compositions of simulated and natural populations.
 These forecasts capture dynamics of influenza clades and enable the identification of optimal individual strains for vaccine composition.
-Using this framework and a modern machine learning approach to model testing, we found that phenotypic measures of antigenic drift were more consistently predictive of future influenza populations than sequence-only measures while the opposite was true for measures of mutational load.
+Using this framework and a modern machine learning approach to model testing, we found that phenotypic measures of antigenic drift were more consistently predictive of future influenza populations than sequence-only measures while the opposite was true for measures of functional constraint.
 Importantly, the combination of antigenic drift and mutational load was the most predictive of future success.

### 2019-12-30 manuscript/main.tex
Copyedit main and abstract text
Correct spelling and clarify wording.

@@
     Corresponding values per test timepoint are in the top-right.
-    For reference, the naive model's distance to the future is also shown per validation and test timepoint in grey.
+    For reference, the naive model's distance to the future is also shown per validation and test timepoint in gray.
     The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 9.07 $\pm$ 1.71 AAs for test timepoints.
@@
   The estimated best strain was in the top 20th percentile of observed closest strains for 89\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
-  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
@@
 Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
-These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
+These results confirm that our approach of minimizing the distance between yearly populations can simultaneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
 These results also highlight a well-established yet critical point that forecasts on truly out-of-sample data are less accurate than those based on training and validation data.
@@
   The estimated best strain was in the top 20th percentile of observed closest strains for 88\% of timepoints.
-  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
@@
 We have further introduced novel fitness metrics based on experimental measurements of antigenic drift and functional constraint.
-We demonstrated how these metrics can be effectively integrated with previously published sequence-only metrics.
+We demonstrated that the integration of these phenotypic metrics with previously published sequence-only metrics produces more accurate forecasts than sequence-only models.
 We have added this framework as a component of seasonal influenza analyses on nextstrain.org where it provides real-time forecasts for influenza researchers, decision makers, and the public.
@@
 When we applied these measurements to predict the success of global populations, they were less informative on average than the naive model.
-To benefit from the more comprehensive fitness costs measured by DMS data, future models will need to synthesize DMS measurements across multiple A/H3N2 strains from distinct historical contexts.
+To benefit from the more comprehensive fitness costs measured by DMS data, future models will need to synthesize DMS measurements across multiple A/H3N2 strains from distinct genetic contexts.
 We anticipate that these measurements could be used to define and continually update a modern set of sites contributing to mutational load in natural populations.
@@
 Despite this model's excellent performance historically, it was unable to account for rare yet important events such as reassortment.
-To avoid such overfitting in the future, models will need fitness metrics that can represent the evolutionary contributions of neuraminidase, specifically, or of reassortment, generally.
+To avoid such overfitting in the future, models will need fitness metrics that can balance the fitness benefits of evolution in other genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment itself \cite{Villa:2017iw}.
 
@@
 By forecasting the composition of future A/H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
-While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AA per year for validation data and 0.66 AA per year for test data.
+While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AAs per year for validation data and 0.66 AAs per year for test data.
 However, these results are consistent with the observed dynamics of A/H3N2.
@@
 
-There are still several clear opportunities to improve forecasts.
+Several clear opportunities to improve forecasts still remain.
 Integration of more recent experimental data may improve estimates of antigenic drift.
@@
 This bandwidth also roughly corresponds to the median lag time between sample collection and submission to the GISAID database.
-We estimated the frequency of each sample at each timepoint by calculating the probabilitiy density function of each KDE at that timepoint and normalizing the resulting values to sum to one.
+We estimated the frequency of each sample at each timepoint by calculating the probability density function of each KDE at that timepoint and normalizing the resulting values to sum to one.
 We implemented this frequency estimation logic in the augur frequencies command.
@@
 Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
-This solution requires both a ``ground distance'' between pairs fo samples from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.
+This solution requires both a ``ground distance'' between pairs of samples from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.
 

### 2019-12-30 manuscript/main.tex
More copy editing including past tense consistency
Also corrects an old conclusion in the introduction written prior to the test
data results for natural populations.

@@
 With this framework, we show that HI phenotypes enable more accurate long-term forecasts of A/H3N2 populations compared to previous metrics based on epitope mutations alone.
-However, we also find that phylogenetic fitness metrics based on recent growth of circulating clades consistently outperform any combination of genotypic or phenotypic metrics, suggesting that existing mechanistic models of seasonal influenza evolution are missing critical components.
+However, we also find that composite models based on phenotypic measures of antigenic drift and genotypic measures of functional constraint consistently outperform any fitness models based on individual genotypic or phenotypic metrics.
 
@@
 
-Here, we present a model of seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
-As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
-We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
+We developed a framework to forecast seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
+We estimated the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
 
-We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
+We estimated viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
 All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.

### 2019-12-30 src/forecast_model.py
Revert to simpler best composite model for practical forecasting and annotate models in forecast output for debugging

@@ -167,4 +167,5 @@ if __name__ == "__main__":
     # Save forecasts table, if requested.
     if args.output_table:
         all_forecasts = pd.concat(forecasts, ignore_index=True)
+        all_forecasts["model"] = "-".join(predictors)
         all_forecasts.to_csv(args.output_table, sep="\t", index=False, header=True, na_rep="N/A")

### 2019-12-31 manuscript/cover-letter/cover_letter.tex
Add cover letter

\documentclass[stdletter,letterpaper,addrfromright,orderfromdateto,dateleft,11pt,noaddrto,sigleft]{newlfm}
\topmarginskip{-0.0in}
\bottommarginskip{-0.0in}
\leftmarginsize{1.25in}
\rightmarginsize{1.5in}
\sigskipbefore{0.2in}
\sigskipafter{0in}
\noLines
\nolines
\noHeadline
\noheadline
\signature{John Huddleston}

\namefrom{}
\addrfrom{
	\includegraphics[width=6.5cm]{figures/fhcrc_logo} \\
  Vaccines and Infectious Disease Division \\
  Fred Hutchinson Cancer Research Center \\
  1100 Fairview Ave N \\
  Seattle, WA 98109, USA}
\emailfrom{jlhudd@uw.edu}

\greetto{Dear Editor,}
\closeline{\\Sincerely,}

% comments
\usepackage{color}
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}}
\def\tbc#1{\textcolor{purple}{[#1]}}

\begin{document}

\begin{newlfm}
  Please find attached our manuscript entitled ``Integrating genotypes and phenotypes improves long-term forecasts of seasonal influenza A/H3N2 evolution''.
  We would be grateful if you considered it for publication in \textit{eLife}.
  We believe that \textit{eLife} is an appropriate venue, as this work provides the most comprehensive analysis of fitness metrics for seasonal influenza to date and a generalized forecasting framework that will be a valuable tool for other biological researchers.

  The selection of vaccine strains for seasonal influenza A/H3N2 remains a daunting challenge due to constant accumulation of mutations in the hemagglutinin (HA) surface protein.
  Until recently, experts selected vaccine strains by manually synthesizing experimental measurements of antigenic drift provided by hemagglutination inhibition (HI) assays.
  Two landmark papers in 2014 showed that H3N2 evolution may be predictable from HA sequences alone.
  The first paper, ``A predictive fitness model for influenza'' by {\L}uksza and L\"assig in \textit{Nature}, established a mechanistic fitness model that predicted the frequencies of phylogenetic clades one year in advance.
  This model defined three biological fitness metrics including antigenic drift measured by mutations at epitope sites, functional constraint measured by mutations at non-epitope sites, and recent clade growth measured by synonymous mutations.
  The second paper, ``Predicting evolution from the shape of genealogical trees'' by Neher et al. in \textit{eLife}, developed a phylogenetic metric of recent clade fitness called the local branching index (LBI).
  This metric accurately identified high fitness clades without providing a biological basis for that success.
  Predictions from both of these models are now routinely considered in the World Health Organization's vaccine composition meetings.

  Several substantial questions remained unanswered by these models.
  Are genetic estimates of antigenic drift sufficient replacements for the gold standard phenotypic measures provided by HI assays?
  What mechanistic components of influenza fitness does LBI measure?
  Can simple genetic estimates of functional constraint be improved upon using modern deep mutational scanning (DMS) assays that measure the functional effects of every single amino acid mutation to a given strain's HA?
  Here, we investigate for the first time the relationships between all of these genotypic and phenotypic estimates of influenza fitness using a collection of HA sequences and HI measurements of an unprecedented breadth and depth.
  To this end, we developed a novel approach to forecasting influenza evolution by estimating the sequence composition of future populations and rigorously tested this framework with simulated and natural H3N2 populations.
  We implemented new experimentally-informed fitness metrics including antigenic drift by HI assays and functional constraint by DMS assays.
  We applied modern machine learning methods for model training, validation, and testing to thoroughly evaluate the predictive accuracy of these new metrics as well as {\L}uksza and L\"assig's epitope and non-epitope metrics and Neher et al.'s LBI.
  Importantly, we verified that this framework can both capture dynamics of phylogenetic clades and identify optimal candidate strains for vaccine composition.

  We find that phenotypic measures of antigenic drift consistently outperform genetic estimates based on epitope sites.
  We further show that the predictive value of these epitope sites in {\L}uksza and L\"assig's mechanistic model was most likely due to inadvertantly borrowing information from the future during model fitting.
  This surprising finding has important ramifications for future attempts to map influenza genotypes to fitness, the use of epitope sites in practical forecasts, and the continued value of HI assays.
  Additionally, we find that DMS-based estimates of functional constraint produced models that overfit to the genetic background of the strain, A/Perth/16/2009, used in the experiments.
  Instead, we find that {\L}uksza and L\"assig's original minimal model of mutational load at non-epitope sites is more robust.
  Composite models that integrate mutational load with either HI phenotypes or LBI performed the best and indicate that LBI may be measuring similar fitness effects as HI assays.
  We have integrated these models into nextstrain.org, enabling real-time forecasts for H3N2 for influenza researchers, decision makers, and the public.

  Our findings challenge the widely held assumption about the power of epitope sites to predict the success of influenza populations.
  Correspondingly, we validate the predictive value of HI assays to measure antigenic drift and provide a framework that synthesizes these measurements into forecasts for expert interpretation.
  Beyond influenza, our framework enables similar forecasting efforts for other viral and bacterial pathogens by allowing researchers to define sample-specific fitness metrics in tidy data frames and fit models to those metrics without relying on phylogenetic inference.
  We are confident this open source framework will provide a foundation on which future forecasting efforts can be built or inspired.
  For these reasons, our software and findings are relevant not only to the specific audience of influenza researchers, but also to the more general audience of infectious disease researchers, evolutionary biologists, and bioinformaticians.
\end{newlfm}

\end{document}

### 2019-12-31 manuscript/main.tex
Update TB funding
Acknowledge Bing, simplify funding text

@@
 
-We thank Jesse Bloom, Erick Matsen, Harmit Malik, Sidney Bell, Alli Black, Lola Arakaki, and members of the Bedford lab for useful advice and discussions.
-JH is a Graduate Research Fellow and is supported by the National Institute Of Allergy And Infectious Diseases under NIH F31AI140714.
-TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.
+We thank Jesse Bloom, Erick Matsen, Bing Brunton, Harmit Malik, Sidney Bell, Alli Black, Lola Arakaki, and members of the Bedford lab for useful advice and discussions.
+JH is a Graduate Research Fellow and is supported by the NIH grant NIAID F31AI140714.
+TB is a Pew Biomedical Scholar and is supported by NIH grants NIGMS R35 GM119774-01, NIAID U19 AI117891-01 and NIAID R01 AI127893-01.
 \jhc{Add RN support here.}

### 2020-01-02 manuscript/main.tex
Use H3N2 after initial use of A/H3N2, declare competing interests

@@
   \caption{
-    Schematic representation of the fitness model for simulated A/H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
+    Schematic representation of the fitness model for simulated H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
     Samples are colored by their amino acid sequence composition such that genetically similar samples have similar colors (Methods).
@@
 
-\subsection*{Models accurately forecast evolution of A/H3N2-like viruses}
+\subsection*{Models accurately forecast evolution of H3N2-like viruses}
 
-The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Fig.~\ref{sup_fig:cross_validation_for_simulated_populations}).
-These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Fig.
+The long-term evolution of influenza H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Fig.~\ref{sup_fig:cross_validation_for_simulated_populations}).
+These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural H3N2 HA (Supplemental Fig.
 \ref{sup_fig:simulated_h3n2_ha_phylogeny}, Supplemental Tables~\ref{sup_table:mutations_by_trunk_status_for_simulated_populations} and \ref{sup_table:mutations_by_trunk_status}).
@@
 
-\subsection*{Models reflect historical patterns of A/H3N2 evolution}
+\subsection*{Models reflect historical patterns of H3N2 evolution}
 
@@
 
-Next, we trained and validated models for individual fitness predictors using 25 years of natural A/H3N2 populations spanning from October 1, 1990 to October 1, 2015.
+Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
 We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
@@
 We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
-As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than mutational load or DMS mutational effects.
+As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than mutational load or DMS mutational effects.
 Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
@@
   \caption{
-  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity and mutational load.
+  Test of best model for natural populations of H3N2 viruses, the composite model of HI cross-immunity and mutational load.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
   \caption{
-    Observed distance to natural A/H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated (orange) closest strains to the future at the corresponding timepoints.
+    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated (orange) closest strains to the future at the corresponding timepoints.
     Vaccine strains were assigned to the validation or test timepoint closest to the date they were selected by the WHO.
@@
 
-We further evaluated our best model's apparent success at estimating the closest strain to the next season's A/H3N2 population by comparing our model's selection to the WHO's vaccine strain selection.
+We further evaluated our best model's apparent success at estimating the closest strain to the next season's H3N2 population by comparing our model's selection to the WHO's vaccine strain selection.
 For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed and estimated closest strains.
@@
 
-To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}, and used our best model to forecast A/H3N2 populations in September 2020.
+To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}, and used our best model to forecast H3N2 populations in September 2020.
 Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning coefficients per timepoint from the test data.
@@
   \caption{
-    Snapshot of the last two years of seasonal influenza A/H3N2 evolution on nextstrain.org showing the estimated distance per sample to the future population.
+    Snapshot of the last two years of seasonal influenza H3N2 evolution on nextstrain.org showing the estimated distance per sample to the future population.
     Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating samples weighted by the other sample's projected frequencies under the best fitness model.
@@
 
-We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza A/H3N2 by estimating the sequence composition of future populations.
+We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza H3N2 by estimating the sequence composition of future populations.
 A key innovation of this framework is its ability to directly compare viral populations between seasons using the Earth Mover's Distance metric \cite{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
-The best models from this framework still effectively capture clade dynamics and accurately identify optimal vaccine candidates from simulated and natural A/H3N2 populations without relying on clades as model targets.
+The best models from this framework still effectively capture clade dynamics and accurately identify optimal vaccine candidates from simulated and natural H3N2 populations without relying on clades as model targets.
 We have further introduced novel fitness metrics based on experimental measurements of antigenic drift and functional constraint.
@@
 When we applied these measurements to predict the success of global populations, they were less informative on average than the naive model.
-To benefit from the more comprehensive fitness costs measured by DMS data, future models will need to synthesize DMS measurements across multiple A/H3N2 strains from distinct genetic contexts.
+To benefit from the more comprehensive fitness costs measured by DMS data, future models will need to synthesize DMS measurements across multiple H3N2 strains from distinct genetic contexts.
 We anticipate that these measurements could be used to define and continually update a modern set of sites contributing to mutational load in natural populations.
@@
 In the second case, we observed overfitting of sites associated with antigenic drift to historical contexts.
-The fitness benefit of mutations that lead to antigenic drift in A/H3N2 is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}
+The fitness benefit of mutations that lead to antigenic drift in H3N2 is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}
 Although the antigenic importance of seven specific sites in HA has been experimentally validated \cite{Koel:2013jz}, these sites do not explain all antigenic drift observed in natural populations \cite{Neher:2016hy}.
@@
 These results suggest that the set of sites that contribute to antigenic drift at any given time may depend on both the fitness landscape of currently circulating strains and the immune landscape of the hosts these strains need to infect.
-Recent experimental mapping of antigenic escape mutations in A/H3N2 HA with human sera show that the specific sites that confer antigenic escape can vary dramatically between individuals based on their exposure history \cite{Lee2019}.
+Recent experimental mapping of antigenic escape mutations in H3N2 HA with human sera show that the specific sites that confer antigenic escape can vary dramatically between individuals based on their exposure history \cite{Lee2019}.
 In contrast to models based on predefined ``epitope sites'', our model based on experimental measurements of antigenic drift did not suffer from overfitting in the validation or test periods.
@@
 
-By forecasting the composition of future A/H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
+By forecasting the composition of future H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
 While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AAs per year for validation data and 0.66 AAs per year for test data.
-However, these results are consistent with the observed dynamics of A/H3N2.
-First, the one-year forecast horizon is a fraction of the average coalescence time for A/H3N2 populations of about 3--8 years \cite{Rambaut:2008ew}.
+However, these results are consistent with the observed dynamics of H3N2.
+First, the one-year forecast horizon is a fraction of the average coalescence time for H3N2 populations of about 3--8 years \cite{Rambaut:2008ew}.
 Hence, we expect the diversity of circulating strains to persist between seasons.
-Second, the average substitution rate for A/H3N2's HA is 0.0043 substitutions per site per year or an average accumulation of 2.4 AAs per year \cite{}.
+Second, the average substitution rate for H3N2's HA is 0.0043 substitutions per site per year or an average accumulation of 2.4 AAs per year \cite{}.
 This accumulation of substitutions contributes to the distance between annual populations observed by the naive model.
@@
 These assays reveal the specific sites and residues that confer antigenic escape from polyclonal sera obtained from individual humans.
-A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape A/H3N2 strains must navigate to be globally successful.
+A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape H3N2 strains must navigate to be globally successful.
 
 Finally, forecasting models need to account for the geographic distribution of viruses.
-Globally successful clades of influenza A/H3N2 most commonly emerge from Southern Asia, while viruses emerging from Africa, North America, and Europe are less likely to fix globally \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
+Globally successful clades of influenza H3N2 most commonly emerge from Southern Asia, while viruses emerging from Africa, North America, and Europe are less likely to fix globally \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
 Explicitly accounting for these source/sink dynamics would allow models to weight forecasts based on both viral fitness and transmission.
@@
 
-Antigenic drift and mutational load remain the best predictors of future success for seasonal influenza A/H3N2 populations.
+Antigenic drift and mutational load remain the best predictors of future success for seasonal influenza H3N2 populations.
 Our framework enables real-time practical forecasts of these populations by leveraging historical and modern experimental assays and gene sequences.
@@
 
-\subsection*{Simulation of influenza A/H3N2-like populations}
+\subsection*{Simulation of influenza H3N2-like populations}
 
-We simulated the long-term evolution of A/H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 10,000 generations or 50 years where 200 generations was equivalent to 1 year.
+We simulated the long-term evolution of H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 10,000 generations or 50 years where 200 generations was equivalent to 1 year.
 We discarded the first 10 years as a burn-in period, selected the next 30 years for model fitting and validation, and held out the last 9 years as out-of-sample data for model testing.
@@
 
-To avoid overfitting our models to the relatively limited data from natural populations, we used simulated A/H3N2-like populations to tune hyperparameters including the KDE bandwidth for frequency estimates and the L1 penalty for model coefficients.
+To avoid overfitting our models to the relatively limited data from natural populations, we used simulated H3N2-like populations to tune hyperparameters including the KDE bandwidth for frequency estimates and the L1 penalty for model coefficients.
 We simulated populations, as described above, and fit models for each parameter value using the true fitness of samples from the simulator.
@@
 
-We assumed that the evolution seasonal influenza A/H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \cite{Luksza:2014hj}.
+We assumed that the evolution seasonal influenza H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \cite{Luksza:2014hj}.
 Under this model, we estimated the future frequency, $\hat{x}_{i}(t + \Delta{t})$, of each sample $i$ from the sample's current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
@@
 TB planned experiments, implemented the initial forecasting framework, and reviewed the manuscript.
+
+\section*{Competing interests}
+
+The authors declare that no competing interests exist.

### 2020-01-02 manuscript/supplement.tex
Use H3N2 after initial use of A/H3N2, declare competing interests

@@
   \caption{
-  Phylogeny of A/H3N2-like HA sequences sampled between the 24th and 30th years of simulated evolution.
+  Phylogeny of H3N2-like HA sequences sampled between the 24th and 30th years of simulated evolution.
   The phylogenetic structure and rate of accumulated epitope and non-epitope mutations match patterns observed in phylogenies of natural sequences.
@@
   \caption{
-  Validation of best model for simulated populations of A/H3N2-like viruses.
+  Validation of best model for simulated populations of H3N2-like viruses.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
   \caption{
-  Validation of best model for natural populations of A/H3N2 viruses, the composite model of mutational load and LBI.
+  Validation of best model for natural populations of H3N2 viruses, the composite model of mutational load and LBI.
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.

### 2020-01-02 manuscript/abstract.tex
Edit abstract to 150 words or less
This is 149 words according to emacs word count and 144 according to wc.

--- 2019-12-19 manuscript/abstract.tex
@@
-\section*{Key findings}
-
-\begin{itemize}
-\item{Our model accurately forecasts clade frequency trajectories and identifies optimal potential vaccine strains for both simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
-\item{Experimental measurements of antigenic drift based on HI assays are more robust and predictive of viral success than epitope mutations}
-\item{Models based on a predefined set of epitope sites risk overfitting to historical data}
-\item{The combination of non-epitope mutation and LBI metrics outperforms all experimentally-informed metrics on validation data (Oct 1990--Oct 2015) but fails to account for reassortment events in test data (Oct 2015--Oct 2019)}
-\item{The combination of HI assays and non-epitope mutations performs best on test data and is among the best models for validation data}
-\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
-\item{DMS measurements appear to be too background-specific to be predictive for global viral success}
-\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
-\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
-\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
-\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
-\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
-\end{itemize}
+Seasonal influenza virus A/H3N2 is a major cause of death globally.
+Vaccination remains the most effective preventative.
+Rapid mutation of hemagglutinin allows viruses to escape adaptive immunity.
+This antigenic drift necessitates regular vaccine updates.
+Effective vaccine strains need to match H3N2 populations circulating one year after strain selection.
+Experts select strains based on experimental measurements of antigenic drift and predictions made by models from hemagglutinin sequences.
+No modern models use phenotypic measures of viral fitness in their predictions.
+We developed a novel influenza forecasting framework that integrates phenotypic measures of antigenic drift and functional constraint with previously published sequence-only fitness estimates.
+Forecasts informed by phenotypic measures of antigenic drift consistently outperformed previous sequence-only estimates, while sequence-only estimates of functional constraint surpassed more comprehensive experimentally-informed estimates.
+Importantly, the best models integrated estimates of both functional constraint and either antigenic drift phenotypes or recent population growth.

### 2020-01-03 manuscript/main.tex
Include FRA background and results
Restructures main text to introduce FRA assays early on as a more recent
alternative to HI assays, describe the limited availability of FRA data for
training/validation/testing along HI models, include FRA results in the final
results section, and reduce the description of FRAs as future directions in the
discussion. This commit also include an additional citation for the
glycosylation motifs in modern H3N2s (Chambers et al. 2015) and a citation for
the FRAs (Okuno et al. 1990).

@@
 
-Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}.
+Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually with the majority of infections caused by influenza A/H3N2 \cite{flufactsheet}.
 Vaccination remains the most effective public health response available.
@@
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
-The hemagglutination inhibition (HI) assay \cite{hirst1943studies} is used to measure the degree of cross-reactivity between pairs of circulating viruses.
-HI assays are fundamental for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
+Until recently, the hemagglutination inhibition (HI) assay has been the primary experimental measure of antigenic cross-reactivity between pairs of circulating viruses \cite{hirst1943studies}.
+Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \cite{Chambers:2015jt,Zost2017}.
+Correspondingly, many influenza collaborating centers now also measure cross-reactivity by focus reduction assays (FRAs) \cite{Okuno:1990uu}.
+Together, these two assays are the gold standard for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
 As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.
@@
 Recent developments in computational methods and influenza virology have made it feasible to integrate these important metrics of influenza fitness into a single predictive model.
-For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \cite{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning experiments \cite{Lee2018}.
+For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \cite{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning (DMS) experiments \cite{Lee2018}.
 We describe an approach to integrate previously disparate sequence-only models of influenza evolution with high-quality experimental measurements of antigenic drift and functional constraint.
@@
 Here, we provide the first such open source long-term forecasting framework for seasonal influenza.
-With this framework, we show that HI phenotypes enable more accurate long-term forecasts of A/H3N2 populations compared to previous metrics based on epitope mutations alone.
+With this framework, we show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
 However, we also find that composite models based on phenotypic measures of antigenic drift and genotypic measures of functional constraint consistently outperform any fitness models based on individual genotypic or phenotypic metrics.
@@
 Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
+As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA cross-immunity models in parallel with the HI cross-immunity models.
 
@@
 In the last two validation timepoints, this composite model converged to a nearly LBI-only model with the coefficients for the other two metrics converging to near-zero values.
-These last two timepoints span the period when an HA1:160T substitution swept through the global population.
-This substitution introduced a new glycosylation motif and dramatically reduced the efficiency of HI assays \cite{Zost2017}.
-These results reinforce the historical importance of HI assays in measuring viral fitness, while also indicating that these assays may no longer be as predictive as sequence-only metrics.
+These results suggest that HI assays and LBI may measure similar fitness effects.
+However, the dominance of LBI later in this three-metric model also suggests that HI assays may no longer be as predictive as they have been historically.
 
@@
 
-\subsection*{Forecasts predict the rise of A1b/131K and A1b/135K sub-clades in September 2020}
+\subsection*{Historically-trained models enable real-time, actionable forecasts}
 
-To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}, and used our best model to forecast H3N2 populations in September 2020.
-Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning coefficients per timepoint from the test data.
-None of the three models performed better across the test data than the original best model, indicating that the coefficients for our selected best model would be most accurate for live forecasts (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
-With the original coefficients, our combined HI cross-immunity and mutational load model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
-To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
+To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}.
+Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning new coefficients per timepoint from the test data.
+In addition to three composite models described above, we evaluated a composite of FRA cross-immunity and mutational load.
+Since the earliest FRA data are from 2012, we anticipated that there were enough measurements to fit a model across the test data time interval.
+If modern H3N2 strains continue to perform poorly in HI assays, the FRA-based assay will be critical for future forecasting efforts.
+
+None of the three original composite models performed better across the test data than the original best model (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
+These results confirmed that the coefficients for our selected best model would be most accurate for live forecasts.
+Interestingly, the FRA cross-immunity metric received a consistently positive coefficient in its composite with mutational load and this composite ranked among the best models for the test data.
+These results indicate that this FRA-based composite could eventually replace the HI-based composite with mutational load.
+
+After confirming the coefficients for our best model of HI cross-immunity and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021.
+Of the three major currently circulating clades, our model predicted the growth of A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
+To aid with identification of potential vaccine candidates for the next season, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
 
@@
 Integration of more recent experimental data may improve estimates of antigenic drift.
-Influenza collaborating centers now routinely measure antigenic drift by focus reduction assays (FRAs), as an alternative to hemagglutination inhibition (HI) assays that is less affected by glycosylation motifs that alter binding efficacy \cite{Zost2017}.
-While there are not enough historical FRA measurements to extensively validate and test a FRA cross-immunity metric, it may be possible to reuse coefficients from the HI cross-immunity models with modern FRA data.
-Further experiments will be required to fully understand how HI and FRA measurements of the same pairs of strains correlate with each other.
+The strong performance of a composite FRA cross-immunity and mutational load model with recent H3N2 populations indicates that FRAs will continue to be important for future forecasting efforts.
 In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
@@
 A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape H3N2 strains must navigate to be globally successful.
-
 Finally, forecasting models need to account for the geographic distribution of viruses.
@@
 Explicitly accounting for these source/sink dynamics would allow models to weight forecasts based on both viral fitness and transmission.
-For example, a geographically-aware implementation of LBI would most likely have not incorrectly predicted the success of the predominantly North American and European reassortant clade A2/re in 2019.
 

### 2020-01-03 manuscript/supplement.tex
Include FRA background and results
Restructures main text to introduce FRA assays early on as a more recent
alternative to HI assays, describe the limited availability of FRA data for
training/validation/testing along HI models, include FRA results in the final
results section, and reduce the description of FRAs as future directions in the
discussion. This commit also include an additional citation for the
glycosylation motifs in modern H3N2s (Chambers et al. 2015) and a citation for
the FRAs (Okuno et al. 1990).

@@
   \caption{
-    Model coefficients and distances to the future for best composite models fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
     A) Coefficients and B) distances are shown per test timepoint (N=8).

### 2020-01-03 manuscript/main.tex
edits and comments by RN

@@
 However, frequent viral mutation results in viruses that escape previously acquired human immunity.
-The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
+The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine strain selection requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
@@
 Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, these methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
-Here, we provide the first such open source long-term forecasting framework for seasonal influenza.
+Here, we provide the first such open source framework for forcasting the genetic composision of future seasonal influenza populations.
 With this framework, we show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
@@
   \caption{
+  \rnc{missing panel labels!}
     Schematic representation of the fitness model for simulated H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
@@
 As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
-We estimated the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
+We estimated the frequency of virus samples every six months using kernel density estimation (KDE) to represent the frequency of each sample.
 
@@
 Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
+\rnc{Hamming distance or EMD?}
 With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
@@
 
+\rnc{wouldn't we expect the true fitness have coefficent 1.0? A plausible reviewer question would be ``how does the model do with coefficient 1.0?"}
 As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 (97\%) of 33 timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
@@
   Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
+  \rnc{Growth rates have units. Is this 1/year? Also what exactly is $\log_{10}$ growth rate? -2 == 0.01/year?}
   A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
   C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
+  \rnc{it would be interesting to look at this for neg and pos deviations separately.}
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
@@
 Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.
+Note, that the numbers of test and validation data points are small and we therefore expect fluctuations.
 
@@
 While observed and estimated growth rates correlated more weakly for test data (Pearson's $R^2 = 0.14$, $p < 0.001$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
+\rnc{this is not a supplemental figure, right?}
 Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
@@
 Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
+\rnc{the p-value us probably not meaningful due to phylogenetic relatiobships.}
 These results confirm that our approach of minimizing the distance between yearly populations can simultaneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
 These results also highlight a well-established yet critical point that forecasts on truly out-of-sample data are less accurate than those based on training and validation data.
+\rnc{Is the above statement accurate?
+this refers to the true fitness model which has a next to constant coefficient around 9.
+Risk of overfitting is minimal and the only measure that is substantially worse is Pearson's $r^2$ which is sensitive to outliers.
+Furthermore, the clades on which panel A is based are not independent.}
 
@@
 
-The naive model measured a distance of 6.40 $\pm$ 1.36 AAs between natural populations or 71\% of the average distance between yearly simulated populations (Table~\ref{table_natural_model_selection}).
+\rnc{Just a suggestion:}
+Across our training period, HA sequences of H3N2 influenza populations one year apart was 6.40 $\pm$ 1.36 AAs.
+This distance corresponds to the ``naive model'' in Table~\ref{table_natural_model_selection}.
 Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
@@
 Indeed, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict to October 2005.
+\rnc{I thought this was predictive until 2008.}
 To further test our hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
-Inspired by the original approach of Shih et al.
-\cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population, including 12 that were also present in the original set of 49 sites.
+Inspired by the original approach of Shih et al.~\cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population, including 12 that were also present in the original set of 49 sites.
 We fit a cross-immunity model to these 25 sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefited from knowledge of the future in its forecasts.
 In contrast to the epitope cross-immunity model, the oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
-These results support our hypothesis that the success of previous epitope models based on a fixed set of 49 sites is due to historical contingency rather than a constant contribution of these sites to antigenic drift.
+These results support our hypothesis that fitness benefit of mutations at the original 49 sites was due to historical contingency and that the success of previous epitope models based on these sites partly due to ``borrowing from the future''.
 We suspect that our HI cross-immunity model benefits from its ability to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
@@
 TB is a Pew Biomedical Scholar and is supported by NIH grants NIGMS R35 GM119774-01, NIAID U19 AI117891-01 and NIAID R01 AI127893-01.
-\jhc{Add RN support here.}
+RAN is supported by NIAID R01 AI127893-01 and institutional core funding.
 

### 2020-01-03 manuscript/main.tex
edit discussion:

@@
 We demonstrated that the integration of these phenotypic metrics with previously published sequence-only metrics produces more accurate forecasts than sequence-only models.
-We have added this framework as a component of seasonal influenza analyses on nextstrain.org where it provides real-time forecasts for influenza researchers, decision makers, and the public.
+We have added this framework as a component of seasonal influenza analyses on \href{https://nextstrain.org/flu}{nextstrain.org} where it provides real-time forecasts for influenza researchers, decision makers, and the public.
 
@@
 
-We observed two different cases of overfitting to general historical contexts.
-In the first case, the model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
-Specifically, we found that this model failed to predict the sudden decline of a dominant reassortant clade, A2/re, in 2019.
-Despite this model's excellent performance historically, it was unable to account for rare yet important events such as reassortment.
-To avoid such overfitting in the future, models will need fitness metrics that can balance the fitness benefits of evolution in other genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment itself \cite{Villa:2017iw}.
-
-In the second case, we observed overfitting of sites associated with antigenic drift to historical contexts.
+We observed the latter type of overfitting in the context of sites associated with antigenic drift.
 The fitness benefit of mutations that lead to antigenic drift in H3N2 is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}
@@
 We suspect that this model was able to minimize overfitting by continuously updating its antigenic model with recent experimental data and assigning antigenic weight to branches of a phylogeny rather than specific positions in HA.
+
+Even well performing models with few parameters will sometimes fail due to the probabilistic nature of evolution: The model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
+Specifically, we found that this model failed to predict the sudden decline of a dominant reassortant clade, A2/re, in 2019.
+Despite this model's excellent performance historically, it was unable to account for rare yet important events such as reassortment.
 
@@
 Integration of more recent experimental data may improve estimates of antigenic drift.
+Models should integrate information from multiple segments of the influenza genome and will need to balance the fitness benefits of evolution in genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment \cite{Villa:2017iw}.
 The strong performance of a composite FRA cross-immunity and mutational load model with recent H3N2 populations indicates that FRAs will continue to be important for future forecasting efforts.
@@
 A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape H3N2 strains must navigate to be globally successful.
-Finally, forecasting models need to account for the geographic distribution of viruses.
-Globally successful clades of influenza H3N2 most commonly emerge from Southern Asia, while viruses emerging from Africa, North America, and Europe are less likely to fix globally \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
-Explicitly accounting for these source/sink dynamics would allow models to weight forecasts based on both viral fitness and transmission.
+Finally, forecasting models need to account for the geographic distribution of viruses and the vastly different sampling intensities across the globe.
+Most influenza sequence data come from highly developed countries that account for a small fraction of the global population, while globally successful clades of influenza H3N2 often emerge in less well sampled regions \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
+Explicitly accounting for these sampling biases and the associated migration dynamics would allow models to weight forecasts based on both viral fitness and transmission.
 
@@
 
-Antigenic drift and mutational load remain the best predictors of future success for seasonal influenza H3N2 populations.
+Prediction of future influenza virus populations are intrinsically limited by the small number of data points available to train and test models.
+Increasing the complexity therefore makes models prone to overfitting.
+Across the training and validation period, we found that antigenic drift and mutational load are the most robust predictors of future success for seasonal influenza H3N2 populations.
+
 Our framework enables real-time practical forecasts of these populations by leveraging historical and modern experimental assays and gene sequences.

### 2020-01-03 manuscript/main.tex
Address most comments from Richard
Makes the following minor changes in response to Richard's edits from the
previous commits:

 - spellcheck and edit text for clarity
 - note that distance is earth mover's distance (no need to capitalize this)
 - remove p-values as they are not meaningful for the non-independent clades
 - clarify that growth rates are actually growth ratios of frequencies

@@
 Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, these methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
-Here, we provide the first such open source framework for forcasting the genetic composision of future seasonal influenza populations.
+Here, we provide the first such open source framework for forecasting the genetic composition of future seasonal influenza populations.
 With this framework, we show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
@@
 the emergence of the light blue cluster from within the successful dark blue cluster.
+    Model coefficients minimize the earth mover's distance between amino acid sequences in the observed and estimated future populations across all training windows.
   }
@@
 After selecting optimal models from training and validation, we evaluated the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
-Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
-\rnc{Hamming distance or EMD?}
+Importantly, our models find fitness coefficients that minimize the earth mover's distance (EMD) between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
 With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
@@
   Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
-  \rnc{Growth rates have units. Is this 1/year? Also what exactly is $\log_{10}$ growth rate? -2 == 0.01/year?}
-  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  A) The correlation of log estimated clade growth ratio, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade growth ratio, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
   B) The rank of the estimated best strain based on its distance to the future in the best model for 18 timepoints shows how often the model makes a good choice when forced to select a single representative strain for the future population.
@@
 First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed and estimated clade growth rates.
-The model's estimated growth rates correlated well with observed growth rates (Pearson's $R^2 = 0.52$, $p < 0.001$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+The model's estimated growth rates correlated well with observed growth rates (Pearson's $R^2 = 0.52$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
 The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
@@
 The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
-Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, $p < 0.001$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
+Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
 
@@
 As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
-While observed and estimated growth rates correlated more weakly for test data (Pearson's $R^2 = 0.14$, $p < 0.001$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
-\rnc{this is not a supplemental figure, right?}
+While observed and estimated growth rates correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
 Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
-The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
-Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
-\rnc{the p-value us probably not meaningful due to phylogenetic relatiobships.}
+The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
+Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
 These results confirm that our approach of minimizing the distance between yearly populations can simultaneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
-These results also highlight a well-established yet critical point that forecasts on truly out-of-sample data are less accurate than those based on training and validation data.
-\rnc{Is the above statement accurate?
-this refers to the true fitness model which has a next to constant coefficient around 9.
-Risk of overfitting is minimal and the only measure that is substantially worse is Pearson's $r^2$ which is sensitive to outliers.
-Furthermore, the clades on which panel A is based are not independent.}
 
@@
 
-\rnc{Just a suggestion:}
-Across our training period, HA sequences of H3N2 influenza populations one year apart was 6.40 $\pm$ 1.36 AAs.
+Across our training period, the distance between HA sequences of H3N2 influenza populations one year apart was 6.40 $\pm$ 1.36 AAs.
 This distance corresponds to the ``naive model'' in Table~\ref{table_natural_model_selection}.
@@
 Indeed, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict to October 2005.
-\rnc{I thought this was predictive until 2008.}
 To further test our hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
@@
 In contrast to the epitope cross-immunity model, the oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
-These results support our hypothesis that fitness benefit of mutations at the original 49 sites was due to historical contingency and that the success of previous epitope models based on these sites partly due to ``borrowing from the future''.
+These results support our hypothesis that the fitness benefit of mutations at the original 49 sites was due to historical contingency and that the success of previous epitope models based on these sites was partly due to ``borrowing from the future''.
 We suspect that our HI cross-immunity model benefits from its ability to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
@@
 As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
-The composite model of mutational load and LBI effectively captured clade dynamics with a growth correlation of $R^2 = 0.35$~$(p < 0.001)$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
+The composite model of mutational load and LBI effectively captured clade dynamics with a growth correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
 Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}C).
 The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}B).
-This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, $p < 0.001$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}D).
+This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}D).
 
@@
 After identifying the composite HI cross-immunity and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
-The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$~$(p < 0.001)$ between observed and estimated growth rates and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
+The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$ between observed and estimated growth rates and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
 The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Fig.~\ref{fig:test_of_best_model_for_natural_populations}C).
 The estimated best strain from this model was in the top second percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for seven (88\%) of eight timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}B).
-Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.69$, $p < 0.001$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
+Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.69$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
 
@@
 
-Even well performing models with few parameters will sometimes fail due to the probabilistic nature of evolution: The model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
+Even the most accurate models with few parameters will sometimes fail due to the probabilistic nature of evolution.
+For example, the model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
 Specifically, we found that this model failed to predict the sudden decline of a dominant reassortant clade, A2/re, in 2019.
@@
 Finally, forecasting models need to account for the geographic distribution of viruses and the vastly different sampling intensities across the globe.
-Most influenza sequence data come from highly developed countries that account for a small fraction of the global population, while globally successful clades of influenza H3N2 often emerge in less well sampled regions \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
+Most influenza sequence data come from highly developed countries that account for a small fraction of the global population, while globally successful clades of influenza H3N2 often emerge in less well-sampled regions \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
 Explicitly accounting for these sampling biases and the associated migration dynamics would allow models to weight forecasts based on both viral fitness and transmission.
@@
 Prediction of future influenza virus populations are intrinsically limited by the small number of data points available to train and test models.
-Increasing the complexity therefore makes models prone to overfitting.
-Across the training and validation period, we found that antigenic drift and mutational load are the most robust predictors of future success for seasonal influenza H3N2 populations.
+Increasingly more complex models are therefore prone to overfitting.
+Across the validation and test periods, we found that antigenic drift and mutational load were the most robust predictors of future success for seasonal influenza H3N2 populations.
 
@@
 By releasing our framework as an open source tool based on modern data science standards like tidy data frames, we hope to encourage continued development of this tool by the influenza research community.
-We additionally anticipate that the ability to forecast the sequence composition of populations with Earth Mover's Distance will enable future forecasting research with pathogens whose genomes cannot be analyzed by traditional phylogenetic methods including recombinant viruses, bacteria, and fungi.
+We additionally anticipate that the ability to forecast the sequence composition of populations with earth mover's distance will enable future forecasting research with pathogens whose genomes cannot be analyzed by traditional phylogenetic methods including recombinant viruses, bacteria, and fungi.
 
@@
 For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the weighted Hamming distance between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of samples at time $t$ by their estimated fitnesses.
-We measured distance between populations using Earth Mover's Distance (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{Rubner1998,Kusner2015}.
+We measured distance between populations using earth mover's distance (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{Rubner1998,Kusner2015}.
 Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.

### 2020-01-03 manuscript/main.tex
Include GISRS and GISAID groups in acknowledgements
Also updates the citation for GISAID.

@@
 
-Prior to our analyses, we downloaded all HA sequences and metadata from GISAID \cite{Elbe2017}.
+Prior to our analyses, we downloaded all HA sequences and metadata from GISAID \cite{shu2017gisaid}.
 For model training and validation, we selected 15,583 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1990 and October 1, 2015.
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
+
+We thank the Influenza Division at the US Centers for Disease Control and Prevention, the Victorian Infectious Diseases Reference Laboratory at the Australian Peter Doherty Institute for Infection and Immunity, the Influenza Virus Research Center at the Japan National Institute of Infectious Diseases, the Crick Worldwide Influenza Centre at the UK Francis Crick Institute for sharing HI and FRA data.
+
+We gratefully acknowledge the authors, originating and submitting laboratories of the sequences from the GISAID EpiFlu Database \cite{shu2017gisaid} on which this research is based.
+The list is detailed in the Supplemental Material.
 

### 2020-01-03 manuscript/supplement.tex
Include GISRS and GISAID groups in acknowledgements
Also updates the citation for GISAID.

@@
 \end{figure*}
+
+\pagebreak
+
+\subsection*{GISAID Acknowledgements}
+
+WHO Collaborating Centre for Reference and Research on Influenza, Victorian Infectious Diseases Reference Laboratory, Australia; WHO Collaborating Centre for Reference and Research on Influenza, Chinese National Influenza Center, China; WHO Collaborating Centre for Reference and Research on Influenza, National Institute of Infectious Diseases, Japan; The Crick Worldwide Influenza Centre, The Francis Crick Institute, United Kingdom; WHO Collaborating Centre for the Surveillance, Epidemiology and Control of Influenza, Centers for Disease Control and Prevention, United States; ADImmune Corporation, Taiwan; ADPH Bureau of Clinical Laboratories, United States; Aichi Prefectural Institute of Public Health, Japan; Akershus University Hospital, Norway; Akita Research Center for Public Health and Environment, Japan; Alabama State Laboratory, United States; Alaska State Public Health Laboratory, United States; Alaska State Virology Lab, United States; Aomori Prefectural Institute of Public Health and Environment, Japan; Aristotelian University of Thessaloniki, Greece; Arizona Department of Health Services, United States; Arkansas Children's Hospital, United States; Arkansas Department of Health, United States; Auckland Healthcare, New Zealand; Auckland Hospital, New Zealand; Austin Health, Australia; Baylor College of Medicine, United States; California Department of Health Services, United States; Canberra Hospital, Australia; Cantacuzino Institute, Romania; Canterbury Health Services, New Zealand; Caribbean Epidemiology Center, Trinidad and Tobago; CDC GAP Nigeria, Nigeria; CDC-Kenya, Kenya; CEMIC University Hospital, Argentina; CENETROP, Bolivia, Plurinationial State of; Center for Disease Control, Taiwan; Center for Public Health and Environment, Hiroshima Prefectural Technology Research Institute, Japan; Central Health Laboratory, Mauritius; Central Laboratory of Public Health, Paraguay; Central Public Health Laboratory, Ministry of Health, Oman; Central Public Health Laboratory, Palestinian Territory; Central Public Health Laboratory, Papua New Guinea; Central Research Institute for Epidemiology, Russian Federation; Centre for Diseases Control and Prevention, Armenia; Centre for Infections, Health Protection Agency, United Kingdom; Centre Pasteur du Cameroun, Cameroon; Chiba City Institute of Health and Environment, Japan; Chiba Prefectural Institute of Public Health, Japan; Childrens Hospital Westmead, Australia; Chuuk State Hospital, Micronesia, Federated States of; City of El Paso Dept of Public Health, United States; Clinical Virology Unit, CDIM, Australia; Colorado Department of Health Lab, United States; Connecticut Department.
+of Public Health, United States; Contiguo a Hospital Rosales, El Salvador; Croatian Institute of Public Health , Croatia; CRR virus Influenza region Sud, France; CRR virus Influenza region Sud, Guyana; CSL Ltd, United States; Dallas County Health and Human Services, United States; DC Public Health Lab, United States; Delaware Public Health Lab, United States; Departamento de Laboratorio de Salud Publica, Uruguay; Department of Virology, Medical University Vienna, Austria; Disease Investigation Centre Wates (BBVW), Australia; Drammen Hospital / Vestreviken HF, Norway; Ehime Prefecture Institute of Public Health and Environmental Science, Japan; Erasmus Medical Center, Netherlands; Erasmus University of Rotterdam, Netherlands; Ethiopian Health and Nutrition Research Institute (EHNRI), Ethiopia; Evanston Hospital and NorthShore University, United States; Facultad de Medicina, Spain; Fiji Centre for Communicable Disease Control, Fiji; Florida Department of Health, United States; Fukui Prefectural Institute of Public Health, Japan; Fukuoka City Institute for Hygiene and the Environment, Japan; Fukuoka Institute of Public Health and Environmental Sciences, Japan; Fukushima Prefectural Institute of Public Health, Japan; Gart Naval General Hospital, United Kingdom; Georgia Public Health Laboratory, United States; Gifu Municipal Institute of Public Health, Japan; Gifu Prefectural Institute of Health and Environmental Sciences, Japan; Government Virus Unit, Hong Kong; Gunma Prefectural Institute of Public Health and Environmental Sciences, Japan; Hamamatsu City Health Environment Research Center, Japan; Haukeland University Hospital, Dept.
+of Microbiology , Norway; Headquarters British Gurkhas Nepal, Nepal; Health Forde, Department of Microbiology, Norway; Health Protection Agency, United Kingdom; Health Protection Inspectorate, Estonia; Hellenic Pasteur Institute, Greece; Hiroshima City Institute of Public Health, Japan; Hokkaido Institute of Public Health, Japan; Hopital Cantonal Universitaire de Geneves, Switzerland; Hopital Charles Nicolle, Tunisia; Hospital Clinic de Barcelona, Spain; Hospital Universitari Vall d'Hebron, Spain; Houston Department of Health and Human Services, United States; Hyogo Prefectural Institute of Public Health and Consumer Sciences, Japan; Ibaraki Prefectural Institute of Public Health, Japan; Illinois Department of Public Health, United States; Indiana State Department of Health Laboratories, United States; Infectology Center of Latvia, Latvia; Innlandet Hospital Trust, Division Lillehammer, Department for Microbiology, Norway; INSA National Institute of Health Portugal, Portugal; Institut National d'Hygiene, Morocco; Institut Pasteur d'Algerie, Algeria; Institut Pasteur de Dakar, Senegal; Institut Pasteur de Madagascar, Madagascar; Institut Pasteur in Cambodia, Cambodia; Institut Pasteur New Caledonia, New Caledonia; Institut Pasteur, France; Institut Pasteur, Saudi Arabia; Institut Penyelidikan Perubatan, Malaysia; Institute National D'Hygiene, Togo; Institute of Environmental Science and Research, New Zealand; Institute of Environmental Science and Research, Tonga; Institute of Epidemiology and Infectious Diseases, Ukraine; Institute of Epidemiology Disease Control and Research, Bangladesh; Institute of Immunology and Virology Torlak, Serbia; Institute of Medical and Veterinary Science (IMVS), Australia; Institute of Public Health, Serbia; Institute of Public Health, Albania; Institute of Public Health, Montenegro; Institute Pasteur du Cambodia, Cambodia; Instituto Adolfo Lutz, Brazil; Instituto Conmemorativo Gorgas de Estudios de la Salud, Panama; Instituto de Salud Carlos III, Spain; Instituto de Salud Publica de Chile, Chile; Instituto Nacional de Enfermedades Infecciosas, Argentina; Instituto Nacional de Higiene Rafael Rangel, Venezuela, Bolivia; Instituto Nacional de Laboratoriosde Salud (INLASA), Bolivia; Instituto Nacional de Salud de Columbia, Colombia; Instituto Nacional de Saude, Portugal; Iowa State Hygienic Laboratory, United States; IRSS, Burkina Faso; Ishikawa Prefectural Institute of Public Health and Environmental Science, Japan; ISS, Italy; Istanbul University, Turkey; Istituto Superiore di Sanità, Italy; Ivanovsky Research Institute of Virology RAMS, Russian Federation; Jiangsu Provincial Center for Disease Control and Prevention, China; John Hunter Hospital, Australia; Kagawa Prefectural Research Institute for Environmental Sciences and Public Health, Japan; Kagoshima Prefectural Institute for Environmental Research and Public Health, Japan; Kanagawa Prefectural Institute of Public Health, Japan; Kansas Department of Health and Environment, United States; Kawasaki City Institute of Public Health , Japan; Kentucky Division of Laboratory Services, United States; Kitakyusyu City Institute of Enviromental Sciences, Japan; Kobe Institute of Health, Japan; Kochi Public Health and Sanitation Institute, Japan; Kumamoto City Environmental Research Center, Japan; Kumamoto Prefectural Institute of Public Health and Environmental Science, Japan; Kyoto City Institute of Health and Environmental Sciences, Japan; Kyoto Prefectural Institute of Public Health and Environment, Japan; Laboratoire National de Sante Publique, Haiti; Laboratoire National de Sante, Luxembourg; Laboratório Central do Estado do Paraná, Brazil; Laboratorio Central do Estado do Rio de Janeiro, Brazil; Laboratorio de Investigacion / Centro de Educacion Medica y Amistad Dominico Japones (CEMADOJA), Dominican Republic; Laboratorio De Saude Publico, Macao; Laboratorio de Virologia, Direccion de Microbiologia, Nicaragua; Laboratorio de Virus Respiratorio, Mexico; Laboratorio Nacional de Influenza, Costa Rica; Laboratorio Nacional De Salud Guatemala, Guatemala; Laboratorio Nacional de Virologia, Honduras; Laboratory Directorate, Jordan; Laboratory for Virology, National Institute of Public Health, Slovenia; Laboratory of Influenza and ILI, Belarus; LACEN/RS - Laboratório Central de Saúde Pública do Rio Grande do Sul, Brazil; Landspitali - University Hospital, Iceland; Lithuanian AIDS Center Laboratory, Lithuania; Los Angeles Quarantine Station, CDC Quarantine Epidemiology and Surveillance Team, United States; Louisiana Department of Health and Hospitals, United States; Maine Health and Environmental Testing Laboratory, United States; Malbran, Argentina; Marshfield Clinic Research Foundation, United States; Maryland Department of Health and Mental Hygiene, United States; Massachusetts Department of Public Health, United States; Mater Dei Hospital, Malta; Medical Research Institute, Sri Lanka; Medical University Vienna, Austria; Melbourne Pathology, Australia; Michigan Department of Community Health, United States; Mie Prefecture Health and Environment Research Institute, Japan; Mikrobiologisk laboratorium, Sykehuset i Vestfold, Norway; Ministry of Health and Population, Egypt; Ministry of Health of Ukraine, Ukraine; Ministry of Health, Bahrain; Ministry of Health, Kiribati; Ministry of Health, Lao, People's Democratic Republic; Ministry of Health, NIHRD, Indonesia; Ministry of Health, Oman; Minnesota Department of Health, United States; Mississippi Public Health Laboratory, United States; Missouri Department.
+of Health and Senior Services, United States; Miyagi Prefectural Institute of Public Health and Environment, Japan; Miyazaki Prefectural Institute for Public Health and Environment, Japan; Molde Hospital, Laboratory for Medical Microbiology, Norway; Molecular Diagnostics Unit , United Kingdom; Monash Medical Centre, Australia; Montana Laboratory Services Bureau, United States; Montana Public Health Laboratory, United States; Nagano City Health Center, Japan; Nagano Environmental Conservation Research Institute, Japan; Nagoya City Public Health Research Institute, Japan; Nara Prefectural Institute for Hygiene and Environment, Japan; National Center for Communicable Diseases, Mongolia; National Center for Laboratory and Epidemiology, Laos; National Centre for Disease Control (NCDC), Mongolia; National Centre for Disease Control and Public Health, Georgia; National Centre for Preventive Medicine, Moldova, Republic of; National Centre for Scientific Services for Virology and Vector Borne Diseases, Fiji; National Health Laboratory, Japan; National Health Laboratory, Myanmar; National Influenza Center French Guiana and French Indies, French Guiana; National Influenza Center, Brazil; National Influenza Center, Mongolia; National Influenza Centre for Northern Greece, Greece; National Influenza Centre of Iraq, Iraq; National Influenza Lab, Tanzania, United Republic of; National Influenza Reference Laboratory, Nigeria; National Insitut of Hygien, Morocco; National Institute for Biological Standards and Control (NIBSC), United States; National Institute for Communicable Disease, South Africa; National Institute for Health and Welfare, Finland; National Institute of Health Research and Development, Indonesia; National Institute of Health, Korea, Republic of; National Institute of Health, Pakistan; National Institute of Hygien, Morocco; National Institute of Hygiene and Epidemiology, Vietnam; National Institute of Public Health - National Institute of Hygiene, Poland; National Institute of Public Health, Czech Republic; National Institute of Virology, India; National Microbiology Laboratory, Health Canada, Canada; National Public Health Institute of Slovakia, Slovakia; National Public Health Laboratory, Cambodia; National Public Health Laboratory, Ministry of Health, Singapore, Singapore; National Public Health Laboratory, Nepal; National Public Health Laboratory, Singapore; National Reference Laboratory, Kazakhstan; National University Hospital, Singapore; National Virology Laboratory, Center Microbiological Investigations, Kyrgyzstan; National Virus Reference Laboratory, Ireland; Naval Health Research Center, United States; Nebraska Public Health Lab, United States; Nevada State Health Laboratory, United States; New Hampshire Public Health Laboratories, United States; New Jersey Department of Health and Senior Services, United States; New Mexico Department of Health, United States; New York City Department of Health, United States; New York Medical College, United States; New York State Department of Health, United States; Nicosia General Hospital, Cyprus; Niigata City Institute of Public Health and Environment, Japan; Niigata Prefectural Institute of Public Health and Environmental Sciences, Japan; Niigata University, Japan; Nordlandssykehuset, Norway; North Carolina State Laboratory of Public Health, United States; North Dakota Department of Health, United States; Norwegian Institute of Public Health, Norway; Norwegian Institute of Public Health, Svalbard and Jan Mayen; Ohio Department of Health Laboratories, United States; Oita Prefectural Institute of Health and Environment, Japan; Okayama Prefectural Institute for Environmental Science and Public Health, Japan; Okinawa Prefectural Institute of Health and Environment, Japan; Oklahoma State Department of Health, United States; Ontario Agency for Health Protection and Promotion (OAHPP), Canada; Oregon Public Health Laboratory, United States; Osaka City Institute of Public Health and Environmental Sciences, Japan; Osaka Prefectural Institute of Public Health, Japan; Oslo University Hospital, Ulleval Hospital, Dept.
+of Microbiology, Norway; Ostfold Hospital - Fredrikstad, Dept.
+of Microbiology, Norway; Oswaldo Cruz Institute - FIOCRUZ - Laboratory of Respiratory Viruses and Measles (LVRS), Brazil; Papua New Guinea Institute of Medical Research, Papua New Guinea; Pasteur Institut of Cote d'Ivoire, Cote d'Ivoire; Pasteur Institute, Influenza Laboratory, Vietnam; Pathwest QE II Medical Centre, Australia; Pennsylvania Department of Health, United States; Prince of Wales Hospital, Australia; Princess Margaret Hospital for Children, Australia; Public Health Laboratory Services Branch, Centre for Health Protection, Hong Kong; Public Health Laboratory, Barbados; Puerto Rico Department of Health, Puerto Rico; Qasya Diagnostic Services Sdn Bhd, Brunei; Queensland Health Scientific Services, Australia; Refik Saydam National Public Health Agency, Turkey; Regent Seven Seas Cruises, United States; Royal Victoria Hospital, United Kingdom; Republic Institute for Health Protection, Macedonia, the former Yogoslav Republic of; Republic of Nauru Hospital, Nauru; Research Institute for Environmental Sciences and Public Health of Iwate Prefecture, Japan; Research Institute of Tropical Medicine, Philippines; Rhode Island Department of Health, United States; RIVM National Institute for Public Health and Environment, Netherlands; Robert-Koch-Institute, Germany; Royal Chidrens Hospital, Australia; Royal Darwin Hospital, Australia; Royal Hobart Hospital, Australia; Royal Melbourne Hospital, Australia; Russian Academy of Medical Sciences, Russian Federation; Rwanda Biomedical Center, National Reference Laboratory, Rwanda; Saga Prefectural Institute of Public Health and Pharmaceutical Research, Japan; Sagamihara City Laboratory of Public Health, Japan; Saitama City Institute of Health Science and Research, Japan; Saitama Institute of Public Health, Japan; Sakai City Institute of Public Health, Japan; San Antonio Metropolitan Health, United States; Sandringham, National Institute for Communicable D, South Africa; Sapporo City Institute of Public Health, Japan; Scientific Institute of Public Health, Belgium; Seattle and King County Public Health Lab, United States; Sendai City Institute of Public Health, Japan; Servicio de Microbiología Clínica Universidad de Navarra, Spain; Servicio de Microbiología Complejo Hospitalario de Navarra, Spain; Servicio de Microbiología Hospital Central Universitario de Asturias, Spain; Servicio de Microbiología Hospital Donostia, Spain; Servicio de Microbiología Hospital Meixoeiro, Spain; Servicio de Microbiología Hospital Miguel Servet, Spain; Servicio de Microbiología Hospital Ramón y Cajal, Spain; Servicio de Microbiología Hospital San Pedro de Alcántara, Spain; Servicio de Microbiología Hospital Santa María Nai, Spain; Servicio de Microbiología Hospital Universitario de Gran Canaria Doctor Negrín, Spain; Servicio de Microbiología Hospital Universitario Son Espases, Spain; Servicio de Microbiología Hospital Virgen de la Arrixaca, Spain; Servicio de Microbiología Hospital Virgen de las Nieves, Spain; Servicio de Virosis Respiratorias INEI-ANLIS Carlos G.
+Malbran, Argentina; Shiga Prefectural Institute of Public Health, Japan; Shimane Prefectural Institute of Public Health and Environmental Science, Japan; Shizuoka City Institute of Environmental Sciences and Public Health , Japan; Shizuoka Institute of Environment and Hygiene, Japan; Singapore General Hospital, Singapore; Sorlandet Sykehus HF, Dept.
+of Medical Microbiology, Norway; South Carolina Department of Health, United States; South Dakota Public Health Lab, United States; Southern Nevada Public Health Lab, United States; Spokane Regional Health District, United States; St.\ Judes Childrens Research Hospital, United States; St.
+Olavs Hospital HF, Dept.
+of Medical Microbiology, Norway; State Agency, Infectology Center of Latvia, Latvia; State of Hawaii Department of Health, United States; State of Idaho Bureau of Laboratories, United States; State Research Center of Virology and Biotechnology Vector, Russian Federation; Statens Serum Institute, Denmark; Stavanger Universitetssykehus, Avd.
+for Medisinsk Mikrobiologi, Norway; Subdireccion General de Epidemiologia y Vigilancia de la Salud, Spain; Subdirección General de Epidemiología y Vigilancia de la Salud, Spain; Swedish Institute for Infectious Disease Control, Sweden; Swedish National Institute for Communicable Disease Control, Sweden; Taiwan CDC, Taiwan; Tan Tock Seng Hospital, Singapore; Tehran University of Medical Sciences, Iran; Tennessee Department of Health Laboratory-Nashville, United States; Texas Childrens Hospital, United States; Texas Department of State Health Services, United States; Thai National Influenza Center, Thailand; Thailand MOPH-U.S.
+CDC Collaboration (IEIP), Thailand; The Nebraska Medical Center, United States; Tochigi Prefectural Institute of Public Health and Environmental Science, Japan; Tokushima Prefectural Centre for Public Health and Environmental Sciences, Japan; Tokyo Metropolitan Institute of Public Health, Japan; Tottori Prefectural Institute of Public Health and Environmental Science, Japan; Toyama Institute of Health, Japan; U.S.
+Air Force School of Aerospace Medicine, United States; U.S.
+Naval Medical Research Unit No.3, Egypt; Uganda Virus Research Institute (UVRI), National Influenza Center, Uganda; Universidad de Valladolid, Spain; Università Cattolica del Sacro Cuore, Italy; Universitetssykehuset Nord-Norge HF, Norway; University Malaya, Malaysia; University of Florence, Italy; University of Genoa, Italy; University of Ghana, Ghana; University of Michigan SPH EPID, United States; University of Parma, Italy; University of Perugia, Italy; University of Pittsburgh Medical Center Microbiology Lab, United States; University of Sarajevo, Bosnia and Herzegovina; University of Sassari, Italy; University of the West Indies, Jamaica; University of Vienna, Austria; University of Virginia, Medical Labs/Microbiology, United States; University Teaching Hospital, Zambia; UPMC-CLB Dept of Microbiology, United States; US Army Medical Research Unit - Kenya (USAMRU-K), GEIS Human Influenza Program, Kenya; USAMC-AFRIMS Department of Virology, Cambodia; Utah Department of Health, United States; Utah Public Health Laboratory, United States; Utsunomiya City Institute of Public Health and Environment Science, Japan; VACSERA, Egypt; Vermont Department of Health Laboratory, United States; Victorian Infectious Diseases Reference Laboratory, Australia; Virginia Division of Consolidated Laboratories, United States; Wakayama City Institute of Public Health, Japan; Wakayama Prefectural Research Center of Environment and Public Health, Japan; Washington State Public Health Laboratory, United States; West Virginia Office of Laboratory Services, United States; Westchester County Department of Laboratories and Research, United States; Westmead Hospital, Australia; WHO National Influenza Centre Russian Federation, Russian Federation; WHO National Influenza Centre, National Institute of Medical Research (NIMR), Thailand; WHO National Influenza Centre, Norway; Wisconsin State Laboratory of Hygiene, United States; Wyoming Public Health Laboratory, United States; Yamagata Prefectural Institute of Public Health, Japan; Yamaguchi Prefectural Institute of Public Health and Environment, Japan; Yamanashi Institute for Public Health, Japan; Yap State Hospital, Micronesia; Yokohama City Institute of Health, Japan; Yokosuka Institute of Public Health, Japan

### 2020-01-03 manuscript/main.tex
Add panel labels to model schematic figure
Also replace PNG of figure with a PDF.

@@
   \begin{center}
-  \includegraphics[width=\columnwidth]{figures/distance-based-fitness-model.png}
+  \includegraphics[width=\columnwidth]{figures/distance-based-fitness-model.pdf}
   \caption{
-  \rnc{missing panel labels!}
     Schematic representation of the fitness model for simulated H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.

### 2020-01-03 manuscript/main.tex
Use consistent initials for authors
Remove Richard's comments that require further analyses
These have been noted elsewhere.

@@
 However, frequent viral mutation results in viruses that escape previously acquired human immunity.
-The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine strain selection requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
+The World Health Organization (WHO) Global Influenza Surveillance and Response System (GISRS) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine strain selection requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
 Until recently, the hemagglutination inhibition (HI) assay has been the primary experimental measure of antigenic cross-reactivity between pairs of circulating viruses \cite{hirst1943studies}.
-Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \cite{Chambers:2015jt,Zost2017}.
-Correspondingly, many influenza collaborating centers now also measure cross-reactivity by focus reduction assays (FRAs) \cite{Okuno:1990uu}.
+Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \cite{Chambers:2015jt,Zost2017}, prompting the increased use of focus reduction assays (FRAs) \cite{Okuno:1990uu}.
 Together, these two assays are the gold standard for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
@@
 Here, we provide the first such open source framework for forecasting the genetic composition of future seasonal influenza populations.
+We apply this framework to HA sequence data shared via the GISAID EpiFlu database \cite{shu2017gisaid} and to HI and FRA titer data shared by WHO GISRS Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
 With this framework, we show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
@@
 
-\rnc{wouldn't we expect the true fitness have coefficent 1.0? A plausible reviewer question would be ``how does the model do with coefficient 1.0?"}
 As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 (97\%) of 33 timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
@@
   C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
-  \rnc{it would be interesting to look at this for neg and pos deviations separately.}
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
@@
 
-Hemagglutination inhibition (HI) measurements were provided by collaborating centers from the WHO Global Influenza Surveillance and Response System.
+Hemagglutination inhibition (HI) measurements were provided by WHO Global Influenza Surveillance and Response System (GISRS) Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
 We converted these raw two-fold dilution measurements to $log_{2}$ titer drops normalized by the corresponding $log_{2}$ autologous measurements as previously described \cite{Neher:2016hy}.
@@
 JH planned experiments, implemented the final forecasting framework, analyzed results, and wrote the manuscript.
-RN planned experiments and reviewed the manuscript.
+RAN planned experiments and reviewed the manuscript.
 TB planned experiments, implemented the initial forecasting framework, and reviewed the manuscript.

### 2020-01-03 manuscript/main.tex
Refer to growth ratios and not rates

@@
 We validated the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
-First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed and estimated clade growth rates.
-The model's estimated growth rates correlated well with observed growth rates (Pearson's $R^2 = 0.52$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed and estimated clade growth ratios.
+The model's estimated growth ratios correlated well with observed growth ratios (Pearson's $R^2 = 0.52$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
 The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
@@
 As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
-While observed and estimated growth rates correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
+While observed and estimated growth ratios correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
 Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
@@
 
-As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
+As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth ratios and the ranking of estimated best strains compared to the observed closest strains to future populations.
 The composite model of mutational load and LBI effectively captured clade dynamics with a growth correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
@@
   Test of best model for natural populations of H3N2 viruses, the composite model of HI cross-immunity and mutational load.
-  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
   B) The rank of the estimated best strain based on its distance to the future for 8 timepoints.
@@
 After identifying the composite HI cross-immunity and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
-The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$ between observed and estimated growth rates and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
+The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$ between observed and estimated growth ratios and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
 The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Fig.~\ref{fig:test_of_best_model_for_natural_populations}C).
@@
 \end{figure*}
-
-\pagebreak
 

### 2020-01-03 manuscript/supplement.tex
Refer to growth ratios and not rates

@@
 \renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\Roman{table}}
+\renewcommand{\thetable}{S\arabic{table}}
 
@@
 
-\begin{figure*}[h]
+\subsection*{Supplemental Figures}
+
+\begin{figure}[H]
   \begin{center}
@@
   \end{center}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[h]
+\begin{figure}[H]
   \begin{center}
@@
   \end{center}
-\end{figure*}
+\end{figure}
 
-\begin{table*}[ht]
-  \begin{center}
-    \input{tables/mutations_by_trunk_status_for_simulated_populations.tex}
-    \caption{
-    Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
-    Epitope sites were defined previously described \cite{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
-    Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
-    }
-    \label{sup_table:mutations_by_trunk_status_for_simulated_populations}
-  \end{center}
-\end{table*}
-
-\begin{table*}[ht]
-  \begin{center}
-    \input{tables/mutations_by_trunk_status.tex}
-    \caption{
-    Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
-    Epitope sites were defined previously described \cite{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
-    Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.
-    }
-    \label{sup_table:mutations_by_trunk_status}
-  \end{center}
-\end{table*}
-
-\begin{figure*}[t]
+\begin{figure}[H]
   \begin{center}
@@
   \end{center}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[t]
+\begin{figure}[H]
   \begin{center}
@@
   Validation of best model for simulated populations of H3N2-like viruses.
-  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
   B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
@@
   \end{center}
-\end{figure*}
+\end{figure}
 
-\begin{table*}[ht]
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.png}
+  \caption{
+  Model coefficients and distances to the future for cross-immunity models fit to natural populations.
+  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+  The epitope cross-immunity model relies on previously published epitope sites \cite{Luksza:2014hj}.
+  The ``oracle'' cross-immunity model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
+  The improved performance of the ``oracle'' model indicates that the sequence-based cross-immunity metric can be effective when sites of beneficial mutations are known prior to forecasting.
+  }
+  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
+  \caption{
+  Validation of best model for natural populations of H3N2 viruses, the composite model of mutational load and LBI.
+  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.
+  The estimated best strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+  }
+  \label{sup_fig:validation_of_best_model_for_natural_populations}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.png}
+  \caption{
+    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    A) Coefficients and B) distances are shown per test timepoint (N=8).
+    In contrast to the results for these models based on fixed coefficients from training/validation, these coefficients were learned for each six-year window prior to the corresponding test timepoint.
+    The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
+    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.
+  }
+  \label{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
+  \end{center}
+\end{figure}
+
+\subsection*{Supplemental Tables}
+
+\begin{table}[H]
+  \begin{center}
+    \input{tables/mutations_by_trunk_status_for_simulated_populations.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
+    Epitope sites were defined previously described \cite{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
+    }
+    \label{sup_table:mutations_by_trunk_status_for_simulated_populations}
+  \end{center}
+\end{table}
+
+\begin{table}[H]
+  \begin{center}
+    \input{tables/mutations_by_trunk_status.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
+    Epitope sites were defined previously described \cite{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.
+    }
+    \label{sup_table:mutations_by_trunk_status}
+  \end{center}
+\end{table}
+
+\begin{table}[H]
   \begin{center}
@@
   \end{center}
-\end{table*}
-
-\begin{figure*}[h]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.png}
-  \caption{
-  Model coefficients and distances to the future for cross-immunity models fit to natural populations.
-  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-  The epitope cross-immunity model relies on previously published epitope sites \cite{Luksza:2014hj}.
-  The ``oracle'' cross-immunity model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
-  The improved performance of the ``oracle'' model indicates that the sequence-based cross-immunity metric can be effective when sites of beneficial mutations are known prior to forecasting.
-  }
-  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
-  \end{center}
-\end{figure*}
-
-\begin{figure*}[ht]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
-  \caption{
-  Validation of best model for natural populations of H3N2 viruses, the composite model of mutational load and LBI.
-  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.
-  The estimated best strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
-  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
-  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  }
-  \label{sup_fig:validation_of_best_model_for_natural_populations}
-  \end{center}
-\end{figure*}
-
-\begin{figure*}[ht]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.png}
-  \caption{
-    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-    A) Coefficients and B) distances are shown per test timepoint (N=8).
-    In contrast to the results for these models based on fixed coefficients from training/validation, these coefficients were learned for each six-year window prior to the corresponding test timepoint.
-    The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
-    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.
-  }
-  \label{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
-  \end{center}
-\end{figure*}
+\end{table}
 
@@
 
-\subsection*{GISAID Acknowledgements}
+\subsection*{Supplemental Text}
+
+\subsubsection*{GISAID Acknowledgements}
 

### 2020-01-03 manuscript/supplement.tex
Reorder supp material I'm most used to Supp Material being strictly divided into sections. Here, Supp Figures, Supp Tables and Supp Text. The previous use of [h] had resulted in GISAID acknowledgments flowing alongside figures.

@@
   Validation of best model for simulated populations of H3N2-like viruses.
-  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
   B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
@@
   Validation of best model for natural populations of H3N2 viruses, the composite model of mutational load and LBI.
-  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
   B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.

### 2020-01-03 manuscript/main.tex
Add panel labels to all figures and export figures as PDFs
Multi-panel figures now include the appropriate labels to guide readers through
the captions and figures are now vector graphics.

--- 2020-01-03 manuscript/main.tex
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations-controls.png}
+  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations-controls.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
+  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.png}
+  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
+  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/best-composite-unadjusted-model-accuracy-and-coefficients-for-natural-populations.png}
+  \includegraphics[width=\textwidth]{figures/best-composite-unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-natural-populations.png}
+  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-natural-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/vaccine-comparison.png}
+  \includegraphics[width=\textwidth]{figures/vaccine-comparison.pdf}
   \caption{

### 2020-01-03 manuscript/supplement.tex
Add panel labels to all figures and export figures as PDFs
Multi-panel figures now include the appropriate labels to guide readers through
the captions and figures are now vector graphics.

--- 2020-01-03 manuscript/supplement.tex
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/cross-validation-for-simulated-populations.png}
+  \includegraphics[width=\textwidth]{figures/cross-validation-for-simulated-populations.pdf}
   \caption{
@@
   After fixing hyperparameters, model coefficients were fixed at the mean values across all training windows.
-  Fixed coefficients were applied to 10 years of new out-of-sample test data (open circles) to estimate true forecast errors.
+  Fixed coefficients were applied to 9 years of new out-of-sample test data (open circles) to estimate true forecast errors.
   }
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.png}
+  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-simulated-populations.png}
+  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-simulated-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/cross-validation-for-natural-populations.png}
+  \includegraphics[width=\textwidth]{figures/cross-validation-for-natural-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.png}
+  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
+  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.png}
+  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.pdf}
   \caption{

### 2020-01-04 manuscript/main.tex
Swap "samples" for "strains"

@@
   \caption{
-    Schematic representation of the fitness model for simulated H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
-    Samples are colored by their amino acid sequence composition such that genetically similar samples have similar colors (Methods).
-    A) Samples at timepoint $t$, $\mathbf{x}(t)$, are shown in their phylogenetic context and sized by their frequency at that timepoint.
-    The estimated future population at timepoint $u$, $\mathbf{\hat{x}}(u)$, is projected to the right with samples scaled in size by their projected frequency based on the true fitness model.
-    B) The frequency trajectories of samples at timepoint $t$ to $u$ represent the predicted the growth of the dark blue samples to the detriment of the pink samples.
-    C) Samples at timepoint $u$, $\mathbf{x}(u)$, are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
-    D) The observed frequency trajectories of samples at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate, e.g.
+    Schematic representation of the fitness model for simulated H3N2-like populations wherein the fitness of strains at timepoint $t$ determines the estimated frequency of strains with similar sequences one year in the future at timepoint $u$.
+    Strains are colored by their amino acid sequence composition such that genetically similar strains have similar colors (Methods).
+    A) Strains at timepoint $t$, $\mathbf{x}(t)$, are shown in their phylogenetic context and sized by their frequency at that timepoint.
+    The estimated future population at timepoint $u$, $\mathbf{\hat{x}}(u)$, is projected to the right with strains scaled in size by their projected frequency based on the true fitness model.
+    B) The frequency trajectories of strains at timepoint $t$ to $u$ represent the predicted the growth of the dark blue strains to the detriment of the pink strains.
+    C) Strains at timepoint $u$, $\mathbf{x}(u)$, are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
+    D) The observed frequency trajectories of strains at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate, e.g.
 the emergence of the light blue cluster from within the successful dark blue cluster.
@@
 As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
-We estimated the frequency of virus samples every six months using kernel density estimation (KDE) to represent the frequency of each sample.
+We estimated the frequency of virus strains every six months using kernel density estimation (KDE) to represent the frequency of each sample.
 
@@
 All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.
-The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating samples based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
-Increased antigenic drift relative to previously circulating samples is expected to correspond to increased viral fitness.
+The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating strains based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
+Increased antigenic drift relative to previously circulating strains is expected to correspond to increased viral fitness.
 Mutational load is the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each sample since their ancestor in the previous season.
 DMS mutational effects provide a more comprehensive biophysical model of functional constraint by measuring the beneficial or deleterious effect of each possible single amino acid mutation in HA from the background of a previous vaccine strain, A/Perth/16/2009.
-The growth metrics estimate how successful populations of samples have been in the last six months based on either rapid branching in the phylogeny (LBI) or the change in clade frequencies over time (delta frequency).
+The growth metrics estimate how successful populations of strains have been in the last six months based on either rapid branching in the phylogeny (LBI) or the change in clade frequencies over time (delta frequency).
 
@@
 With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
-However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.
+However, we retain the benefits of fitting models to highly similar strains found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.
 
@@
 However, when we fit a composite model of the epitope ancestor and mutational load metrics, this model outperformed both the individual mutational load model and the composite cross-immunity and mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
+From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated strains was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
 
@@
 Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
-We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
+We held out strains collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
@@
 Of the three major currently circulating clades, our model predicted the growth of A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
-To aid with identification of potential vaccine candidates for the next season, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
+To aid with identification of potential vaccine candidates for the next season, we annotated strains in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
 
@@
     Snapshot of the last two years of seasonal influenza H3N2 evolution on nextstrain.org showing the estimated distance per sample to the future population.
-    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating samples weighted by the other sample's projected frequencies under the best fitness model.
+    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating strains weighted by the other sample's projected frequencies under the best fitness model.
   }
@@
 To avoid overfitting our models to the relatively limited data from natural populations, we used simulated H3N2-like populations to tune hyperparameters including the KDE bandwidth for frequency estimates and the L1 penalty for model coefficients.
-We simulated populations, as described above, and fit models for each parameter value using the true fitness of samples from the simulator.
+We simulated populations, as described above, and fit models for each parameter value using the true fitness of strains from the simulator.
 
@@
 We set the L1 lambda penalty to zero, to reduce variables in the analysis and avoid interactions between the coefficients and the KDE bandwidths.
-Higher bandwidths completely wash out dynamics of populations by making all samples appear to exist for long time periods.
+Higher bandwidths completely wash out dynamics of populations by making all strains appear to exist for long time periods.
 This flattening of frequency trajectories means that as bandwidths increase, the naive model gets more accurate and less informative.
@@
 To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 90 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
-We excluded all egg-passaged samples and all samples with ambiguous year, month, and day annotations.
-We prioritized samples with more available HI titer measurements.
+We excluded all egg-passaged strains and all strains with ambiguous year, month, and day annotations.
+We prioritized strains with more available HI titer measurements.
 For model testing, we selected an additional 7,171 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and October 1, 2019.
@@
 
-For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the weighted Hamming distance between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of samples at time $t$ by their estimated fitnesses.
+For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the weighted Hamming distance between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of strains at time $t$ by their estimated fitnesses.
 We measured distance between populations using earth mover's distance (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{Rubner1998,Kusner2015}.
 Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
-This solution requires both a ``ground distance'' between pairs of samples from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.
+This solution requires both a ``ground distance'' between pairs of strains from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.
 
-For each timepoint $t$ and corresponding timepoint $u = t + 1$, we defined the ground distance as the Hamming distance between HA amino acid sequences for all pairs of samples between timepoints.
-For samples with less than full length nucleotide sequences, we inferred missing nucleotides through TreeTime's ancestral sequence reconstruction analysis.
-We defined weights for samples at timepoint $t$ based on their projected future frequencies.
-We defined weights for samples at timepoint $u$ based on their observed frequencies.
+For each timepoint $t$ and corresponding timepoint $u = t + 1$, we defined the ground distance as the Hamming distance between HA amino acid sequences for all pairs of strains between timepoints.
+For strains with less than full length nucleotide sequences, we inferred missing nucleotides through TreeTime's ancestral sequence reconstruction analysis.
+We defined weights for strains at timepoint $t$ based on their projected future frequencies.
+We defined weights for strains at timepoint $u$ based on their observed frequencies.
 We then identified the fitness coefficients that provided projected future frequencies that minimized the EMD between the estimated and observed future populations.
@@
 Specifically, we selected a representative tree from our simulated population of viruses at month 10 of year 30.
-From this tree, we selected all samples with a collection date in the previous two years.
-We calculated the pairwise Hamming distance between the full-length HA amino acid sequences for all selected samples and applied t-SNE dimensionality reduction \cite{vanDerMaaten2008} to the resulting distance matrix (n=2 components, perplexity=30.0, and learning rate=400).
-We assigned each sample to a cluster based on its two-dimensional t-SNE embedding using DBSCAN \cite{Ester1996} with a maximum neighborhood distance of 10 AAs and a minimum of 20 samples per cluster.
-Despite known limitations of applying hierarchical clustering to manifold projections that do not preserve sample density, this approach allowed us to effectively assign samples to qualitative genetic clusters for the purposes of visualization.
+From this tree, we selected all strains with a collection date in the previous two years.
+We calculated the pairwise Hamming distance between the full-length HA amino acid sequences for all selected strains and applied t-SNE dimensionality reduction \cite{vanDerMaaten2008} to the resulting distance matrix (n=2 components, perplexity=30.0, and learning rate=400).
+We assigned each sample to a cluster based on its two-dimensional t-SNE embedding using DBSCAN \cite{Ester1996} with a maximum neighborhood distance of 10 AAs and a minimum of 20 strains per cluster.
+Despite known limitations of applying hierarchical clustering to manifold projections that do not preserve sample density, this approach allowed us to effectively assign strains to qualitative genetic clusters for the purposes of visualization.
 

### 2020-01-04 scripts/plot_validation_figure_by_population.py
Plot validation figures as PDFs with panel labels
Updates plotting script to include panel labels and correct axes titles for
clade frequency growth ratios. Updates Snakefiles to output PDFs just for builds
with full trees (those for which we want to build validation figures).

@@ -332,8 +332,8 @@ if __name__ == "__main__":
         transform=clade_ax.transAxes
     )
 
-    clade_ax.set_xlabel("Observed $log_{10}$ growth rate")
-    clade_ax.set_ylabel("Estimated $log_{10}$ growth rate")
+    clade_ax.set_xlabel("Observed $log_{10}$ growth ratio")
+    clade_ax.set_ylabel("Estimated $log_{10}$ growth ratio")
 
     growth_rate_ticks = np.arange(-6, 4, 1)
     clade_ax.set_xticks(growth_rate_ticks)
@@ -439,6 +439,16 @@ if __name__ == "__main__":
     all_rank_ax.set_xlabel("Observed percentile rank")
     all_rank_ax.set_ylabel("Estimated percentile rank")
     all_rank_ax.set_aspect("equal")
+
+    # Annotate panel labels.
+    panel_labels_dict = {
+        "weight": "bold",
+        "size": 14
+    }
+    plt.figtext(0.0, 0.97, "A", **panel_labels_dict)
+    plt.figtext(0.5, 0.97, "B", **panel_labels_dict)
+    plt.figtext(0.0, 0.47, "C", **panel_labels_dict)
+    plt.figtext(0.5, 0.47, "D", **panel_labels_dict)
 
     gs.tight_layout(fig)
     plt.savefig(args.output)

### 2020-01-04 scripts/pairwise_titer_tree_distances.py
Add FRA cross-immunity predictor to builds
This requires some duplication of Snakemake rules, for now, to download titers,
fit a titer model with FRA data instead of HI data, and handle all of the
downstream steps to calculate cross-immunty from FRA pairwise tree
distances. Part of the changes here include a modification to the pairwise
distance script to allow the user to specify which attribute to use from the
titer model JSON for pairwise distance calculations.

@@ -18,7 +18,8 @@ if __name__ == "__main__":
     parser = argparse.ArgumentParser()
     parser.add_argument("--tree", help="Newick tree", required=True)
     parser.add_argument("--frequencies", help="frequencies JSON", required=True)
-    parser.add_argument("--attribute-name", help="name to store distances associated with the given distance map; multiple attribute names are linked to corresponding positional comparison method and distance map arguments", required=True)
+    parser.add_argument("--model-attribute-name", help="name of attribute to use from titer model file", default="dTiter")
+    parser.add_argument("--attribute-name", help="name to store distances", required=True)
     parser.add_argument("--model", help="JSON providing the titer tree model", required=True)
     parser.add_argument("--date-annotations", help="JSON of branch lengths and date annotations from augur refine for samples in the given tree; required for comparisons to earliest or latest date", required=True)
     parser.add_argument("--years-back-to-compare", type=int, help="number of years prior to the current season to search for samples to calculate pairwise comparisons with", required=True)
@@ -84,7 +85,8 @@ if __name__ == "__main__":
                 distances_by_node[current_sample][args.attribute_name][past_sample] = get_titer_distance_between_nodes(
                     tree,
                     tips_by_sample[past_sample],
-                    tips_by_sample[current_sample]
+                    tips_by_sample[current_sample],
+                    args.model_attribute_name
                 )
 
             comparisons += 1

### 2020-01-08 manuscript/main.tex
Update Nextstrain figures of forecasts and associated results
These figures reflect the integration of the HI cross-immunity and mutational
load model into nextstrain.org for seasonal flu builds.

@@
 After confirming the coefficients for our best model of HI cross-immunity and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021.
-Of the three major currently circulating clades, our model predicted the growth of A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
+Of the three major currently circulating clades, our model predicted growth of the clades A1b/131K and 3c3.A and decline of all other clades (Fig.~\ref{fig:nextstrain_forecasts}).
 To aid with identification of potential vaccine candidates for the next season, we annotated strains in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/nextstrain-forecasts-for-september-2020.png}
+  \includegraphics[width=\textwidth]{figures/nextstrain-forecasts.png}
   \caption{
-    Snapshot of live forecasts on nextstrain.org from our best model for September 2020.
-    The observed frequency trajectories for currently circulating clades are shown up to September 2019.
-    Our model favors the A1b/131K subclade, A1b/197R, to grow over the next year and estimates that the A1b/135K subclade, A1b/137F, will remain stable.
+    Snapshot of live forecasts on nextstrain.org from our best model (HI cross-immunity and mutational load) for January 2021.
+    The observed frequency trajectories for currently circulating clades are shown up to January 2020.
+    Our model forecasts growth of the clades A1b/131K and 3c3.A and decline of all other major clades.
   }
@@
     Snapshot of the last two years of seasonal influenza H3N2 evolution on nextstrain.org showing the estimated distance per sample to the future population.
-    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating strains weighted by the other sample's projected frequencies under the best fitness model.
+    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating strains weighted by the other sample's projected frequencies under the best fitness model (HI cross-immunity and mutational load).
   }

### 2020-01-14 scripts/pairwise_titer_tree_distances.py
Remove waning term from cross-immunity and use all pairwise comparisons for titer tree model
Updates logic for pairwise titer tree distances and cross-immunity based on
visual inspection of cross-immunity (really, antigenic novelty) values from live
seasonal flu builds. This inspection revealed that waning was producing an
unexpected secular trend such that more recent tips had higher antigenic novelty
than earlier tips regardless of their antigenic distance. We also observed
higher than expected heterogeneity of HI antigenic novelty values between
samples in these live builds.

Based on these

@@ -1,17 +1,39 @@
 """Calculate the distance between sequences between seasons.
 """
 import argparse
-from augur.frequency_estimators import TreeKdeFrequencies
+from augur.frequency_estimators import float_to_datestring, timestamp_to_float
 from augur.utils import annotate_parents_for_tree, read_node_data, write_json
 import Bio.Phylo
 import json
 import numpy as np
+import pandas as pd
 
 
 def get_titer_distance_between_nodes(tree, past_node, current_node, titer_attr="dTiter"):
-    ancestor = tree.common_ancestor([past_node, current_node])
-    nodes_between_tips = ancestor.get_path(past_node) + ancestor.get_path(current_node)
-    return sum([node.attr[titer_attr] for node in nodes_between_tips])
+    # Find MRCA of tips from one tip up. Sum the titer attribute of interest
+    # while walking up to the MRCA, to avoid an additional pass later. The loop
+    # below stops when the past node is found in the list of the candidate
+    # MRCA's terminals. This test should always evaluate to true when the MRCA
+    # is the root node, so we should not have to worry about trying to find the
+    # parent of the root.
+    current_node_branch_sum = 0.0
+    mrca = current_node
+    while past_node.name not in mrca.terminals:
+        current_node_branch_sum += mrca.attr[titer_attr]
+        mrca = mrca.parent
+
+    # Sum the node weights for the other tip from the bottom up until we reach
+    # the MRCA. The value of the MRCA is intentionally excluded here, as it
+    # would represent the branch leading to the MRCA and would be outside the
+    # path between the two tips.
+    past_node_branch_sum = 0.0
+    current_node = past_node
+    while current_node != mrca:
+        past_node_branch_sum += current_node.attr[titer_attr]
+        current_node = current_node.parent
+
+    final_sum = past_node_branch_sum + current_node_branch_sum
+    return final_sum
 
 
 if __name__ == "__main__":
@@ -22,6 +44,7 @@ if __name__ == "__main__":
     parser.add_argument("--attribute-name", help="name to store distances", required=True)
     parser.add_argument("--model", help="JSON providing the titer tree model", required=True)
     parser.add_argument("--date-annotations", help="JSON of branch lengths and date annotations from augur refine for samples in the given tree; required for comparisons to earliest or latest date", required=True)
+    parser.add_argument("--months-back-for-current-samples", type=int, help="number of months prior to the last date with estimated frequencies to include samples as current", required=True)
[…]

### 2020-01-14 src/cross_immunity.py

@@ -28,9 +28,11 @@ if __name__ == '__main__':
 
     # Identify maximum frequency per sample.
     max_frequency_per_sample = {
-        sample: float(max(sample_frequencies))
-        for sample, sample_frequencies in frequencies["data"]["frequencies"].items()
+        sample: float(max(sample_frequencies["frequencies"]))
+        for sample, sample_frequencies in frequencies.items()
+        if sample not in ["pivots", "generated_by"] and not sample.startswith("count")
     }
+    current_timepoint = frequencies["pivots"][-1]
 
     # Load distances.
     with open(args.distances, "r") as fh:
@@ -51,6 +53,11 @@ if __name__ == '__main__':
    "rb": 3
   },
     """
+    if args.years_to_wane is not None:
+        print("Waning effect with max years of %i" % args.years_to_wane)
+    else:
+        print("No waning effect")
+
     # Calculate cross-immunity for distances defined by the given attributes.
     cross_immunities = {}
     for sample, sample_distances in distances.items():
@@ -68,15 +75,16 @@ if __name__ == '__main__':
             for past_sample, distance in sample_distances[distance_attribute].items():
                 # Calculate effect of waning immunity.
                 if args.years_to_wane is not None:
-                    waning_effect = max(1 - ((date_by_node_name[sample] - date_by_node_name[past_sample]) / args.years_to_wane), 0)
+                    waning_effect = max(1 - ((current_timepoint - date_by_node_name[past_sample]) / args.years_to_wane), 0)
                 else:
                     waning_effect = 1.0
 
                 # Calculate cost of cross-immunity with waning.
-                cross_immunity += waning_effect * max_frequency_per_sample[past_sample] * cross_immunity_cost(
-                    distance,
-                    decay_factor
-                )
+                if waning_effect > 0:
+                    cross_immunity += waning_effect * max_frequency_per_sample[past_sample] * cross_immunity_cost(
+                        distance,
+                        decay_factor
+                    )
 
             cross_immunities[sample][immunity_attribute] = -1 * cross_immunity
 

### 2020-01-15 manuscript/main.tex
Cite Snakemake

@@
 
-We have provided all source code, configuration files, and datasets at https://github.com/blab/flu-forecasting.
+The entire workflow for our analyses was implemented with Snakemake \cite{Snakemake}.
+We have provided all source code, configuration files, and datasets at \href{https://github.com/blab/flu-forecasting}{https://github.com/blab/flu-forecasting}.
 

### 2020-01-17 manuscript/main.tex
Rename "cross-immunity" to "antigenic novelty"
Antigenic novelty is a more appropriate name for a metric for which higher
numbers lead to higher fitness.

@@
 
-We estimated viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
-All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.
-The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating strains based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
+We estimated viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope antigenic novelty and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) antigenic novelty \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
+All of these metrics except for HI antigenic novelty and DMS mutational effects rely only on HA sequences.
+The antigenic novelty metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating strains based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
 Increased antigenic drift relative to previously circulating strains is expected to correspond to increased viral fitness.
@@
 
-We hypothesized that fitness metrics associated with viral success such as true fitness, epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like mutational load, would receive negative coefficients.
+We hypothesized that fitness metrics associated with viral success such as true fitness, epitope antigenic novelty, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like mutational load, would receive negative coefficients.
 We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
-Correspondingly, we expected that a composite model of epitope cross-immunity and mutational load would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.
+Correspondingly, we expected that a composite model of epitope antigenic novelty and mutational load would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.
 
@@
 In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future.
-With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
+With the exception of epitope antigenic novelty, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
 LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
@@
 
-Surprisingly, the composite model of epitope cross-immunity and mutational load did not perform better than the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
-The cross-immunity fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
+Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+While epitope antigenic novelty was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
+The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
 To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
-By itself, this linear fitness metric did not outperform the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
-However, when we fit a composite model of the epitope ancestor and mutational load metrics, this model outperformed both the individual mutational load model and the composite cross-immunity and mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+By itself, this linear fitness metric did not outperform the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
+However, when we fit a composite model of the epitope ancestor and mutational load metrics, this model outperformed both the individual mutational load model and the composite antigenic novelty and mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated strains was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
@@
 We held out strains collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
-In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
+In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI antigenic novelty and DMS mutational effects.
 We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
-As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than mutational load or DMS mutational effects.
+As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI antigenic novelty would be individually more predictive than mutational load or DMS mutational effects.
 Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
-As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA cross-immunity models in parallel with the HI cross-immunity models.
+As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA antigenic novelty models in parallel with the HI antigenic novelty models.
 
@@
 This distance corresponds to the ``naive model'' in Table~\ref{table_natural_model_selection}.
-Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope antigenic novelty and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
 The average improvement of the sequence-only models over the naive model was consistently lower than the same models in simulated populations.
@@
 
-Of the two metrics for antigenic drift, HI cross-immunity consistently outperformed epitope cross-immunity (Table~\ref{table_natural_model_selection}).
-HI cross-immunity estimated an average distance to the future of 6.04 $\pm$ 1.57 AAs and outperformed the naive model at 17 of 23 timepoints (74\%).
-After the first three training windows where HI measurements were relatively sparse, the HI cross-immunity coefficient was stable across all remaining timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-In contrast, epitope cross-immunity estimated a distance of 7.06 $\pm$ 1.37 AAs and only outperformed the naive model at six timepoints (26\%).
-Epitope cross-immunity was also the only metric whose coefficient started at a positive value (1.45 $\pm$ 0.10 on average prior to October 2009) and transitioned to a negative value through the validation period (-0.13 $\pm$ 0.32 on average for October 2009 and after).
-This strong coefficient for the first half of training windows indicated that, unlike the results for simulated populations, the nonlinear cross-immunity metric was historically an effective measure of antigenic drift.
+Of the two metrics for antigenic drift, HI antigenic novelty consistently outperformed epitope antigenic novelty (Table~\ref{table_natural_model_selection}).
+HI antigenic novelty estimated an average distance to the future of 6.04 $\pm$ 1.57 AAs and outperformed the naive model at 17 of 23 timepoints (74\%).
+After the first three training windows where HI measurements were relatively sparse, the HI antigenic novelty coefficient was stable across all remaining timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+In contrast, epitope antigenic novelty estimated a distance of 7.06 $\pm$ 1.37 AAs and only outperformed the naive model at six timepoints (26\%).
+Epitope antigenic novelty was also the only metric whose coefficient started at a positive value (1.45 $\pm$ 0.10 on average prior to October 2009) and transitioned to a negative value through the validation period (-0.13 $\pm$ 0.32 on average for October 2009 and after).
+This strong coefficient for the first half of training windows indicated that, unlike the results for simulated populations, the nonlinear antigenic novelty metric was historically an effective measure of antigenic drift.
 The historical importance of the epitope sites used for this metric was further supported by the relative enrichment of mutations at these sites for the most successful ``trunk'' lineages of natural populations compared to side branch lineages (Supplemental Table~\ref{sup_table:mutations_by_trunk_status}).
@@
 If the beneficial effects of mutations at these sites were due to historical contingency rather than a constant contribution to antigenic drift, we would expect models based on these sites to perform well until 2005 and then overfit relative to future data.
-Indeed, the epitope cross-immunity model is strongly predictive for the first two validation timepoints until it has to predict to October 2005.
+Indeed, the epitope antigenic novelty model is strongly predictive for the first two validation timepoints until it has to predict to October 2005.
 To further test our hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
 Inspired by the original approach of Shih et al.~\cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population, including 12 that were also present in the original set of 49 sites.
-We fit a cross-immunity model to these 25 sites across the complete validation period and dubbed this the ``oracle cross-immunity'' model, as it benefited from knowledge of the future in its forecasts.
-In contrast to the epitope cross-immunity model, the oracle cross-immunity model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
+We fit a antigenic novelty model to these 25 sites across the complete validation period and dubbed this the ``oracle antigenic novelty'' model, as it benefited from knowledge of the future in its forecasts.
+In contrast to the epitope antigenic novelty model, the oracle antigenic novelty model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
 These results support our hypothesis that the fitness benefit of mutations at the original 49 sites was due to historical contingency and that the success of previous epitope models based on these sites was partly due to ``borrowing from the future''.
-We suspect that our HI cross-immunity model benefits from its ability to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.
+We suspect that our HI antigenic novelty model benefits from its ability to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope antigenic novelty metric is forced to give a constant weight to the same 49 sites throughout time.
 
@@
 
-LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
-Delta frequency performed worse than LBI and HI cross-immunity and only performed slightly better than mutational load.
+LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI antigenic novelty and mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
+Delta frequency performed worse than LBI and HI antigenic novelty and only performed slightly better than mutational load.
 While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
@@
 To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
-Specifically, we fit models based on HI cross-immunity and mutational load, mutational load and LBI, and all three of these metrics together.
+Specifically, we fit models based on HI antigenic novelty and mutational load, mutational load and LBI, and all three of these metrics together.
 We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.
@@
 The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
-In contrast, the three-metric model strongly preferred the HI cross-immunity and mutational load metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
+In contrast, the three-metric model strongly preferred the HI antigenic novelty and mutational load metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
 In the last two validation timepoints, this composite model converged to a nearly LBI-only model with the coefficients for the other two metrics converging to near-zero values.
@@
   \caption{
-  Test of best model for natural populations of H3N2 viruses, the composite model of HI cross-immunity and mutational load.
+  Test of best model for natural populations of H3N2 viruses, the composite model of HI antigenic novelty and mutational load.
   A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
 However, these increases were mostly consistent with an increased average distance between years of 6.82 $\pm$ 1.74 AAs as measured by the naive model.
-The composite of HI cross-immunity, mutational load, and LBI outperformed all other models on the test data with an average distance of 6.12 $\pm$ 1.42 AAs.
-The simpler composite of HI cross-immunity and mutational load performed nearly as well, however, with more stable coefficients per metric over time.
+The composite of HI antigenic novelty, mutational load, and LBI outperformed all other models on the test data with an average distance of 6.12 $\pm$ 1.42 AAs.
+The simpler composite of HI antigenic novelty and mutational load performed nearly as well, however, with more stable coefficients per metric over time.
 For these reasons, we selected this two-metric model for our practical forecasts.
@@
 
-After identifying the composite HI cross-immunity and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
+After identifying the composite HI antigenic novelty and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
 The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$ between observed and estimated growth ratios and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
@@
     The distance to the future of each vaccine strain was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
-    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI cross-immunity and mutational load).
+    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI antigenic novelty and mutational load).
   }
@@
 Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning new coefficients per timepoint from the test data.
-In addition to three composite models described above, we evaluated a composite of FRA cross-immunity and mutational load.
+In addition to three composite models described above, we evaluated a composite of FRA antigenic novelty and mutational load.
 Since the earliest FRA data are from 2012, we anticipated that there were enough measurements to fit a model across the test data time interval.
@@
 These results confirmed that the coefficients for our selected best model would be most accurate for live forecasts.
-Interestingly, the FRA cross-immunity metric received a consistently positive coefficient in its composite with mutational load and this composite ranked among the best models for the test data.
+Interestingly, the FRA antigenic novelty metric received a consistently positive coefficient in its composite with mutational load and this composite ranked among the best models for the test data.
 These results indicate that this FRA-based composite could eventually replace the HI-based composite with mutational load.
 
-After confirming the coefficients for our best model of HI cross-immunity and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021.
+After confirming the coefficients for our best model of HI antigenic novelty and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021.
 Of the three major currently circulating clades, our model predicted growth of the clades A1b/131K and 3c3.A and decline of all other clades (Fig.~\ref{fig:nextstrain_forecasts}).
@@
   \caption{
-    Snapshot of live forecasts on nextstrain.org from our best model (HI cross-immunity and mutational load) for January 2021.
+    Snapshot of live forecasts on nextstrain.org from our best model (HI antigenic novelty and mutational load) for January 2021.
     The observed frequency trajectories for currently circulating clades are shown up to January 2020.
@@
     Snapshot of the last two years of seasonal influenza H3N2 evolution on nextstrain.org showing the estimated distance per sample to the future population.
-    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating strains weighted by the other sample's projected frequencies under the best fitness model (HI cross-immunity and mutational load).
+    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating strains weighted by the other sample's projected frequencies under the best fitness model (HI antigenic novelty and mutational load).
   }
@@
 Models should integrate information from multiple segments of the influenza genome and will need to balance the fitness benefits of evolution in genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment \cite{Villa:2017iw}.
-The strong performance of a composite FRA cross-immunity and mutational load model with recent H3N2 populations indicates that FRAs will continue to be important for future forecasting efforts.
+The strong performance of a composite FRA antigenic novelty and mutational load model with recent H3N2 populations indicates that FRAs will continue to be important for future forecasting efforts.
 In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
@@
 We defined purifying selection across all three domains, allowing the preferred amino acid at each site to change at a fixed rate over time.
-We additionally defined exposure-dependent selection for 49 putative epitope sites in HA1 \cite{Luksza:2014hj} to impose an effect of cross-immunity that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
+We additionally defined exposure-dependent selection for 49 putative epitope sites in HA1 \cite{Luksza:2014hj} to impose an effect of antigenic novelty that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
 We modified the SANTA-SIM source code to enable the inclusion of true fitness values for each sample in the FASTA header of the sampled sequences from each generation.
@@
 We estimated antigenic drift for each strain using either genetic or HI data.
-To estimate antigenic drift with genetic data, we implemented the cross-immunity metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+To estimate antigenic drift with genetic data, we implemented the antigenic novelty metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
 Briefly, for each pair of strains in adjacent seasons, we counted the number of amino acid differences between the strains' HA sequences at 49 epitope sites.
@@
 For each individual strain $i$ at each timepoint $t$, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as in Equation~\ref{equation_epitope_cross_immunity}.
-We defined the constant $D_{0} = 14$, as in the original definition of cross-immunity \cite{Luksza:2014hj}.
+We defined the constant $D_{0} = 14$, as in the original definition of antigenic novelty \cite{Luksza:2014hj}.
 
@@
 Finally, we calculated each strain's ability to escape cross-immunity using Equation~\ref{equation_epitope_cross_immunity} with the pairwise distances between epitope sequences replaced with pairwise antigenic distance from HI data.
-As with the original epitope cross-immunity described above, this HI cross-immunity metric produces higher values for strains that are more antigenically distinct from previously circulating strains.
+As with the original epitope antigenic novelty described above, this HI antigenic novelty metric produces higher values for strains that are more antigenically distinct from previously circulating strains.
 

### 2020-01-17 manuscript/supplement.tex
Rename "cross-immunity" to "antigenic novelty"
Antigenic novelty is a more appropriate name for a metric for which higher
numbers lead to higher fitness.

@@
   \caption{
-  Model coefficients and distances to the future for cross-immunity models fit to natural populations.
+  Model coefficients and distances to the future for antigenic novelty models fit to natural populations.
   A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-  The epitope cross-immunity model relies on previously published epitope sites \cite{Luksza:2014hj}.
-  The ``oracle'' cross-immunity model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
-  The improved performance of the ``oracle'' model indicates that the sequence-based cross-immunity metric can be effective when sites of beneficial mutations are known prior to forecasting.
+  The epitope antigenic novelty model relies on previously published epitope sites \cite{Luksza:2014hj}.
+  The ``oracle'' antigenic novelty model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
+  The improved performance of the ``oracle'' model indicates that the sequence-based antigenic novelty metric can be effective when sites of beneficial mutations are known prior to forecasting.
   }

### 2020-01-17 manuscript/main.tex
Start updating text summary of antigenic novelty results

@@
 Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-While epitope antigenic novelty was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
 The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
 To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
-By itself, this linear fitness metric did not outperform the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
-However, when we fit a composite model of the epitope ancestor and mutational load metrics, this model outperformed both the individual mutational load model and the composite antigenic novelty and mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+This linear fitness metric slightly outperformed the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
+Importantly, a composite model of the epitope ancestor and mutational load metrics outperformed all other epitope-based models and the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated strains was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
@@
 A standard expectation from machine learning is that models will perform worse on test data than on validation data.
-Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-The composite model of mutational load and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_selection}).
+Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
+The composite model of mutational load and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively.
 Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.
@@
 Of the two metrics for antigenic drift, HI antigenic novelty consistently outperformed epitope antigenic novelty (Table~\ref{table_natural_model_selection}).
-HI antigenic novelty estimated an average distance to the future of 6.04 $\pm$ 1.57 AAs and outperformed the naive model at 17 of 23 timepoints (74\%).
-After the first three training windows where HI measurements were relatively sparse, the HI antigenic novelty coefficient was stable across all remaining timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-In contrast, epitope antigenic novelty estimated a distance of 7.06 $\pm$ 1.37 AAs and only outperformed the naive model at six timepoints (26\%).
-Epitope antigenic novelty was also the only metric whose coefficient started at a positive value (1.45 $\pm$ 0.10 on average prior to October 2009) and transitioned to a negative value through the validation period (-0.13 $\pm$ 0.32 on average for October 2009 and after).
+HI antigenic novelty estimated an average distance to the future of 6.01 $\pm$ 1.50 AAs and outperformed the naive model at 16 of 23 timepoints (70\%).
+The coefficient for HI antigenic novelty remained stable across all timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+In contrast, epitope antigenic novelty estimated a distance of 7.13 $\pm$ 1.47 AAs and only outperformed the naive model at seven timepoints (30\%).
+Epitope antigenic novelty was also the only metric whose coefficient started at a positive value (1.17 $\pm$ 0.03 on average prior to October 2009) and transitioned to a negative value through the validation period (-0.19 $\pm$ 0.34 on average for October 2009 and after).
 This strong coefficient for the first half of training windows indicated that, unlike the results for simulated populations, the nonlinear antigenic novelty metric was historically an effective measure of antigenic drift.

### 2020-01-17 manuscript/main.tex
Text and figure revisions
Updates remaining main text to reflect new results from cross-immunity-based
models without waning effect. Also includes numerous minor changes like renaming
"growth ratios" to "fold changes", adding a reference and correct expected amino
acid changes per year to discussion, resizing supplmental tables, and other
prose edits.

@@
 
-We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
+We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and selected the best of these models using time-series cross-validation.
 After selecting optimal models from training and validation, we evaluated the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
@@
   Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
-  A) The correlation of log estimated clade growth ratio, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade growth ratio, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future in the best model for 18 timepoints shows how often the model makes a good choice when forced to select a single representative strain for the future population.
-  The estimated best strain was in the top 20th percentile of observed closest strains for 89\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  A) The correlation of log estimated clade frequency fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade frequency fold change, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  B) The rank of the estimated best strain based on its distance to the future in the best model was in the top 20th percentile for 89\% of 18 timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
   C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
@@
 We validated the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
-First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed and estimated clade growth ratios.
-The model's estimated growth ratios correlated well with observed growth ratios (Pearson's $R^2 = 0.52$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed fold change in clade frequencies, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$ and estimated fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$.
+The model's estimated fold changes correlated well with observed fold changes (Pearson's $R^2 = 0.52$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
 The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
@@
 Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.
-Note, that the numbers of test and validation data points are small and we therefore expect fluctuations.
+Note, that the numbers of test and validation data points are small and we therefore expect fluctuations in model performance.
 
 As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
-While observed and estimated growth ratios correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
-Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
+While observed and estimated clade frequency fold changes correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
+We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
 The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
@@
     \caption{
-      Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
       Validation results are based on 23 timepoints.
@@
 Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
-We held out strains collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
+We held out strains collected after October 1, 2015 up through October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI antigenic novelty and DMS mutational effects.
@@
 If the beneficial effects of mutations at these sites were due to historical contingency rather than a constant contribution to antigenic drift, we would expect models based on these sites to perform well until 2005 and then overfit relative to future data.
-Indeed, the epitope antigenic novelty model is strongly predictive for the first two validation timepoints until it has to predict to October 2005.
-To further test our hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
+Indeed, the epitope antigenic novelty model outperforms the naive model for the first three validation timepoints until it has to predict to April 2006.
+To test this hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
 Inspired by the original approach of Shih et al.~\cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population, including 12 that were also present in the original set of 49 sites.
-We fit a antigenic novelty model to these 25 sites across the complete validation period and dubbed this the ``oracle antigenic novelty'' model, as it benefited from knowledge of the future in its forecasts.
-In contrast to the epitope antigenic novelty model, the oracle antigenic novelty model produced a strong positive coefficient across all training windows (1.06 $\pm$ 0.23) and consistently outperformed the original model with an average distance to the future of 5.59 $\pm$ 1.30 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
+We fit an antigenic novelty model to these 25 sites across the complete validation period and dubbed this the ``oracle antigenic novelty'' model, as it benefited from knowledge of the future in its forecasts.
+The oracle model produced a consistently positive coefficient across all training windows (0.80 $\pm$ 0.21) and consistently outperformed the original epitope model with an average distance to the future of 5.71 $\pm$ 1.27 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
 These results support our hypothesis that the fitness benefit of mutations at the original 49 sites was due to historical contingency and that the success of previous epitope models based on these sites was partly due to ``borrowing from the future''.
@@
 
-LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI antigenic novelty and mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
+LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
 Delta frequency performed worse than LBI and HI antigenic novelty and only performed slightly better than mutational load.
-While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.
+While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results show that the current implementations of these metrics represent qualitatively different fitness components.
 
@@
 Both two-metric composite models performed better than their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
-Although the three-metric composite model performed slightly worse than the individual LBI model based on average distance to the future, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
 The composite of mutational load and LBI performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
+The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
+In contrast, the three-metric model strongly preferred the HI antigenic novelty and mutational load metrics over LBI for the entire validation period, producing an average LBI coefficient of -0.04 $\pm$ 0.09.
+These results suggest that HI assays and LBI may measure similar fitness effects.
 
-The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
-In contrast, the three-metric model strongly preferred the HI antigenic novelty and mutational load metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
-In the last two validation timepoints, this composite model converged to a nearly LBI-only model with the coefficients for the other two metrics converging to near-zero values.
-These results suggest that HI assays and LBI may measure similar fitness effects.
-However, the dominance of LBI later in this three-metric model also suggests that HI assays may no longer be as predictive as they have been historically.
-
-As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth ratios and the ranking of estimated best strains compared to the observed closest strains to future populations.
-The composite model of mutational load and LBI effectively captured clade dynamics with a growth correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
+As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade frequency fold changes and the ranking of estimated best strains compared to the observed closest strains to future populations.
+The composite model of mutational load and LBI effectively captured clade dynamics with a fold change correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
 Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}C).
@@
   Test of best model for natural populations of H3N2 viruses, the composite model of HI antigenic novelty and mutational load.
-  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future for 8 timepoints.
-  The estimated best strain was in the top 20th percentile of observed closest strains for 88\% of timepoints.
+  A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  B) The rank of the estimated best strain based on its distance to the future for eight timepoints.
+  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints.
   C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
@@
 However, these increases were mostly consistent with an increased average distance between years of 6.82 $\pm$ 1.74 AAs as measured by the naive model.
-The composite of HI antigenic novelty, mutational load, and LBI outperformed all other models on the test data with an average distance of 6.12 $\pm$ 1.42 AAs.
-The simpler composite of HI antigenic novelty and mutational load performed nearly as well, however, with more stable coefficients per metric over time.
-For these reasons, we selected this two-metric model for our practical forecasts.
+The composite of HI antigenic novelty and mutational load outperformed all other models on the test data with an average distance of 5.97 $\pm$ 1.47 AAs.
 Surprisingly, the best model for the validation data -- mutational load and LBI -- was one of the worst models for the test data with an average distance to the future of 7.70 $\pm$ 3.53 AAs.
@@
 However, by October 2018, this clade had declined in frequency to approximately 30\% and, by October 2019, it had gone extinct.
-These results highlight a major limitation of growth-based fitness metrics like LBI and a corresponding benefit of more mechanistic metrics that explicitly measure antigenic drift and functional constraint.
+That LBI incorrectly predicted the success of this reassortant clade highlights a major limitation of growth-based fitness metrics and a corresponding benefit of more mechanistic metrics that explicitly measure antigenic drift and functional constraint.
 
 After identifying the composite HI antigenic novelty and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
-The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.52$ between observed and estimated growth ratios and growth and decline accuracies of 67\% and 63\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
+The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.46$ between observed and estimated growth ratios and growth and decline accuracies of 52\% and 58\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
 The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Fig.~\ref{fig:test_of_best_model_for_natural_populations}C).
-The estimated best strain from this model was in the top second percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for seven (88\%) of eight timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}B).
-Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.69$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
+The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}B).
+Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.72$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
 
@@
 For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed and estimated closest strains.
-Our models selected a strain that was closer to the future than the selected vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
-In the two seasons where the vaccine strain was closer to the future than the model's selection, the difference between the two strains' distances to the future was less than one amino acid.
+Our models selected a strain that was closer to the future than the selected vaccine strain for 11 (92\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
+In the one season where the vaccine strain was closer to the future than the model's selection, the difference between the two strains' distances to the future was 0.19 AAs.
 
@@
 Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning new coefficients per timepoint from the test data.
-In addition to three composite models described above, we evaluated a composite of FRA antigenic novelty and mutational load.
-Since the earliest FRA data are from 2012, we anticipated that there were enough measurements to fit a model across the test data time interval.
+Additionally, we evaluated a composite of FRA antigenic novelty and mutational load.
+Since the earliest FRA data were from 2012, we anticipated that there were enough measurements to fit a model across the test data time interval.
 If modern H3N2 strains continue to perform poorly in HI assays, the FRA-based assay will be critical for future forecasting efforts.
 
-None of the three original composite models performed better across the test data than the original best model (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
+Two of three models performed worse after refitting coefficients to the test data than their original fixed coefficient implementations (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
+While, the mutational load and LBI model improved considerably over its original performance, it still performed worse than the naive model on average.
 These results confirmed that the coefficients for our selected best model would be most accurate for live forecasts.
-Interestingly, the FRA antigenic novelty metric received a consistently positive coefficient in its composite with mutational load and this composite ranked among the best models for the test data.
-These results indicate that this FRA-based composite could eventually replace the HI-based composite with mutational load.
+Interestingly, the FRA antigenic novelty metric received a consistently positive coefficient of 1.40 $\pm$ 0.24 in its composite with mutational load.
+Unfortunately, this model performed considerably worse than the corresponding HI-based model.
+These results suggest that we may need more FRA data across a longer historical timespan to train a model that could replace the HI-based model.
 
@@
 We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza H3N2 by estimating the sequence composition of future populations.
-A key innovation of this framework is its ability to directly compare viral populations between seasons using the Earth Mover's Distance metric \cite{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
+A key innovation of this framework is its ability to directly compare viral populations between seasons using the earth mover's distance metric \cite{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
 The best models from this framework still effectively capture clade dynamics and accurately identify optimal vaccine candidates from simulated and natural H3N2 populations without relying on clades as model targets.
@@
 
-We observed the latter type of overfitting in the context of sites associated with antigenic drift.
+We observed overfitting to historical context in sequence-based models of antigenic drift.
 The fitness benefit of mutations that lead to antigenic drift in H3N2 is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}
-Although the antigenic importance of seven specific sites in HA has been experimentally validated \cite{Koel:2013jz}, these sites do not explain all antigenic drift observed in natural populations \cite{Neher:2016hy}.
+Although the antigenic importance of seven specific sites in HA were experimentally validated by Koel et al.
+2013 \cite{Koel:2013jz}, these sites do not explain all antigenic drift observed in natural populations \cite{Neher:2016hy}.
 Other attempts to define these so-called ``epitope sites'' have relied on either aggregation of results from antigenic escape assays \cite{Wolf:2006da} or retrospective computational analyses of sites with beneficial mutations \cite{Shih:2007bd,Luksza:2014hj}.
@@
 These results are consistent with previous work that found improved performance by integrating components of antigenic drift, functional constraint, and clade growth \cite{Luksza:2014hj}.
-However, the convergence of our three-metric model to a LBI-only model during the validation period (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}) reveals the limitations of our current additive approach to composite models.
+However, the effective elimination of LBI from our three-metric model during the validation period (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}) reveals the limitations of our current additive approach to composite models.
 The recent success of weighted ensembles for short-term influenza forecasting \cite{Reich:2019bq} suggests that long-term forecasting may benefit from a similar approach.
@@
 By forecasting the composition of future H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
-While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AAs per year for validation data and 0.66 AAs per year for test data.
+While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AAs per year for validation data and 0.85 AAs per year for test data.
 However, these results are consistent with the observed dynamics of H3N2.
@@
 Hence, we expect the diversity of circulating strains to persist between seasons.
-Second, the average substitution rate for H3N2's HA is 0.0043 substitutions per site per year or an average accumulation of 2.4 AAs per year \cite{}.
-This accumulation of substitutions contributes to the distance between annual populations observed by the naive model.
-In this context, our model gains of 0.96 and 0.66 AAs per year correspond to an explanation of 40\% and 28\% of the expected additional distance between annual populations, respectively.
+Second, H3N2 hemagglutinin accumulates 3.6 amino acid changes per year \cite{Smith:2004jc}.
+This accumulation of amino acid substitutions contributes to the distance between annual populations observed by the naive model.
+In this context, our model gains of 0.96 and 0.85 AAs per year correspond to an explanation of 27\% and 24\% of the expected additional distance between annual populations, respectively.
 
@@
 Integration of more recent experimental data may improve estimates of antigenic drift.
-Models should integrate information from multiple segments of the influenza genome and will need to balance the fitness benefits of evolution in genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment \cite{Villa:2017iw}.
-The strong performance of a composite FRA antigenic novelty and mutational load model with recent H3N2 populations indicates that FRAs will continue to be important for future forecasting efforts.
+Despite the weak performance of our FRA antigenic novelty model on recent data, continued accumulation of FRA measurements over time should eventually enable models as accurate as the current HI-based models.
 In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
@@
 A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape H3N2 strains must navigate to be globally successful.
+Models should also integrate information from multiple segments of the influenza genome and will need to balance the fitness benefits of evolution in genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment \cite{Villa:2017iw}.
 Finally, forecasting models need to account for the geographic distribution of viruses and the vastly different sampling intensities across the globe.
@@
 
-Prediction of future influenza virus populations are intrinsically limited by the small number of data points available to train and test models.
+Prediction of future influenza virus populations is intrinsically limited by the small number of data points available to train and test models.
 Increasingly more complex models are therefore prone to overfitting.
@@
 
-For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the weighted Hamming distance between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of strains at time $t$ by their estimated fitnesses.
-We measured distance between populations using earth mover's distance (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{Rubner1998,Kusner2015}.
+For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the earth mover's distance (EMD) \cite{Rubner1998,Kusner2015} between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of strains at time $t$ by their estimated fitnesses.
 Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
@@
 We then identified the fitness coefficients that provided projected future frequencies that minimized the EMD between the estimated and observed future populations.
-With this metric, an optimal estimate of the future would produce a distance of zero.
+With this metric, a perfect estimate of the future's strain sequence composition and frequencies would produce a distance of zero.
 However, the inevitable accumulation of substitutions between the two populations prevents this outcome.
@@
 To calculate validation error for each training window, we applied the resulting model coefficients to estimate the future frequencies for the year after the last timepoint in the training window.
-These validation errors informed our tuning of hyperparameters including a L1 regularization of the fitness coefficients, the LBI neighborhood parameter $\tau$, and the length of the training window itself.
+These validation errors informed our tuning of hyperparameters.
 Finally, we fixed the coefficients for each model at the mean values across all training windows and applied these fixed models to the test data to estimate the true forecasting accuracy of each model on previously unobserved data.
@@
 We estimated antigenic drift for each strain using either genetic or HI data.
-To estimate antigenic drift with genetic data, we implemented the antigenic novelty metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+To estimate antigenic drift with genetic data, we implemented an antigenic novelty metric based on the ``cross-immunity'' metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
 Briefly, for each pair of strains in adjacent seasons, we counted the number of amino acid differences between the strains' HA sequences at 49 epitope sites.
@@
 For each individual strain $i$ at each timepoint $t$, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as in Equation~\ref{equation_epitope_cross_immunity}.
-We defined the constant $D_{0} = 14$, as in the original definition of antigenic novelty \cite{Luksza:2014hj}.
+We defined the constant $D_{0} = 14$, as in the original definition of cross-immunity \cite{Luksza:2014hj}.
+To compare these epitope sites with other previously published sites, we fit epitope antigenic novelty models based on sites defined by Wolf et al.
+2006 \cite{Wolf:2006da} and Koel et al.
+2013 \cite{Koel:2013jz}.
 
 \begin{equation}
-    f_{i,\mathrm{ep}}(t) = \sum_{j: t_{j} < t_{i}}{x_{j}\exp{(-D_{\mathrm{ep}}(a_{i}, a_{j}) / D_{0})}}
+    f_{i,\mathrm{ep}}(t) = \sum_{j: t_{j} < t_{i}}{-\max(x_{j})\exp{(-D_{\mathrm{ep}}(a_{i}, a_{j}) / D_{0})}}
     \label{equation_epitope_cross_immunity}
@@
 We required mutations to emerge from below 5\% global frequency and reach $>$90\% frequency.
-Although we did not require sweeps to complete within a fixed amount of time, we observed that they required no longer than 1-3 years to complete.
+Although we did not require sweeps to complete within a fixed amount of time, we observed that they required no longer than one to three years to complete.
 To minimize false positives, we eliminated any sites where one or more mutations rose above 20\% frequency and subsequently died out.
@@
 This method effectively estimates the antigenic drift per branch in units of $log_{2}$ titer change.
-Next, we calculated the pairwise antigenic distance between strains as the sum of antigenic drift weights per branch on the phylogenetic path between each pair of strains.
+We selected all strains with nonzero frequencies in the last six months as ``current strains'' and all strains sampled five years prior to that threshold as ``past strains''.
+Next, we calculated the pairwise antigenic distance between all current and past strains as the sum of antigenic drift weights per branch on the phylogenetic path between each pair of strains.
 Finally, we calculated each strain's ability to escape cross-immunity using Equation~\ref{equation_epitope_cross_immunity} with the pairwise distances between epitope sequences replaced with pairwise antigenic distance from HI data.
@@
 JH is a Graduate Research Fellow and is supported by the NIH grant NIAID F31AI140714.
+RAN is supported by NIAID R01 AI127893-01 and institutional core funding.
 TB is a Pew Biomedical Scholar and is supported by NIH grants NIGMS R35 GM119774-01, NIAID U19 AI117891-01 and NIAID R01 AI127893-01.
-RAN is supported by NIAID R01 AI127893-01 and institutional core funding.
 
@@
 JH planned experiments, implemented the final forecasting framework, analyzed results, and wrote the manuscript.
-RAN planned experiments and reviewed the manuscript.
-TB planned experiments, implemented the initial forecasting framework, and reviewed the manuscript.
+RAN planned experiments and edited the manuscript.
+TB planned experiments, implemented the initial forecasting framework, and edited the manuscript.
 

### 2020-01-17 manuscript/supplement.tex
Text and figure revisions
Updates remaining main text to reflect new results from cross-immunity-based
models without waning effect. Also includes numerous minor changes like renaming
"growth ratios" to "fold changes", adding a reference and correct expected amino
acid changes per year to discussion, resizing supplmental tables, and other
prose edits.

@@
   Validation of best model for simulated populations of H3N2-like viruses.
-  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
   B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
@@
   Validation of best model for natural populations of H3N2 viruses, the composite model of mutational load and LBI.
-  A) The correlation of estimated and observed clade growth ratios shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+  A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
   B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.
@@
       Validation results are based on 23 timepoints.
-      Test results are based on 8 timepoints not observed during model training and validation.
+      Test results are based on eight timepoints not observed during model training and validation.
       Model results for additional variants of fitness metrics including those based on epitope mutations and DMS preferences are included for reference.

### 2020-01-17 scripts/plot_validation_figure_by_population.py
Rename growth ratio to fold change in validation figures

@@ -332,8 +332,8 @@ if __name__ == "__main__":
         transform=clade_ax.transAxes
     )
 
-    clade_ax.set_xlabel("Observed $log_{10}$ growth ratio")
-    clade_ax.set_ylabel("Estimated $log_{10}$ growth ratio")
+    clade_ax.set_xlabel("Observed $log_{10}$ fold change")
+    clade_ax.set_ylabel("Estimated $log_{10}$ fold change")
 
     growth_rate_ticks = np.arange(-6, 4, 1)
     clade_ax.set_xticks(growth_rate_ticks)

### 2020-01-25 manuscript/cover-letter/cover_letter.tex
Revise cover letter
Reduce length and directly address four primary questions asked by the the
journal.

@@
   We would be grateful if you considered it for publication in \textit{eLife}.
-  We believe that \textit{eLife} is an appropriate venue, as this work provides the most comprehensive analysis of fitness metrics for seasonal influenza to date and a generalized forecasting framework that will be a valuable tool for other biological researchers.
 
-  The selection of vaccine strains for seasonal influenza A/H3N2 remains a daunting challenge due to constant accumulation of mutations in the hemagglutinin (HA) surface protein.
-  Until recently, experts selected vaccine strains by manually synthesizing experimental measurements of antigenic drift provided by hemagglutination inhibition (HI) assays.
-  Two landmark papers in 2014 showed that H3N2 evolution may be predictable from HA sequences alone.
-  The first paper, ``A predictive fitness model for influenza'' by {\L}uksza and L\"assig in \textit{Nature}, established a mechanistic fitness model that predicted the frequencies of phylogenetic clades one year in advance.
-  This model defined three biological fitness metrics including antigenic drift measured by mutations at epitope sites, functional constraint measured by mutations at non-epitope sites, and recent clade growth measured by synonymous mutations.
-  The second paper, ``Predicting evolution from the shape of genealogical trees'' by Neher et al.
-in \textit{eLife}, developed a phylogenetic metric of recent clade fitness called the local branching index (LBI).
-  This metric accurately identified high fitness clades without providing a biological basis for that success.
-  Predictions from both of these models are now routinely considered in the World Health Organization's vaccine composition meetings.
+  The World Health Organization selects vaccine strains for seasonal influenza A/H3N2 based on experimental measurements of antigenic drift and computational models of hemagglutinin (HA) sequence evolution.
+  Hemagglutination inhibition (HI) assays provide phenotypic measurements of antigenic drift that have been the gold standard for decades.
+  In their 2014 \textit{Nature} paper, {\L}uksza and L\"assig provided the first mechanistic fitness model to predict the success of H3N2 populations one year into the future from HA sequences alone.
+  That same year in \textit{eLife}, Neher et al. presented the local branching index (LBI), a fitness estimate based on rapid branching in HA phylogenies.
+  No studies to date have integrated these genetic and phenotypic measures of HA into a single model to understand their relative contributions to influenza fitness.
 
-  Several substantial questions remained unanswered by these models.
-  Are genetic estimates of antigenic drift sufficient replacements for the gold standard phenotypic measures provided by HI assays?
-  What mechanistic components of influenza fitness does LBI measure?
-  Can simple genetic estimates of functional constraint be improved upon using modern deep mutational scanning (DMS) assays that measure the functional effects of every single amino acid mutation to a given strain's HA?
-  Here, we investigate for the first time the relationships between all of these genotypic and phenotypic estimates of influenza fitness using a collection of HA sequences and HI measurements of an unprecedented breadth and depth.
-  To this end, we developed a novel approach to forecasting influenza evolution by estimating the sequence composition of future populations and rigorously tested this framework with simulated and natural H3N2 populations.
-  We implemented new experimentally-informed fitness metrics including antigenic drift by HI assays and functional constraint by DMS assays.
-  We applied modern machine learning methods for model training, validation, and testing to thoroughly evaluate the predictive accuracy of these new metrics as well as {\L}uksza and L\"assig's epitope and non-epitope metrics and Neher et al.'s LBI.
-  Importantly, we verified that this framework can both capture dynamics of phylogenetic clades and identify optimal candidate strains for vaccine composition.
+  To address this gap, we present the first fitness framework to integrate genetic and phenotypic measures of HA antigenic drift and functional constraint and phylogenetic measures of clade growth.
+  We implemented novel measures of antigenic drift by HI assays, functional constraint measured by deep mutational scanning (DMS) experiments, and clade growth based on recent changes in clade frequencies.
+  We also reimplemented {\L}uksza and L\"assig's sequence-based measures of antigenic drift and functional constraint, and Neher et al.'s clade growth metric, LBI.
+  To identify the most predictive models, we tested each of these fitness metrics individually and in relevant combinations using time-series cross-validation.
+  The most accurate models included a combination of sequence-based functional constraint and either HI-based measures of antigenic drift or LBI-based measures of clade growth.
 
-  We find that phenotypic measures of antigenic drift consistently outperform genetic estimates based on epitope sites.
-  We further show that the predictive value of these epitope sites in {\L}uksza and L\"assig's mechanistic model was most likely due to inadvertantly borrowing information from the future during model fitting.
-  This surprising finding has important ramifications for future attempts to map influenza genotypes to fitness, the use of epitope sites in practical forecasts, and the continued value of HI assays.
-  Additionally, we find that DMS-based estimates of functional constraint produced models that overfit to the genetic background of the strain, A/Perth/16/2009, used in the experiments.
-  Instead, we find that {\L}uksza and L\"assig's original minimal model of mutational load at non-epitope sites is more robust.
-  Composite models that integrate mutational load with either HI phenotypes or LBI performed the best and indicate that LBI may be measuring similar fitness effects as HI assays.
-  We have integrated these models into nextstrain.org, enabling real-time forecasts for H3N2 for influenza researchers, decision makers, and the public.
+  Our findings answer several open questions in the field and point the way for future forecasting efforts.
+  We find that genetic estimates of antigenic drift are insufficient replacements for the gold standard phenotypic measures provided by HI assays.
+  We further show that the success of previous models based on these genetic estimates is due, in part, to inadvertantly borrowing information from the future.
+  In contrast, we show that simple genetic estimates of functional constraint are superior to more detailed estimates provided by deep mutational scanning (DMS) measurements from a single background strain.
+  The results suggest that future forecasting models should strive not to replace genetic fitness metrics with phenotypic metrics but to identify metrics that are the least likely to be affected by historical contingency.
 
-  Our findings challenge the widely held assumption about the power of epitope sites to predict the success of influenza populations.
-  Correspondingly, we validate the predictive value of HI assays to measure antigenic drift and provide a framework that synthesizes these measurements into forecasts for expert interpretation.
-  Beyond influenza, our framework enables similar forecasting efforts for other viral and bacterial pathogens by allowing researchers to define sample-specific fitness metrics in tidy data frames and fit models to those metrics without relying on phylogenetic inference.
-  We are confident this open source framework will provide a foundation on which future forecasting efforts can be built or inspired.
-  For these reasons, our software and findings are relevant not only to the specific audience of influenza researchers, but also to the more general audience of infectious disease researchers, evolutionary biologists, and bioinformaticians.
+  We designed our fitness framework as a reproducible workflow built from open source tools, released as open source itself, and integrated into the real-time pathogen surveillance ecosystem of Nextstrain.
+  Our models estimate the sequence composition of future populations instead of phylogenetic clade frequencies, enabling subsequent forecasting efforts for rapidly evolving organisms whose genomes are not suited for phylogenetic inference such as recombining viruses and bacteria.
+  Importantly, fitness metrics can be readily defined in tidy data frames from any continuous trait of a sample.
+  Altogether, this manuscript and the associated fitness framework are immediately relevant to a general audience of evolutionary biologists, infectious disease researchers, virologists, and bioinformaticians.
+  For these reasons, we believe that \textit{eLife} is an appropriate venue for this work.
 \end{newlfm}

### 2020-01-26 manuscript/cover-letter/cover_letter.tex
Minor edit to cover letter

@@
   In contrast, we show that simple genetic estimates of functional constraint are superior to more detailed estimates provided by deep mutational scanning (DMS) measurements from a single background strain.
-  The results suggest that future forecasting models should strive not to replace genetic fitness metrics with phenotypic metrics but to identify metrics that are the least likely to be affected by historical contingency.
+  These results suggest that future forecasting models should strive not to replace genetic fitness metrics with phenotypic metrics but to identify metrics that are the least likely to be affected by historical contingency.
 

### 2020-04-04 scripts/standardize_simulated_sequence_dates.py
Update standardize simulated dates scripts
This is part of the migration from run to shell blocks.

@@ -1,12 +1,34 @@
-"""Concatenate two or more tables as data frames.
+"""Standardize simulated sequence dates to human readable format (YYYY-MM-DD).
 """
 import argparse
+import datetime
 import pandas as pd
+
+
+def float_to_datestring(time):
+    """Convert a floating point date from TreeTime `numeric_date` to a date string
+    """
+    # Extract the year and remainder from the floating point date.
+    year = int(time)
+    remainder = time - year
+
+    # Calculate the day of the year (out of 365 + 0.25 for leap years).
+    tm_yday = int(remainder * 365.25)
+    if tm_yday == 0:
+        tm_yday = 1
+
+    # Construct a date object from the year and day of the year.
+    date = datetime.datetime.strptime("%s-%s" % (year, tm_yday), "%Y-%j")
+
+    # Build the date string with zero-padded months and days.
+    date_string = "%s-%.2i-%.2i" % (date.year, date.month, date.day)
+
+    return date_string
 
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser()
-    parser.add_argument("--metadata", nargs="+", help="")
+    parser.add_argument("--metadata", help="metadata for simulated sequences")
     parser.add_argument("--start-year", default=2000.0, type=float, help="year to start simulated dates from")
     parser.add_argument("--generations-per-year", default=200.0, type=float, help="number of generations to map to a single yeasr")
     parser.add_argument("--output", help="metadata with standardized dates and nonzero fitness records")

### 2020-04-04 scripts/filter_simulated_metadata.py
Filter simulated metadata with a script instead of a run block

@@ -1,0 +1,23 @@
+import argparse
+import Bio.SeqIO
+import pandas as pd
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--sequences", help="simulated sequences that have already been filtered")
+    parser.add_argument("--metadata", help="original metadata table for simulated sequences")
+    parser.add_argument("--output", help="filtered metadata where only samples present in the given sequences are included")
+
+    args = parser.parse_args()
+
+    # Get a list of all samples that passed the sequence filtering step.
+    sequences = Bio.SeqIO.parse(args.sequences, "fasta")
+    sample_ids = [sequence.id for sequence in sequences]
+
+    # Load all metadata.
+    metadata = pd.read_csv(args.metadata, sep="\t")
+    filtered_metadata = metadata[metadata["strain"].isin(sample_ids)].copy()
+
+    # Save only the metadata records that have entries in the filtered sequences.
+    filtered_metadata.to_csv(args.output, sep="\t", header=True, index=False)

### 2020-04-04 scripts/get_titers_by_passage.py
Move logic to filter titers by passage into a script

@@ -1,0 +1,26 @@
+import argparse
+import pandas as pd
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--titers", required=True, help="JSON of complete titer records to be filtered by passage type")
+    parser.add_argument("--passage-type", required=True, help="type of passage for viruses used in titer assays")
+    parser.add_argument("--output", required=True, help="table of filtered titer records by passage type")
+
+    args = parser.parse_args()
+
+    df = pd.read_json(args.titers)
+    passaged = (df["serum_passage_category"] == args.passage_type)
+    tdb_passaged = df["index"].apply(lambda index: isinstance(index, list) and args.passage_type in index)
+    tsv_fields = [
+        "virus_strain",
+        "serum_strain",
+        "serum_id",
+        "source",
+        "titer",
+        "assay_type"
+    ]
+
+    titers_df = df.loc[(passaged | tdb_passaged), tsv_fields]
+    titers_df.to_csv(args.output, sep="\t", header=False, index=False)

### 2020-04-04 scripts/filter_strains_with_ambiguous_dates.py
Convert metadata filters in run blocks to scripts

@@ -1,0 +1,16 @@
+import argparse
+import pandas as pd
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument("--metadata", required=True, help="table of metadata to be filtered based on a date column")
+    parser.add_argument("--date-field", default="date", help="name of date column in the metadata")
+    parser.add_argument("--output", required=True, help="table of filtered metadata")
+
+    args = parser.parse_args()
+
+    df = pd.read_csv(args.metadata, sep="\t")
+
+    # Exclude strains with ambiguous collection dates.
+    df[~df[args.date_field].str.contains("XX")].to_csv(args.output, sep="\t", header=True, index=False)

### 2020-04-04 scripts/frequencies_to_table.py
Merge node data and frequencies with a script instead of a run block

@@ -34,7 +34,7 @@ if __name__ == '__main__':
         frequencies = {
             node_name: region_frequencies["global"]
             for node_name, region_frequencies in frequencies_json.items()
-            if node_name not in ["pivots", "counts"]
+            if node_name not in ["pivots", "counts", "generated_by"]
         }
 
     # Collect the last frequency for each node keeping only terminal nodes

### 2020-04-04 scripts/merge_node_data_and_frequencies.py

@@ -1,0 +1,34 @@
+import argparse
+import pandas as pd
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument("--node-data", required=True, help="table of node data from one or more timepoints")
+    parser.add_argument("--kde-frequencies", required=True, help="table of KDE frequencies by strain name and timepoint")
+    parser.add_argument("--diffusion-frequencies", required=True, help="table of diffusion frequencies by strain name and timepoint")
+    parser.add_argument("--preferred-frequency-method", choices=["kde", "diffusion"], help="specify which frequency method should be used for the primary frequency column")
+    parser.add_argument("--output", required=True, help="table of merged node data and frequencies")
+
+    args = parser.parse_args()
+
+    node_data = pd.read_table(args.node_data)
+    kde_frequencies = pd.read_table(args.kde_frequencies)
+    diffusion_frequencies = pd.read_table(args.diffusion_frequencies)
+    df = node_data.merge(
+        kde_frequencies,
+        how="inner",
+        on=["strain", "timepoint", "is_terminal"]
+    ).merge(
+        diffusion_frequencies,
+        how="inner",
+        on=["strain", "timepoint", "is_terminal"]
+    )
+
+    # Annotate frequency by the preferred method if there isn't already a
+    # frequency column defined.
+    if "frequency" not in df.columns:
+        df["frequency"] = df["%s_frequency" % args.preferred_frequency_method]
+
+    df = df[df["frequency"] > 0.0].copy()
+    df.to_csv(args.output, sep="\t", index=False, header=True)

### 2020-04-05 scripts/project_frequencies.py
Remove unused Snakemake files and Python scripts

@@ -1,239 +1,0 @@
-"""
-Project the frequency trajectories of clades including:
-
-  - observed frequencies over all time periods
-  - censored frequencies at a specific timepoint
-  - predicted frequencies from a specific timepoint and delta t
-
-Inputs to these plots include:
-
-  - a tree to estimate frequencies from
-  - a set of fitness model parameters for a given set of predictors
-"""
-import argparse
-from collections import defaultdict
-import json
-import numpy as np
-import os
-import pandas as pd
-import sys
-
-import matplotlib
-matplotlib.use("Agg")
-import matplotlib.pyplot as plt
-
-# augur imports.
-augur_path = os.path.join(os.path.dirname(os.path.dirname(os.path.abspath(__file__))), "dist", "augur")
-sys.path.append(augur_path)
-from base.fitness_model import fitness_model as FitnessModel
-from base.frequencies import KdeFrequencies
-
-from utils import load_tree_from_json_filename, load_frequencies_from_json_filename
-
-
-def project_clade_frequencies_by_delta_from_time(tree, model, time, delta, delta_steps_per_year=12):
-    """
-    Project clade frequencies from a given time to the future by a given delta.
-    """
-    # Calculate the steps between the projection date and delta time into the
-    # future. First, find the frequency pivot that is closest to the requested
-    # projection date.
-    max_date = model.timepoints[np.searchsorted(model.timepoints, time)]
-    future_date = max_date + delta
-
-    # Then, calculate a fixed number of steps between that pivot and delta time
-    # into the future.
-    projected_pivots = np.linspace(max_date, future_date, int(delta_steps_per_year * delta))
-    deltas = projected_pivots - max_date
-
-    # Identify tip predictors and frequencies at the current time point.
-    all_pred = model.predictor_arrays[max_date]
-    all_freqs = model.freq_arrays[max_date]
-
-    # For each requested delta, project current tip frequencies using the model
-    # and calculate the corresponding projected clade frequencies.
-    projected_clade_frequencies = defaultdict(list)
-
-    for delta in deltas:
-        # Project all tip frequencies.
-        pred_freq = model.projection(model.model_params, all_pred, all_freqs, delta)
-
-        # Normalize projected frequencies.
-        pred_freq = pred_freq / pred_freq.sum()
-
-        # Store projected frequencies by clade id.
-        for i, tip in enumerate(model.tips):
[…]

### 2020-04-05 scripts/utils.py

@@ -1,33 +1,0 @@
-"""
-Utilities for working with augur.
-"""
-import json
-import os
-import sys
-
-# augur imports.
-sys.path.append(os.path.join(os.path.dirname(os.path.dirname(os.path.abspath(__file__))), "augur"))
-from base.frequencies import KdeFrequencies
-from base.io_util import json_to_tree
-
-
-def load_tree_from_json_filename(filename):
-    # Load JSON tree.
-    with open(filename, "r") as json_fh:
-         json_tree = json.load(json_fh)
-
-    # Convert JSON tree layout to a Biopython Clade instance.
-    tree = json_to_tree(json_tree)
-
-    return tree
-
-
-def load_frequencies_from_json_filename(filename):
-    # Load JSON frequencies.
-    with open(filename, "r") as json_fh:
-         json_frequencies = json.load(json_fh)
-
-    # Create frequencies instance from JSON.
-    frequencies = KdeFrequencies.from_json(json_frequencies)
-
-    return frequencies

### 2020-04-07 manuscript/main.tex
Add authors from NIID
Add authors from Crick and VIDRL CCs and update contributions
Note that our modification to SANTA-SIM is in the official repo now

@@
 We modified the SANTA-SIM source code to enable the inclusion of true fitness values for each sample in the FASTA header of the sampled sequences from each generation.
-This modified implementation is available at \url{https://github.com/huddlej/santa-sim/tree/emit-fitness}.
+This modified implementation has been integrated into the official SANTA-SIM code repository at \url{https://github.com/santa-dev/santa-sim} as of commit e2b3ea3.
 For our full analysis of model performance, we sampled 90 viruses per month to match the sampling density of natural populations.
@@
 JH planned experiments, implemented the final forecasting framework, analyzed results, and wrote the manuscript.
+JB, TR, XX, RK, DEW, LW, BE, RSD, JWM, SF, KN, NK, SW, HH, IB, and KS performed and provided data from serological assays.
 RAN planned experiments and edited the manuscript.

### 2020-04-08 manuscript/main.tex
Clarify wording and captions

@@
 However, frequent viral mutation results in viruses that escape previously acquired human immunity.
-The World Health Organization (WHO) Global Influenza Surveillance and Response System (GISRS) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine strain selection requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
+The World Health Organization (WHO) Global Influenza Surveillance and Response System (GISRS) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine design requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
@@
     A) Strains at timepoint $t$, $\mathbf{x}(t)$, are shown in their phylogenetic context and sized by their frequency at that timepoint.
-    The estimated future population at timepoint $u$, $\mathbf{\hat{x}}(u)$, is projected to the right with strains scaled in size by their projected frequency based on the true fitness model.
+    The estimated future population at timepoint $u$, $\mathbf{\hat{x}}(u)$, is projected to the right with strains scaled in size by their projected frequency based on the known fitness of each simulated strain.
     B) The frequency trajectories of strains at timepoint $t$ to $u$ represent the predicted the growth of the dark blue strains to the detriment of the pink strains.
@@
 the emergence of the light blue cluster from within the successful dark blue cluster.
-    Model coefficients minimize the earth mover's distance between amino acid sequences in the observed and estimated future populations across all training windows.
+    Model coefficients minimize the earth mover's distance between amino acid sequences in the observed, $\mathbf{x}(u)$,  and estimated, $\mathbf{\hat{x}}(u)$, future populations across all training windows.
   }

### 2020-04-08 manuscript/main.tex
Use normal figure environments to fix pandoc conversion to Word
The references to "figure*" environments don't convert properly with
pandoc-crossref, but the normal "figure" environments do.

@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
   \begin{center}
@@
   \end{center}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}[ht!]
+\begin{figure}[ht!]
   \begin{center}
@@
 
-\begin{table*}[ht]
+\begin{table}[ht]
   \begin{center}
@@
   \end{center}
-\end{table*}
+\end{table}
 
@@
 
-\begin{figure*}[p]
+\begin{figure}[p]
   \begin{center}
@@
   \end{center}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
   \begin{center}
@@
   \caption{
-    Snapshot of the last two years of seasonal influenza H3N2 evolution on nextstrain.org showing the estimated distance per sample to the future population.
-    Distance to the future is calculated for each sample as the Hamming distance of HA amino acid sequences to all other circulating strains weighted by the other sample's projected frequencies under the best fitness model (HI antigenic novelty and mutational load).
+    Snapshot of the last two years of seasonal influenza H3N2 evolution on nextstrain.org showing the estimated distance per strain to the future population.
+    Distance to the future is calculated for each strain as the Hamming distance of HA amino acid sequences to all other circulating strains weighted by the other strain's projected frequencies under the best fitness model (HI antigenic novelty and mutational load).
   }

### 2020-04-08 manuscript/main.tex
Replace "sample" with "strain"

@@
 We developed a framework to forecast seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
-As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
-We estimated the frequency of virus strains every six months using kernel density estimation (KDE) to represent the frequency of each sample.
+As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus strain an exponential growth factor scaled by an estimate of the strain's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
+We estimated the frequency of virus strains every six months using kernel density estimation (KDE) to represent the frequency of each strain.
 
@@
 All of these metrics except for HI antigenic novelty and DMS mutational effects rely only on HA sequences.
-The antigenic novelty metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating strains based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
+The antigenic novelty metrics estimate how antigenically distinct each strain at time $t$ is from previously circulating strains based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
 Increased antigenic drift relative to previously circulating strains is expected to correspond to increased viral fitness.
-Mutational load is the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each sample since their ancestor in the previous season.
+Mutational load is the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each strain since their ancestor in the previous season.
 DMS mutational effects provide a more comprehensive biophysical model of functional constraint by measuring the beneficial or deleterious effect of each possible single amino acid mutation in HA from the background of a previous vaccine strain, A/Perth/16/2009.
@@
 We fit models to these simulated populations using all sequence-only fitness metrics.
-As a positive control for our model framework, we also fit a model based on the true fitness of each sample as measured by the simulator.
+As a positive control for our model framework, we also fit a model based on the true fitness of each strain as measured by the simulator.
 We evaluated the performance of each model by the absolute distance between projected and observed future populations.
@@
 The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
-To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
+To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each strain's ancestor in the previous season.
 This linear fitness metric slightly outperformed the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
@@
 We additionally defined exposure-dependent selection for 49 putative epitope sites in HA1 \cite{Luksza:2014hj} to impose an effect of antigenic novelty that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
-We modified the SANTA-SIM source code to enable the inclusion of true fitness values for each sample in the FASTA header of the sampled sequences from each generation.
+We modified the SANTA-SIM source code to enable the inclusion of true fitness values for each strain in the FASTA header of the sampled sequences from each generation.
 This modified implementation has been integrated into the official SANTA-SIM code repository at \url{https://github.com/santa-dev/santa-sim} as of commit e2b3ea3.
@@
 
-To account for uncertainty in collection date and sampling error, we applied a kernel density estimation (KDE) approach to calculate global sample frequencies.
-Specifically, we constructed a Gaussian kernel for each sample with the mean at the reported collection date and a variance (or KDE bandwidth) of two months.
+To account for uncertainty in collection date and sampling error, we applied a kernel density estimation (KDE) approach to calculate global strain frequencies.
+Specifically, we constructed a Gaussian kernel for each strain with the mean at the reported collection date and a variance (or KDE bandwidth) of two months.
 The bandwidth was identified by cross-validation, as described above.
-This bandwidth also roughly corresponds to the median lag time between sample collection and submission to the GISAID database.
-We estimated the frequency of each sample at each timepoint by calculating the probability density function of each KDE at that timepoint and normalizing the resulting values to sum to one.
+This bandwidth also roughly corresponds to the median lag time between strain collection and submission to the GISAID database.
+We estimated the frequency of each strain at each timepoint by calculating the probability density function of each KDE at that timepoint and normalizing the resulting values to sum to one.
 We implemented this frequency estimation logic in the augur frequencies command.
@@
 We assumed that the evolution seasonal influenza H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \cite{Luksza:2014hj}.
-Under this model, we estimated the future frequency, $\hat{x}_{i}(t + \Delta{t})$, of each sample $i$ from the sample's current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
+Under this model, we estimated the future frequency, $\hat{x}_{i}(t + \Delta{t})$, of each strain $i$ from the strain's current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
 
@@
 
-We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
-For example, Equation~\ref{equation_fitness_estimation} estimates fitness per sample by mutational load ($\mathrm{ml}$) and local branching index ($\mathrm{lbi}$).
+We defined the fitness of each strain at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
+For example, Equation~\ref{equation_fitness_estimation} estimates fitness per strain by mutational load ($\mathrm{ml}$) and local branching index ($\mathrm{lbi}$).
 
@@
 Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
-This solution requires both a ``ground distance'' between pairs of strains from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.
+This solution requires both a ``ground distance'' between pairs of strains from both populations and weights assigned to each strain that determine how much that strain contributes to the overall distance.
 
@@
 We calculated the pairwise Hamming distance between the full-length HA amino acid sequences for all selected strains and applied t-SNE dimensionality reduction \cite{vanDerMaaten2008} to the resulting distance matrix (n=2 components, perplexity=30.0, and learning rate=400).
-We assigned each sample to a cluster based on its two-dimensional t-SNE embedding using DBSCAN \cite{Ester1996} with a maximum neighborhood distance of 10 AAs and a minimum of 20 strains per cluster.
+We assigned each strain to a cluster based on its two-dimensional t-SNE embedding using DBSCAN \cite{Ester1996} with a maximum neighborhood distance of 10 AAs and a minimum of 20 strains per cluster.
 Despite known limitations of applying hierarchical clustering to manifold projections that do not preserve sample density, this approach allowed us to effectively assign strains to qualitative genetic clusters for the purposes of visualization.

### 2020-05-05 manuscript/main.tex
Add funding sources for Crick CC
Add funding source for VIDRL CC

@@
 JH is a Graduate Research Fellow and is supported by the NIH grant NIAID F31AI140714.
+The work done at the Crick Worldwide Influenza Centre was supported by the Francis Crick Institute receiving core funding from Cancer Research UK (FC001030), the Medical Research Council (FC001030) and the Wellcome Trust (FC001030).
+The Melbourne WHO Collaborating Centre for Reference and Research on Influenza is supported by the Australian Government Department of Health.
 RAN is supported by NIAID R01 AI127893-01 and institutional core funding.

### 2020-05-16 manuscript/main.tex
Merge edits after clearing CDC review
One major change here is the clarification that forecasts made by the model from
January 2020 to 2021 were made on January 17. This addresses a concern of Xiyan
Xu that A1b/131K as it is currently circulating would need to accumulate more
mutations to be successful. Relative to the perspective of January 17, these
changes may have already happened by May 16. I added a brief description of how
A1b/131K and 3c3.A have changed in frequency in the four months since those
initial predictions.

@@
 However, frequent viral mutation results in viruses that escape previously acquired human immunity.
-The World Health Organization (WHO) Global Influenza Surveillance and Response System (GISRS) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine design requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
+The World Health Organization (WHO) Global Influenza Surveillance and Response System (GISRS) selects vaccine viruses to represent circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine design requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
 Until recently, the hemagglutination inhibition (HI) assay has been the primary experimental measure of antigenic cross-reactivity between pairs of circulating viruses \cite{hirst1943studies}.
-Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \cite{Chambers:2015jt,Zost2017}, prompting the increased use of focus reduction assays (FRAs) \cite{Okuno:1990uu}.
-Together, these two assays are the gold standard for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
+Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \cite{Chambers:2015jt,Zost2017}, prompting the increased use of virus neutralization assays including the neutralization-based focus reduction assay (FRA) \cite{Okuno:1990uu}.
+Together, these two assays are the gold standard in virus antigenic characterizations for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
 As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.
@@
 
-After confirming the coefficients for our best model of HI antigenic novelty and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021.
+After confirming the coefficients for our best model of HI antigenic novelty and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021 using data available up through January 17, 2020.
 Of the three major currently circulating clades, our model predicted growth of the clades A1b/131K and 3c3.A and decline of all other clades (Fig.~\ref{fig:nextstrain_forecasts}).
+Four months after those forecasts, the subclades of A1b/131K represented 40\% of the global H3N2 population an increase from an estimated 35\% frequency in January.
+The 3c3.A clade increased slightly in global frequency from 13\% to 15\%, in that time.
 To aid with identification of potential vaccine candidates for the next season, we annotated strains in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
@@
 
+The findings and conclusions in this report are those of the author(s) and do not necessarily represent the official position of the Centers for Disease Control and Prevention.
+
 \section*{Author contributions}

### 2020-05-16 manuscript/main.tex
Revisions to clarify model interpretations
Integrates edits recommended by Richard with some minimal rewriting by jlhudd.

@@
 Together, these two assays are the gold standard in virus antigenic characterizations for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
-As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.
+As a result, researchers have developed computational methods to predict influenza evolution from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.
 
@@
 
-\subsection*{Models accurately forecast evolution of H3N2-like viruses}
+\subsection*{Models accurately forecast evolution of simulated H3N2-like viruses}
 
@@
 With the exception of epitope antigenic novelty, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
-LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
+LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs consistent with the fact that the LBI is a correlate of fitness in models of rapidly adapting populations \cite{Neher:2014eu}.
 Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
@@
 The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
-These results support our hypothesis that multiple complementary metrics can produce more accurate models.
+This small gain supports our hypothesis that multiple complementary metrics can produce more accurate models.
 
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+  \jhc{Add and reference supplementary figure with naive model results to compare with these.}
   }
@@
 Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
-These results confirm that our approach of minimizing the distance between yearly populations can simultaneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
+These results confirm that our approach of minimizing the distance between yearly populations can simultaneously capture clade-level dynamics of simulated influenza populations and identify individual strains that are most representative of future populations.
 
@@
 As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI antigenic novelty would be individually more predictive than mutational load or DMS mutational effects.
-Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
+Previous research \cite{Neher:2014eu} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
 As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA antigenic novelty models in parallel with the HI antigenic novelty models.
@@
 The average improvement of the sequence-only models over the naive model was consistently lower than the same models in simulated populations.
-This reduced performance may have been caused by both the relatively reduced diversity between years in natural populations and the increased complexity of the fitness constraints on these real populations.
+This reduced performance may have been caused by both the relatively reduced diversity between years in natural populations and the fact that our simple models do not capture all drivers of evolution in natural H3N2 populations.
 
@@
 
-LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
-Delta frequency performed worse than LBI and HI antigenic novelty and only performed slightly better than mutational load.
+LBI was the best individual metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
+Delta frequency performed worse than LBI and HI antigenic novelty and was comparable to mutational load.
 While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results show that the current implementations of these metrics represent qualitatively different fitness components.
+The LBI and mutational load might also be predictive for reasons other than correlation with fitness, see Discussion.
 
@@
 In contrast, the three-metric model strongly preferred the HI antigenic novelty and mutational load metrics over LBI for the entire validation period, producing an average LBI coefficient of -0.04 $\pm$ 0.09.
-These results suggest that HI assays and LBI may measure similar fitness effects.
+Overall, the gain by combining multiple predictors was limited and the sensitivity of coefficients to the set of metrics included in the model suggests that there is substantial overlap in predictive value of different metrics.
 
@@
 We observed overfitting to historical context in sequence-based models of antigenic drift.
-The fitness benefit of mutations that lead to antigenic drift in H3N2 is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}
+The fitness benefit of mutations that led to antigenic drift in H3N2 in the past is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}.
 Although the antigenic importance of seven specific sites in HA were experimentally validated by Koel et al.
@@
 
-\subsection*{Conclusions}
+\subsection*{The nature of the predictive power of individual metrics remains unclear}
 
@@
 Across the validation and test periods, we found that antigenic drift and mutational load were the most robust predictors of future success for seasonal influenza H3N2 populations.
+
+Several metrics like the rate of frequency change or epitope mutations are naively expected to have predictive power but do not.
+Others like the mutational load are not expected to pick up adaptation, but are predictive.
+This points to one aspect that often overlooked when comparing the genetic make-up of an asexual population at two time points:
+The future population is unlikely to descend from any of the sampled tips but ancestral lineages of the future population merge with those of the present population in the past.
+Optimal representatives of the future therefore tend to be tips in the present that tend to be basal and less evolved.
+The LBI and the mutational load metric have the tendency to assign low fitness to evolved tips.
+The LBI in particular assigns high fitness to the base of large clades.
+Much the predictive power, in the sense of a reduced distance between the predicted and observed populations, might be due to putting more weight on less evolved strains rather than \emph{bona fide} prediction of fitness.
+In a companion manuscript, Barrat-Charlaix et al.~show that LBI has little predictive power for fixation probabilities of mutations in H3N2.
 

### 2020-05-19 manuscript/main.tex
Use simpler LaTeX figure and table placement directives
Let LaTeX do its best to place figures instead of forcing placement.
Copy edits to the new conclusions paragraph

@@
 
-\begin{figure}[ht]
+\begin{figure}[htb]
   \begin{center}
@@
 
-\begin{figure}[ht!]
+\begin{figure}[htb]
   \begin{center}
@@
 
-\begin{table}[ht]
+\begin{table}[htb]
   \begin{center}
@@
 Several metrics like the rate of frequency change or epitope mutations are naively expected to have predictive power but do not.
-Others like the mutational load are not expected to pick up adaptation, but are predictive.
-This points to one aspect that often overlooked when comparing the genetic make-up of an asexual population at two time points:
+Others metrics like the mutational load are not expected to measure adaptation but are predictive.
+These results point to one aspect that often overlooked when comparing the genetic make-up of an asexual population at two time points:
 The future population is unlikely to descend from any of the sampled tips but ancestral lineages of the future population merge with those of the present population in the past.
@@
 The LBI in particular assigns high fitness to the base of large clades.
-Much the predictive power, in the sense of a reduced distance between the predicted and observed populations, might be due to putting more weight on less evolved strains rather than \emph{bona fide} prediction of fitness.
+Much of the predictive power, in the sense of a reduced distance between the predicted and observed populations, might be due to putting more weight on less evolved strains rather than \emph{bona fide} prediction of fitness.
 In a companion manuscript, Barrat-Charlaix et al.~show that LBI has little predictive power for fixation probabilities of mutations in H3N2.

### 2020-05-22 manuscript/main.tex
Add bootstrap analysis to text and tables
Updates Tables 1 and 2 and Supplemental Table S3 and their captions to include
indicators for models that were significantly closer to the future than the
corresponding populations' naive model. Adds a description of the bootstrap
hypothesis tests to the methods and adds figures displaying actual p-values and
corresponding bootstraps of different models to the supplementary material.

@@
       Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap hypothesis tests (see Methods).
       The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
@@
       Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap hypothesis tests (see Methods).
       Validation results are based on 23 timepoints.
@@
 
+\subsubsection*{Model comparison by bootstrap hypothesis testing}
+
+We compared the performance of each biologically-informed model to the corresponding population's naive model using bootstrap hypothesis tests \cite{EfronBradley1998}.
+For each model's empirical distribution of earth mover's distances to the future, $\mathbf{y}$, and the corresponding naive model's distribution, $\mathbf{x}$, we applied the following procedure to test whether their means were significantly different.
+We calculated the test statistic $t$ as follows where $n$ and $m$ were the respective lengths of the naive and biologically-informed model's distributions.
+
+$$
+t = \frac{\bar{x}-\bar{y}}{\sqrt{\sigma_x^2/n + \sigma_y^2/m}}
+$$
+
+Next, we combined the two empirical distributions of distances into a single distribution, $\mathbf{z}$, and calculated the mean of this distribution.
+For each original distribution, we created a new distribution that was recentered on the mean of $\mathbf{z}$.
+For example, we recentered the $\mathbf{x}$ distribution as $\mathbf{x^{'}} = \mathbf{x} - \bar{x} + \bar{z}$.
+We drew $n$ and $m$ samples, respectively, from the resulting distributions, $\mathbf{x^{'}}$ and $\mathbf{y^{'}}$, with replacement for 10,000 bootstrap iterations.
+For each iteration, we recalculated the test statistic $t^{'}$ for the means and variances of the bootstrap samples.
+We calculated the p-value as the proportion of bootstrap samples for which the $t^{'}$ statistic was greater than or equal to the observed $t$ for the empirical distributions.
+To assist with the interpretation of these p-values, we further bootstrapped each model's empirical distance distribution 10,000 times, calculated the mean of each bootstrap sample, and visualized the distributions of each model annotated by its p-value from the test above (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+
 \subsection*{Fitness metrics}

### 2020-05-22 manuscript/supplement.tex
Add bootstrap analysis to text and tables
Updates Tables 1 and 2 and Supplemental Table S3 and their captions to include
indicators for models that were significantly closer to the future than the
corresponding populations' naive model. Adds a description of the bootstrap
hypothesis tests to the methods and adds figures displaying actual p-values and
corresponding bootstraps of different models to the supplementary material.

@@
   \begin{center}
+  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_simulated_sample_3.pdf}
+  \caption{
+  Bootstrap distributions of the mean distance to the future for simulated populations.
+  Empirical distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
+  The horizontal gray line indicates the median of the bootstrap distribution for the naive model.
+  Each model is annotated by the p-value from a bootstrap hypothesis test between the mean distances for that model and the naive model (see Methods).
+  }
+  \label{sup_fig:bootstrap_distributions_for_simulated_sample_3}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
   \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-simulated-populations.pdf}
@@
   \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding.pdf}
+  \caption{
+  Bootstrap distributions of the mean distance to the future for natural populations.
+  Empirical distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
+  The horizontal gray line indicates the median of the bootstrap distribution for the naive model.
+  Each model is annotated by the p-value from a bootstrap hypothesis test between the mean distances for that model and the naive model (see Methods).
+  }
+  \label{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}
   \end{center}
@@
       All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap hypothesis tests (see Methods).
       Validation results are based on 23 timepoints.

### 2020-05-23 src/fit_model.py
Calculate lower bound for EMD between validation and test timepoints
When calculating the distance to the future for the naive model, also calculated
the lowest bound for the EMD between each timepoint and the future. We
calculate the lower bound using a greedy optimization strategy that maps each
strain in the future to its closest strain by Hamming distance in the current
timepoint and assigns each future strain's frequency to its corresponding
current strain's estimated frequencies. Then we calculate the EMD between these
new estimated frequencies for the current timepoint strains and the o

@@ -398,7 +398,7 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
         super().__init__(predictors, delta_time, l1_lambda, cost_function)
         self.distances = distances
 
-    def _fit(self, coefficients, X, y, use_l1_penalty=True):
+    def _fit(self, coefficients, X, y, use_l1_penalty=True, calculate_optimal_distance=False):
         """Calculate the error between observed and estimated values for the given
         parameters and data.
 
@@ -443,6 +443,35 @@ class DistanceExponentialGrowthModel(ExponentialGrowthModel):
                 samples_b,
                 self.distances
             ).astype(np.float32)
+
+            # Calculate the optimal distance to the future timepoint by mapping
+            # the frequency of each future strain to the closest strain in the
+            # current timepoint.
+            if calculate_optimal_distance:
+                # For each strain in the future timepoint, identify the closest
+                # strain in the current timepoint. This is an array of current
+                # strain indices (one index per future strain).
+                closest_strain_to_future = np.argmin(distance_matrix, axis=0)
+
+                # Sum the frequencies of the future strains across each closest
+                # strain in the current timepoint. This can and will often
+                # result in a few current strains accuring most of the future
+                # frequencies.
+                estimated_frequencies = np.zeros_like(sample_a_frequencies)
+                for i in range(sample_b_frequencies.shape[0]):
+                    estimated_frequencies[closest_strain_to_future[i]] += sample_b_frequencies[i]
+
+                # Calculate earth mover's distance to the future based on this
+                # optimal (or, at least, greedy) mapping of strains between
+                # timepoints. The resulting EMD value should be the best any
+                # model can hope to perform and establishes a lower bound for
+                # all models.
+                self.optimal_model_emd, _, optimal_model_flow = cv2.EMD(
+                    estimated_frequencies,
+                    sample_b_frequencies,
+                    cv2.DIST_USER,
+                    cost=distance_matrix
+                )
 
             # Estimate the distance between the model's estimated future and the
             # observed future populations.
@@ -699,16 +728,18 @@ def cross_validate(model_class, model_kwargs, data, targets, train_validate_time
 
         # Calculate the model score for the validation data.
[…]

### 2020-06-03 manuscript/main.tex
Update caption and main text for vaccine strain comparisons
Updates text to reflect inclusion of two models in the vaccine comparison figure
and summarize performance of both models across all timepoints.

Closes #70.

@@
   \caption{
-    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated (orange) closest strains to the future at the corresponding timepoints.
+    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated closest strains to the future by the mutational load and LBI model (orange) and the HI antigenic novelty and mutational load model (purple).
     Vaccine strains were assigned to the validation or test timepoint closest to the date they were selected by the WHO.
-    The distance to the future of each vaccine strain was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
-    The estimated closest strain to the future was identified by either the best model in the validation period (mutational load and LBI) or the best model in the test period (HI antigenic novelty and mutational load).
+    The weighted distance to the future for each strain was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
   }
@@
 
-We further evaluated our best model's apparent success at estimating the closest strain to the next season's H3N2 population by comparing our model's selection to the WHO's vaccine strain selection.
-For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed and estimated closest strains.
-Our models selected a strain that was closer to the future than the selected vaccine strain for 11 (92\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
-In the one season where the vaccine strain was closer to the future than the model's selection, the difference between the two strains' distances to the future was 0.19 AAs.
+We further evaluated our models' ability to estimate the closest strain to the next season's H3N2 population by comparing our best models' selections to the WHO's vaccine strain selection.
+For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed closest strains.
+We compared these distances to those of the closest strains to the future as estimated by our best models for the validation period (mutational load and LBI) and the test period (HI antigenic novelty and mutational load).
+The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
+For the two seasons that the model selected more distant strains than the vaccine strain, the mean distance relative to the vaccine strain was 1.58 AAs.
+The HI antigenic novelty and mutational load model performed similarly by identifying strains as close or closer to the future for 11 (92\%) seasons.
+For the one season that the model selected a more distant strain, that selected strain was 0.75 AAs farther from the future than the vaccine strain.
 

### 2020-06-04 scripts/concatenate_tables.py
Explicitly ask pandas to sort columns during concatenation

@@ -16,6 +16,6 @@ if __name__ == "__main__":
     df = pd.concat([
         pd.read_csv(table_file, sep=args.separator)
         for table_file in args.tables
-    ], ignore_index=True)
+    ], ignore_index=True, sort=True)
 
     df.to_csv(args.output, sep=args.separator, header=True, index=False)

### 2020-06-05 manuscript/main.tex
Update bootstrap methods and results
Adds latest bootstrap p values, updates tables in main and supplement, and
updates the methods section for model comparison.

@@
       Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap hypothesis tests (see Methods).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3}).
       The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
@@
       Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap hypothesis tests (see Methods).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
       Validation results are based on 23 timepoints.
@@
 
-\subsubsection*{Model comparison by bootstrap hypothesis testing}
+\subsubsection*{Model comparison by bootstrap tests}
 
-We compared the performance of each biologically-informed model to the corresponding population's naive model using bootstrap hypothesis tests \cite{EfronBradley1998}.
-For each model's empirical distribution of earth mover's distances to the future, $\mathbf{y}$, and the corresponding naive model's distribution, $\mathbf{x}$, we applied the following procedure to test whether their means were significantly different.
-We calculated the test statistic $t$ as follows where $n$ and $m$ were the respective lengths of the naive and biologically-informed model's distributions.
-
-$$
-t = \frac{\bar{x}-\bar{y}}{\sqrt{\sigma_x^2/n + \sigma_y^2/m}}
-$$
-
-Next, we combined the two empirical distributions of distances into a single distribution, $\mathbf{z}$, and calculated the mean of this distribution.
-For each original distribution, we created a new distribution that was recentered on the mean of $\mathbf{z}$.
-For example, we recentered the $\mathbf{x}$ distribution as $\mathbf{x^{'}} = \mathbf{x} - \bar{x} + \bar{z}$.
-We drew $n$ and $m$ samples, respectively, from the resulting distributions, $\mathbf{x^{'}}$ and $\mathbf{y^{'}}$, with replacement for 10,000 bootstrap iterations.
-For each iteration, we recalculated the test statistic $t^{'}$ for the means and variances of the bootstrap samples.
-We calculated the p-value as the proportion of bootstrap samples for which the $t^{'}$ statistic was greater than or equal to the observed $t$ for the empirical distributions.
-To assist with the interpretation of these p-values, we further bootstrapped each model's empirical distance distribution 10,000 times, calculated the mean of each bootstrap sample, and visualized the distributions of each model annotated by its p-value from the test above (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+We compared the performance of each biologically-informed model to the corresponding population's naive model using bootstrap tests.
+For each model's empirical distribution of earth mover's distances to the future at a specific timepoint $t$, $y(t)$, and the corresponding naive model's distribution, $x(t)$, we calculated the difference between the two models as $y(t) - x(t)$ such that values less than zero represented timepoints when the given model outperformed the naive model.
+To determine whether each model generally estimated populations that were closer to the future than the naive model, we bootstrapped the empirical difference distributions for n=10,000 samples and calculated the mean difference of each bootstrap sample (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+We calculated an empirical p value for each model as the proportion of bootstrap samples with mean values less than zero.
+We identified models as significant if their empirical p value was less than 0.05.
 

### 2020-06-05 manuscript/supplement.tex
Update bootstrap methods and results
Adds latest bootstrap p values, updates tables in main and supplement, and
updates the methods section for model comparison.

--- 2020-01-17 manuscript/supplement.tex
@@
   \begin{center}
+  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_simulated_sample_3.pdf}
+  \caption{
+  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for simulated populations.
+  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
+  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
+  Each model is annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
+  }
+  \label{sup_fig:bootstrap_distributions_for_simulated_sample_3}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
   \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-simulated-populations.pdf}
@@
   \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding.pdf}
+  \caption{
+  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for natural populations.
+  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
+  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
+  Each model is annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
+  }
+  \label{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}
   \end{center}
@@
       All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
       Validation results are based on 23 timepoints.

### 2020-06-05 manuscript/main.tex
Describe naive model and lower bound in first results section
Moves the description of the naive model from the simulated populations section
up to the first results section, adds a description of the lower bound for EMD,
and generally clarifies the model fit and evaluation text in the first results
section.

@@
 
-We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and selected the best of these models using time-series cross-validation.
-After selecting optimal models from training and validation, we evaluated the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
-Importantly, our models find fitness coefficients that minimize the earth mover's distance (EMD) between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
-With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
-However, we retain the benefits of fitting models to highly similar strains found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.
+We fit models for individual fitness metrics and combinations of metrics that we anticipated would be mutually beneficial.
+For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
+We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Supplemental Figs~\ref{sup_fig:cross_validation_for_simulated_populations} and \ref{sup_fig:cross_validation_for_natural_populations}).
+The earth mover's distance to the future can never be zero, because each model makes predictions based on sequences available at the time of prediction and cannot account for new mutations that occur during the prediction interval.
+We calculated the lower bound for each model's performance as the optimal distance to the future possible given the current sequences at each timepoint.
+As an additional reference, we evaluated the performance of a ``naive'' model that predicted the future population would be identical to the current population.
+We expected that the best models would consistently outperform the naive model and perform as close as possible to the lower bound.
 
@@
 \ref{sup_fig:simulated_h3n2_ha_phylogeny}, Supplemental Tables~\ref{sup_table:mutations_by_trunk_status_for_simulated_populations} and \ref{sup_table:mutations_by_trunk_status}).
-
 We fit models to these simulated populations using all sequence-only fitness metrics.
 As a positive control for our model framework, we also fit a model based on the true fitness of each strain as measured by the simulator.
-We evaluated the performance of each model by the absolute distance between projected and observed future populations.
-As each model only projects existing sequences forward one year without trying to predict which mutations occur in the future, the model's absolute distance to the future can never be zero.
-For this reason, we additionally calculated how often each model outperformed a naive model that estimated no change in populations between years and, therefore, measured the observed distance between annual populations.
 

### 2020-06-05 manuscript/main.tex
Readability edits

@@
 As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus strain an exponential growth factor scaled by an estimate of the strain's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
-We estimated the frequency of virus strains every six months using kernel density estimation (KDE) to represent the frequency of each strain.
+We estimated the frequency of virus strains every six months using kernel density estimation (KDE).
 
-We estimated viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope antigenic novelty and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) antigenic novelty \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
+We estimated viral fitness with biologically-informed metrics including those originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj} of epitope antigenic novelty and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) antigenic novelty \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
 All of these metrics except for HI antigenic novelty and DMS mutational effects rely only on HA sequences.
@@
 Increased antigenic drift relative to previously circulating strains is expected to correspond to increased viral fitness.
-Mutational load is the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each strain since their ancestor in the previous season.
+Mutational load estimates functional constraint by measuring the number of putatively deleterious mutations that have accumulated in each strain since their ancestor in the previous season.
 DMS mutational effects provide a more comprehensive biophysical model of functional constraint by measuring the beneficial or deleterious effect of each possible single amino acid mutation in HA from the background of a previous vaccine strain, A/Perth/16/2009.

### 2020-06-05 manuscript/main.tex
Add description of lower bound for EMD
Adds methods and results for the lower bound on EMD between populations
including updated model results figures showing the lower bound in dark gray and
the naive model now in light gray.

@@
     Corresponding values per test timepoint are in the top-right.
-    For reference, the naive model's distance to the future is also shown per validation and test timepoint in gray.
-    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 9.07 $\pm$ 1.71 AAs for test timepoints.
+    The naive model's distances to the future for validation and test timepoints (light gray) were 8.97 $\pm$ 1.35 AAs and 9.07 $\pm$ 1.71 AAs, respectively.
+    The corresponding lower bounds on the estimated distance to the future (dark gray) were 4.57 $\pm$ 0.61 AAs and 4.85 $\pm$ 0.82 AAs.
   }
@@
 As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 (97\%) of 33 timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
-In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future.
+Although the true fitness model performed better than the naive model's average distance of 8.97 $\pm$ 1.35 AAs, it did not reach the closest possible distance between populations of 4.57 $\pm$ 0.61 AAs.
 With the exception of epitope antigenic novelty, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
@@
     A) Coefficients and B) distances are shown per validation and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 8.90 $\pm$ 1.69 AAs for test timepoints.
   }
@@
     A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-    The naive model's distance to the future was 6.40 $\pm$ 1.36 AAs for validation timepoints and 6.82 $\pm$ 1.74 AAs for test timepoints.
+    The naive model's distance to the future (light gray) was 6.40 $\pm$ 1.36 AAs for validation timepoints and 6.82 $\pm$ 1.74 AAs for test timepoints.
+    The corresponding lower bounds on the estimated distance to the future (dark gray) were 2.60 $\pm$ 0.89 AAs and 2.28 $\pm$ 0.61 AAs.
   }
@@
 This distance corresponds to the ``naive model'' in Table~\ref{table_natural_model_selection}.
+The lower bound for all models was 2.60 $\pm$ 0.89 AAs.
 Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope antigenic novelty and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
@@
 
+\subsubsection*{Lower bound on earth mover's distance}
+
+The minimum distance to the future between any two timepoints cannot be zero due to the accumulation of mutations between populations.
+We estimated the lower bound on earth mover's distance between timepoints using the following greedy solution to the optimal transport problem.
+For each timepoint $t$, we initialized the optimal frequency of each current strain to zero.
+For each strain in the future timepoint $u$, we identified the closest strain in the current timepoint by Hamming distance and added the frequency of the future strain to the optimal frequency of the corresponding current strain.
+This approach allows each strain from timepoint $t$ to accumulate frequencies from multiple strains at timepoint $u$.
+We calculated the minimum distance between populations as the earth mover's distance between the resulting optimal frequencies for current strains, the observed frequencies of future strains, and the original distance matrix between those two populations.
+
 \subsubsection*{Strain-specific distance to the future}

### 2020-06-05 manuscript/main.tex
Fixed sentence fragment

@@
 Others metrics like the mutational load are not expected to measure adaptation but are predictive.
-These results point to one aspect that often overlooked when comparing the genetic make-up of an asexual population at two time points:
-The future population is unlikely to descend from any of the sampled tips but ancestral lineages of the future population merge with those of the present population in the past.
+These results point to one aspect that often overlooked when comparing the genetic make-up of an asexual population at two time points: the future population is unlikely to descend from any of the sampled tips but ancestral lineages of the future population merge with those of the present population in the past.
 Optimal representatives of the future therefore tend to be tips in the present that tend to be basal and less evolved.

### 2020-06-05 manuscript/main.tex
Improve readability of figure/table captions and bootstrap methods

@@
   \caption{
-    Model coefficients and distances between projected and observed future populations as measured in amino acids (AAs) for simulated populations.
+    Simulated population model coefficients and distances between projected and observed future populations as measured in amino acids (AAs).
     A) Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
@@
     \caption{
-      Model coefficients and performance on validation and test data for simulated populations ordered from best to worst by distance to the future in the validation analysis.
+      Simulated population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis.
       Coefficients are the mean $\pm$ standard deviation for each metric in a given model across 33 training windows.
@@
   \caption{
-    Model coefficients and distances to the future for individual biologically-informed fitness metrics and the best composite model fit to simulated populations.
+    Simulated population model coefficients and distances to the future for individual biologically-informed fitness metrics and the best composite model.
     A) Coefficients and B) distances are shown per validation and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
@@
     \caption{
-      Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
+      Natural population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
       Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
@@
   \caption{
-    Model coefficients and distances to the future for individual biologically-informed fitness metrics fit to natural populations.
+    Natural population model coefficients and distances to the future for individual biologically-informed fitness metrics.
     A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
@@
   \caption{
-    Model coefficients and distances to the future for composite fitness metrics fit to natural populations.
+    Natural population model coefficients and distances to the future for composite fitness metrics.
     A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
@@
 We compared the performance of each biologically-informed model to the corresponding population's naive model using bootstrap tests.
-For each model's empirical distribution of earth mover's distances to the future at a specific timepoint $t$, $y(t)$, and the corresponding naive model's distribution, $x(t)$, we calculated the difference between the two models as $y(t) - x(t)$ such that values less than zero represented timepoints when the given model outperformed the naive model.
+First, we calculated an empirical distribution of differences between each model's earth mover's distance to the future at a given timepoint and the corresponding naive model's distance at the same timepoint.
+Values less than zero in the resulting distribution represented timepoints when the given model outperformed the naive model.
 To determine whether each model generally estimated populations that were closer to the future than the naive model, we bootstrapped the empirical difference distributions for n=10,000 samples and calculated the mean difference of each bootstrap sample (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).

### 2020-06-06 manuscript/main.tex
Add bootstrap comparisons between composite and individual models

@@
 To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
-This composite model outperformed all individual metrics with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
+This composite model outperformed both of its corresponding individual metric models with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}, Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
 The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
@@
 Finally, we tested all of our models on out-of-sample data.
-Specifically, we fixed the coefficients of each model to the average values across the training/validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
-A standard expectation from machine learning is that models will perform worse on test data than on validation data.
-Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
-The composite model of mutational load and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively.
-Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.
-Note, that the numbers of test and validation data points are small and we therefore expect fluctuations in model performance.
+Specifically, we fixed the coefficients of each model to the average values across the validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
+A standard expectation from machine learning is that models will perform worse on test data than on validation data due to overfitting to training data.
+Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
+The composite model of mutational load and LBI appeared to outperform the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively.
+However, we did not find a significant difference between these models by bootstrap testing (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}) and could not rule out fluctuations in model performance across a relatively small number of data points.
 
@@
 
-Across our training period, the distance between HA sequences of H3N2 influenza populations one year apart was 6.40 $\pm$ 1.36 AAs.
-This distance corresponds to the ``naive model'' in Table~\ref{table_natural_model_selection}.
-The lower bound for all models was 2.60 $\pm$ 0.89 AAs.
-Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope antigenic novelty and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope antigenic novelty and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations} and Table~\ref{table_natural_model_selection}).
+The naive model estimated an average distance between natural H3N2 populations of 6.40 $\pm$ 1.36 AAs.
+The lower bound for how well any model could perform, 2.60 $\pm$ 0.89 AAs, was considerably lower than the corresponding bounds for simulated populations.
 The average improvement of the sequence-only models over the naive model was consistently lower than the same models in simulated populations.
@@
 
-Both two-metric composite models performed better than their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
+Both two-metric composite models modestly outperformed their corresponding individual models (Table~\ref{table_natural_model_selection}, Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}, and Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
 The composite of mutational load and LBI performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
-The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
+The relative stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
 In contrast, the three-metric model strongly preferred the HI antigenic novelty and mutational load metrics over LBI for the entire validation period, producing an average LBI coefficient of -0.04 $\pm$ 0.09.
@@
 We anticipated that most models would perform worse on truly out-of-sample data than on validation data.
-As expected, all models estimated higher average distances to the future across these eight test timepoints (Table~\ref{table_natural_model_selection}).
-However, these increases were mostly consistent with an increased average distance between years of 6.82 $\pm$ 1.74 AAs as measured by the naive model.
-The composite of HI antigenic novelty and mutational load outperformed all other models on the test data with an average distance of 5.97 $\pm$ 1.47 AAs.
+Correspondingly, only the three models with the HI antigenic novelty metric significantly outperformed the naive model on the test data (Table~\ref{table_natural_model_selection}).
+The composite of HI antigenic novelty and mutational load performed modestly, although not significantly, better than the individual HI antigenic novelty model (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
 Surprisingly, the best model for the validation data -- mutational load and LBI -- was one of the worst models for the test data with an average distance to the future of 7.70 $\pm$ 3.53 AAs.
@@
 
-We compared the performance of each biologically-informed model to the corresponding population's naive model using bootstrap tests.
-First, we calculated an empirical distribution of differences between each model's earth mover's distance to the future at a given timepoint and the corresponding naive model's distance at the same timepoint.
-Values less than zero in the resulting distribution represented timepoints when the given model outperformed the naive model.
-To determine whether each model generally estimated populations that were closer to the future than the naive model, we bootstrapped the empirical difference distributions for n=10,000 samples and calculated the mean difference of each bootstrap sample (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
-We calculated an empirical p value for each model as the proportion of bootstrap samples with mean values less than zero.
-We identified models as significant if their empirical p value was less than 0.05.
+We compared the performance of different pairs of models using bootstrap tests.
+For each timepoint, we calculated the difference between one model's earth mover's distance to the future and the other model's distance.
+The resulting empirical distribution values less than zero when the first model outperformed the second model.
+To determine whether the first model generally outperformed the second model, we bootstrapped the empirical difference distributions for n=10,000 samples and calculated the mean difference of each bootstrap sample.
+We calculated an empirical p value for the first model as the proportion of bootstrap samples with mean values greater than or equal to zero.
+We performed pairwise model comparisons for all biologically-informed models against the naive model (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+We also compared a subset of composite models to their respective individual models (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
 

### 2020-06-06 manuscript/supplement.tex
Add bootstrap comparisons between composite and individual models

@@
 
+\begin{table}[H]
+  \begin{center}
+    \scalebox{0.8}{
+        \input{tables/composite_vs_individual_model_comparison.tex}
+    }
+    \caption{
+      Comparison of composite and individual model distances to the future by bootstrap test (see Methods).
+      Empirical differences represent the mean difference between the composite and individual models' distances to the future at each timepoint.
+      The p values represent the proportion of n=10,000 bootstrap samples where the mean difference was greater than or equal to zero.
+    }
+    \label{sup_table:composite_vs_individual_model_comparison}
+  \end{center}
+\end{table}
+
 \pagebreak

### 2020-06-06 manuscript/tables/composite_vs_individual_model_comparison.tex
Add bootstrap comparisons between composite and individual models

\begin{tabular}{llllrr}
\toprule
    sample &  error\_type &      individual\_model &                         composite\_model &  empirical\_difference &  p\_value \\
\midrule
 simulated &  validation &          true fitness &                  mutational load +  LBI &                  0.42 &   0.9678 \\
 simulated &  validation &       mutational load &                  mutational load +  LBI &                 -1.03 &   0.0000 \\
 simulated &  validation &                   LBI &                  mutational load +  LBI &                 -0.33 &   0.0096 \\
 simulated &        test &          true fitness &                  mutational load +  LBI &                 -0.28 &   0.1421 \\
 simulated &        test &       mutational load &                  mutational load +  LBI &                 -1.11 &   0.0000 \\
 simulated &        test &                   LBI &                  mutational load +  LBI &                 -0.41 &   0.0002 \\
   natural &  validation &       mutational load &                  mutational load +  LBI &                 -0.69 &   0.0046 \\
   natural &  validation &                   LBI &                  mutational load +  LBI &                 -0.23 &   0.0027 \\
   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &                 -0.31 &   0.0410 \\
   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.18 &   0.0559 \\
   natural &        test &       mutational load &                  mutational load +  LBI &                  1.17 &   0.9400 \\
   natural &        test &                   LBI &                  mutational load +  LBI &                 -0.70 &   0.0000 \\
   natural &        test &       mutational load &  mutational load + HI antigenic novelty &                 -0.56 &   0.0135 \\
   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.24 &   0.0948 \\
\bottomrule
\end{tabular}

### 2020-06-06 manuscript/main.tex
Clarify bootstrap p value results
Set the y-axis range of the bootstrap distribution figures to the same values
for both validation and test data to more clearly represent the higher variance
of the test data points for natural populations. Also update the methods text
for bootstrap tests to explicitly state the meaning of the p value.

@@
 For each timepoint, we calculated the difference between one model's earth mover's distance to the future and the other model's distance.
-The resulting empirical distribution values less than zero when the first model outperformed the second model.
+Values less than zero in the resulting empirical distribution represent when the first model outperformed the second model.
 To determine whether the first model generally outperformed the second model, we bootstrapped the empirical difference distributions for n=10,000 samples and calculated the mean difference of each bootstrap sample.
 We calculated an empirical p value for the first model as the proportion of bootstrap samples with mean values greater than or equal to zero.
+This p value represents how likely the mean difference between the models' distances to the future is to be zero or greater.
 We performed pairwise model comparisons for all biologically-informed models against the naive model (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).

### 2020-06-06 manuscript/tables/composite_vs_individual_model_comparison.tex
Clarify bootstrap p value results
Set the y-axis range of the bootstrap distribution figures to the same values
for both validation and test data to more clearly represent the higher variance
of the test data points for natural populations. Also update the methods text
for bootstrap tests to explicitly state the meaning of the p value.

@@
 \midrule
- simulated &  validation &          true fitness &                  mutational load +  LBI &                  0.42 &   0.9678 \\
+ simulated &  validation &          true fitness &                  mutational load +  LBI &                  0.42 &   0.9672 \\
  simulated &  validation &       mutational load &                  mutational load +  LBI &                 -1.03 &   0.0000 \\
- simulated &  validation &                   LBI &                  mutational load +  LBI &                 -0.33 &   0.0096 \\
- simulated &        test &          true fitness &                  mutational load +  LBI &                 -0.28 &   0.1421 \\
+ simulated &  validation &                   LBI &                  mutational load +  LBI &                 -0.33 &   0.0097 \\
+ simulated &        test &          true fitness &                  mutational load +  LBI &                 -0.28 &   0.1395 \\
  simulated &        test &       mutational load &                  mutational load +  LBI &                 -1.11 &   0.0000 \\
- simulated &        test &                   LBI &                  mutational load +  LBI &                 -0.41 &   0.0002 \\
-   natural &  validation &       mutational load &                  mutational load +  LBI &                 -0.69 &   0.0046 \\
-   natural &  validation &                   LBI &                  mutational load +  LBI &                 -0.23 &   0.0027 \\
-   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &                 -0.31 &   0.0410 \\
-   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.18 &   0.0559 \\
-   natural &        test &       mutational load &                  mutational load +  LBI &                  1.17 &   0.9400 \\
+ simulated &        test &                   LBI &                  mutational load +  LBI &                 -0.41 &   0.0001 \\
+   natural &  validation &       mutational load &                  mutational load +  LBI &                 -0.69 &   0.0034 \\
+   natural &  validation &                   LBI &                  mutational load +  LBI &                 -0.23 &   0.0026 \\
+   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &                 -0.31 &   0.0420 \\
+   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.18 &   0.0560 \\
+   natural &        test &       mutational load &                  mutational load +  LBI &                  1.17 &   0.9379 \\
    natural &        test &                   LBI &                  mutational load +  LBI &                 -0.70 &   0.0000 \\
-   natural &        test &       mutational load &  mutational load + HI antigenic novelty &                 -0.56 &   0.0135 \\
-   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.24 &   0.0948 \\
+   natural &        test &       mutational load &  mutational load + HI antigenic novelty &                 -0.56 &   0.0113 \\
+   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.24 &   0.0962 \\
 \bottomrule

### 2020-06-06 manuscript/main.tex
Update forecast snapshots to make forecasts from latest data
As of June 6, the latest H3N2 data were collected up to the beginning of
April. This commit uses forecasts made from April 1, 2020 to April 1, 2021 for a
more realistic comparison with how forecasts are made elsewhere in the
manuscript (i.e., from timepoints where all data are available).

@@
 
-After confirming the coefficients for our best model of HI antigenic novelty and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021 using data available up through January 17, 2020.
-Of the three major currently circulating clades, our model predicted growth of the clades A1b/131K and 3c3.A and decline of all other clades (Fig.~\ref{fig:nextstrain_forecasts}).
-Four months after those forecasts, the subclades of A1b/131K represented 40\% of the global H3N2 population an increase from an estimated 35\% frequency in January.
-The 3c3.A clade increased slightly in global frequency from 13\% to 15\%, in that time.
+After confirming the coefficients for our best model of HI antigenic novelty and mutational load, we inspected forecasts of H3N2 clades using all data available up through June 6, 2020.
+Consistent with an average two-month lag between data collection and submission, the most recent data were collected up to April 1, 2020 and made our forecasts from this timepoint to April 1, 2021.
+Of the five major currently circulating clades, our model predicted growth of the clades 3c3.A and A1b/94N and decline of clades A1b/135K, A1b/137F, and A1b/197R (Fig.~\ref{fig:nextstrain_forecasts}).
 To aid with identification of potential vaccine candidates for the next season, we annotated strains in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
@@
   \caption{
-    Snapshot of live forecasts on nextstrain.org from our best model (HI antigenic novelty and mutational load) for January 2021.
-    The observed frequency trajectories for currently circulating clades are shown up to January 2020.
-    Our model forecasts growth of the clades A1b/131K and 3c3.A and decline of all other major clades.
+    Snapshot of live forecasts on nextstrain.org from our best model (HI antigenic novelty and mutational load) for April 1, 2021.
+    The observed frequency trajectories for currently circulating clades are shown up to April 1, 2020.
+    Our model forecasts growth of the clades 3c3.A and A1b/94N and decline of all other major clades.
   }

### 2020-06-06 manuscript/main.tex
Minor copy edits for readability

@@
 We fit models for individual fitness metrics and combinations of metrics that we anticipated would be mutually beneficial.
-For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
+For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model} and Eq.~\ref{equation_fitness_estimation}).
 We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Supplemental Figs~\ref{sup_fig:cross_validation_for_simulated_populations} and \ref{sup_fig:cross_validation_for_natural_populations}).
@@
 With the exception of epitope antigenic novelty, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
-LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs consistent with the fact that the LBI is a correlate of fitness in models of rapidly adapting populations \cite{Neher:2014eu}.
+LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
+This result is consistent with the fact that the LBI is a correlate of fitness in models of rapidly adapting populations \cite{Neher:2014eu}.
 Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
@@
 Specifically, we fixed the coefficients of each model to the average values across the validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
-A standard expectation from machine learning is that models will perform worse on test data than on validation data due to overfitting to training data.
+A standard expectation from machine learning is that models will perform worse on test data due to overfitting to training data.
 Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).

### 2020-06-06 manuscript/main.tex
Note alternate possibility that LBI overfit to natural training data

@@
 By April 2018, the A2/re clade had risen to a global frequency over 50\% from less than 15\% the previous year, despite an absence of antigenic drift.
-However, by October 2018, this clade had declined in frequency to approximately 30\% and, by October 2019, it had gone extinct.
+By October 2018, this clade had declined in frequency to approximately 30\% and, by October 2019, it had gone extinct.
 That LBI incorrectly predicted the success of this reassortant clade highlights a major limitation of growth-based fitness metrics and a corresponding benefit of more mechanistic metrics that explicitly measure antigenic drift and functional constraint.
+However, we cannot rule out the alternate possibility that the LBI model was overfit to the training data.
 

### 2020-06-06 manuscript/main.tex
Add Duncan to the acknowledgments
Especially since Duncan's comments led to the display of a lower bound for EMD
in model results figures.

@@
 
-We thank Jesse Bloom, Erick Matsen, Bing Brunton, Harmit Malik, Sidney Bell, Alli Black, Lola Arakaki, and members of the Bedford lab for useful advice and discussions.
+We thank Jesse Bloom, Erick Matsen, Bing Brunton, Harmit Malik, Sidney Bell, Allison Black, Lola Arakaki, Duncan Ralph, and members of the Bedford lab for useful advice and discussions.
 JH is a Graduate Research Fellow and is supported by the NIH grant NIAID F31AI140714.

### 2020-06-07 manuscript/main.tex
Reference naive model in validation/test figures
Adds validation/test figures for the naive model to the supplemental material
for simulated and natural populations and adds references to these figures to
the corresponding best model figure captions.

@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-normalized_fitness.pdf}
   \caption{
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  \jhc{Add and reference supplementary figure with naive model results to compare with these.}
+  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_simulated_populations}.
   }
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-natural-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-cTiter_x-ne_star.pdf}
   \caption{
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_natural_populations}.
   }

### 2020-06-07 manuscript/supplement.tex
Reference naive model in validation/test figures
Adds validation/test figures for the naive model to the supplemental material
for simulated and natural populations and adds references to these figures to
the corresponding best model figure captions.

@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-simulated-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-normalized_fitness.pdf}
   \caption{
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}.
   }
   \label{sup_fig:validation_of_best_model_for_simulated_populations}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-naive.pdf}
+  \caption{
+  Validation of naive model for simulated populations of H3N2-like viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
+  }
+  \label{sup_fig:validation_of_naive_model_for_simulated_populations}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-naive.pdf}
+  \caption{
+  Test of naive model for simulated populations of H3N2-like viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
+  }
+  \label{sup_fig:test_of_naive_model_for_simulated_populations}
   \end{center}
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-ne_star-lbi.pdf}
   \caption{
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_natural_populations}.
   }
   \label{sup_fig:validation_of_best_model_for_natural_populations}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-naive.pdf}
+  \caption{
+  Validation of naive model for natural populations of H3N2 viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
+  }
+  \label{sup_fig:validation_of_naive_model_for_natural_populations}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-naive.pdf}
+  \caption{
+  Test of naive model for natural populations of H3N2 viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
+  }
+  \label{sup_fig:test_of_naive_model_for_natural_populations}
   \end{center}

### 2020-06-08 manuscript/supplement.tex
Use "gray" instead of "grey"

@@
   Time-series cross-validation scheme for simulated populations.
-  Models were trained in six-year sliding windows (grey lines) and validated on out-of-sample data from validation timepoints (filled circles).
+  Models were trained in six-year sliding windows (gray lines) and validated on out-of-sample data from validation timepoints (filled circles).
   Validation results from 30 years of data were used to iteratively tune model hyperparameters.
@@
   The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
-  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
@@
   Time-series cross-validation scheme for natural populations.
-  Models were trained in six-year sliding windows (grey lines) and validated on out-of-sample data from validation timepoints (filled circles).
+  Models were trained in six-year sliding windows (gray lines) and validated on out-of-sample data from validation timepoints (filled circles).
   Validation results from 25 years of data were used to iteratively tune model hyperparameters.
@@
   The estimated best strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
-  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.

### 2020-06-09 manuscript/supplement.tex
Report effect sizes for bootstrap analyses
Adds mean +/- standard deviation of bootstrap distributions to violin plots of
distributions to better communicate the effect sizes of model improvements over
the naive.

@@
   The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
-  Each model is annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
+  Each model is annotated by the mean $\pm$ the standard deviation of the bootstrap distribution.
+  Models are also annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
   }

### 2020-06-09 manuscript/tables/composite_vs_individual_model_comparison.tex
Report effect sizes for bootstrap analyses
Adds mean +/- standard deviation of bootstrap distributions to violin plots of
distributions to better communicate the effect sizes of model improvements over
the naive.

@@
 \midrule
- simulated &  validation &          true fitness &                  mutational load +  LBI &                  0.42 &   0.9672 \\
+ simulated &  validation &          true fitness &                  mutational load +  LBI &                  0.42 &   0.9670 \\
  simulated &  validation &       mutational load &                  mutational load +  LBI &                 -1.03 &   0.0000 \\
- simulated &  validation &                   LBI &                  mutational load +  LBI &                 -0.33 &   0.0097 \\
- simulated &        test &          true fitness &                  mutational load +  LBI &                 -0.28 &   0.1395 \\
+ simulated &  validation &                   LBI &                  mutational load +  LBI &                 -0.33 &   0.0098 \\
+ simulated &        test &          true fitness &                  mutational load +  LBI &                 -0.28 &   0.1391 \\
  simulated &        test &       mutational load &                  mutational load +  LBI &                 -1.11 &   0.0000 \\
  simulated &        test &                   LBI &                  mutational load +  LBI &                 -0.41 &   0.0001 \\
-   natural &  validation &       mutational load &                  mutational load +  LBI &                 -0.69 &   0.0034 \\
-   natural &  validation &                   LBI &                  mutational load +  LBI &                 -0.23 &   0.0026 \\
-   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &                 -0.31 &   0.0420 \\
-   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.18 &   0.0560 \\
-   natural &        test &       mutational load &                  mutational load +  LBI &                  1.17 &   0.9379 \\
+   natural &  validation &       mutational load &                  mutational load +  LBI &                 -0.69 &   0.0028 \\
+   natural &  validation &                   LBI &                  mutational load +  LBI &                 -0.23 &   0.0037 \\
+   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &                 -0.31 &   0.0396 \\
+   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.18 &   0.0534 \\
+   natural &        test &       mutational load &                  mutational load +  LBI &                  1.17 &   0.9365 \\
    natural &        test &                   LBI &                  mutational load +  LBI &                 -0.70 &   0.0000 \\
-   natural &        test &       mutational load &  mutational load + HI antigenic novelty &                 -0.56 &   0.0113 \\
-   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.24 &   0.0962 \\
+   natural &        test &       mutational load &  mutational load + HI antigenic novelty &                 -0.56 &   0.0133 \\
+   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.24 &   0.0965 \\
 \bottomrule

### 2020-06-09 manuscript/tables/composite_vs_individual_model_comparison.tex
Report effect sizes from bootstrap distributions in tables, too

@@
-\begin{tabular}{llllrr}
+\begin{tabular}{llllrrl}
 \toprule
-    sample &  error\_type &      individual\_model &                         composite\_model &  empirical\_difference &  p\_value \\
+    sample &  error\_type &      individual\_model &                         composite\_model &  bootstrap\_mean &  bootstrap\_std &    p\_value \\
 \midrule
- simulated &  validation &          true fitness &                  mutational load +  LBI &                  0.42 &   0.9670 \\
- simulated &  validation &       mutational load &                  mutational load +  LBI &                 -1.03 &   0.0000 \\
- simulated &  validation &                   LBI &                  mutational load +  LBI &                 -0.33 &   0.0098 \\
- simulated &        test &          true fitness &                  mutational load +  LBI &                 -0.28 &   0.1391 \\
- simulated &        test &       mutational load &                  mutational load +  LBI &                 -1.11 &   0.0000 \\
- simulated &        test &                   LBI &                  mutational load +  LBI &                 -0.41 &   0.0001 \\
-   natural &  validation &       mutational load &                  mutational load +  LBI &                 -0.69 &   0.0028 \\
-   natural &  validation &                   LBI &                  mutational load +  LBI &                 -0.23 &   0.0037 \\
-   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &                 -0.31 &   0.0396 \\
-   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.18 &   0.0534 \\
-   natural &        test &       mutational load &                  mutational load +  LBI &                  1.17 &   0.9365 \\
-   natural &        test &                   LBI &                  mutational load +  LBI &                 -0.70 &   0.0000 \\
-   natural &        test &       mutational load &  mutational load + HI antigenic novelty &                 -0.56 &   0.0133 \\
-   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &                 -0.24 &   0.0965 \\
+ simulated &  validation &          true fitness &                  mutational load +  LBI &            0.43 &           0.23 &     0.9667 \\
+ simulated &  validation &       mutational load &                  mutational load +  LBI &           -1.03 &           0.21 &  $<$0.0001 \\
+ simulated &  validation &                   LBI &                  mutational load +  LBI &           -0.33 &           0.14 &     0.0096 \\
+ simulated &        test &          true fitness &                  mutational load +  LBI &           -0.28 &           0.26 &     0.1396 \\
+ simulated &        test &       mutational load &                  mutational load +  LBI &           -1.11 &           0.25 &  $<$0.0001 \\
+ simulated &        test &                   LBI &                  mutational load +  LBI &           -0.41 &           0.16 &     0.0001 \\
+   natural &  validation &       mutational load &                  mutational load +  LBI &           -0.69 &           0.27 &     0.0028 \\
+   natural &  validation &                   LBI &                  mutational load +  LBI &           -0.23 &           0.10 &     0.0028 \\
+   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &           -0.31 &           0.18 &     0.0409 \\
+   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &           -0.18 &           0.11 &     0.0586 \\
+   natural &        test &       mutational load &                  mutational load +  LBI &            1.18 &           0.80 &     0.9406 \\
+   natural &        test &                   LBI &                  mutational load +  LBI &           -0.69 &           0.24 &  $<$0.0001 \\
+   natural &        test &       mutational load &  mutational load + HI antigenic novelty &           -0.57 &           0.32 &     0.0123 \\
+   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &           -0.24 &           0.18 &     0.0976 \\
 \bottomrule

### 2020-06-10 manuscript/supplement.tex
Report effect sizes from bootstrap distributions in tables, too
Flag models that were not tested for significance
Adds a caret to models that were not tested for significance compared to the
naive model. These are only models for natural populations. This commit also
adds a missing bootstrap test for the epitope ancestor model of the simulated
populations.

@@
       Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+      Distances annotated with carets ($\wedge$) were not tested for significance relative to the naive model.
       Validation results are based on 23 timepoints.
@@
   \begin{center}
-    \scalebox{0.8}{
+    \scalebox{0.7}{
         \input{tables/composite_vs_individual_model_comparison.tex}
@@
       Comparison of composite and individual model distances to the future by bootstrap test (see Methods).
-      Empirical differences represent the mean difference between the composite and individual models' distances to the future at each timepoint.
+      The effect size of differences between models in amino acids is given by the mean and standard deviation of the bootstrap distributions.
       The p values represent the proportion of n=10,000 bootstrap samples where the mean difference was greater than or equal to zero.

### 2020-06-10 src/weighted_distances.py
Correct usage documentation for weighted distances script

@@ -76,7 +76,7 @@ if __name__ == '__main__':
         formatter_class=argparse.ArgumentDefaultsHelpFormatter
     )
     parser.add_argument("--tip-attributes", required=True, help="a tab-delimited file describing tip attributes at one or more timepoints")
-    parser.add_argument("--distances", required=True, help="tab-delimited output file with pairwise distances between samples")
+    parser.add_argument("--distances", required=True, help="tab-delimited file with pairwise distances between samples")
     parser.add_argument("--delta-months", required=True, type=int, help="number of months to project clade frequencies into the future")
     parser.add_argument("--output", required=True, help="tab-delimited output file with mean and standard deviation used to standardize each predictor")
     args = parser.parse_args()

### 2020-06-11 manuscript/main.tex
Report effect sizes for bootstrap analyses
Adds mean +/- standard deviation of bootstrap distributions to violin plots of
distributions to better communicate the effect sizes of model improvements over
the naive.
Acknowledge funding sources for NIID authors

@@
 This p value represents how likely the mean difference between the models' distances to the future is to be zero or greater.
+We measured the effect size of each comparison as the mean $\pm$ the standard deviation of the bootstrap distributions.
 We performed pairwise model comparisons for all biologically-informed models against the naive model (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
@@
 The work done at the Crick Worldwide Influenza Centre was supported by the Francis Crick Institute receiving core funding from Cancer Research UK (FC001030), the Medical Research Council (FC001030) and the Wellcome Trust (FC001030).
+SF, KN, KN, SW and HH were supported by the Ministry of Health, Labour and Welfare, Japan (10110400).
+SW was supported by the Japan Agency for Medical Research and Development (JPfk0108118).
 The Melbourne WHO Collaborating Centre for Reference and Research on Influenza is supported by the Australian Government Department of Health.

### 2020-06-12 manuscript/main.tex
Correct statements about existing antigenic models

@@
 The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods into an extensible, open source framework that can be used by professional virologists in their vaccine design process \cite{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
-Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, these methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
+Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, published methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
+However, unpublished work by {\L}uksza and L\"assig submitted to the WHO GISRS network incorporates antigenic phenotypes into fitness-based predictions \cite{luksza_personal_communication}.
 Here, we provide the first such open source framework for forecasting the genetic composition of future seasonal influenza populations.

### 2020-06-12 manuscript/main.tex
Fix typo in caption for simulated population naive model distances

@@
     Corresponding values per test timepoint are in the top-right.
-    The naive model's distances to the future for validation and test timepoints (light gray) were 8.97 $\pm$ 1.35 AAs and 9.07 $\pm$ 1.71 AAs, respectively.
+    The naive model's distances to the future for validation and test timepoints (light gray) were 8.97 $\pm$ 1.35 AAs and 9.07 $\pm$ 1.70 AAs, respectively.
     The corresponding lower bounds on the estimated distance to the future (dark gray) were 4.57 $\pm$ 0.61 AAs and 4.85 $\pm$ 0.82 AAs.

### 2020-06-13 manuscript/main.tex
Add citation for unpublished work on antigenic models
Also, removes unhelpful precedence statement.

@@
 Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, published methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
-However, unpublished work by {\L}uksza and L\"assig submitted to the WHO GISRS network incorporates antigenic phenotypes into fitness-based predictions \cite{luksza_personal_communication}.
-Here, we provide the first such open source framework for forecasting the genetic composition of future seasonal influenza populations.
+However, unpublished work by {\L}uksza and L\"assig submitted to the WHO GISRS network incorporates antigenic phenotypes into fitness-based predictions \cite{Morris:2017ea,luksza_personal_communication}.
+Here, we provide an open source framework for forecasting the genetic composition of future seasonal influenza populations using genotypic and phenotypic fitness estimates.
 We apply this framework to HA sequence data shared via the GISAID EpiFlu database \cite{shu2017gisaid} and to HI and FRA titer data shared by WHO GISRS Collaborating Centers in London, Melbourne, Atlanta and Tokyo.

### 2020-06-13 manuscript/main.tex
Revise state of the field paragraph

@@
 
-The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods into an extensible, open source framework that can be used by professional virologists in their vaccine design process \cite{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
+The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods to inform the vaccine design process \cite{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
 Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, published methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
@@
 We apply this framework to HA sequence data shared via the GISAID EpiFlu database \cite{shu2017gisaid} and to HI and FRA titer data shared by WHO GISRS Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
-With this framework, we show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
-However, we also find that composite models based on phenotypic measures of antigenic drift and genotypic measures of functional constraint consistently outperform any fitness models based on individual genotypic or phenotypic metrics.
+We systematically compare potential predictors and show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
+We also find that composite models based on phenotypic measures of antigenic drift and genotypic measures of functional constraint consistently outperform any fitness models based on individual genotypic or phenotypic metrics.
 

### 2020-06-17 manuscript/cover-letter/cover_letter.tex
Revise cover letter
Minor edits for clarity.

--- 2020-01-25 manuscript/cover-letter/cover_letter.tex
@@
 
-\greetto{Dear Editor,}
+\greetto{Dear Editors,}
 \closeline{\\Sincerely,}
@@
 
-  The World Health Organization selects vaccine strains for seasonal influenza A/H3N2 based on experimental measurements of antigenic drift and computational models of hemagglutinin (HA) sequence evolution.
+  The World Health Organization (WHO) selects vaccine strains for seasonal influenza A/H3N2 based on experimental measurements of antigenic drift and computational models of hemagglutinin (HA) sequence evolution.
   Hemagglutination inhibition (HI) assays provide phenotypic measurements of antigenic drift that have been the gold standard for decades.
-  In their 2014 \textit{Nature} paper, {\L}uksza and L\"assig provided the first mechanistic fitness model to predict the success of H3N2 populations one year into the future from HA sequences alone.
-  That same year in \textit{eLife}, Neher et al. presented the local branching index (LBI), a fitness estimate based on rapid branching in HA phylogenies.
-  No studies to date have integrated these genetic and phenotypic measures of HA into a single model to understand their relative contributions to influenza fitness.
+  The WHO augments these data with computational fitness models based on HA sequences including a mechanistic model by {\L}uksza and L{\"a}ssig (2014) that predicts the success of H3N2 populations one year into the future and a phylogenetic model by Neher et al.
+(2014) called the local branching index (LBI) that estimates H3N2 strain fitness based on rapid branching in HA phylogenies.
+  No published studies to date have integrated all of these genetic and phenotypic measures of HA into a single model to understand their relative contributions to influenza fitness.
 
-  To address this gap, we present the first fitness framework to integrate genetic and phenotypic measures of HA antigenic drift and functional constraint and phylogenetic measures of clade growth.
-  We implemented novel measures of antigenic drift by HI assays, functional constraint measured by deep mutational scanning (DMS) experiments, and clade growth based on recent changes in clade frequencies.
-  We also reimplemented {\L}uksza and L\"assig's sequence-based measures of antigenic drift and functional constraint, and Neher et al.'s clade growth metric, LBI.
-  To identify the most predictive models, we tested each of these fitness metrics individually and in relevant combinations using time-series cross-validation.
+  To address this gap, we developed a fitness framework that integrates genetic and phenotypic measures of antigenic drift and functional constraint with phylogenetic measures of clade growth.
+  We implemented novel measures of antigenic drift from HI assays, functional constraint from deep mutational scanning (DMS) experiments, and clade growth from recent changes in clade frequencies.
+  We also reimplemented {\L}uksza and L{\"a}ssig's sequence-based measures of antigenic drift and functional constraint and Neher et al.'s LBI metric.
+  We rigorously tested the performance of these fitness metrics individually and in relevant combinations using time-series cross-validation.
   The most accurate models included a combination of sequence-based functional constraint and either HI-based measures of antigenic drift or LBI-based measures of clade growth.
 
-  Our findings answer several open questions in the field and point the way for future forecasting efforts.
+  Our findings address several open questions in the field of long-term influenza forecasting and point the way for future forecasting efforts.
   We find that genetic estimates of antigenic drift are insufficient replacements for the gold standard phenotypic measures provided by HI assays.
-  We further show that the success of previous models based on these genetic estimates is due, in part, to inadvertantly borrowing information from the future.
-  In contrast, we show that simple genetic estimates of functional constraint are superior to more detailed estimates provided by deep mutational scanning (DMS) measurements from a single background strain.
-  The results suggest that future forecasting models should strive not to replace genetic fitness metrics with phenotypic metrics but to identify metrics that are the least likely to be affected by historical contingency.
+  We further show that the success of previous models based on these genetic estimates is due, in part, to inadvertently borrowing information from the future.
+  In contrast, simple genetic estimates of functional constraint are superior to more detailed estimates provided by deep mutational scanning (DMS) measurements from a single background strain.
+  These results suggest that future forecasting models should identify metrics that are the least likely to be affected by historical contingency.
 
-  We designed our fitness framework as a reproducible workflow built from open source tools, released as open source itself, and integrated into the real-time pathogen surveillance ecosystem of Nextstrain.
-  Our models estimate the sequence composition of future populations instead of phylogenetic clade frequencies, enabling subsequent forecasting efforts for rapidly evolving organisms whose genomes are not suited for phylogenetic inference such as recombining viruses and bacteria.
-  Importantly, fitness metrics can be readily defined in tidy data frames from any continuous trait of a sample.
-  Altogether, this manuscript and the associated fitness framework are immediately relevant to a general audience of evolutionary biologists, infectious disease researchers, virologists, and bioinformaticians.
-  For these reasons, we believe that \textit{eLife} is an appropriate venue for this work.
+  We designed our fitness framework as an open source, reproducible workflow built on tidy data frames and integrated it into Nextstrain, a real-time pathogen surveillance tool.
+  As our models estimate the sequence composition of future populations instead of phylogenetic clade frequencies, they could enable forecasting efforts for other rapidly evolving organisms whose genomes are not suited for phylogenetic inference such as recombining viruses and bacteria.
+  This manuscript is relevant to an audience of evolutionary biologists, infectious disease researchers, and computational biologists.
+  For this reason, we would like to submit this manuscript for consideration in the Special Issue on Evolutionary Medicine.
 \end{newlfm}

### 2020-06-21 manuscript/main.tex
Include GISAID submitting labs where available
Also, fixes an author initials typo in the funding section and expands the data
availability section to explicitly reference the Supplemental File S1.

@@
 
-Hemagglutination inhibition (HI) measurements were provided by WHO Global Influenza Surveillance and Response System (GISRS) Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
+Hemagglutination inhibition (HI) and focus reduction assay (FRA) measurements were provided by WHO Global Influenza Surveillance and Response System (GISRS) Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
 We converted these raw two-fold dilution measurements to $log_{2}$ titer drops normalized by the corresponding $log_{2}$ autologous measurements as previously described \cite{Neher:2016hy}.
@@
 
-All source code, configuration files, and datasets are available at https://github.com/blab/flu-forecasting.
+Sequence data are available from GISAID using accession ids provided in Supplemental File S1.
+Source code, derived data from serological measurements, fitness metric annotations, and resulting fitness model performance data are available in the project's GitHub repository (https://github.com/blab/flu-forecasting).
+Raw serological measurements are restricted from public distribution by previous data sharing agreements.
 
@@
 The work done at the Crick Worldwide Influenza Centre was supported by the Francis Crick Institute receiving core funding from Cancer Research UK (FC001030), the Medical Research Council (FC001030) and the Wellcome Trust (FC001030).
-SF, KN, KN, SW and HH were supported by the Ministry of Health, Labour and Welfare, Japan (10110400).
+SF, KN, NK, SW and HH were supported by the Ministry of Health, Labour and Welfare, Japan (10110400).
 SW was supported by the Japan Agency for Medical Research and Development (JPfk0108118).

### 2020-07-20 manuscript/main.tex
Explicitly define strains, clades, and earth mover's distance
Restructures some introductory text to clarify terms and how earth mover's
distance actually works.

@@
 However, frequent viral mutation results in viruses that escape previously acquired human immunity.
-The World Health Organization (WHO) Global Influenza Surveillance and Response System (GISRS) selects vaccine viruses to represent circulating viruses, but because the process of vaccine development and distribution requires several months to complete, optimal vaccine design requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
+The World Health Organization (WHO) Global Influenza Surveillance and Response System (GISRS) monitors influenza evolution by sampling currently circulating viruses, or strains, and analyzing these strains with genome sequencing and serological assays.
+The WHO GISRS uses these data to select vaccine viruses that should best represent circulating viruses in the next influenza season.
+However, because the process of vaccine development and distribution requires several months to complete, optimal vaccine design requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
@@
 As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus strain an exponential growth factor scaled by an estimate of the strain's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
-We estimated the frequency of virus strains every six months using kernel density estimation (KDE).
+{\L}uksza and L\"assig \cite{Luksza:2014hj} measured model performance by identifying clades -- groups of strains that all share a recent common ancestor -- and comparing observed and estimated future clade frequencies.
+However, as clade definitions are inherently unstable between seasons, we evaluated our models by comparing the genetic composition of observed and estimated future populations with the earth mover's distance metric.
+The earth mover's distance calculates the minimum distance between two populations, given the frequency of each individual within a population and a pairwise ``ground distance'' between individuals \cite{Rubner1998}.
+We defined distinct amino acid haplotypes as individuals in our observed and estimated future populations.
+For frequencies of individuals, we used the observed frequencies of haplotypes in the future and our model's estimated frequencies.
+We calculated the ground distance between individuals as the Hamming distance between haplotypes.
+With this implementation, more accurate projections of the future population's composition produce smaller earth mover's distances between the observed and estimated future (Fig.~\ref{fig:model}).
 
@@
 We fit models for individual fitness metrics and combinations of metrics that we anticipated would be mutually beneficial.
-For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model} and Eq.~\ref{equation_fitness_estimation}).
+For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Eq.~\ref{equation_fitness_estimation}).
 We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Supplemental Figs~\ref{sup_fig:cross_validation_for_simulated_populations} and \ref{sup_fig:cross_validation_for_natural_populations}).

### 2020-07-21 manuscript/main.tex
Clarify the source of HI titers
Describe technical and scientific feasibility of multi-gene models

@@
 Models should also integrate information from multiple segments of the influenza genome and will need to balance the fitness benefits of evolution in genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment \cite{Villa:2017iw}.
+Our forecasting framework makes the inclusion of fitness metrics based on additional gene segments technically straightforward.
+However, the definition of appropriate fitness metrics for neuraminidase and other genes remains an important scientific challenge.
+An additional challenge to model training is a relative lack of historical strains for which all genes have been sequenced.
+Of the 34,312 H3N2 strains in GISAID with all eight primary gene segments and collection dates between October 1, 1990 and 2019, the majority (24,466 or 71\%) were collected after October 1, 2015.
+Data availability will therefore inform which gene segments are prioritized for inclusion in future models.
 Finally, forecasting models need to account for the geographic distribution of viruses and the vastly different sampling intensities across the globe.
@@
 We excluded all egg-passaged strains and all strains with ambiguous year, month, and day annotations.
-We prioritized strains with more available HI titer measurements.
+We prioritized strains with more available HI titer measurements provided by the WHO GISRS Collaborating Centers.
 For model testing, we selected an additional 7,171 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and October 1, 2019.

### 2020-07-21 manuscript/main.tex
Explicitly reference CDC's FluSight network for short-term forecasts

@@
 However, the effective elimination of LBI from our three-metric model during the validation period (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}) reveals the limitations of our current additive approach to composite models.
-The recent success of weighted ensembles for short-term influenza forecasting \cite{Reich:2019bq} suggests that long-term forecasting may benefit from a similar approach.
+The recent success of weighted ensembles for short-term influenza forecasting through the CDC's FluSight network \cite{Reich:2019bq} suggests that long-term forecasting may benefit from a similar approach.
 

### 2020-07-21 manuscript/main.tex
Describe interaction between overfitting and evolutionary processes
Expands a paragraph in the Discussion to mention the increased diversity of H3N2
clades during our test period and the effect that diversity and other
evolutionary events may have had on our models. These effects are framed in the
context of model overfitting.

@@
 For example, the model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
-Specifically, we found that this model failed to predict the sudden decline of a dominant reassortant clade, A2/re, in 2019.
-Despite this model's excellent performance historically, it was unable to account for rare yet important events such as reassortment.
+Although we cannot rule out the role of overfitting, this model's poor performance coincided with unusual evolutionary circumstances.
+The diversity of H3N2 lineages during our test period was higher than the historical average \cite{Koelle:2006kf}, with the most recent common ancestor of all circulating strains dating eight years back.
+This persistence of diversity may have reduced the effectiveness of the LBI metric that assumes relatively rapid population turnover.
+Additionally, this model's poorest performance occurred in 2019 when it failed to predict the sudden decline of a dominant reassortant clade, A2/re.
+Only our models based on HI antigenic novelty and mutational load continued to perform as well or better than the naive model during the same time period.
+These results highlight the challenge of identifying models that remain robust to stochastic evolutionary events by avoiding overfitting to the past.
 
-Finally, we observed that composite models of multiple orthogonal fitness metrics often outperformed models based on their individual components.
+Correspondingly, we observed that composite models of multiple orthogonal fitness metrics often outperformed models based on their individual components.
 These results are consistent with previous work that found improved performance by integrating components of antigenic drift, functional constraint, and clade growth \cite{Luksza:2014hj}.

### 2020-07-27 manuscript/main.tex
Add a summary table of primary models used in the paper

@@
 
-We estimated viral fitness with biologically-informed metrics including those originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj} of epitope antigenic novelty and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) antigenic novelty \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
+\begin{table}[htb]
+  \begin{center}
+    \scalebox{0.7}{
+      \input{tables/model_summary.tex}
+    }
+    \caption{
+      Summary of models used with simulated and natural populations.
+      Models are labeled by the type of population they were applied to, the type of data they were based on, and the component of influenza fitness they represent.
+    }
+    \label{table_model_summary}
+  \end{center}
+\end{table}
+
+We estimated viral fitness with biologically-informed metrics including those originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj} of epitope antigenic novelty and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) antigenic novelty \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency) (Table~\ref{table_model_summary}).
 All of these metrics except for HI antigenic novelty and DMS mutational effects rely only on HA sequences.

### 2020-07-27 manuscript/tables/model_summary.tex
Add a summary table of primary models used in the paper

\begin{tabular*}{1.4\textwidth}{lllll}
\toprule
Model                                                        & Populations        & Data type             & Fitness category      & Originally implemented by                   \\
\midrule
true fitness                                                 & simulated          & simulated populations & positive control      & Huddleston et al. 2020                      \\
naive                                                        & simulated, natural & HA sequences          & negative control      & Huddleston et al. 2020                      \\
epitope antigenic novelty                                    & simulated, natural & HA sequences          & antigenic drift       & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
epitope ancestor                                             & simulated, natural & HA sequences          & antigenic drift       & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
HI antigenic novelty                                         & natural            & serological assays    & antigenic drift       & Huddleston et al. 2020                      \\
mutational load                                              & simulated, natural & HA sequences          & functional constraint & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
\makecell{deep mutational scanning (DMS) \\ mutational effects} & natural            & DMS assays            & functional constraint & Lee et al. 2018 \cite{Lee2018}              \\
local branching index (LBI)                                  & simulated, natural & HA sequences          & clade growth          & Neher et al. 2014 \cite{Neher:2014eu}       \\
delta frequency                                              & simulated, natural & HA sequences          & clade growth          & Huddleston et al. 2020                      \\
\bottomrule
\end{tabular*}

### 2020-07-28 manuscript/main.tex
Clarify and expand comparison of models to vaccine strains
Adds best estimated strains by naive model to vaccine comparison figure, adds a
new supplemental figure that shows the distance to the future of observed and
estimated best strains relative to the corresponding vaccine strains, and
quantifies improvement of models over the vaccine strains in the results text.

This commit also changes the vaccine strain name annotation in the vaccine
comparison figure to use an abbreviation with a larger font and improve overall
readability. The abbreviations are now explained in the figure caption.

@@
   \caption{
-    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated closest strains to the future by the mutational load and LBI model (orange) and the HI antigenic novelty and mutational load model (purple).
+    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated closest strains to the future by the mutational load and LBI model (orange), the HI antigenic novelty and mutational load model (purple), and the naive model (gray).
     Vaccine strains were assigned to the validation or test timepoint closest to the date they were selected by the WHO.
     The weighted distance to the future for each strain was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
+    Vaccine strain names are abbreviated from A/Fujian/411/2002, A/Wellington/1/2004, A/California/7/2004, A/Wisconsin/67/2005, A/Brisbane/10/2007, A/Perth/16/2009, A/Victoria/361/2011, A/Texas/50/2012, A/Switzerland/9715293/2013, A/HongKong/4801/2014, A/Singapore/Infimh-16-0019/2016, and A/Switzerland/8060/2017.
   }
@@
 The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
+On average, the strains selected by this model were closer to future than the vaccine strain by 1.93 AAs (Supplemental Fig.~\ref{sup_fig:vaccine_comparison_relative_distance}).
 For the two seasons that the model selected more distant strains than the vaccine strain, the mean distance relative to the vaccine strain was 1.58 AAs.
-The HI antigenic novelty and mutational load model performed similarly by identifying strains as close or closer to the future for 11 (92\%) seasons.
+The HI antigenic novelty and mutational load model performed similarly by identifying strains as close or closer to the future for 11 (92\%) seasons with an average improvement over the vaccine strains of 2.33 AAs.
 For the one season that the model selected a more distant strain, that selected strain was 0.75 AAs farther from the future than the vaccine strain.
+Interestingly, the strains selected by the naive model were always better than the selected vaccine strain.
+With an average improvement over the vaccine strains of 2.19 AAs, the naive model performed consistently better than the LBI-based model and nearly as well as the HI-based model.
 

### 2020-07-28 manuscript/supplement.tex
Add title and caption for supplemental file S1
Clarify and expand comparison of models to vaccine strains
Adds best estimated strains by naive model to vaccine comparison figure, adds a
new supplemental figure that shows the distance to the future of observed and
estimated best strains relative to the corresponding vaccine strains, and
quantifies improvement of models over the vaccine strains in the results text.

This commit also changes the vaccine strain name annotation in the vaccine
comparison figure to use an abbreviation with a larger font and improve overall
readability. The abbreviati

@@
   \begin{center}
+  \includegraphics[width=\textwidth]{figures/vaccine-comparison-relative-distance.pdf}
+  \caption{
+    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Fig.~\ref{fig:vaccine_comparison}.
+    Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.
+  }
+  \label{sup_fig:vaccine_comparison_relative_distance}
+  \end{center}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
   \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.pdf}
@@
 
+\subsection*{Supplemental Files}
+
+\textbf{Supplemental File S1.} Tab-delimited GISAID accessions and metadata including originating and submitting labs for natural strains used across all timepoints.
+
 \pagebreak

### 2020-07-28 manuscript/main.tex
Summarize forecasts made by the naive models

@@
 Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
+In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}B).
+However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}C and D).
+These results suggested that estimating a single closest strain to the future is a more tractable problem than estimating the future frequencies of clades.
 
@@
 Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
-These results confirm that our approach of minimizing the distance between yearly populations can simultaneously capture clade-level dynamics of simulated influenza populations and identify individual strains that are most representative of future populations.
+The naive model performed comparatively well on these test data with all its estimated closest strains to the future in the top 20th percentile and a slightly higher correlation between observed and estimated percentile ranks than the true fitness model (Spearman's $\rho^2 = 0.82$, Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_simulated_populations}).
+These results confirmed that our approach of minimizing the distance between yearly populations could simultaneously capture clade-level dynamics of simulated influenza populations and identify individual strains that are most representative of future populations.
+However, they also supported the earlier finding that clade frequency forecasts may be inherently more challenging than identification of the closest strain to the future.
 
@@
 This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}D).
+The naive model's performance repeated the pattern we observed with simulated populations: it made poor forecasts of absolute clade frequencies, but its estimated closest strains to the future were consistently highly ranked among the observed closest strains (Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_natural_populations}B and C).
 
@@
 Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.72$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
+The estimated strain ranks of the naive model were not as well correlated (Spearman's $\rho^2 = 0.56$), but seven of its eight estimates for the closest strain to the future (88\%) were in the top fifth percentile of observed closest strains (Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_natural_populations}B and D).
 
@@
 With an average improvement over the vaccine strains of 2.19 AAs, the naive model performed consistently better than the LBI-based model and nearly as well as the HI-based model.
+These results were consistent with our earlier observations that the naive model often performs as well as biologically-informed models when estimating a single closest strain to the future.
 
@@
 We demonstrated that the integration of these phenotypic metrics with previously published sequence-only metrics produces more accurate forecasts than sequence-only models.
+Interestingly, we found that a naive model that predicts no change over the course of one year can often identify a single representative strain of the future despite its inability to accurately forecast clade frequencies.
 We have added this framework as a component of seasonal influenza analyses on \href{https://nextstrain.org/flu}{nextstrain.org} where it provides real-time forecasts for influenza researchers, decision makers, and the public.

### 2020-07-29 manuscript/tables/model_summary.tex
Replace reference to this study with "this study"

@@
 \midrule
-true fitness                                                 & simulated          & simulated populations & positive control      & Huddleston et al. 2020                      \\
-naive                                                        & simulated, natural & HA sequences          & negative control      & Huddleston et al. 2020                      \\
+true fitness                                                 & simulated          & simulated populations & positive control      & this study                      \\
+naive                                                        & simulated, natural & HA sequences          & negative control      & this study                      \\
 epitope antigenic novelty                                    & simulated, natural & HA sequences          & antigenic drift       & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
 epitope ancestor                                             & simulated, natural & HA sequences          & antigenic drift       & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
-HI antigenic novelty                                         & natural            & serological assays    & antigenic drift       & Huddleston et al. 2020                      \\
+HI antigenic novelty                                         & natural            & serological assays    & antigenic drift       & this study                      \\
 mutational load                                              & simulated, natural & HA sequences          & functional constraint & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
@@
 local branching index (LBI)                                  & simulated, natural & HA sequences          & clade growth          & Neher et al. 2014 \cite{Neher:2014eu}       \\
-delta frequency                                              & simulated, natural & HA sequences          & clade growth          & Huddleston et al. 2020                      \\
+delta frequency                                              & simulated, natural & HA sequences          & clade growth          & this study                      \\
 \bottomrule

### 2020-07-29 manuscript/main.tex
Summarize antigenic distances in vaccine comparison
Adds analyses to the vaccine comparison notebook that include a comparison of
relative distance to the future for closest strains vs. vaccine strains and
the antigenic distance between those corresponding strains. Also, improves the
aesthetics and readability of the vaccine comparison figure and adds the
supplemental figure showing relative distance to the future for closest strains
vs. vaccine strains.

@@
   \caption{
-    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated closest strains to the future by the mutational load and LBI model (orange), the HI antigenic novelty and mutational load model (purple), and the naive model (gray).
+    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated closest strains to the future by the mutational load and LBI model (orange), the HI antigenic novelty and mutational load model (purple), and the naive model (black).
     Vaccine strains were assigned to the validation or test timepoint closest to the date they were selected by the WHO.

### 2020-07-31 manuscript/main.tex
Summarize H3N2 vaccine effectiveness and causes for poor VE

@@
 However, because the process of vaccine development and distribution requires several months to complete, optimal vaccine design requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
+
+Historically, the effectiveness of the H3N2 vaccine component has been much lower than the other seasonal influenza subtypes.
+For example, H3N2's mean vaccine effectiveness from 2004--2015 was 33\% compared to 61\% for H1N1pdm and 54\% for influenza B viruses \cite{Belongia2016}.
+Multiple factors can reduce vaccine effectiveness including selection of a vaccine strain that is not antigenically representative of future populations \cite{Belongia2016,gouma_antigenic_2020} and adaptations of the selected strain to egg-passaging during vaccine production that alter the antigenicity of the resulting vaccine component \cite{zost_contemporary_2017}.
+Even when vaccine strains are well-matched antigenically, they may fail to induce a strong immune response due to previous infection history of vaccine recipients \cite{Cobey2018}.
+While all of these factors must be addressed to increase vaccine effectiveness, substantial effort has focused on the selection of the most representative strain for the next season's vaccine.
+
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.

### 2020-08-03 manuscript/main.tex
Use eLife template and citation styles
Updates most citations to use the parenthetical form, adds the eLife document
type and bibliography style, updates the model summary table to use the full
citation style, and adds a citation for Barrat-Charlaix et al. 2020 which is now
available as a preprint.

--- 2020-07-29 manuscript/main.tex
@@
 
-Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually with the majority of infections caused by influenza A/H3N2 \cite{flufactsheet}.
+Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually with the majority of infections caused by influenza A/H3N2 \citep{flufactsheet}.
 Vaccination remains the most effective public health response available.
@@
 However, because the process of vaccine development and distribution requires several months to complete, optimal vaccine design requires an accurate prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
+
+Historically, the effectiveness of the H3N2 vaccine component has been much lower than the other seasonal influenza subtypes.
+For example, H3N2's mean vaccine effectiveness from 2004--2015 was 33\% compared to 61\% for H1N1pdm and 54\% for influenza B viruses \citep{Belongia2016}.
+Multiple factors can reduce vaccine effectiveness including selection of a vaccine strain that is not antigenically representative of future populations \citep{Belongia2016,gouma_antigenic_2020} and adaptations of the selected strain to egg-passaging during vaccine production that alter the antigenicity of the resulting vaccine component \citep{zost_contemporary_2017}.
+Even when vaccine strains are well-matched antigenically, they may fail to induce a strong immune response due to previous infection history of vaccine recipients \citep{Cobey2018}.
+While all of these factors must be addressed to increase vaccine effectiveness, substantial effort has focused on the selection of the most representative strain for the next season's vaccine.
+
 Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
-Until recently, the hemagglutination inhibition (HI) assay has been the primary experimental measure of antigenic cross-reactivity between pairs of circulating viruses \cite{hirst1943studies}.
-Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \cite{Chambers:2015jt,Zost2017}, prompting the increased use of virus neutralization assays including the neutralization-based focus reduction assay (FRA) \cite{Okuno:1990uu}.
-Together, these two assays are the gold standard in virus antigenic characterizations for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
-As a result, researchers have developed computational methods to predict influenza evolution from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.
+Until recently, the hemagglutination inhibition (HI) assay has been the primary experimental measure of antigenic cross-reactivity between pairs of circulating viruses \citep{hirst1943studies}.
+Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \citep{Chambers:2015jt,Zost2017}, prompting the increased use of virus neutralization assays including the neutralization-based focus reduction assay (FRA) \citep{Okuno:1990uu}.
+Together, these two assays are the gold standard in virus antigenic characterizations for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \citep{Wood:2012ii}.
+As a result, researchers have developed computational methods to predict influenza evolution from sequence data alone \citep{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.
 
@@
 Recent developments in computational methods and influenza virology have made it feasible to integrate these important metrics of influenza fitness into a single predictive model.
-For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \cite{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning (DMS) experiments \cite{Lee2018}.
+For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \citep{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning (DMS) experiments \citep{Lee2018}.
 We describe an approach to integrate previously disparate sequence-only models of influenza evolution with high-quality experimental measurements of antigenic drift and functional constraint.
 
-The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods to inform the vaccine design process \cite{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
-Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, published methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
-However, unpublished work by {\L}uksza and L\"assig submitted to the WHO GISRS network incorporates antigenic phenotypes into fitness-based predictions \cite{Morris:2017ea,luksza_personal_communication}.
+The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods to inform the vaccine design process \citep{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
+Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \citep{Steinbruck:2014kq,Neher:2016hy}, published methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
+However, unpublished work by {\L}uksza and L\"assig submitted to the WHO GISRS network incorporates antigenic phenotypes into fitness-based predictions \citep{Morris:2017ea,luksza_personal_communication}.
 Here, we provide an open source framework for forecasting the genetic composition of future seasonal influenza populations using genotypic and phenotypic fitness estimates.
-We apply this framework to HA sequence data shared via the GISAID EpiFlu database \cite{shu2017gisaid} and to HI and FRA titer data shared by WHO GISRS Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
+We apply this framework to HA sequence data shared via the GISAID EpiFlu database \citep{shu2017gisaid} and to HI and FRA titer data shared by WHO GISRS Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
 We systematically compare potential predictors and show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
@@
 
-We developed a framework to forecast seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+We developed a framework to forecast seasonal influenza evolution inspired by the Malthusian growth fitness model of \cite{Luksza:2014hj}.
 As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus strain an exponential growth factor scaled by an estimate of the strain's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
-{\L}uksza and L\"assig \cite{Luksza:2014hj} measured model performance by identifying clades -- groups of strains that all share a recent common ancestor -- and comparing observed and estimated future clade frequencies.
+\cite{Luksza:2014hj} measured model performance by identifying clades -- groups of strains that all share a recent common ancestor -- and comparing observed and estimated future clade frequencies.
 However, as clade definitions are inherently unstable between seasons, we evaluated our models by comparing the genetic composition of observed and estimated future populations with the earth mover's distance metric.
-The earth mover's distance calculates the minimum distance between two populations, given the frequency of each individual within a population and a pairwise ``ground distance'' between individuals \cite{Rubner1998}.
+The earth mover's distance calculates the minimum distance between two populations, given the frequency of each individual within a population and a pairwise ``ground distance'' between individuals \citep{Rubner1998}.
 We defined distinct amino acid haplotypes as individuals in our observed and estimated future populations.
@@
 
-We estimated viral fitness with biologically-informed metrics including those originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj} of epitope antigenic novelty and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) antigenic novelty \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency) (Table~\ref{table_model_summary}).
+We estimated viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope antigenic novelty and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) antigenic novelty \citep{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \citep{Lee2018}, local branching index (LBI) \citep{Neher:2014eu}, and change in clade frequency over time (delta frequency) (Table~\ref{table_model_summary}).
 All of these metrics except for HI antigenic novelty and DMS mutational effects rely only on HA sequences.
@@
 
-The long-term evolution of influenza H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
+The long-term evolution of influenza H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \citep{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
 To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Fig.~\ref{sup_fig:cross_validation_for_simulated_populations}).
@@
 LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
-This result is consistent with the fact that the LBI is a correlate of fitness in models of rapidly adapting populations \cite{Neher:2014eu}.
+This result is consistent with the fact that the LBI is a correlate of fitness in models of rapidly adapting populations \citep{Neher:2014eu}.
 Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
@@
 Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
-The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
+The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \citep{Luksza:2014hj} that are not explicitly present in our simulations.
 To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each strain's ancestor in the previous season.
@@
 We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
-As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI antigenic novelty would be individually more predictive than mutational load or DMS mutational effects.
-Previous research \cite{Neher:2014eu} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
+As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \citep{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI antigenic novelty would be individually more predictive than mutational load or DMS mutational effects.
+Previous research \citep{Neher:2014eu} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
 As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA antigenic novelty models in parallel with the HI antigenic novelty models.
@@
 These results led us to hypothesize that the contribution of these specific epitope sites to antigenic drift has weakened over time.
-Importantly, these 49 epitope sites were originally selected by {\L}uksza and L\"assig \cite{Luksza:2014hj} from a previous historical survey of sites with beneficial mutations between 1968--2005 \cite{Shih:2007bd}.
+Importantly, these 49 epitope sites were originally selected by \cite{Luksza:2014hj} from a previous historical survey of sites with beneficial mutations between 1968--2005 \citep{Shih:2007bd}.
 If the beneficial effects of mutations at these sites were due to historical contingency rather than a constant contribution to antigenic drift, we would expect models based on these sites to perform well until 2005 and then overfit relative to future data.
@@
 To test this hypothesis, we identified a new set of beneficial sites across our entire validation period of October 1990 through October 2015.
-Inspired by the original approach of Shih et al.~\cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population, including 12 that were also present in the original set of 49 sites.
+Inspired by the original approach of \cite{Shih:2007bd}, we identified 25 sites in HA1 where mutations rapidly swept through the global population, including 12 that were also present in the original set of 49 sites.
 We fit an antigenic novelty model to these 25 sites across the complete validation period and dubbed this the ``oracle antigenic novelty'' model, as it benefited from knowledge of the future in its forecasts.
@@
 Of the two metrics for functional constraint, mutational load outperformed DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
-In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the mutational load metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
+In contrast to the original \cite{Luksza:2014hj} model, where the coefficient of the mutational load metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
 Notably, the best performance of the DMS mutational effects model was forecasting from April 2007 to April 2008 when the major clade containing A/Perth/16/2009 was first emerging.
@@
 LBI performed especially poorly in the last two test timepoints of April and October 2018 (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
-These timepoints correspond to the dominance and sudden decline of a reassortant clade named A2/re \cite{Potter2019}.
+These timepoints correspond to the dominance and sudden decline of a reassortant clade named A2/re \citep{Potter2019}.
 By April 2018, the A2/re clade had risen to a global frequency over 50\% from less than 15\% the previous year, despite an absence of antigenic drift.
@@
 
-To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}.
+To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \citep{Hadfield2018}.
 Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning new coefficients per timepoint from the test data.
@@
 We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza H3N2 by estimating the sequence composition of future populations.
-A key innovation of this framework is its ability to directly compare viral populations between seasons using the earth mover's distance metric \cite{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
+A key innovation of this framework is its ability to directly compare viral populations between seasons using the earth mover's distance metric \citep{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
 The best models from this framework still effectively capture clade dynamics and accurately identify optimal vaccine candidates from simulated and natural H3N2 populations without relying on clades as model targets.
@@
 A clear example of the former was the poor performance of our DMS-based fitness metric compared to a simpler mutational load metric.
-Although the DMS experiments provided detailed estimates of which amino acids were preferred at which positions in HA, these measurements were specific to a single strain, A/Perth/16/2009 \cite{Lee2018}.
+Although the DMS experiments provided detailed estimates of which amino acids were preferred at which positions in HA, these measurements were specific to a single strain, A/Perth/16/2009 \citep{Lee2018}.
 When we applied these measurements to predict the success of global populations, they were less informative on average than the naive model.
@@
 We observed overfitting to historical context in sequence-based models of antigenic drift.
-The fitness benefit of mutations that led to antigenic drift in H3N2 in the past is well-documented \cite{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}.
-Although the antigenic importance of seven specific sites in HA were experimentally validated by Koel et al.
-2013 \cite{Koel:2013jz}, these sites do not explain all antigenic drift observed in natural populations \cite{Neher:2016hy}.
-Other attempts to define these so-called ``epitope sites'' have relied on either aggregation of results from antigenic escape assays \cite{Wolf:2006da} or retrospective computational analyses of sites with beneficial mutations \cite{Shih:2007bd,Luksza:2014hj}.
+The fitness benefit of mutations that led to antigenic drift in H3N2 in the past is well-documented \citep{Wiley:1981bc,Smith:2004jc,Wolf:2006da,Koel:2013jz}.
+Although the antigenic importance of seven specific sites in HA were experimentally validated by \cite{Koel:2013jz}, these sites do not explain all antigenic drift observed in natural populations \citep{Neher:2016hy}.
+Other attempts to define these so-called ``epitope sites'' have relied on either aggregation of results from antigenic escape assays \citep{Wolf:2006da} or retrospective computational analyses of sites with beneficial mutations \citep{Shih:2007bd,Luksza:2014hj}.
 We found that models based on all of these definitions except for the seven Koel epitope sites overfit to the historical context from which they were identified (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
 These results suggest that the set of sites that contribute to antigenic drift at any given time may depend on both the fitness landscape of currently circulating strains and the immune landscape of the hosts these strains need to infect.
-Recent experimental mapping of antigenic escape mutations in H3N2 HA with human sera show that the specific sites that confer antigenic escape can vary dramatically between individuals based on their exposure history \cite{Lee2019}.
+Recent experimental mapping of antigenic escape mutations in H3N2 HA with human sera show that the specific sites that confer antigenic escape can vary dramatically between individuals based on their exposure history \citep{Lee2019}.
 In contrast to models based on predefined ``epitope sites'', our model based on experimental measurements of antigenic drift did not suffer from overfitting in the validation or test periods.
@@
 Although we cannot rule out the role of overfitting, this model's poor performance coincided with unusual evolutionary circumstances.
-The diversity of H3N2 lineages during our test period was higher than the historical average \cite{Koelle:2006kf}, with the most recent common ancestor of all circulating strains dating eight years back.
+The diversity of H3N2 lineages during our test period was higher than the historical average \citep{Koelle:2006kf}, with the most recent common ancestor of all circulating strains dating eight years back.
 This persistence of diversity may have reduced the effectiveness of the LBI metric that assumes relatively rapid population turnover.
@@
 Correspondingly, we observed that composite models of multiple orthogonal fitness metrics often outperformed models based on their individual components.
-These results are consistent with previous work that found improved performance by integrating components of antigenic drift, functional constraint, and clade growth \cite{Luksza:2014hj}.
+These results are consistent with previous work that found improved performance by integrating components of antigenic drift, functional constraint, and clade growth \citep{Luksza:2014hj}.
 However, the effective elimination of LBI from our three-metric model during the validation period (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}) reveals the limitations of our current additive approach to composite models.
-The recent success of weighted ensembles for short-term influenza forecasting through the CDC's FluSight network \cite{Reich:2019bq} suggests that long-term forecasting may benefit from a similar approach.
+The recent success of weighted ensembles for short-term influenza forecasting through the CDC's FluSight network \citep{Reich:2019bq} suggests that long-term forecasting may benefit from a similar approach.
 
@@
 By forecasting the composition of future H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
-While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AAs per year for validation data and 0.85 AAs per year for test data.
+While this performance confirms previously demonstrated potential for long-term influenza forecasting \citep{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AAs per year for validation data and 0.85 AAs per year for test data.
 However, these results are consistent with the observed dynamics of H3N2.
-First, the one-year forecast horizon is a fraction of the average coalescence time for H3N2 populations of about 3--8 years \cite{Rambaut:2008ew}.
+First, the one-year forecast horizon is a fraction of the average coalescence time for H3N2 populations of about 3--8 years \citep{Rambaut:2008ew}.
 Hence, we expect the diversity of circulating strains to persist between seasons.
-Second, H3N2 hemagglutinin accumulates 3.6 amino acid changes per year \cite{Smith:2004jc}.
+Second, H3N2 hemagglutinin accumulates 3.6 amino acid changes per year \citep{Smith:2004jc}.
 This accumulation of amino acid substitutions contributes to the distance between annual populations observed by the naive model.
@@
 Despite the weak performance of our FRA antigenic novelty model on recent data, continued accumulation of FRA measurements over time should eventually enable models as accurate as the current HI-based models.
-In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
+In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \citep{Lee2019}.
 These assays reveal the specific sites and residues that confer antigenic escape from polyclonal sera obtained from individual humans.
 A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape H3N2 strains must navigate to be globally successful.
-Models should also integrate information from multiple segments of the influenza genome and will need to balance the fitness benefits of evolution in genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment \cite{Villa:2017iw}.
+Models should also integrate information from multiple segments of the influenza genome and will need to balance the fitness benefits of evolution in genes such as neuraminidase \citep{Chen:2018kp} with the costs of reassortment \citep{Villa:2017iw}.
 Our forecasting framework makes the inclusion of fitness metrics based on additional gene segments technically straightforward.
@@
 Finally, forecasting models need to account for the geographic distribution of viruses and the vastly different sampling intensities across the globe.
-Most influenza sequence data come from highly developed countries that account for a small fraction of the global population, while globally successful clades of influenza H3N2 often emerge in less well-sampled regions \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
+Most influenza sequence data come from highly developed countries that account for a small fraction of the global population, while globally successful clades of influenza H3N2 often emerge in less well-sampled regions \citep{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
 Explicitly accounting for these sampling biases and the associated migration dynamics would allow models to weight forecasts based on both viral fitness and transmission.
@@
 Much of the predictive power, in the sense of a reduced distance between the predicted and observed populations, might be due to putting more weight on less evolved strains rather than \emph{bona fide} prediction of fitness.
-In a companion manuscript, Barrat-Charlaix et al.~show that LBI has little predictive power for fixation probabilities of mutations in H3N2.
+In a companion manuscript, \cite{Barrat-Charlaix2020} show that LBI has little predictive power for fixation probabilities of mutations in H3N2.
 
@@
 
-The entire workflow for our analyses was implemented with Snakemake \cite{Snakemake}.
+The entire workflow for our analyses was implemented with Snakemake \citep{Snakemake}.
 We have provided all source code, configuration files, and datasets at \href{https://github.com/blab/flu-forecasting}{https://github.com/blab/flu-forecasting}.
@@
 
-We simulated the long-term evolution of H3N2-like viruses with SANTA-SIM \cite{Jariani2019} for 10,000 generations or 50 years where 200 generations was equivalent to 1 year.
+We simulated the long-term evolution of H3N2-like viruses with SANTA-SIM \citep{Jariani2019} for 10,000 generations or 50 years where 200 generations was equivalent to 1 year.
 We discarded the first 10 years as a burn-in period, selected the next 30 years for model fitting and validation, and held out the last 9 years as out-of-sample data for model testing.
@@
 We defined purifying selection across all three domains, allowing the preferred amino acid at each site to change at a fixed rate over time.
-We additionally defined exposure-dependent selection for 49 putative epitope sites in HA1 \cite{Luksza:2014hj} to impose an effect of antigenic novelty that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
+We additionally defined exposure-dependent selection for 49 putative epitope sites in HA1 \citep{Luksza:2014hj} to impose an effect of antigenic novelty that would allow mutations at those sites to increase viral fitness despite underlying purifying selection.
 We modified the SANTA-SIM source code to enable the inclusion of true fitness values for each strain in the FASTA header of the sampled sequences from each generation.
@@
 Hemagglutination inhibition (HI) and focus reduction assay (FRA) measurements were provided by WHO Global Influenza Surveillance and Response System (GISRS) Collaborating Centers in London, Melbourne, Atlanta and Tokyo.
-We converted these raw two-fold dilution measurements to $log_{2}$ titer drops normalized by the corresponding $log_{2}$ autologous measurements as previously described \cite{Neher:2016hy}.
+We converted these raw two-fold dilution measurements to $log_{2}$ titer drops normalized by the corresponding $log_{2}$ autologous measurements as previously described \citep{Neher:2016hy}.
 
@@
 
-Prior to our analyses, we downloaded all HA sequences and metadata from GISAID \cite{shu2017gisaid}.
+Prior to our analyses, we downloaded all HA sequences and metadata from GISAID \citep{shu2017gisaid}.
 For model training and validation, we selected 15,583 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1990 and October 1, 2015.
@@
 For each timepoint in model training, validation, and testing, we selected the subsampled HA sequences with collection dates up to that timepoint.
-We aligned sequences with the augur align command \cite{Hadfield2018} and MAFFT v7.407 \cite{Katoh2002}.
-We inferred initial phylogenies for HA sequences at each timepoint with IQ-TREE v1.6.10 \cite{Nguyen2014}.
-To reconstruct time-resolved phylogenies, we applied TreeTime v0.5.6 \cite{Sagulenko2018} with the augur refine command.
+We aligned sequences with the augur align command \citep{Hadfield2018} and MAFFT v7.407 \citep{Katoh2002}.
+We inferred initial phylogenies for HA sequences at each timepoint with IQ-TREE v1.6.10 \citep{Nguyen2014}.
+To reconstruct time-resolved phylogenies, we applied TreeTime v0.5.6 \citep{Sagulenko2018} with the augur refine command.
 
@@
 
-We assumed that the evolution seasonal influenza H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \cite{Luksza:2014hj}.
+We assumed that the evolution seasonal influenza H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \citep{Luksza:2014hj}.
 Under this model, we estimated the future frequency, $\hat{x}_{i}(t + \Delta{t})$, of each strain $i$ from the strain's current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
@@
 
-For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the earth mover's distance (EMD) \cite{Rubner1998,Kusner2015} between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of strains at time $t$ by their estimated fitnesses.
+For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the earth mover's distance (EMD) \citep{Rubner1998,Kusner2015} between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of strains at time $t$ by their estimated fitnesses.
 Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
@@
 However, the inevitable accumulation of substitutions between the two populations prevents this outcome.
-We calculated EMD with the Python bindings for the OpenCV 3.4.1 implementation \cite{opencv_library}.
-We applied the Nelder-Mead minimization algorithm as implemented in SciPy \cite{SciPy} to learn fitness coefficients that minimize the average of this distance metric over all timepoints in a given training window.
+We calculated EMD with the Python bindings for the OpenCV 3.4.1 implementation \citep{opencv_library}.
+We applied the Nelder-Mead minimization algorithm as implemented in SciPy \citep{SciPy} to learn fitness coefficients that minimize the average of this distance metric over all timepoints in a given training window.
 
@@
 We estimated antigenic drift for each strain using either genetic or HI data.
-To estimate antigenic drift with genetic data, we implemented an antigenic novelty metric based on the ``cross-immunity'' metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+To estimate antigenic drift with genetic data, we implemented an antigenic novelty metric based on the ``cross-immunity'' metric originally defined by \cite{Luksza:2014hj}.
 Briefly, for each pair of strains in adjacent seasons, we counted the number of amino acid differences between the strains' HA sequences at 49 epitope sites.
@@
 For each individual strain $i$ at each timepoint $t$, we estimated that strain's ability to escape cross-immunity by summing the exponentially-scaled epitope distances between previously circulating strains and the given strain as in Equation~\ref{equation_epitope_cross_immunity}.
-We defined the constant $D_{0} = 14$, as in the original definition of cross-immunity \cite{Luksza:2014hj}.
-To compare these epitope sites with other previously published sites, we fit epitope antigenic novelty models based on sites defined by Wolf et al.
-2006 \cite{Wolf:2006da} and Koel et al.
-2013 \cite{Koel:2013jz}.
+We defined the constant $D_{0} = 14$, as in the original definition of cross-immunity \citep{Luksza:2014hj}.
+To compare these epitope sites with other previously published sites, we fit epitope antigenic novelty models based on sites defined by \cite{Wolf:2006da} and \cite{Koel:2013jz}.
 
@@
 To test the historical contingency of the epitope sites defined above, we additionally identified a new set of sites with beneficial mutations across the training/validation period of October 1990 through October 2015.
-Following the general approach of Shih et al. \cite{Shih:2007bd}, we manually identified 25 sites in HA1 where mutations rapidly swept through the global population.
+Following the general approach of \cite{Shih:2007bd}, we manually identified 25 sites in HA1 where mutations rapidly swept through the global population.
 We required mutations to emerge from below 5\% global frequency and reach $>$90\% frequency.
@@
 
-To estimate antigenic drift with HI data, we first applied the titer tree model to the phylogeny at a given timepoint and the corresponding HI data for its strains, as previously described by Neher et al.
-2016 \cite{Neher:2016hy}.
+To estimate antigenic drift with HI data, we first applied the titer tree model to the phylogeny at a given timepoint and the corresponding HI data for its strains, as previously described by \cite{Neher:2016hy}.
 This method effectively estimates the antigenic drift per branch in units of $log_{2}$ titer change.
@@
 We estimated functional constraint for each strain using either genetic or deep mutational scanning (DMS) data.
-To estimate functional constraint with genetic data, we implemented the non-epitope mutation metric originally defined by {\L}uksza and L\"assig \cite{Luksza:2014hj}.
+To estimate functional constraint with genetic data, we implemented the non-epitope mutation metric originally defined by \cite{Luksza:2014hj}.
 This metric counts the number of amino acid differences at 517 non-epitope sites in HA sequences between each strain $i$ at timepoint $t$ and that strain's most recent inferred ancestral sequence in the previous season ($t - 1$).
 
-We estimated functional constraint using mutational preferences from DMS data as previously defined \cite{Lee2018}.
+We estimated functional constraint using mutational preferences from DMS data as previously defined \citep{Lee2018}.
 Briefly, mutational effects were defined as the log ratio of DMS preferences, $\pi$, at site $r$ for the derived amino acid, $a_{i}$, and the ancestral amino acid, $a_{j}$.
@@
 We estimated clade growth for each strain using local branching index (LBI) and the change in frequency over time (delta frequency).
-To calculate LBI for each strain at each timepoint, we applied the LBI heuristic algorithm as originally described \cite{Neher:2014eu} to the phylogenetic tree constructed at each timepoint.
+To calculate LBI for each strain at each timepoint, we applied the LBI heuristic algorithm as originally described \citep{Neher:2014eu} to the phylogenetic tree constructed at each timepoint.
 We set the neighborhood parameter, $\tau$, to 0.3 and only considered viruses sampled in the last 6 months of each phylogeny as contributing to recent clade growth.
 
-We estimated the change in frequency over time by calculating clade frequencies under a Brownian motion diffusion process as previously described \cite{Lee2018}.
+We estimated the change in frequency over time by calculating clade frequencies under a Brownian motion diffusion process as previously described \citep{Lee2018}.
 These frequency calculations allowed us to assign a partial clade frequency to each strain within nested clades.
@@
 From this tree, we selected all strains with a collection date in the previous two years.
-We calculated the pairwise Hamming distance between the full-length HA amino acid sequences for all selected strains and applied t-SNE dimensionality reduction \cite{vanDerMaaten2008} to the resulting distance matrix (n=2 components, perplexity=30.0, and learning rate=400).
-We assigned each strain to a cluster based on its two-dimensional t-SNE embedding using DBSCAN \cite{Ester1996} with a maximum neighborhood distance of 10 AAs and a minimum of 20 strains per cluster.
+We calculated the pairwise Hamming distance between the full-length HA amino acid sequences for all selected strains and applied t-SNE dimensionality reduction \citep{vanDerMaaten2008} to the resulting distance matrix (n=2 components, perplexity=30.0, and learning rate=400).
+We assigned each strain to a cluster based on its two-dimensional t-SNE embedding using DBSCAN \citep{Ester1996} with a maximum neighborhood distance of 10 AAs and a minimum of 20 strains per cluster.
 Despite known limitations of applying hierarchical clustering to manifold projections that do not preserve sample density, this approach allowed us to effectively assign strains to qualitative genetic clusters for the purposes of visualization.
@@
 
-We gratefully acknowledge the authors, originating and submitting laboratories of the sequences from the GISAID EpiFlu Database \cite{shu2017gisaid} on which this research is based.
+We gratefully acknowledge the authors, originating and submitting laboratories of the sequences from the GISAID EpiFlu Database \citep{shu2017gisaid} on which this research is based.
 The list is detailed in the Supplemental Material.

### 2020-08-03 manuscript/tables/model_summary.tex
Use eLife template and citation styles
Updates most citations to use the parenthetical form, adds the eLife document
type and bibliography style, updates the model summary table to use the full
citation style, and adds a citation for Barrat-Charlaix et al. 2020 which is now
available as a preprint.

@@
 naive                                                        & simulated, natural & HA sequences          & negative control      & this study                      \\
-epitope antigenic novelty                                    & simulated, natural & HA sequences          & antigenic drift       & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
-epitope ancestor                                             & simulated, natural & HA sequences          & antigenic drift       & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
+epitope antigenic novelty                                    & simulated, natural & HA sequences          & antigenic drift       & \cite{Luksza:2014hj} \\
+epitope ancestor                                             & simulated, natural & HA sequences          & antigenic drift       & \cite{Luksza:2014hj} \\
 HI antigenic novelty                                         & natural            & serological assays    & antigenic drift       & this study                      \\
-mutational load                                              & simulated, natural & HA sequences          & functional constraint & {\L}uksza and L\"assig \cite{Luksza:2014hj} \\
-\makecell{deep mutational scanning (DMS) \\ mutational effects} & natural            & DMS assays            & functional constraint & Lee et al. 2018 \cite{Lee2018}              \\
-local branching index (LBI)                                  & simulated, natural & HA sequences          & clade growth          & Neher et al. 2014 \cite{Neher:2014eu}       \\
+mutational load                                              & simulated, natural & HA sequences          & functional constraint & \cite{Luksza:2014hj} \\
+\makecell{deep mutational scanning (DMS) \\ mutational effects} & natural            & DMS assays            & functional constraint & \cite{Lee2018}              \\
+local branching index (LBI)                                  & simulated, natural & HA sequences          & clade growth          & \cite{Neher:2014eu}       \\
 delta frequency                                              & simulated, natural & HA sequences          & clade growth          & this study                      \\

### 2020-08-03 manuscript/supplement.tex
Update citation style for supplemental material

@@
   A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-  The epitope antigenic novelty model relies on previously published epitope sites \cite{Luksza:2014hj}.
+  The epitope antigenic novelty model relies on previously published epitope sites \citep{Luksza:2014hj}.
   The ``oracle'' antigenic novelty model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
@@
     Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
-    Epitope sites were defined previously described \cite{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Epitope sites were defined previously described \citep{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \citep{Bedford:2015fj}.
     Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
@@
     Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
-    Epitope sites were defined previously described \cite{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \cite{Bedford:2015fj}.
+    Epitope sites were defined previously described \citep{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \citep{Bedford:2015fj}.
     Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.

### 2020-08-04 manuscript/main.tex
Don't force large figures to have their own page
The eLife template handles placement of these figures without our intervention.
Reposition relative vaccine comparison as a figure supplement
Moves the relative vaccine comparison figure from the supplemental material to a
figure supplement, to make comparison between the absolute and relative
distances easier.

@@
 
-\begin{figure}[p]
+\begin{figure}
   \begin{center}
@@
 \begin{figure}[htb]
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/vaccine-comparison.pdf}
@@
   \label{fig:vaccine_comparison}
-  \end{center}
+
+  \figsupp[Relative improvement of model selections over vaccine strains.]{
+    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Fig.~\ref{fig:vaccine_comparison}.
+    Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.}{\includegraphics[width=\textwidth]{figures/vaccine-comparison-relative-distance.pdf}}\label{figsupp:vaccine_comparison_relative_distance}
 \end{figure}
@@
 The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
-On average, the strains selected by this model were closer to future than the vaccine strain by 1.93 AAs (Supplemental Fig.~\ref{sup_fig:vaccine_comparison_relative_distance}).
+On average, the strains selected by this model were closer to future than the vaccine strain by 1.93 AAs (Fig.~\ref{fig:vaccine_comparison}-Figure supplement~\ref{figsupp:vaccine_comparison_relative_distance}).
 For the two seasons that the model selected more distant strains than the vaccine strain, the mean distance relative to the vaccine strain was 1.58 AAs.

### 2020-08-04 manuscript/supplement.tex
Reposition relative vaccine comparison as a figure supplement
Moves the relative vaccine comparison figure from the supplemental material to a
figure supplement, to make comparison between the absolute and relative
distances easier.

@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/vaccine-comparison-relative-distance.pdf}
-  \caption{
-    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Fig.~\ref{fig:vaccine_comparison}.
-    Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.
-  }
-  \label{sup_fig:vaccine_comparison_relative_distance}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.pdf}

### 2020-08-04 manuscript/main.tex
Update SciPy citation and simply Potter et al. bibtex entry

@@
 We calculated EMD with the Python bindings for the OpenCV 3.4.1 implementation \citep{opencv_library}.
-We applied the Nelder-Mead minimization algorithm as implemented in SciPy \citep{SciPy} to learn fitness coefficients that minimize the average of this distance metric over all timepoints in a given training window.
+We applied the Nelder-Mead minimization algorithm as implemented in SciPy \citep{virtanen_scipy_2020} to learn fitness coefficients that minimize the average of this distance metric over all timepoints in a given training window.
 

### 2020-08-04 manuscript/main.tex
Expand figure and equation titles (e.g., "Fig." to Figure)

@@
 We developed a framework to forecast seasonal influenza evolution inspired by the Malthusian growth fitness model of \cite{Luksza:2014hj}.
-As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus strain an exponential growth factor scaled by an estimate of the strain's fitness (Fig.~\ref{fig:model} and Eq.~\ref{equation_exponential_growth_model}).
+As with this original model, we forecasted the frequencies of viral populations one year in advance by applying to each virus strain an exponential growth factor scaled by an estimate of the strain's fitness (Figure~\ref{fig:model} and Equation~\ref{equation_exponential_growth_model}).
 \cite{Luksza:2014hj} measured model performance by identifying clades -- groups of strains that all share a recent common ancestor -- and comparing observed and estimated future clade frequencies.
@@
 We calculated the ground distance between individuals as the Hamming distance between haplotypes.
-With this implementation, more accurate projections of the future population's composition produce smaller earth mover's distances between the observed and estimated future (Fig.~\ref{fig:model}).
+With this implementation, more accurate projections of the future population's composition produce smaller earth mover's distances between the observed and estimated future (Figure~\ref{fig:model}).
 
@@
 We fit models for individual fitness metrics and combinations of metrics that we anticipated would be mutually beneficial.
-For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Eq.~\ref{equation_fitness_estimation}).
-We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Supplemental Figs~\ref{sup_fig:cross_validation_for_simulated_populations} and \ref{sup_fig:cross_validation_for_natural_populations}).
+For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Equation~\ref{equation_fitness_estimation}).
+We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Supplemental Figures~\ref{sup_fig:cross_validation_for_simulated_populations} and \ref{sup_fig:cross_validation_for_natural_populations}).
 The earth mover's distance to the future can never be zero, because each model makes predictions based on sequences available at the time of prediction and cannot account for new mutations that occur during the prediction interval.
@@
 The long-term evolution of influenza H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \citep{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Fig.~\ref{sup_fig:cross_validation_for_simulated_populations}).
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Figure~\ref{sup_fig:cross_validation_for_simulated_populations}).
 These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural H3N2 HA (Supplemental Fig.
@@
       Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3}).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Figure~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3}).
       The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
@@
 
-As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 (97\%) of 33 timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
+As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 (97\%) of 33 timepoints (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
 Although the true fitness model performed better than the naive model's average distance of 8.97 $\pm$ 1.35 AAs, it did not reach the closest possible distance between populations of 4.57 $\pm$ 0.61 AAs.
-With the exception of epitope antigenic novelty, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
+With the exception of epitope antigenic novelty, all biologically-informed models consistently outperformed the naive model (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
 LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
@@
     Simulated population model coefficients and distances to the future for individual biologically-informed fitness metrics and the best composite model.
-    A) Coefficients and B) distances are shown per validation and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    A) Coefficients and B) distances are shown per validation and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }
@@
 
-Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \citep{Luksza:2014hj} that are not explicitly present in our simulations.
@@
 This linear fitness metric slightly outperformed the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
-Importantly, a composite model of the epitope ancestor and mutational load metrics outperformed all other epitope-based models and the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Importantly, a composite model of the epitope ancestor and mutational load metrics outperformed all other epitope-based models and the individual mutational load model (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated strains was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
@@
 To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
-This composite model outperformed both of its corresponding individual metric models with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}, Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
+This composite model outperformed both of its corresponding individual metric models with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}, Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
 The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_simulated_populations}.
+  The corresponding results for the naive model are shown in Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_simulated_populations}.
   }
@@
 First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed fold change in clade frequencies, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$ and estimated fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$.
-The model's estimated fold changes correlated well with observed fold changes (Pearson's $R^2 = 0.52$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+The model's estimated fold changes correlated well with observed fold changes (Pearson's $R^2 = 0.52$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
 The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
-Model forecasts were increasingly more accurate with increasing initial clade frequencies (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}C).
+Model forecasts were increasingly more accurate with increasing initial clade frequencies (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}C).
 Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
-The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
-Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
-In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}B).
-However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}C and D).
+The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
+Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
+In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}B).
+However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}C and D).
 These results suggested that estimating a single closest strain to the future is a more tractable problem than estimating the future frequencies of clades.
@@
 A standard expectation from machine learning is that models will perform worse on test data due to overfitting to training data.
-Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
+Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
 The composite model of mutational load and LBI appeared to outperform the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively.
@@
 As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
-While observed and estimated clade frequency fold changes correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
-We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
-The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
-Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
-The naive model performed comparatively well on these test data with all its estimated closest strains to the future in the top 20th percentile and a slightly higher correlation between observed and estimated percentile ranks than the true fitness model (Spearman's $\rho^2 = 0.82$, Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_simulated_populations}).
+While observed and estimated clade frequency fold changes correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Figure~\ref{fig:test_of_best_model_for_simulated_populations}A).
+We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Figure~\ref{fig:test_of_best_model_for_simulated_populations}C).
+The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}B).
+Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}D).
+The naive model performed comparatively well on these test data with all its estimated closest strains to the future in the top 20th percentile and a slightly higher correlation between observed and estimated percentile ranks than the true fitness model (Spearman's $\rho^2 = 0.82$, Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_simulated_populations}).
 These results confirmed that our approach of minimizing the distance between yearly populations could simultaneously capture clade-level dynamics of simulated influenza populations and identify individual strains that are most representative of future populations.
@@
       Natural population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Figure~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
       Validation results are based on 23 timepoints.
@@
 Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
-We held out strains collected after October 1, 2015 up through October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
+We held out strains collected after October 1, 2015 up through October 1, 2019 for model testing (Supplemental Figure~\ref{sup_fig:cross_validation_for_natural_populations}).
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI antigenic novelty and DMS mutational effects.
@@
     Natural population model coefficients and distances to the future for individual biologically-informed fitness metrics.
-    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
     The naive model's distance to the future (light gray) was 6.40 $\pm$ 1.36 AAs for validation timepoints and 6.82 $\pm$ 1.74 AAs for test timepoints.
@@
 
-Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope antigenic novelty and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations} and Table~\ref{table_natural_model_selection}).
+Biologically-informed metrics generally performed better than the naive model with the exceptions of the epitope antigenic novelty and DMS mutational effects (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations} and Table~\ref{table_natural_model_selection}).
 The naive model estimated an average distance between natural H3N2 populations of 6.40 $\pm$ 1.36 AAs.
@@
 HI antigenic novelty estimated an average distance to the future of 6.01 $\pm$ 1.50 AAs and outperformed the naive model at 16 of 23 timepoints (70\%).
-The coefficient for HI antigenic novelty remained stable across all timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+The coefficient for HI antigenic novelty remained stable across all timepoints (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
 In contrast, epitope antigenic novelty estimated a distance of 7.13 $\pm$ 1.47 AAs and only outperformed the naive model at seven timepoints (30\%).
@@
 We fit an antigenic novelty model to these 25 sites across the complete validation period and dubbed this the ``oracle antigenic novelty'' model, as it benefited from knowledge of the future in its forecasts.
-The oracle model produced a consistently positive coefficient across all training windows (0.80 $\pm$ 0.21) and consistently outperformed the original epitope model with an average distance to the future of 5.71 $\pm$ 1.27 AAs (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
+The oracle model produced a consistently positive coefficient across all training windows (0.80 $\pm$ 0.21) and consistently outperformed the original epitope model with an average distance to the future of 5.71 $\pm$ 1.27 AAs (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
 These results support our hypothesis that the fitness benefit of mutations at the original 49 sites was due to historical contingency and that the success of previous epitope models based on these sites was partly due to ``borrowing from the future''.
@@
 
-LBI was the best individual metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
+LBI was the best individual metric by average distance to the future (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied mutational load by outperforming the naive model at 17 (74\%) timepoints (Table~\ref{table_natural_model_selection}).
 Delta frequency performed worse than LBI and HI antigenic novelty and was comparable to mutational load.
@@
     Natural population model coefficients and distances to the future for composite fitness metrics.
-    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }
@@
 
-Both two-metric composite models modestly outperformed their corresponding individual models (Table~\ref{table_natural_model_selection}, Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}, and Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
+Both two-metric composite models modestly outperformed their corresponding individual models (Table~\ref{table_natural_model_selection}, Figure~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}, and Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
 The composite of mutational load and LBI performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
@@
 As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade frequency fold changes and the ranking of estimated best strains compared to the observed closest strains to future populations.
-The composite model of mutational load and LBI effectively captured clade dynamics with a fold change correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
-Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}C).
-The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}B).
-This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_natural_populations}D).
-The naive model's performance repeated the pattern we observed with simulated populations: it made poor forecasts of absolute clade frequencies, but its estimated closest strains to the future were consistently highly ranked among the observed closest strains (Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_natural_populations}B and C).
+The composite model of mutational load and LBI effectively captured clade dynamics with a fold change correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
+Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}C).
+The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}B).
+This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}D).
+The naive model's performance repeated the pattern we observed with simulated populations: it made poor forecasts of absolute clade frequencies, but its estimated closest strains to the future were consistently highly ranked among the observed closest strains (Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_natural_populations}B and C).
 
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_natural_populations}.
+  The corresponding results for the naive model are shown in Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_natural_populations}.
   }
@@
 The individual LBI model was the worst model, while mutational load continued to perform well with test data.
-LBI performed especially poorly in the last two test timepoints of April and October 2018 (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
+LBI performed especially poorly in the last two test timepoints of April and October 2018 (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
 These timepoints correspond to the dominance and sudden decline of a reassortant clade named A2/re \citep{Potter2019}.
@@
 After identifying the composite HI antigenic novelty and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
-The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.46$ between observed and estimated growth ratios and growth and decline accuracies of 52\% and 58\%, respectively (Fig.~\ref{fig:test_of_best_model_for_natural_populations}A).
-The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Fig.~\ref{fig:test_of_best_model_for_natural_populations}C).
-The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}B).
-Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.72$) across all strains and test timepoints (Fig.~\ref{fig:test_of_best_model_for_natural_populations}D).
-The estimated strain ranks of the naive model were not as well correlated (Spearman's $\rho^2 = 0.56$), but seven of its eight estimates for the closest strain to the future (88\%) were in the top fifth percentile of observed closest strains (Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_natural_populations}B and D).
+The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.46$ between observed and estimated growth ratios and growth and decline accuracies of 52\% and 58\%, respectively (Figure~\ref{fig:test_of_best_model_for_natural_populations}A).
+The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Figure~\ref{fig:test_of_best_model_for_natural_populations}C).
+The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}B).
+Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.72$) across all strains and test timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}D).
+The estimated strain ranks of the naive model were not as well correlated (Spearman's $\rho^2 = 0.56$), but seven of its eight estimates for the closest strain to the future (88\%) were in the top fifth percentile of observed closest strains (Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_natural_populations}B and D).
 
@@
   \figsupp[Relative improvement of model selections over vaccine strains.]{
-    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Fig.~\ref{fig:vaccine_comparison}.
+    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Figure~\ref{fig:vaccine_comparison}.
     Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.}{\includegraphics[width=\textwidth]{figures/vaccine-comparison-relative-distance.pdf}}\label{figsupp:vaccine_comparison_relative_distance}
@@
 We compared these distances to those of the closest strains to the future as estimated by our best models for the validation period (mutational load and LBI) and the test period (HI antigenic novelty and mutational load).
-The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
-On average, the strains selected by this model were closer to future than the vaccine strain by 1.93 AAs (Fig.~\ref{fig:vaccine_comparison}-Figure supplement~\ref{figsupp:vaccine_comparison_relative_distance}).
+The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Figure~\ref{fig:vaccine_comparison}).
+On average, the strains selected by this model were closer to future than the vaccine strain by 1.93 AAs (Figure~\ref{fig:vaccine_comparison}-Figure supplement~\ref{figsupp:vaccine_comparison_relative_distance}).
 For the two seasons that the model selected more distant strains than the vaccine strain, the mean distance relative to the vaccine strain was 1.58 AAs.
@@
 
-Two of three models performed worse after refitting coefficients to the test data than their original fixed coefficient implementations (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
+Two of three models performed worse after refitting coefficients to the test data than their original fixed coefficient implementations (Supplemental Figure~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
 While, the mutational load and LBI model improved considerably over its original performance, it still performed worse than the naive model on average.
@@
 Consistent with an average two-month lag between data collection and submission, the most recent data were collected up to April 1, 2020 and made our forecasts from this timepoint to April 1, 2021.
-Of the five major currently circulating clades, our model predicted growth of the clades 3c3.A and A1b/94N and decline of clades A1b/135K, A1b/137F, and A1b/197R (Fig.~\ref{fig:nextstrain_forecasts}).
-To aid with identification of potential vaccine candidates for the next season, we annotated strains in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).
+Of the five major currently circulating clades, our model predicted growth of the clades 3c3.A and A1b/94N and decline of clades A1b/135K, A1b/137F, and A1b/197R (Figure~\ref{fig:nextstrain_forecasts}).
+To aid with identification of potential vaccine candidates for the next season, we annotated strains in the phylogeny by their estimated distance to the future based on our best model (Figure~\ref{fig:nextstrain_distance_to_future}).
 
@@
 These results are consistent with previous work that found improved performance by integrating components of antigenic drift, functional constraint, and clade growth \citep{Luksza:2014hj}.
-However, the effective elimination of LBI from our three-metric model during the validation period (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}) reveals the limitations of our current additive approach to composite models.
+However, the effective elimination of LBI from our three-metric model during the validation period (Figure~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}) reveals the limitations of our current additive approach to composite models.
 The recent success of weighted ensembles for short-term influenza forecasting through the CDC's FluSight network \citep{Reich:2019bq} suggests that long-term forecasting may benefit from a similar approach.
@@
 We measured the effect size of each comparison as the mean $\pm$ the standard deviation of the bootstrap distributions.
-We performed pairwise model comparisons for all biologically-informed models against the naive model (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+We performed pairwise model comparisons for all biologically-informed models against the naive model (Supplemental Figures~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
 We also compared a subset of composite models to their respective individual models (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
@@
 
-For the purpose of visualizing related amino acid sequences in Fig.~\ref{fig:model}, we applied dimensionality reduction to pairwise amino acid distances followed by hierarchical clustering.
+For the purpose of visualizing related amino acid sequences in Figure~\ref{fig:model}, we applied dimensionality reduction to pairwise amino acid distances followed by hierarchical clustering.
 Specifically, we selected a representative tree from our simulated population of viruses at month 10 of year 30.

### 2020-08-04 manuscript/supplement.tex
Expand figure and equation titles (e.g., "Fig." to Figure)

@@
   Composite model coefficients and distances to the future for models fit to simulated populations.
-  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   }
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}.
+  The corresponding results for the naive model are shown in Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}.
   }
@@
   \caption{
-  Validation of naive model for simulated populations of H3N2-like viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Validation of naive model for simulated populations of H3N2-like viruses as in Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
   Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
@@
   \caption{
-  Test of naive model for simulated populations of H3N2-like viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Test of naive model for simulated populations of H3N2-like viruses as in Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
   Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
@@
   Model coefficients and distances to the future for antigenic novelty models fit to natural populations.
-  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
   The epitope antigenic novelty model relies on previously published epitope sites \citep{Luksza:2014hj}.
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_natural_populations}.
+  The corresponding results for the naive model are shown in Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_natural_populations}.
   }
@@
   \caption{
-  Validation of naive model for natural populations of H3N2 viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Validation of naive model for natural populations of H3N2 viruses as in Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
   Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
@@
   \caption{
-  Test of naive model for natural populations of H3N2 viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
+  Test of naive model for natural populations of H3N2 viruses as in Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
   Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
@@
   \caption{
-    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
     A) Coefficients and B) distances are shown per test timepoint (N=8).
@@
       All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Figure~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
       Distances annotated with carets ($\wedge$) were not tested for significance relative to the naive model.

### 2020-08-04 manuscript/main.tex
Move supplemental figures and tables into main text
Promotes some supplemental figures to main figures and moves others to figure
supplements for the appropriate main figures. Also promotes most supplemental
tables to main tables.

@@
 For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Equation~\ref{equation_fitness_estimation}).
-We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Supplemental Figures~\ref{sup_fig:cross_validation_for_simulated_populations} and \ref{sup_fig:cross_validation_for_natural_populations}).
+We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Figures~\ref{fig:cross_validation_for_simulated_populations} and \ref{fig:cross_validation_for_natural_populations}).
 The earth mover's distance to the future can never be zero, because each model makes predictions based on sequences available at the time of prediction and cannot account for new mutations that occur during the prediction interval.
@@
 The long-term evolution of influenza H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \citep{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
-To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Figure~\ref{sup_fig:cross_validation_for_simulated_populations}).
-These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural H3N2 HA (Supplemental Fig.
-\ref{sup_fig:simulated_h3n2_ha_phylogeny}, Supplemental Tables~\ref{sup_table:mutations_by_trunk_status_for_simulated_populations} and \ref{sup_table:mutations_by_trunk_status}).
+To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Figure~\ref{fig:cross_validation_for_simulated_populations}).
+These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural H3N2 HA (Figure \ref{fig:simulated_h3n2_ha_phylogeny}, Tables~\ref{table:mutations_by_trunk_status_for_simulated_populations} and \ref{table:mutations_by_trunk_status}).
 We fit models to these simulated populations using all sequence-only fitness metrics.
@@
       Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Figure~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3}).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Figure~\ref{fig:bootstrap_distributions_for_simulated_sample_3}).
       The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
@@
 \begin{figure}[htb]
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.pdf}
@@
   \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
-  \end{center}
+
+  \figsupp[Composite model coefficients and distances to the future for models fit to simulated populations.]{
+    A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.}{\includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}
+
 \end{figure}
 
-Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}-Figure supplement~\ref{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \citep{Luksza:2014hj} that are not explicitly present in our simulations.
@@
 This linear fitness metric slightly outperformed the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
-Importantly, a composite model of the epitope ancestor and mutational load metrics outperformed all other epitope-based models and the individual mutational load model (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
+Importantly, a composite model of the epitope ancestor and mutational load metrics outperformed all other epitope-based models and the individual mutational load model (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}-Figure supplement~\ref{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
 From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated strains was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.
@@
 To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
-This composite model outperformed both of its corresponding individual metric models with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}, Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
+This composite model outperformed both of its corresponding individual metric models with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}, Table~\ref{table:composite_vs_individual_model_comparison}).
 The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
@@
 \begin{figure}
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-normalized_fitness.pdf}
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_simulated_populations}.
+  The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:test_of_naive_model_for_simulated_populations}.
   }
   \label{fig:test_of_best_model_for_simulated_populations}
-  \end{center}
+
+  \figsupp[Validation of best model for simulated populations of H3N2-like viruses.]{
+    A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+    B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
+  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
+  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+  The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-normalized_fitness.pdf}}\label{figsupp:validation_of_best_model_for_simulated_populations}
+
+  \figsupp[Validation of naive model for simulated populations of H3N2-like viruses.]{
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-naive.pdf}}\label{figsupp:validation_of_naive_model_for_simulated_populations}
+
+  \figsupp[Test of naive model for simulated populations of H3N2-like viruses.]{
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-naive.pdf}}\label{figsupp:test_of_naive_model_for_simulated_populations}
+
 \end{figure}
@@
 First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed fold change in clade frequencies, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$ and estimated fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$.
-The model's estimated fold changes correlated well with observed fold changes (Pearson's $R^2 = 0.52$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
+The model's estimated fold changes correlated well with observed fold changes (Pearson's $R^2 = 0.52$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}A).
 The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
-Model forecasts were increasingly more accurate with increasing initial clade frequencies (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}C).
+Model forecasts were increasingly more accurate with increasing initial clade frequencies (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}C).
 Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
-The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
-Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
-In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}B).
-However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}C and D).
+The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}B).
+Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}D).
+In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}B).
+However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}C and D).
 These results suggested that estimating a single closest strain to the future is a more tractable problem than estimating the future frequencies of clades.
@@
 A standard expectation from machine learning is that models will perform worse on test data due to overfitting to training data.
-Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
+Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}-Figure supplement~\ref{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
 The composite model of mutational load and LBI appeared to outperform the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively.
-However, we did not find a significant difference between these models by bootstrap testing (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}) and could not rule out fluctuations in model performance across a relatively small number of data points.
+However, we did not find a significant difference between these models by bootstrap testing (Table~\ref{table:composite_vs_individual_model_comparison}) and could not rule out fluctuations in model performance across a relatively small number of data points.
 
@@
 While observed and estimated clade frequency fold changes correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Figure~\ref{fig:test_of_best_model_for_simulated_populations}A).
-We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Figure~\ref{fig:test_of_best_model_for_simulated_populations}C).
+We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Figure~\ref{fig:test_of_best_model_for_simulated_populations}C).
 The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}B).
 Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}D).
-The naive model performed comparatively well on these test data with all its estimated closest strains to the future in the top 20th percentile and a slightly higher correlation between observed and estimated percentile ranks than the true fitness model (Spearman's $\rho^2 = 0.82$, Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_simulated_populations}).
+The naive model performed comparatively well on these test data with all its estimated closest strains to the future in the top 20th percentile and a slightly higher correlation between observed and estimated percentile ranks than the true fitness model (Spearman's $\rho^2 = 0.82$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:test_of_naive_model_for_simulated_populations}).
 These results confirmed that our approach of minimizing the distance between yearly populations could simultaneously capture clade-level dynamics of simulated influenza populations and identify individual strains that are most representative of future populations.
@@
       Natural population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Figure~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Figure~\ref{fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
       Validation results are based on 23 timepoints.
@@
 Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
-We held out strains collected after October 1, 2015 up through October 1, 2019 for model testing (Supplemental Figure~\ref{sup_fig:cross_validation_for_natural_populations}).
+We held out strains collected after October 1, 2015 up through October 1, 2019 for model testing (Figure~\ref{fig:cross_validation_for_natural_populations}).
 In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI antigenic novelty and DMS mutational effects.
@@
 \begin{figure}[htb]
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
@@
   \label{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}
-  \end{center}
+
+  \figsupp[Comparison of epitope-based models with knowledge of the future.]{
+    Model coefficients and distances to the future for antigenic novelty models fit to natural populations.
+    A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    The epitope antigenic novelty model relies on previously published epitope sites \citep{Luksza:2014hj}.
+    The ``oracle'' antigenic novelty model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
+    The improved performance of the ``oracle'' model indicates that the sequence-based antigenic novelty metric can be effective when sites of beneficial mutations are known prior to forecasting.}{\includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
+
 \end{figure}
@@
 This strong coefficient for the first half of training windows indicated that, unlike the results for simulated populations, the nonlinear antigenic novelty metric was historically an effective measure of antigenic drift.
-The historical importance of the epitope sites used for this metric was further supported by the relative enrichment of mutations at these sites for the most successful ``trunk'' lineages of natural populations compared to side branch lineages (Supplemental Table~\ref{sup_table:mutations_by_trunk_status}).
+The historical importance of the epitope sites used for this metric was further supported by the relative enrichment of mutations at these sites for the most successful ``trunk'' lineages of natural populations compared to side branch lineages (Table~\ref{table:mutations_by_trunk_status}).
 
@@
 We fit an antigenic novelty model to these 25 sites across the complete validation period and dubbed this the ``oracle antigenic novelty'' model, as it benefited from knowledge of the future in its forecasts.
-The oracle model produced a consistently positive coefficient across all training windows (0.80 $\pm$ 0.21) and consistently outperformed the original epitope model with an average distance to the future of 5.71 $\pm$ 1.27 AAs (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
+The oracle model produced a consistently positive coefficient across all training windows (0.80 $\pm$ 0.21) and consistently outperformed the original epitope model with an average distance to the future of 5.71 $\pm$ 1.27 AAs (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}-Figure supplement~\ref{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}).
 These results support our hypothesis that the fitness benefit of mutations at the original 49 sites was due to historical contingency and that the success of previous epitope models based on these sites was partly due to ``borrowing from the future''.
@@
 \begin{figure}[htb]
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/best-composite-unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
@@
   \label{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}
-  \end{center}
+
+  \figsupp[Composite models fit to most recent data from natural populations.]{
+    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
+    A) Coefficients and B) distances are shown per test timepoint (N=8).
+    In contrast to the results for these models based on fixed coefficients from training/validation, these coefficients were learned for each six-year window prior to the corresponding test timepoint.
+    The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
+    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.}{\includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.pdf}}\label{figsupp:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
+
 \end{figure}
@@
 
-Both two-metric composite models modestly outperformed their corresponding individual models (Table~\ref{table_natural_model_selection}, Figure~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}, and Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
+Both two-metric composite models modestly outperformed their corresponding individual models (Table~\ref{table_natural_model_selection}, Figure~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}, and Table~\ref{table:composite_vs_individual_model_comparison}).
 The composite of mutational load and LBI performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.
@@
 As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade frequency fold changes and the ranking of estimated best strains compared to the observed closest strains to future populations.
-The composite model of mutational load and LBI effectively captured clade dynamics with a fold change correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}A).
-Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}C).
-The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}B).
-This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_natural_populations}D).
-The naive model's performance repeated the pattern we observed with simulated populations: it made poor forecasts of absolute clade frequencies, but its estimated closest strains to the future were consistently highly ranked among the observed closest strains (Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_natural_populations}B and C).
+The composite model of mutational load and LBI effectively captured clade dynamics with a fold change correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}A).
+Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}C).
+The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}B).
+This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}D).
+The naive model's performance repeated the pattern we observed with simulated populations: it made poor forecasts of absolute clade frequencies, but its estimated closest strains to the future were consistently highly ranked among the observed closest strains (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_natural_populations}B and C).
 
 \begin{figure}
-  \begin{center}
   \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-cTiter_x-ne_star.pdf}
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_natural_populations}.
+  The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:test_of_naive_model_for_natural_populations}.
   }
   \label{fig:test_of_best_model_for_natural_populations}
-  \end{center}
+
+  \figsupp[Validation of best model for natural populations of H3N2 viruses.]{
+    Validation results for the composite model of mutational load and LBI.
+    A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
+    B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.
+    The estimated best strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+    C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
+    D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
+    The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_natural_populations}.}{\includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-ne_star-lbi.pdf}}\label{figsupp:validation_of_best_model_for_natural_populations}
+
+  \figsupp[Validation of naive model for natural populations of H3N2 viruses.]{
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-naive.pdf}}\label{figsupp:validation_of_naive_model_for_natural_populations}
+
+  \figsupp[Test of naive model for natural populations of H3N2 viruses.]{
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-naive.pdf}}\label{figsupp:test_of_naive_model_for_natural_populations}
+
 \end{figure}
@@
 Correspondingly, only the three models with the HI antigenic novelty metric significantly outperformed the naive model on the test data (Table~\ref{table_natural_model_selection}).
-The composite of HI antigenic novelty and mutational load performed modestly, although not significantly, better than the individual HI antigenic novelty model (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
+The composite of HI antigenic novelty and mutational load performed modestly, although not significantly, better than the individual HI antigenic novelty model (Table~\ref{table:composite_vs_individual_model_comparison}).
 Surprisingly, the best model for the validation data -- mutational load and LBI -- was one of the worst models for the test data with an average distance to the future of 7.70 $\pm$ 3.53 AAs.
@@
 Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.72$) across all strains and test timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}D).
-The estimated strain ranks of the naive model were not as well correlated (Spearman's $\rho^2 = 0.56$), but seven of its eight estimates for the closest strain to the future (88\%) were in the top fifth percentile of observed closest strains (Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_natural_populations}B and D).
+The estimated strain ranks of the naive model were not as well correlated (Spearman's $\rho^2 = 0.56$), but seven of its eight estimates for the closest strain to the future (88\%) were in the top fifth percentile of observed closest strains (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:test_of_naive_model_for_natural_populations}B and D).
 
@@
 
-Two of three models performed worse after refitting coefficients to the test data than their original fixed coefficient implementations (Supplemental Figure~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
+Two of three models performed worse after refitting coefficients to the test data than their original fixed coefficient implementations (Figure~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}-Figure supplement~\ref{figsupp:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
 While, the mutational load and LBI model improved considerably over its original performance, it still performed worse than the naive model on average.
@@
 We simulated the long-term evolution of H3N2-like viruses with SANTA-SIM \citep{Jariani2019} for 10,000 generations or 50 years where 200 generations was equivalent to 1 year.
-We discarded the first 10 years as a burn-in period, selected the next 30 years for model fitting and validation, and held out the last 9 years as out-of-sample data for model testing.
+We discarded the first 10 years as a burn-in period, selected the next 30 years for model fitting and validation, and held out the last 9 years as out-of-sample data for model testing (Figure~\ref{fig:cross_validation_for_simulated_populations}).
 Each simulated population was seeded with the full length HA from A/Beijing/32/1992 (NCBI accession: U26830.1) such that all simulated sequences contained signal peptide, HA1, and HA2 domains.
@@
 For tuning of hyperparameters, we sampled 10 viruses per month to enable rapid exploration of hyperparameter space.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/cross-validation-for-simulated-populations.pdf}
+  \caption{
+  Time-series cross-validation scheme for simulated populations.
+  Models were trained in six-year sliding windows (gray lines) and validated on out-of-sample data from validation timepoints (filled circles).
+  Validation results from 30 years of data were used to iteratively tune model hyperparameters.
+  After fixing hyperparameters, model coefficients were fixed at the mean values across all training windows.
+  Fixed coefficients were applied to 9 years of new out-of-sample test data (open circles) to estimate true forecast errors.
+  }
+  \label{fig:cross_validation_for_simulated_populations}
+  \end{center}
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/simulated-h3n2-ha-phylogeny.png}
+  \caption{
+  Phylogeny of H3N2-like HA sequences sampled between the 24th and 30th years of simulated evolution.
+  The phylogenetic structure and rate of accumulated epitope and non-epitope mutations match patterns observed in phylogenies of natural sequences.
+  Sample dates were annotated as the generation in the simulation divided by 200 and added to 2000, to acquire realistic date ranges that were compatible with our modeling machinery.
+  }
+  \label{fig:simulated_h3n2_ha_phylogeny}
+  \end{center}
+\end{figure}
+
+\begin{table}
+  \begin{center}
+    \input{tables/mutations_by_trunk_status_for_simulated_populations.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
+    Epitope sites were defined previously described \citep{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \citep{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
+    }
+    \label{table:mutations_by_trunk_status_for_simulated_populations}
+  \end{center}
+\end{table}
+
+\begin{table}
+  \begin{center}
+    \input{tables/mutations_by_trunk_status.tex}
+    \caption{
+    Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
+    Epitope sites were defined previously described \citep{Luksza:2014hj}.
+    Annotation of trunk and side branch was performed as previously described \citep{Bedford:2015fj}.
+    Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.
+    }
+    \label{table:mutations_by_trunk_status}
+  \end{center}
+\end{table}
 
@@
 To obtain unbiased estimates for the out-of-sample errors of our models, we adopted the standard cross-validation strategy of training, validation, and testing.
-We divided our available data into an initial training and validation set spanning October 1990 to October 2015 and an additional testing set spanning October 2015 to October 2019.
+We divided our available data into an initial training and validation set spanning October 1990 to October 2015 and an additional testing set spanning October 2015 to October 2019 (Figure~\ref{fig:cross_validation_for_natural_populations}).
 We partitioned our training and validation data into six month seasons corresponding to winter in the Northern Hemisphere (October--April) and the Southern Hemisphere (April--October) and trained models to estimate frequencies of populations one year into the future from each season in six-year sliding windows.
@@
 Finally, we fixed the coefficients for each model at the mean values across all training windows and applied these fixed models to the test data to estimate the true forecasting accuracy of each model on previously unobserved data.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/cross-validation-for-natural-populations.pdf}
+  \caption{
+  Time-series cross-validation scheme for natural populations.
+  Models were trained in six-year sliding windows (gray lines) and validated on out-of-sample data from validation timepoints (filled circles).
+  Validation results from 25 years of data were used to iteratively tune model hyperparameters.
+  After fixing hyperparameters, model coefficients were fixed at the mean values across all training windows.
+  Fixed coefficients were applied to four years of new out-of-sample test data (open circles) to estimate true forecast errors.
+  }
+  \label{fig:cross_validation_for_natural_populations}
+  \end{center}
+\end{figure}
 
@@
 We measured the effect size of each comparison as the mean $\pm$ the standard deviation of the bootstrap distributions.
-We performed pairwise model comparisons for all biologically-informed models against the naive model (Supplemental Figures~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
-We also compared a subset of composite models to their respective individual models (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
+We performed pairwise model comparisons for all biologically-informed models against the naive model (Figures~\ref{fig:bootstrap_distributions_for_simulated_sample_3} and \ref{fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+We also compared a subset of composite models to their respective individual models (Table~\ref{table:composite_vs_individual_model_comparison}).
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_simulated_sample_3.pdf}
+  \caption{
+  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for simulated populations.
+  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
+  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
+  Each model is annotated by the mean $\pm$ the standard deviation of the bootstrap distribution.
+  Models are also annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
+  }
+  \label{fig:bootstrap_distributions_for_simulated_sample_3}
+  \end{center}
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding.pdf}
+  \caption{
+  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for natural populations.
+  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
+  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
+  Each model is annotated by the mean $\pm$ the standard deviation of the bootstrap distribution.
+  Models are also annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
+  }
+  \label{fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}
+  \end{center}
+\end{figure}
+
+\begin{table}[H]
+  \begin{center}
+    \scalebox{0.7}{
+        \input{tables/composite_vs_individual_model_comparison.tex}
+    }
+    \caption{
+      Comparison of composite and individual model distances to the future by bootstrap test (see Methods).
+      The effect size of differences between models in amino acids is given by the mean and standard deviation of the bootstrap distributions.
+      The p values represent the proportion of n=10,000 bootstrap samples where the mean difference was greater than or equal to zero.
+    }
+    \label{table:composite_vs_individual_model_comparison}
+  \end{center}
+\end{table}
 

### 2020-08-04 manuscript/supplement.tex
Move supplemental figures and tables into main text
Promotes some supplemental figures to main figures and moves others to figure
supplements for the appropriate main figures. Also promotes most supplemental
tables to main tables.

--- 2020-08-04 manuscript/supplement.tex
@@
 
-\subsection*{Supplemental Figures}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/cross-validation-for-simulated-populations.pdf}
-  \caption{
-  Time-series cross-validation scheme for simulated populations.
-  Models were trained in six-year sliding windows (gray lines) and validated on out-of-sample data from validation timepoints (filled circles).
-  Validation results from 30 years of data were used to iteratively tune model hyperparameters.
-  After fixing hyperparameters, model coefficients were fixed at the mean values across all training windows.
-  Fixed coefficients were applied to 9 years of new out-of-sample test data (open circles) to estimate true forecast errors.
-  }
-  \label{sup_fig:cross_validation_for_simulated_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/simulated-h3n2-ha-phylogeny.png}
-  \caption{
-  Phylogeny of H3N2-like HA sequences sampled between the 24th and 30th years of simulated evolution.
-  The phylogenetic structure and rate of accumulated epitope and non-epitope mutations match patterns observed in phylogenies of natural sequences.
-  Sample dates were annotated as the generation in the simulation divided by 200 and added to 2000, to acquire realistic date ranges that were compatible with our modeling machinery.
-  }
-  \label{sup_fig:simulated_h3n2_ha_phylogeny}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.pdf}
-  \caption{
-  Composite model coefficients and distances to the future for models fit to simulated populations.
-  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-  }
-  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_simulated_sample_3.pdf}
-  \caption{
-  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for simulated populations.
-  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
-  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
-  Each model is annotated by the mean $\pm$ the standard deviation of the bootstrap distribution.
-  Models are also annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
-  }
-  \label{sup_fig:bootstrap_distributions_for_simulated_sample_3}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-normalized_fitness.pdf}
-  \caption{
-  Validation of best model for simulated populations of H3N2-like viruses.
-  A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
-  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
-  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
-  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}.
-  }
-  \label{sup_fig:validation_of_best_model_for_simulated_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-naive.pdf}
-  \caption{
-  Validation of naive model for simulated populations of H3N2-like viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
-  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
-  }
-  \label{sup_fig:validation_of_naive_model_for_simulated_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-naive.pdf}
-  \caption{
-  Test of naive model for simulated populations of H3N2-like viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
-  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
-  }
-  \label{sup_fig:test_of_naive_model_for_simulated_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/cross-validation-for-natural-populations.pdf}
-  \caption{
-  Time-series cross-validation scheme for natural populations.
-  Models were trained in six-year sliding windows (gray lines) and validated on out-of-sample data from validation timepoints (filled circles).
-  Validation results from 25 years of data were used to iteratively tune model hyperparameters.
-  After fixing hyperparameters, model coefficients were fixed at the mean values across all training windows.
-  Fixed coefficients were applied to four years of new out-of-sample test data (open circles) to estimate true forecast errors.
-  }
-  \label{sup_fig:cross_validation_for_natural_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.pdf}
-  \caption{
-  Model coefficients and distances to the future for antigenic novelty models fit to natural populations.
-  A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-  The epitope antigenic novelty model relies on previously published epitope sites \citep{Luksza:2014hj}.
-  The ``oracle'' antigenic novelty model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
-  The improved performance of the ``oracle'' model indicates that the sequence-based antigenic novelty metric can be effective when sites of beneficial mutations are known prior to forecasting.
-  }
-  \label{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding.pdf}
-  \caption{
-  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for natural populations.
-  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
-  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
-  Each model is annotated by the mean $\pm$ the standard deviation of the bootstrap distribution.
-  Models are also annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
-  }
-  \label{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-ne_star-lbi.pdf}
-  \caption{
-  Validation of best model for natural populations of H3N2 viruses, the composite model of mutational load and LBI.
-  A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.
-  The estimated best strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
-  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
-  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:validation_of_naive_model_for_natural_populations}.
-  }
-  \label{sup_fig:validation_of_best_model_for_natural_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-naive.pdf}
-  \caption{
-  Validation of naive model for natural populations of H3N2 viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
-  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
-  }
-  \label{sup_fig:validation_of_naive_model_for_natural_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-naive.pdf}
-  \caption{
-  Test of naive model for natural populations of H3N2 viruses as in Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}.
-  Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.
-  }
-  \label{sup_fig:test_of_naive_model_for_natural_populations}
-  \end{center}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.pdf}
-  \caption{
-    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
-    A) Coefficients and B) distances are shown per test timepoint (N=8).
-    In contrast to the results for these models based on fixed coefficients from training/validation, these coefficients were learned for each six-year window prior to the corresponding test timepoint.
-    The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
-    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.
-  }
-  \label{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
-  \end{center}
-\end{figure}
-
 \subsection*{Supplemental Tables}
-
-\begin{table}[H]
-  \begin{center}
-    \input{tables/mutations_by_trunk_status_for_simulated_populations.tex}
-    \caption{
-    Number of epitope and non-epitope mutations per branch by trunk or side branch status for simulated populations.
-    Epitope sites were defined previously described \citep{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \citep{Bedford:2015fj}.
-    Mutations were calculated for the full validation tree for simulated sequences samples between October of years 10 and 40.
-    }
-    \label{sup_table:mutations_by_trunk_status_for_simulated_populations}
-  \end{center}
-\end{table}
-
-\begin{table}[H]
-  \begin{center}
-    \input{tables/mutations_by_trunk_status.tex}
-    \caption{
-    Number of epitope and non-epitope mutations per branch by trunk or side branch status for natural populations.
-    Epitope sites were defined previously described \citep{Luksza:2014hj}.
-    Annotation of trunk and side branch was performed as previously described \citep{Bedford:2015fj}.
-    Mutations were calculated for the full validation tree for natural sequences samples between 1990 and 2015.
-    }
-    \label{sup_table:mutations_by_trunk_status}
-  \end{center}
-\end{table}
 
@@
       All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Figure~\ref{fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
       Distances annotated with carets ($\wedge$) were not tested for significance relative to the naive model.
@@
     \label{sup_table:complete_natural_model_selection}
-  \end{center}
-\end{table}
-
-\begin{table}[H]
-  \begin{center}
-    \scalebox{0.7}{
-        \input{tables/composite_vs_individual_model_comparison.tex}
-    }
-    \caption{
-      Comparison of composite and individual model distances to the future by bootstrap test (see Methods).
-      The effect size of differences between models in amino acids is given by the mean and standard deviation of the bootstrap distributions.
-      The p values represent the proportion of n=10,000 bootstrap samples where the mean difference was greater than or equal to zero.
-    }
-    \label{sup_table:composite_vs_individual_model_comparison}
   \end{center}

### 2020-08-05 manuscript/main.tex
Fix reference to specific panels in validation figures

@@
 Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}D).
-In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}B).
-However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}C and D).
+In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}C).
+However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}B and D).
 These results suggested that estimating a single closest strain to the future is a more tractable problem than estimating the future frequencies of clades.

### 2020-08-05 manuscript/main.tex
Move complete natural model results table into the methods

@@
 This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
-Alternate implementations of less background-dependent DMS metrics never performed better than the mutational load metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}, Methods).
+Alternate implementations of less background-dependent DMS metrics never performed better than the mutational load metric (Table~\ref{table:complete_natural_model_selection}, Methods).
 Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive biophysical model based on measured mutational effects of a single strain.
@@
 Other attempts to define these so-called ``epitope sites'' have relied on either aggregation of results from antigenic escape assays \citep{Wolf:2006da} or retrospective computational analyses of sites with beneficial mutations \citep{Shih:2007bd,Luksza:2014hj}.
-We found that models based on all of these definitions except for the seven Koel epitope sites overfit to the historical context from which they were identified (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
+We found that models based on all of these definitions except for the seven Koel epitope sites overfit to the historical context from which they were identified (Table~\ref{table:complete_natural_model_selection}).
 These results suggest that the set of sites that contribute to antigenic drift at any given time may depend on both the fitness landscape of currently circulating strains and the immune landscape of the hosts these strains need to infect.
@@
 
+\begin{table}
+  \begin{center}
+    \scalebox{0.9}{
+        \input{tables/complete_natural_model_selection.tex}
+    }
+    \caption{
+      All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
+      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Figure~\ref{fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
+      Distances annotated with carets ($\wedge$) were not tested for significance relative to the naive model.
+      Validation results are based on 23 timepoints.
+      Test results are based on eight timepoints not observed during model training and validation.
+      Model results for additional variants of fitness metrics including those based on epitope mutations and DMS preferences are included for reference.
+    }
+    \label{table:complete_natural_model_selection}
+  \end{center}
+\end{table}
+
 \subsection*{Clustering of amino acid sequences for visualization}

### 2020-08-05 manuscript/supplement.tex
Move complete natural model results table into the methods

@@
 \section*{Supplemental Material}
-
-\subsection*{Supplemental Tables}
-
-\begin{table}[H]
-  \begin{center}
-    \scalebox{0.9}{
-        \input{tables/complete_natural_model_selection.tex}
-    }
-    \caption{
-      All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
-      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Figure~\ref{fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
-      Distances annotated with carets ($\wedge$) were not tested for significance relative to the naive model.
-      Validation results are based on 23 timepoints.
-      Test results are based on eight timepoints not observed during model training and validation.
-      Model results for additional variants of fitness metrics including those based on epitope mutations and DMS preferences are included for reference.
-    }
-    \label{sup_table:complete_natural_model_selection}
-  \end{center}
-\end{table}
 

### 2020-08-06 manuscript/main.tex
Move GISAID acknowledgments to an appendix
Moves acknowledgments into an appendix, moves the supplemental file reference
into the main text, and removes the supplemental text file completely.

@@
 The authors declare that no competing interests exist.
+
+\section*{Supplemental Files}
+
+\textbf{Supplemental File S1.} Tab-delimited GISAID accessions and metadata including originating and submitting labs for natural strains used across all timepoints.
+
+\appendix
+\begin{appendixbox}
+\label{appendix:gisaid_acknowledgments}
+\section{GISAID Acknowledgments}
+
+WHO Collaborating Centre for Reference and Research on Influenza, Victorian Infectious Diseases Reference Laboratory, Australia; WHO Collaborating Centre for Reference and Research on Influenza, Chinese National Influenza Center, China; WHO Collaborating Centre for Reference and Research on Influenza, National Institute of Infectious Diseases, Japan; The Crick Worldwide Influenza Centre, The Francis Crick Institute, United Kingdom; WHO Collaborating Centre for the Surveillance, Epidemiology and Control of Influenza, Centers for Disease Control and Prevention, United States; ADImmune Corporation, Taiwan; ADPH Bureau of Clinical Laboratories, United States; Aichi Prefectural Institute of Public Health, Japan; Akershus University Hospital, Norway; Akita Research Center for Public Health and Environment, Japan; Alabama State Laboratory, United States; Alaska State Public Health Laboratory, United States; Alaska State Virology Lab, United States; Aomori Prefectural Institute of Public Health and Environment, Japan; Aristotelian University of Thessaloniki, Greece; Arizona Department of Health Services, United States; Arkansas Children's Hospital, United States; Arkansas Department of Health, United States; Auckland Healthcare, New Zealand; Auckland Hospital, New Zealand; Austin Health, Australia; Baylor College of Medicine, United States; California Department of Health Services, United States; Canberra Hospital, Australia; Cantacuzino Institute, Romania; Canterbury Health Services, New Zealand; Caribbean Epidemiology Center, Trinidad and Tobago; CDC GAP Nigeria, Nigeria; CDC-Kenya, Kenya; CEMIC University Hospital, Argentina; CENETROP, Bolivia, Plurinationial State of; Center for Disease Control, Taiwan; Center for Public Health and Environment, Hiroshima Prefectural Technology Research Institute, Japan; Central Health Laboratory, Mauritius; Central Laboratory of Public Health, Paraguay; Central Public Health Laboratory, Ministry of Health, Oman; Central Public Health Laboratory, Palestinian Territory; Central Public Health Laboratory, Papua New Guinea; Central Research Institute for Epidemiology, Russian Federation; Centre for Diseases Control and Prevention, Armenia; Centre for Infections, Health Protection Agency, United Kingdom; Centre Pasteur du Cameroun, Cameroon; Chiba City Institute of Health and Environment, Japan; Chiba Prefectural Institute of Public Health, Japan; Childrens Hospital Westmead, Australia; Chuuk State Hospital, Micronesia, Federated States of; City of El Paso Dept of Public Health, United States; Clinical Virology Unit, CDIM, Australia; Colorado Department of Health Lab, United States; Connecticut Department.
+of Public Health, United States; Contiguo a Hospital Rosales, El Salvador; Croatian Institute of Public Health , Croatia; CRR virus Influenza region Sud, France; CRR virus Influenza region Sud, Guyana; CSL Ltd, United States; Dallas County Health and Human Services, United States; DC Public Health Lab, United States; Delaware Public Health Lab, United States; Departamento de Laboratorio de Salud Publica, Uruguay; Department of Virology, Medical University Vienna, Austria; Disease Investigation Centre Wates (BBVW), Australia; Drammen Hospital / Vestreviken HF, Norway; Ehime Prefecture Institute of Public Health and Environmental Science, Japan; Erasmus Medical Center, Netherlands; Erasmus University of Rotterdam, Netherlands; Ethiopian Health and Nutrition Research Institute (EHNRI), Ethiopia; Evanston Hospital and NorthShore University, United States; Facultad de Medicina, Spain; Fiji Centre for Communicable Disease Control, Fiji; Florida Department of Health, United States; Fukui Prefectural Institute of Public Health, Japan; Fukuoka City Institute for Hygiene and the Environment, Japan; Fukuoka Institute of Public Health and Environmental Sciences, Japan; Fukushima Prefectural Institute of Public Health, Japan; Gart Naval General Hospital, United Kingdom; Georgia Public Health Laboratory, United States; Gifu Municipal Institute of Public Health, Japan; Gifu Prefectural Institute of Health and Environmental Sciences, Japan; Government Virus Unit, Hong Kong; Gunma Prefectural Institute of Public Health and Environmental Sciences, Japan; Hamamatsu City Health Environment Research Center, Japan; Haukeland University Hospital, Dept.
+of Microbiology , Norway; Headquarters British Gurkhas Nepal, Nepal; Health Forde, Department of Microbiology, Norway; Health Protection Agency, United Kingdom; Health Protection Inspectorate, Estonia; Hellenic Pasteur Institute, Greece; Hiroshima City Institute of Public Health, Japan; Hokkaido Institute of Public Health, Japan; Hopital Cantonal Universitaire de Geneves, Switzerland; Hopital Charles Nicolle, Tunisia; Hospital Clinic de Barcelona, Spain; Hospital Universitari Vall d'Hebron, Spain; Houston Department of Health and Human Services, United States; Hyogo Prefectural Institute of Public Health and Consumer Sciences, Japan; Ibaraki Prefectural Institute of Public Health, Japan; Illinois Department of Public Health, United States; Indiana State Department of Health Laboratories, United States; Infectology Center of Latvia, Latvia; Innlandet Hospital Trust, Division Lillehammer, Department for Microbiology, Norway; INSA National Institute of Health Portugal, Portugal; Institut National d'Hygiene, Morocco; Institut Pasteur d'Algerie, Algeria; Institut Pasteur de Dakar, Senegal; Institut Pasteur de Madagascar, Madagascar; Institut Pasteur in Cambodia, Cambodia; Institut Pasteur New Caledonia, New Caledonia; Institut Pasteur, France; Institut Pasteur, Saudi Arabia; Institut Penyelidikan Perubatan, Malaysia; Institute National D'Hygiene, Togo; Institute of Environmental Science and Research, New Zealand; Institute of Environmental Science and Research, Tonga; Institute of Epidemiology and Infectious Diseases, Ukraine; Institute of Epidemiology Disease Control and Research, Bangladesh; Institute of Immunology and Virology Torlak, Serbia; Institute of Medical and Veterinary Science (IMVS), Australia; Institute of Public Health, Serbia; Institute of Public Health, Albania; Institute of Public Health, Montenegro; Institute Pasteur du Cambodia, Cambodia; Instituto Adolfo Lutz, Brazil; Instituto Conmemorativo Gorgas de Estudios de la Salud, Panama; Instituto de Salud Carlos III, Spain; Instituto de Salud Publica de Chile, Chile; Instituto Nacional de Enfermedades Infecciosas, Argentina; Instituto Nacional de Higiene Rafael Rangel, Venezuela, Bolivia; Instituto Nacional de Laboratoriosde Salud (INLASA), Bolivia; Instituto Nacional de Salud de Columbia, Colombia; Instituto Nacional de Saude, Portugal; Iowa State Hygienic Laboratory, United States; IRSS, Burkina Faso; Ishikawa Prefectural Institute of Public Health and Environmental Science, Japan; ISS, Italy; Istanbul University, Turkey; Istituto Superiore di Sanità, Italy; Ivanovsky Research Institute of Virology RAMS, Russian Federation; Jiangsu Provincial Center for Disease Control and Prevention, China; John Hunter Hospital, Australia; Kagawa Prefectural Research Institute for Environmental Sciences and Public Health, Japan; Kagoshima Prefectural Institute for Environmental Research and Public Health, Japan; Kanagawa Prefectural Institute of Public Health, Japan; Kansas Department of Health and Environment, United States; Kawasaki City Institute of Public Health , Japan; Kentucky Division of Laboratory Services, United States; Kitakyusyu City Institute of Enviromental Sciences, Japan; Kobe Institute of Health, Japan; Kochi Public Health and Sanitation Institute, Japan; Kumamoto City Environmental Research Center, Japan; Kumamoto Prefectural Institute of Public Health and Environmental Science, Japan; Kyoto City Institute of Health and Environmental Sciences, Japan; Kyoto Prefectural Institute of Public Health and Environment, Japan; Laboratoire National de Sante Publique, Haiti; Laboratoire National de Sante, Luxembourg; Laboratório Central do Estado do Paraná, Brazil; Laboratorio Central do Estado do Rio de Janeiro, Brazil; Laboratorio de Investigacion / Centro de Educacion Medica y Amistad Dominico Japones (CEMADOJA), Dominican Republic; Laboratorio De Saude Publico, Macao; Laboratorio de Virologia, Direccion de Microbiologia, Nicaragua; Laboratorio de Virus Respiratorio, Mexico; Laboratorio Nacional de Influenza, Costa Rica; Laboratorio Nacional De Salud Guatemala, Guatemala; Laboratorio Nacional de Virologia, Honduras; Laboratory Directorate, Jordan; Laboratory for Virology, National Institute of Public Health, Slovenia; Laboratory of Influenza and ILI, Belarus; LACEN/RS - Laboratório Central de Saúde Pública do Rio Grande do Sul, Brazil; Landspitali - University Hospital, Iceland; Lithuanian AIDS Center Laboratory, Lithuania; Los Angeles Quarantine Station, CDC Quarantine Epidemiology and Surveillance Team, United States; Louisiana Department of Health and Hospitals, United States; Maine Health and Environmental Testing Laboratory, United States; Malbran, Argentina; Marshfield Clinic Research Foundation, United States; Maryland Department of Health and Mental Hygiene, United States; Massachusetts Department of Public Health, United States; Mater Dei Hospital, Malta; Medical Research Institute, Sri Lanka; Medical University Vienna, Austria; Melbourne Pathology, Australia; Michigan Department of Community Health, United States; Mie Prefecture Health and Environment Research Institute, Japan; Mikrobiologisk laboratorium, Sykehuset i Vestfold, Norway; Ministry of Health and Population, Egypt; Ministry of Health of Ukraine, Ukraine; Ministry of Health, Bahrain; Ministry of Health, Kiribati; Ministry of Health, Lao, People's Democratic Republic; Ministry of Health, NIHRD, Indonesia; Ministry of Health, Oman; Minnesota Department of Health, United States; Mississippi Public Health Laboratory, United States; Missouri Department.
+of Health and Senior Services, United States; Miyagi Prefectural Institute of Public Health and Environment, Japan; Miyazaki Prefectural Institute for Public Health and Environment, Japan; Molde Hospital, Laboratory for Medical Microbiology, Norway; Molecular Diagnostics Unit , United Kingdom; Monash Medical Centre, Australia; Montana Laboratory Services Bureau, United States; Montana Public Health Laboratory, United States; Nagano City Health Center, Japan; Nagano Environmental Conservation Research Institute, Japan; Nagoya City Public Health Research Institute, Japan; Nara Prefectural Institute for Hygiene and Environment, Japan; National Center for Communicable Diseases, Mongolia; National Center for Laboratory and Epidemiology, Laos; National Centre for Disease Control (NCDC), Mongolia; National Centre for Disease Control and Public Health, Georgia; National Centre for Preventive Medicine, Moldova, Republic of; National Centre for Scientific Services for Virology and Vector Borne Diseases, Fiji; National Health Laboratory, Japan; National Health Laboratory, Myanmar; National Influenza Center French Guiana and French Indies, French Guiana; National Influenza Center, Brazil; National Influenza Center, Mongolia; National Influenza Centre for Northern Greece, Greece; National Influenza Centre of Iraq, Iraq; National Influenza Lab, Tanzania, United Republic of; National Influenza Reference Laboratory, Nigeria; National Insitut of Hygien, Morocco; National Institute for Biological Standards and Control (NIBSC), United States; National Institute for Communicable Disease, South Africa; National Institute for Health and Welfare, Finland; National Institute of Health Research and Development, Indonesia; National Institute of Health, Korea, Republic of; National Institute of Health, Pakistan; National Institute of Hygien, Morocco; National Institute of Hygiene and Epidemiology, Vietnam; National Institute of Public Health - National Institute of Hygiene, Poland; National Institute of Public Health, Czech Republic; National Institute of Virology, India; National Microbiology Laboratory, Health Canada, Canada; National Public Health Institute of Slovakia, Slovakia; National Public Health Laboratory, Cambodia; National Public Health Laboratory, Ministry of Health, Singapore, Singapore; National Public Health Laboratory, Nepal; National Public Health Laboratory, Singapore; National Reference Laboratory, Kazakhstan; National University Hospital, Singapore; National Virology Laboratory, Center Microbiological Investigations, Kyrgyzstan; National Virus Reference Laboratory, Ireland; Naval Health Research Center, United States; Nebraska Public Health Lab, United States; Nevada State Health Laboratory, United States; New Hampshire Public Health Laboratories, United States; New Jersey Department of Health and Senior Services, United States; New Mexico Department of Health, United States; New York City Department of Health, United States; New York Medical College, United States; New York State Department of Health, United States; Nicosia General Hospital, Cyprus; Niigata City Institute of Public Health and Environment, Japan; Niigata Prefectural Institute of Public Health and Environmental Sciences, Japan; Niigata University, Japan; Nordlandssykehuset, Norway; North Carolina State Laboratory of Public Health, United States; North Dakota Department of Health, United States; Norwegian Institute of Public Health, Norway; Norwegian Institute of Public Health, Svalbard and Jan Mayen; Ohio Department of Health Laboratories, United States; Oita Prefectural Institute of Health and Environment, Japan; Okayama Prefectural Institute for Environmental Science and Public Health, Japan; Okinawa Prefectural Institute of Health and Environment, Japan; Oklahoma State Department of Health, United States; Ontario Agency for Health Protection and Promotion (OAHPP), Canada; Oregon Public Health Laboratory, United States; Osaka City Institute of Public Health and Environmental Sciences, Japan; Osaka Prefectural Institute of Public Health, Japan; Oslo University Hospital, Ulleval Hospital, Dept.
+of Microbiology, Norway; Ostfold Hospital - Fredrikstad, Dept.
+of Microbiology, Norway; Oswaldo Cruz Institute - FIOCRUZ - Laboratory of Respiratory Viruses and Measles (LVRS), Brazil; Papua New Guinea Institute of Medical Research, Papua New Guinea; Pasteur Institut of Cote d'Ivoire, Cote d'Ivoire; Pasteur Institute, Influenza Laboratory, Vietnam; Pathwest QE II Medical Centre, Australia; Pennsylvania Department of Health, United States; Prince of Wales Hospital, Australia; Princess Margaret Hospital for Children, Australia; Public Health Laboratory Services Branch, Centre for Health Protection, Hong Kong; Public Health Laboratory, Barbados; Puerto Rico Department of Health, Puerto Rico; Qasya Diagnostic Services Sdn Bhd, Brunei; Queensland Health Scientific Services, Australia; Refik Saydam National Public Health Agency, Turkey; Regent Seven Seas Cruises, United States; Royal Victoria Hospital, United Kingdom; Republic Institute for Health Protection, Macedonia, the former Yogoslav Republic of; Republic of Nauru Hospital, Nauru; Research Institute for Environmental Sciences and Public Health of Iwate Prefecture, Japan; Research Institute of Tropical Medicine, Philippines; Rhode Island Department of Health, United States; RIVM National Institute for Public Health and Environment, Netherlands; Robert-Koch-Institute, Germany; Royal Chidrens Hospital, Australia; Royal Darwin Hospital, Australia; Royal Hobart Hospital, Australia; Royal Melbourne Hospital, Australia; Russian Academy of Medical Sciences, Russian Federation; Rwanda Biomedical Center, National Reference Laboratory, Rwanda; Saga Prefectural Institute of Public Health and Pharmaceutical Research, Japan; Sagamihara City Laboratory of Public Health, Japan; Saitama City Institute of Health Science and Research, Japan; Saitama Institute of Public Health, Japan; Sakai City Institute of Public Health, Japan; San Antonio Metropolitan Health, United States; Sandringham, National Institute for Communicable D, South Africa; Sapporo City Institute of Public Health, Japan; Scientific Institute of Public Health, Belgium; Seattle and King County Public Health Lab, United States; Sendai City Institute of Public Health, Japan; Servicio de Microbiología Clínica Universidad de Navarra, Spain; Servicio de Microbiología Complejo Hospitalario de Navarra, Spain; Servicio de Microbiología Hospital Central Universitario de Asturias, Spain; Servicio de Microbiología Hospital Donostia, Spain; Servicio de Microbiología Hospital Meixoeiro, Spain; Servicio de Microbiología Hospital Miguel Servet, Spain; Servicio de Microbiología Hospital Ramón y Cajal, Spain; Servicio de Microbiología Hospital San Pedro de Alcántara, Spain; Servicio de Microbiología Hospital Santa María Nai, Spain; Servicio de Microbiología Hospital Universitario de Gran Canaria Doctor Negrín, Spain; Servicio de Microbiología Hospital Universitario Son Espases, Spain; Servicio de Microbiología Hospital Virgen de la Arrixaca, Spain; Servicio de Microbiología Hospital Virgen de las Nieves, Spain; Servicio de Virosis Respiratorias INEI-ANLIS Carlos G.
+Malbran, Argentina; Shiga Prefectural Institute of Public Health, Japan; Shimane Prefectural Institute of Public Health and Environmental Science, Japan; Shizuoka City Institute of Environmental Sciences and Public Health , Japan; Shizuoka Institute of Environment and Hygiene, Japan; Singapore General Hospital, Singapore; Sorlandet Sykehus HF, Dept.
+of Medical Microbiology, Norway; South Carolina Department of Health, United States; South Dakota Public Health Lab, United States; Southern Nevada Public Health Lab, United States; Spokane Regional Health District, United States; St.\ Judes Childrens Research Hospital, United States; St.
+Olavs Hospital HF, Dept.
+of Medical Microbiology, Norway; State Agency, Infectology Center of Latvia, Latvia; State of Hawaii Department of Health, United States; State of Idaho Bureau of Laboratories, United States; State Research Center of Virology and Biotechnology Vector, Russian Federation; Statens Serum Institute, Denmark; Stavanger Universitetssykehus, Avd.
+for Medisinsk Mikrobiologi, Norway; Subdireccion General de Epidemiologia y Vigilancia de la Salud, Spain; Subdirección General de Epidemiología y Vigilancia de la Salud, Spain; Swedish Institute for Infectious Disease Control, Sweden; Swedish National Institute for Communicable Disease Control, Sweden; Taiwan CDC, Taiwan; Tan Tock Seng Hospital, Singapore; Tehran University of Medical Sciences, Iran; Tennessee Department of Health Laboratory-Nashville, United States; Texas Childrens Hospital, United States; Texas Department of State Health Services, United States; Thai National Influenza Center, Thailand; Thailand MOPH-U.S.
+CDC Collaboration (IEIP), Thailand; The Nebraska Medical Center, United States; Tochigi Prefectural Institute of Public Health and Environmental Science, Japan; Tokushima Prefectural Centre for Public Health and Environmental Sciences, Japan; Tokyo Metropolitan Institute of Public Health, Japan; Tottori Prefectural Institute of Public Health and Environmental Science, Japan; Toyama Institute of Health, Japan; U.S.
+Air Force School of Aerospace Medicine, United States; U.S.
+Naval Medical Research Unit No.3, Egypt; Uganda Virus Research Institute (UVRI), National Influenza Center, Uganda; Universidad de Valladolid, Spain; Università Cattolica del Sacro Cuore, Italy; Universitetssykehuset Nord-Norge HF, Norway; University Malaya, Malaysia; University of Florence, Italy; University of Genoa, Italy; University of Ghana, Ghana; University of Michigan SPH EPID, United States; University of Parma, Italy; University of Perugia, Italy; University of Pittsburgh Medical Center Microbiology Lab, United States; University of Sarajevo, Bosnia and Herzegovina; University of Sassari, Italy; University of the West Indies, Jamaica; University of Vienna, Austria; University of Virginia, Medical Labs/Microbiology, United States; University Teaching Hospital, Zambia; UPMC-CLB Dept of Microbiology, United States; US Army Medical Research Unit - Kenya (USAMRU-K), GEIS Human Influenza Program, Kenya; USAMC-AFRIMS Department of Virology, Cambodia; Utah Department of Health, United States; Utah Public Health Laboratory, United States; Utsunomiya City Institute of Public Health and Environment Science, Japan; VACSERA, Egypt; Vermont Department of Health Laboratory, United States; Victorian Infectious Diseases Reference Laboratory, Australia; Virginia Division of Consolidated Laboratories, United States; Wakayama City Institute of Public Health, Japan; Wakayama Prefectural Research Center of Environment and Public Health, Japan; Washington State Public Health Laboratory, United States; West Virginia Office of Laboratory Services, United States; Westchester County Department of Laboratories and Research, United States; Westmead Hospital, Australia; WHO National Influenza Centre Russian Federation, Russian Federation; WHO National Influenza Centre, National Institute of Medical Research (NIMR), Thailand; WHO National Influenza Centre, Norway; Wisconsin State Laboratory of Hygiene, United States; Wyoming Public Health Laboratory, United States; Yamagata Prefectural Institute of Public Health, Japan; Yamaguchi Prefectural Institute of Public Health and Environment, Japan; Yamanashi Institute for Public Health, Japan; Yap State Hospital, Micronesia; Yokohama City Institute of Health, Japan; Yokosuka Institute of Public Health, Japan
+
+\end{appendixbox}

### 2020-08-06 manuscript/main.tex
Symlink figures as "Figure_1.pdf" style names

@@
   \begin{center}
-  \includegraphics[width=\columnwidth]{figures/distance-based-fitness-model.pdf}
+  \includegraphics[width=\columnwidth]{figures/Figure_1.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations-controls.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_2.pdf}
   \caption{
@@
 \begin{figure}[htb]
-  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_3.pdf}
   \caption{
@@
   \figsupp[Composite model coefficients and distances to the future for models fit to simulated populations.]{
-    A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.}{\includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}
+    A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.}{\includegraphics[width=\textwidth]{figures/Figure_3-Figure_supplement_1.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}
 
@@
 \begin{figure}
-  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-normalized_fitness.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_4.pdf}
   \caption{
@@
   D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-  The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-normalized_fitness.pdf}}\label{figsupp:validation_of_best_model_for_simulated_populations}
+  The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}.}{\includegraphics[width=\textwidth]{figures/Figure_4-Figure_supplement_1.pdf}}\label{figsupp:validation_of_best_model_for_simulated_populations}
 
   \figsupp[Validation of naive model for simulated populations of H3N2-like viruses.]{
-    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-naive.pdf}}\label{figsupp:validation_of_naive_model_for_simulated_populations}
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_4-Figure_supplement_2.pdf}}\label{figsupp:validation_of_naive_model_for_simulated_populations}
 
   \figsupp[Test of naive model for simulated populations of H3N2-like viruses.]{
-    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-naive.pdf}}\label{figsupp:test_of_naive_model_for_simulated_populations}
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_4-Figure_supplement_3.pdf}}\label{figsupp:test_of_naive_model_for_simulated_populations}
 
@@
 \begin{figure}[htb]
-  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_5.pdf}
   \caption{
@@
     The ``oracle'' antigenic novelty model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
-    The improved performance of the ``oracle'' model indicates that the sequence-based antigenic novelty metric can be effective when sites of beneficial mutations are known prior to forecasting.}{\includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
+    The improved performance of the ``oracle'' model indicates that the sequence-based antigenic novelty metric can be effective when sites of beneficial mutations are known prior to forecasting.}{\includegraphics[width=\textwidth]{figures/Figure_5-Figure_supplement_1.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}
 
@@
 \begin{figure}[htb]
-  \includegraphics[width=\textwidth]{figures/best-composite-unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_6.pdf}
   \caption{
@@
     The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
-    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.}{\includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.pdf}}\label{figsupp:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
+    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.}{\includegraphics[width=\textwidth]{figures/Figure_6-Figure_supplement_1.pdf}}\label{figsupp:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
 
@@
 \begin{figure}
-  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-cTiter_x-ne_star.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_7.pdf}
   \caption{
@@
     D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
-    The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_natural_populations}.}{\includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-ne_star-lbi.pdf}}\label{figsupp:validation_of_best_model_for_natural_populations}
+    The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_natural_populations}.}{\includegraphics[width=\textwidth]{figures/Figure_7-Figure_supplement_1.pdf}}\label{figsupp:validation_of_best_model_for_natural_populations}
 
   \figsupp[Validation of naive model for natural populations of H3N2 viruses.]{
-    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding-naive.pdf}}\label{figsupp:validation_of_naive_model_for_natural_populations}
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_7-Figure_supplement_2.pdf}}\label{figsupp:validation_of_naive_model_for_natural_populations}
 
   \figsupp[Test of naive model for natural populations of H3N2 viruses.]{
-    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-naive.pdf}}\label{figsupp:test_of_naive_model_for_natural_populations}
+    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_7-Figure_supplement_3.pdf}}\label{figsupp:test_of_naive_model_for_natural_populations}
 
@@
 \begin{figure}[htb]
-  \includegraphics[width=\textwidth]{figures/vaccine-comparison.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_8.pdf}
   \caption{
@@
     Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Figure~\ref{fig:vaccine_comparison}.
-    Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.}{\includegraphics[width=\textwidth]{figures/vaccine-comparison-relative-distance.pdf}}\label{figsupp:vaccine_comparison_relative_distance}
+    Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.}{\includegraphics[width=\textwidth]{figures/Figure_8-Figure_supplement_1.pdf}}\label{figsupp:vaccine_comparison_relative_distance}
 \end{figure}
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/nextstrain-forecasts.png}
+  \includegraphics[width=\textwidth]{figures/Figure_9.png}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/nextstrain-weighted-distance-to-future-per-strain.png}
+  \includegraphics[width=\textwidth]{figures/Figure_10.png}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/cross-validation-for-simulated-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_11.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/simulated-h3n2-ha-phylogeny.png}
+  \includegraphics[width=\textwidth]{figures/Figure_12.png}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/cross-validation-for-natural-populations.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_13.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_simulated_sample_3.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_14.pdf}
   \caption{
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding.pdf}
+  \includegraphics[width=\textwidth]{figures/Figure_15.pdf}
   \caption{

### 2020-08-10 manuscript/main.tex
Add Pierre to the manuscript
Clarify figure captions especially for figure supplements
Expands figure supplement captions to be more understandable on their own and
cross-references each validation figure to its corresponding part of the
cross-validation schematic for simulated and natural populations. This latter
change should help readers distinguish between "validation" and "test" results.

@@
   \figsupp[Composite model coefficients and distances to the future for models fit to simulated populations.]{
+    Composite model coefficients and distances to the future for models fit to simulated populations.
     A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.}{\includegraphics[width=\textwidth]{figures/Figure_3-Figure_supplement_1.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}
@@
   \caption{
-  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
+  Test of best model for simulated populations (true fitness) using 9 years (18 timepoints) of previously unobserved test data and fixed model coefficients.
+  These timepoints correspond to the open circles in Figure~\ref{fig:cross_validation_for_simulated_populations}.
   A) The correlation of log estimated clade frequency fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade frequency fold change, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
   \figsupp[Validation of best model for simulated populations of H3N2-like viruses.]{
+    Validation of best model for simulated populations of H3N2-like viruses for 33 timepoints.
+    These timepoints correspond to the closed circles in Figure~\ref{fig:cross_validation_for_simulated_populations}.
     A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
   \figsupp[Validation of naive model for simulated populations of H3N2-like viruses.]{
+    Validation of naive model for simulated populations of H3N2-like viruses for 33 timepoints.
+    These timepoints correspond to the closed circles in Figure~\ref{fig:cross_validation_for_simulated_populations}.
     Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_4-Figure_supplement_2.pdf}}\label{figsupp:validation_of_naive_model_for_simulated_populations}
@@
   \figsupp[Test of naive model for simulated populations of H3N2-like viruses.]{
+    Test of naive model for simulated populations of H3N2-like viruses for 18 timepoints.
+    These timepoints correspond to the open circles in Figure~\ref{fig:cross_validation_for_simulated_populations}.
     Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_4-Figure_supplement_3.pdf}}\label{figsupp:test_of_naive_model_for_simulated_populations}
@@
   \caption{
-  Test of best model for natural populations of H3N2 viruses, the composite model of HI antigenic novelty and mutational load.
+  Test of best model for natural populations of H3N2 viruses, the composite model of HI antigenic novelty and mutational load, across eight timepoints.
+  These timepoints correspond to the open circles in Figure~\ref{fig:cross_validation_for_natural_populations}.
   A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
   \figsupp[Validation of best model for natural populations of H3N2 viruses.]{
-    Validation results for the composite model of mutational load and LBI.
+    Validation of best model for natural populations of H3N2 viruses for 23 timepoints using the composite model of mutational load and LBI.
+    These timepoints correspond to the closed circles in Figure~\ref{fig:cross_validation_for_natural_populations}.
     A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
@@
   \figsupp[Validation of naive model for natural populations of H3N2 viruses.]{
+    Validation of naive model for natural populations of H3N2 viruses for 23 timepoints.
+    These timepoints correspond to the closed circles in Figure~\ref{fig:cross_validation_for_natural_populations}.
     Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_7-Figure_supplement_2.pdf}}\label{figsupp:validation_of_naive_model_for_natural_populations}
@@
   \figsupp[Test of naive model for natural populations of H3N2 viruses.]{
+    Test of naive model for natural populations of H3N2 viruses for eight timepoints.
+    These timepoints correspond to the open circles in Figure~\ref{fig:cross_validation_for_natural_populations}.
     Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/Figure_7-Figure_supplement_3.pdf}}\label{figsupp:test_of_naive_model_for_natural_populations}
@@
 The Melbourne WHO Collaborating Centre for Reference and Research on Influenza is supported by the Australian Government Department of Health.
-RAN is supported by NIAID R01 AI127893-01 and institutional core funding.
+PB and RAN are supported by NIAID R01 AI127893-01 and institutional core funding.
 TB is a Pew Biomedical Scholar and is supported by NIH grants NIGMS R35 GM119774-01, NIAID U19 AI117891-01 and NIAID R01 AI127893-01.
@@
 JB, TR, XX, RK, DEW, LW, BE, RSD, JWM, SF, KN, NK, SW, HH, IB, and KS performed and provided data from serological assays.
+PB analyzed results.
 RAN planned experiments and edited the manuscript.

### 2020-08-10 manuscript/main.tex
Replace "best strain" with "closest strain" and explain
Replaces the "best" value judgement with the quantitative "closest" for each
closest strain to the future, explains how we calculate these closest strains to
the future the first time they are mentioned, and references the equations
used. Also expands the interpretation of these closest strains in the vaccine
comparison section to clarify that they represent centroids of the observed and
estimated future populations.

@@
   A) The correlation of log estimated clade frequency fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade frequency fold change, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future in the best model was in the top 20th percentile for 89\% of 18 timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+  B) The rank of the estimated closest strain based on its distance to the future in the best model was in the top 20th percentile for 89\% of 18 timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
   C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
@@
     A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-    B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
-  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+    B) The rank of the estimated closest strain based on its distance to the future for 33 timepoints.
+  The estimated closest strain was in the top 20th percentile of observed closest strains for 100\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
   C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
@@
 Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
-The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}B).
+We calculated the distance of each present strain to the future as the Hamming distance between the given strain's amino acid sequence and each future strain weighted by the future strain's observed or estimated frequency (Equations~\ref{equation_observed_weighted_distance_to_future} and \ref{equation_estimated_weighted_distance_to_future}).
+The estimated closest strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}B).
 Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}D).
@@
 We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Figure~\ref{fig:test_of_best_model_for_simulated_populations}C).
-The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}B).
+The estimated closest strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}B).
 Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}D).
@@
 
-As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade frequency fold changes and the ranking of estimated best strains compared to the observed closest strains to future populations.
+As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade frequency fold changes and the ranking of estimated closest strains compared to the observed closest strains to future populations.
 The composite model of mutational load and LBI effectively captured clade dynamics with a fold change correlation of $R^2 = 0.35$ and growth and decline accuracies of 87\% and 89\%, respectively (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}A).
 Absolute forecasting error declined noticeably for clades with initial frequencies above 60\%, but generally this error remained below 20\% on average (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}C).
-The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}B).
+The estimated closest strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 20 (87\%) of 23 timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}B).
 This pattern held across all strains and timepoints with a strong correlation between observed and estimated strain ranks (Spearman's $\rho^2 = 0.66$, Figure~\ref{fig:test_of_best_model_for_natural_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_natural_populations}D).
@@
   A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-  B) The rank of the estimated best strain based on its distance to the future for eight timepoints.
-  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints.
+  B) The rank of the estimated closest strain based on its distance to the future for eight timepoints.
+  The estimated closest strain was in the top 20th percentile of observed closest strains for 100\% of timepoints.
   C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
@@
     A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
-    B) The rank of the estimated best strain based on its distance to the future for 23 timepoints.
-    The estimated best strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
+    B) The rank of the estimated closest strain based on its distance to the future for 23 timepoints.
+    The estimated closest strain was in the top 20th percentile of observed closest strains for 87\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
     C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
@@
 
-After identifying the composite HI antigenic novelty and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual best strains for vaccine composition.
+After identifying the composite HI antigenic novelty and mutational load model as the best model on out-of-sample data, we tested this model's ability to detect clade dynamics and select individual closest strains to the future for vaccine composition.
 The composite model partially captured clade dynamics with a Pearson's correlation of $R^2 = 0.46$ between observed and estimated growth ratios and growth and decline accuracies of 52\% and 58\%, respectively (Figure~\ref{fig:test_of_best_model_for_natural_populations}A).
 The mean absolute forecasting error with this model was consistently less than 20\%, regardless of the initial clade frequency (Figure~\ref{fig:test_of_best_model_for_natural_populations}C).
-The estimated best strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}B).
+The estimated closest strain from this model was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}B).
 Similarly, the observed and estimated strain ranks strongly correlated (Spearman's $\rho^2 = 0.72$) across all strains and test timepoints (Figure~\ref{fig:test_of_best_model_for_natural_populations}D).
@@
   \figsupp[Relative improvement of model selections over vaccine strains.]{
-    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Figure~\ref{fig:vaccine_comparison}.
+    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated closest strains at each timepoint as in Figure~\ref{fig:vaccine_comparison}.
     Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.}{\includegraphics[width=\textwidth]{figures/Figure_8-Figure_supplement_1.pdf}}\label{figsupp:vaccine_comparison_relative_distance}
@@
 We further evaluated our models' ability to estimate the closest strain to the next season's H3N2 population by comparing our best models' selections to the WHO's vaccine strain selection.
-For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed closest strains.
-We compared these distances to those of the closest strains to the future as estimated by our best models for the validation period (mutational load and LBI) and the test period (HI antigenic novelty and mutational load).
+For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed closest strains (Equation~\ref{equation_observed_weighted_distance_to_future}).
+We compared these distances to those of the closest strains to the future as estimated by our best models for the validation period (mutational load and LBI) and the test period (HI antigenic novelty and mutational load) using Equation~\ref{equation_estimated_weighted_distance_to_future}.
+The observed closest strain to the future represents the centroid of the observed future population, while the estimated closest strains are the models' predictions of that future population's centroid.
 The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Figure~\ref{fig:vaccine_comparison}).
@@
 Interestingly, the strains selected by the naive model were always better than the selected vaccine strain.
+Since the naive model predicts that the future will be identical to the present, these strains represent the centroid of each current population.
 With an average improvement over the vaccine strains of 2.19 AAs, the naive model performed consistently better than the LBI-based model and nearly as well as the HI-based model.

### 2020-08-17 scripts/plot_validation_figure_by_population.py
Replace references to "best" with "closest" strain in figures
Previous commits replaced this text in the main manuscript, but I missed the
embedded text in these figures.

@@ -137,7 +137,7 @@ if __name__ == "__main__":
     )
 
     # Next, load the complete forecasts data frame for ranking of estimated and
-    # observed best strains.
+    # observed closest strains.
     full_forecasts = pd.read_csv(
         args.forecasts,
         sep="\t",
@@ -364,7 +364,7 @@ if __name__ == "__main__":
     rank_ax.legend(
         frameon=False
     )
-    rank_ax.set_xlabel("Percentile rank by distance\nfor estimated best strain")
+    rank_ax.set_xlabel("Percentile rank by distance\nfor estimated closest strain")
     rank_ax.set_ylabel("Number of timepoints")
 
     #

### 2020-08-17 manuscript/tables_composite_vs_individual_model_comparison.tex

\begin{tabular}{llllrrl}
\toprule
    sample &  error\_type &      individual\_model &                         composite\_model &  bootstrap\_mean &  bootstrap\_std &    p\_value \\
\midrule
 simulated &  validation &          true fitness &                  mutational load +  LBI &            0.42 &           0.23 &     0.9644 \\
 simulated &  validation &       mutational load &                  mutational load +  LBI &           -1.03 &           0.21 &  $<$0.0001 \\
 simulated &  validation &                   LBI &                  mutational load +  LBI &           -0.33 &           0.14 &     0.0091 \\
 simulated &        test &          true fitness &                  mutational load +  LBI &           -0.28 &           0.26 &     0.1392 \\
 simulated &        test &       mutational load &                  mutational load +  LBI &           -1.11 &           0.25 &  $<$0.0001 \\
 simulated &        test &                   LBI &                  mutational load +  LBI &           -0.42 &           0.16 &     0.0001 \\
   natural &  validation &       mutational load &                  mutational load +  LBI &           -0.69 &           0.28 &     0.0036 \\
   natural &  validation &                   LBI &                  mutational load +  LBI &           -0.23 &           0.09 &     0.0025 \\
   natural &  validation &       mutational load &  mutational load + HI antigenic novelty &           -0.31 &           0.18 &     0.0417 \\
   natural &  validation &  HI antigenic novelty &  mutational load + HI antigenic novelty &           -0.18 &           0.11 &     0.0513 \\
   natural &        test &       mutational load &                  mutational load +  LBI &            1.19 &           0.79 &     0.9432 \\
   natural &        test &                   LBI &                  mutational load +  LBI &           -0.70 &           0.24 &  $<$0.0001 \\
   natural &        test &       mutational load &  mutational load + HI antigenic novelty &           -0.56 &           0.33 &     0.0133 \\
   natural &        test &  HI antigenic novelty &  mutational load + HI antigenic novelty &           -0.24 &           0.18 &     0.0999 \\
\bottomrule
\end{tabular}

### 2020-08-19 scripts/plot_validation_figure_by_population.py
Export data frames for validation figures in workflow
For better reproducibility and explorability of results, export the actual data
frames used to make validation figures including the clades and ranks of
strains.

@@ -84,7 +84,12 @@ if __name__ == "__main__":
     parser.add_argument("--forecasts", required=True, help="table of forecasts for the given tips")
     parser.add_argument("--model-errors", required=True, help="annotated validation errors for the model used to make the given forecasts")
     parser.add_argument("--bootstrap-samples", type=int, default=100, help="number of bootstrap samples to generate for confidence intervals around absolute forecast errors")
+    parser.add_argument("--population", help="the population being analyzed (e.g., simulated or natural)")
+    parser.add_argument("--sample", help="sample name for population being analyzed")
+    parser.add_argument("--predictors", help="predictors being analyzed")
     parser.add_argument("--output", required=True, help="validation figure")
+    parser.add_argument("--output-clades-table", help="table of clade frequencies used in left panels")
+    parser.add_argument("--output-ranks-table", help="table of strain ranks used in right panels")
 
     args = parser.parse_args()
 
@@ -462,3 +467,50 @@ if __name__ == "__main__":
             int(np.round((timepoints_better_than_20th_percentile / float(total_timepoints)) * 100))
         )
     )
+
+    if args.output_clades_table:
+        complete_clade_frequencies = complete_clade_frequencies.rename(columns={
+            "frequency": "initial_frequency",
+            "frequency_final": "observed_future_frequency",
+            "projected_frequency": "estimated_future_frequency"
+        })
+        complete_clade_frequencies["population"] = args.population
+        complete_clade_frequencies["predictors"] = args.predictors
+        complete_clade_frequencies["error_type"] = "test" if "test" in args.sample else "validation"
+
+        complete_clade_frequencies.to_csv(
+            args.output_clades_table,
+            sep="\t",
+            header=True,
+            index=False
+        )
+
+    if args.output_ranks_table:
+        sorted_df["observed_distance_to_future"] = sorted_df["weighted_distance_to_future"]
+        sorted_df["estimated_distance_to_future"] = sorted_df["y"]
+        sorted_df["observed_rank"] = sorted_df["timepoint_rank"]
+        sorted_df["estimated_rank"] = sorted_df["timepoint_estimated_rank"]
+
+        sorted_df["population"] = args.population
+        sorted_df["sample"] = args.sample
+        sorted_df["predictors"] = args.predictors
+        sorted_df["error_type"] = "test" if "test" in args.sample else "validation"
+        sorted_df = np.around(sorted_df, 4)
+
+        sorted_df.to_csv(
+            args.output_ranks_table,
+            sep="\t",
+            header=True,
[…]

### 2020-08-19 scripts/plot_validation_figure_by_population.py
Reduce precision for ranks data frame

@@ -495,7 +495,7 @@ if __name__ == "__main__":
         sorted_df["sample"] = args.sample
         sorted_df["predictors"] = args.predictors
         sorted_df["error_type"] = "test" if "test" in args.sample else "validation"
-        sorted_df = np.around(sorted_df, 4)
+        sorted_df = np.around(sorted_df, 2)
 
         sorted_df.to_csv(
             args.output_ranks_table,
